### 2024-12-03 paper.tex
Add paper template

% This is samplepaper.tex, a sample chapter demonstrating the
% LLNCS macro package for Springer Computer Science proceedings;
% Version 2.21 of 2022/01/12
%
\documentclass[runningheads]{llncs}
%
\usepackage[T1]{fontenc}
% T1 fonts will be used to generate the final print and online PDFs,
% so please use T1 fonts in your manuscript whenever possible.
% Other font encondings may result in incorrect characters.
%
\usepackage{graphicx}
% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.
%
% If you use the hyperref package, please uncomment the following two lines
% to display URLs in blue roman font according to Springer's eBook style:
%\usepackage{color}
%\renewcommand\UrlFont{\color{blue}\rmfamily}
%\urlstyle{rm}
%
\begin{document}
%
\title{Contribution Title}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{First Author\inst{1}\orcidID{0000-1111-2222-3333} \and
Second Author\inst{2,3}\orcidID{1111-2222-3333-4444} \and
Third Author\inst{3}\orcidID{2222--3333-4444-5555}}
%
\authorrunning{F. Author et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Princeton University, Princeton NJ 08544, USA \and
Springer Heidelberg, Tiergartenstr. 17, 69121 Heidelberg, Germany
\email{lncs@springer.com}\\
\url{http://www.springer.com/gp/computer-science/lncs} \and
ABC Institute, Rupert-Karls-University Heidelberg, Heidelberg, Germany\\
\email{\{abc,lncs\}@uni-heidelberg.de}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
The abstract should briefly summarize the contents of the paper in
150--250 words.

\keywords{First keyword  \and Second keyword \and Another keyword.}
\end{abstract}
%
%
%
\section{First Section}
\subsection{A Subsection Sample}
Please note that the first paragraph of a section or subsection is
not indented. The first paragraph that follows a table, figure,
equation etc. does not need an indent, either.

Subsequent paragraphs, however, are indented.

\subsubsection{Sample Heading (Third Level)} Only two levels of
headings should be numbered. Lower level headings remain unnumbered;
they are formatted as run-in headings.

\paragraph{Sample Heading (Fourth Level)}
The contribution should contain no more than four levels of
headings. Table~\ref{tab1} gives a summary of all heading levels.

\begin{table}
\caption{Table captions should be placed above the
tables.}\label{tab1}
\begin{tabular}{|l|l|l|}
\hline
Heading level &  Example & Font size and style\\
\hline
Title (centered) &  {\Large\bfseries Lecture Notes} & 14 point, bold\\
1st-level heading &  {\large\bfseries 1 Introduction} & 12 point, bold\\
2nd-level heading & {\bfseries 2.1 Printing Area} & 10 point, bold\\
3rd-level heading & {\bfseries Run-in Heading in Bold.} Text follows & 10 point, bold\\
4th-level heading & {\itshape Lowest Level Heading.} Text follows & 10 point, italic\\
\hline
\end{tabular}
\end{table}

\noindent Displayed equations are centered and set on a separate
line.
\begin{equation}
x + y = z
\end{equation}
Please try to avoid rasterized images for line-art diagrams and
schemas. Whenever possible, use vector graphics instead.

\begin{theorem}
This is a sample theorem. The run-in heading is set in bold, while
the following text appears in italics. Definitions, lemmas,
propositions, and corollaries are styled the same way.
\end{theorem}
%
% the environments 'definition', 'lemma', 'proposition', 'corollary',
% 'remark', and 'example' are defined in the LLNCS documentclass as well.
%
\begin{proof}
Proofs, examples, and remarks have the initial word in italics,
while the following text appears in normal font.
\end{proof}
For citations of references, we prefer the use of square brackets
and consecutive numbers. Citations using labels or the author/year
convention are also acceptable. The following bibliography provides
a sample reference list with entries for journal
articles~\cite{ref_article1}, an LNCS chapter~\cite{ref_lncs1}, a
book~\cite{ref_book1}, proceedings without editors~\cite{ref_proc1},
and a homepage~\cite{ref_url1}. Multiple citations are grouped
\cite{ref_article1,ref_lncs1,ref_book1},
\cite{ref_article1,ref_book1,ref_proc1,ref_url1}.

\begin{credits}
\subsubsection{\ackname} A bold run-in heading in small font size at the end of the paper is
used for general acknowledgments, for example: This study was funded
by X (grant number Y).

\subsubsection{\discintname}
It is now necessary to declare any competing interests or to specifically
state that the authors have no competing interests. Please place the
statement with a bold run-in heading in small font size beneath the
(optional) acknowledgments\footnote{If EquinOCS, our proceedings submission
system, is used, then the disclaimer can be provided directly in the system.},
for example: The authors have no competing interests to declare that are
relevant to the content of this article. Or: Author A has received research
grants from Company W. Author B has received a speaker honorarium from
Company X and owns stock in Company Y. Author C is a member of committee Z.
\end{credits}
%
% ---- Bibliography ----
%
% BibTeX users should specify bibliography style 'splncs04'.
% References will then be sorted and formatted in the correct style.
%
% \bibliographystyle{splncs04}
% \bibliography{mybibliography}
%
\begin{thebibliography}{8}
\bibitem{ref_article1}
Author, F.: Article title. Journal \textbf{2}(5), 99--110 (2016)

\bibitem{ref_lncs1}
Author, F., Author, S.: Title of a proceedings paper. In: Editor,
F., Editor, S. (eds.) CONFERENCE 2016, LNCS, vol. 9999, pp. 1--13.
Springer, Heidelberg (2016). \doi{10.10007/1234567890}

\bibitem{ref_book1}
Author, F., Author, S., Author, T.: Book title. 2nd edn. Publisher,
Location (1999)

\bibitem{ref_proc1}
Author, A.-B.: Contribution title. In: 9th International Proceedings
on Proceedings, pp. 1--2. Publisher, Location (2010)

\bibitem{ref_url1}
LNCS Homepage, \url{http://www.springer.com/lncs}, last accessed 2023/10/25
\end{thebibliography}
\end{document}

### 2025-01-30 paper.tex
wip update

@@
 \usepackage[T1]{fontenc}
+\usepackage{xcolor}
+\usepackage{minted}
+\usepackage{amsmath,amssymb}
 % T1 fonts will be used to generate the final print and online PDFs,
@@
 \usepackage{graphicx}
+\usepackage{pgfplots}
+
+\newtheorem{lem}{Lemma}
+
+\setminted[lean4]{extrakeywords={aesop cases add aesop?
+intro simp simp_all only split apply on_goal next rename_i safe unsafe norm constructors forward destruct norm_num done add_aesop_rules rfl subst ext}}
+\newmintinline[lean]{lean4}{bgcolor={},ignorelexererrors=true}
+\newminted[leancode]{lean4}{bgcolor={},ignorelexererrors=true,fontsize=\footnotesize,autogobble}
+\BeforeBeginEnvironment{leancode}{\begin{tcolorbox}}
+\AfterEndEnvironment{leancode}{\end{tcolorbox}}
+\usemintedstyle{xcode}
+
+\newcommand{\xcom}[1]{{\color{cyan}{Xavier: #1}} }
+\newcommand{\jcom}[1]{{\color{orange}{Jannis: #1}} }
 % Used for displaying a sample figure. If possible, figure files should
@@
 %
-\title{Contribution Title}
+\title{Stateful Forward Reasoning}
 %
@@
 %
-\institute{Princeton University, Princeton NJ 08544, USA \and
-Springer Heidelberg, Tiergartenstr. 17, 69121 Heidelberg, Germany
-\email{lncs@springer.com}\\
-\url{http://www.springer.com/gp/computer-science/lncs} \and
-ABC Institute, Rupert-Karls-University Heidelberg, Heidelberg, Germany\\
-\email{\{abc,lncs\}@uni-heidelberg.de}}
+\institute{University of Munich (LMU), Munich, Germany\\
+\email{\{xavier.genereux,jannis.limpberg\}@lmu.de}}
 %
@@
 
-\keywords{First keyword  \and Second keyword \and Another keyword.}
+\keywords{Forward Reasoning \and Proof Automation \and Theorem Proving.}
 \end{abstract}
@@
 %
+
+\section{Introduction}
+\xcom{TODO}
+
+\xcom{ Specify the problem (user defined rules)
+ + deal with changes with local context}
+\xcom{ Talk about aesop}
+\xcom{ Do we outline the solution?}
+
+An important contextualization is that the forward reasoning procedure in this paper takes place
+within aesop's routine procedure. We will focus on examples that highlight this style of proof
+while noting that in practice, a combination of both forwards and backwards reasoning are used 
+\section{Preliminaries}
+
+\xcom{Adapt for CADE.}
+
+In what follows, we write $\Gamma \models T$ for a goal where $\Gamma$ represents the context and
+$T$ the target. In a mathematical problem, $\Gamma$ is the list of hypotheses and variable from
+which we wish to show $T$.
+
+\paragraph{Forward Rules.}
+Consider the following lemma :
+\begin{lem}[\lean{le_trans}]
+  Given a preorder on some type $\alpha$ and elements $a,b,c$ of type $\alpha$, the following holds :
+  $$a \leq b \rightarrow b \leq c \rightarrow a \leq c$$
+\end{lem}
+
+From such lemmas, we will create what we call \textit{forward rules} that represent
+the theorems on which we would like to reason in a forward manner.
+The premises $\alpha$, $ \text{Preorder } \alpha$ , $a$, $b$, $c$, $a \leq b$ and
+$b \leq c$ are separated between \textit{implied hypotheses} $\text{Preorder } \alpha$,
+$a$, $b$ and $c$ and \textit{maximal hypotheses} $a \leq b$ and $b \leq c$.
+Maximal hypotheses are also called \textit{input hypotheses}.
+
+For the premises $(x_1 : T_1),(x_2 : T_2),\dots,(x_n : T_n)$,
+the input hypotheses consist of the hypotheses for which there are no
+$j > i$ such that $T_j$ depends on $x_i$.
+
+\paragraph{Rules.}
+We write $r : \Phi \to \Psi$ for a \textit{rule} $r$ with input
+hypotheses $\Phi$ and output hypothesis $\Psi$.
+For a given rule, we assign each maximal input hypothesis a unique index
+between $1$ and $n$.
+We call theses indices \textit{slots} of $r$ and define $\text{slots}(r)$ as
+the set of slots of $r$.
+
+Given a slot $i$ and hypothesis $h : T$ with $T \equiv_{[\sigma]} T_i$
+(i.e. $\sigma$ is the substitution that results from unifying $T$ and $T_i$), 
+we say that $h$ is \textit{suitable for slot $i$} and define $\text{sub}(h) ≔ \sigma$.
+
+A \textit{shared variable} of $r$ is an input hypothesis $x_i$ such that there are
+at least two input hypothesis types $T_j$, $T_k$ in which $x_i$ occurs.
+A shared variable is necessarily a non-maximal input hypothesis.
+We also write $?x_i$ to emphasise that we interpret $x_i$ as a variable
+(and, in the code, as a metavariable).
+The set $s_r(i)$ of \textit{shared variables of a slot} $i$ of $r$ are those
+shared variables $?x$ that occur in $T_i$ and in at least one slot $j < i$.
+If the rule in question is clear from the context, we write $s(i)$.
+
+%The typical behavior is that running the procedure in a context containing suitable
+%candidates for the maximal hypotheses will add the \textit{result} of the forward
+%rule $a < c$ to the context.
+%The result is also called the \textit{output hypothesis}.
+%If we find matching hypotheses for all maximal input hypotheses in the local context,
+%we can apply the forward rule, since each non-maximal input hypothesis is uniquely determined
+%by maximal hypotheses.
+
+A \textit{match} $M$ for $r$ is a partial map from slots of $r$ to hypotheses in
+the context.
+It must satisfy the following conditions:
+\begin{itemize}
+\item For each slot $i$ of $r$, if $M(i)$ is defined, then
+    $M(i) \equiv_{[\sigma_i]} T_i$ with a substitution $\sigma_i$.
+    The domain of this substitution is a subset of the non-maximal input hypotheses.
+\item For all slots $i$ and $j$ of $r$ such that $M(i)$ and $M(j)$ are defined,
+    $\sigma_i$ and $\sigma_j$ are compatible.
+\item $M$ is downward-closed: if $M(i)$ is defined, then for any slot $j < i$ of
+    $r$, $M(j)$ is also defined.
+
+\end{itemize}
+A match of $r$ is \textit{complete} if $M(i)$ is defined for every slot $i$ of $r$, and
+\textit{incomplete} otherwise.
+The \textit{level of a match} $M$, $lvl(M)$, is the maximal $i$ such that $M(i)$ 
+is defined.
+The \textit{substitution of a match} $M$, $sub(M)$, is the union of the
+substitutions $\sigma_i$ for each $i$ such that $M_i$ is defined.
+This union is defined since the $\sigma_i$ are pairwise compatible.
+This means that $sub(M)$ contains only one assignement per variable.
+The set of \textit{shared variables of a match} $M$ is $s(lvl(M))$ the
+shared variables of $lvl(M)$.
+
+An output hypothesis is \textit{redundant} if its type is already present in the context.
+A goal is saturated with respect to a set of forward rules if all
+output hypotheses returned by the procedure are redundant.
+
+The following section introduces an algorithm to construct complete matches $M$ 
+from the context $\Gamma$.
+
+\section{Design}
+
+We now focus on the problem of constructing complete matches.
+
+\subsection{A naive design}
+
+A straightforward way to solve this problem is to simply test all the hypotheses present in the
+context against the registered rules.
+
+A few obvious improvements are then available. We can index the rules in terms of one of its
+hypothesis. This index can then be used to guarentee that at least one hypothesis in the context will
+match with a slot of the rule. A heuristic is to take the last slot as it should be more specific
+than the firsts slot of a rule.
+
+If the last slot of a rule matches a hypothesis in the context we can then test
+whether other hypotheses of the context can complete that rule.
+This is done in two steps, first we check whether the hypotheses in the context match the slot.
+If it does, we check if the substitution given by the unification is coherent with
+the preceding matched hypotheses.
+Using the same reasoning as above, we heuristically do this backwards in the slots.
+
+This particular design is fast when the final hypothesis of a rule is not present in the context. 
+However, it otherwise needs to test many potential hypothesis candidate which is quite slow.
+Another weakness of this approach is that in the eventuality that a new hypothesis appears in the
+goal, for example as the result of another rule being completed, all previous work done on one rule
+must be repeated with this new hypothesis.
+
+We call stateless this kind of approach to the problem as no previous work is saved.
+
+\subsection{An improved, stateful design}
+
+With the weaknesses of the last approach in mind, we now consider a
+\textit{stateful} solution to the problem of finding complete matches.
+The main idea is to save partial progress on complete matches.
+
+This partial progress could be simply saved as trees with root the matched
+hypothesis returned by the index.
+
+\paragraph{Indexing Scheme.}
+However, we decided to move towards an index that looks at all the hypotheses of 
+the rule.
+The idea is that we can better track and save information about
+which hypotheses will potentially complete which rules.
+This makes it particularly easy to maintain partial matches which, as we will see,
+are classified in terms of the instantiation of their set of shared variables.
+Another advantage is that we only need to look at each of the context's
+hypotheses once.
+This makes the procedure incremental, which is part of the stateful philosophy.
+
+In other words, the indexing is now a discrimination tree that maps 
+types $T$ to sets of pairs $(r,i)$ where $r : (x_1 : T_1) \dots (x_n : T_n) \to \Psi$
+is a rule, $i$ is a slot of $r$ and $T$ is likely unifies with $T_i$.
+
+Given hypotheses returned by this index, the next challenge comes
+down to finding a complete match of a rule.
+This means identifying compatible hypotheses within the hypotheses returned by the index.
+
+\paragraph{Matches.}
+Compatible hypotheses are hypotheses such that the instantiation of their
+shared variables are the same.
+For example, the hypotheses $h_1 : a \leq b$ and $h_2 : b \leq c$ are compatible
+because they are instantiations of $x \leq y$ and $y \leq z$ where the
+instantiation of their shared variable $\{y\}$, is the same $\{b\}$.
+A complete match is a set of hypotheses such that each of these hypothesis corresponds
+to exactly one slot and they are all compatible.
+We proceed in the following way to find a complete match.
+
+First, we order the slots such that each slot has at least one variable in common
+with the previous slots. This can be proven to always be possible. 
+\xcom{We can add the proof if we want to / have space.}
+
+We then aim to construct matches iteratively : 
+We consider each hypotheses one by one, assign them to their
+respective slots and check whether there exist compatible matches
+which they can extend.
+The comptibility check means that all the shared variables of the considered 
+hypotheses need to be compatible.
+It would be costly to do this comparison everytime and indeed
+we choose instead to maintain what we call \textit{variable maps} whose purpose
+is to save the current partial matches.
+Each variable map is associated to one shared variable, these are the
+variables dictate compatibility.
+
+\paragraph{Variable Maps.}
+Let $x$ ba a shared variable, $t$ an expression that represents an
+instantiation of that variable and $i$ an integer associated to a slot.
+We define two variable maps $\mu_x$ and $\eta_x$ that respectively take the 
+pair $(i,t)$ to $\mathbb{M}$ and $\mathbb{H}$ where 
+\begin{itemize}
+    \item $\mathbb{M}$ is the set that contains exactly those
+    partial matches which already contain assignments for all
+    slots up to $i$ and which instantiate $?x$ with $t$.
+    \item $\mathbb{H}$ is the set of hypotheses which match $T_i$ while
+    instantiating $?x$ with $t$.
+\end{itemize}
+
+Given a hypothesis $h : T$ that matches $T_i$ and $\sigma_h$,
+the map sending the variable of $T_i$ to their specific instantiation for $h$.
+The compatibility check that looks for suitable partial matches is simply computed by
+$$\bigcap_{s(i)} \mu_x \left(i - 1, \sigma_h(x)\right).$$
+
+Similarly, given a match $m$ and $\sigma_m$ the substitution map of the relevant 
+variable in $m$ then we can check for compatible hypotheses that would extend $m$
+by computing 
+$$\bigcap_{s(lvl(M) + 1)} \eta_x \left(i + 1, \sigma_m(x)\right).$$
+
+\paragraph{Example 1.}
+\xcom{WIP}
+To demonstrate the typical behavior of the algorithm we consider the
+following example which has \textbf{Lemma 1} as a rule and the hypotheses
+$(h_1 : 300 \leq 301)$ and $(h_2 : 301 \leq 302)$. \xcom{format this.}
+We will consider the hypothesis $h_2$ first for demonstration purposes.
+
+At the beginning, for the rule \textit{le\_trans}, we notice that the
+only shared variable is ${y}$. Now the hypothesis $(h_2)$ unifies both
+with slot 1 $x \leq y$ and slot 2 $y \leq z$.
+This updates the map $\eta_y(y \mapsto 302, 1)$ with $\{h_2\}$ and 
+$\eta_y(y \mapsto 301, 2)$ with $\{h_2\}$
+
+\begin{tikzpicture}[outer sep=auto, level distance=10mm, scale=0.50]
+    \node[xshift = 0] {$\eta_y$}
+      child[xshift = -20] {node {1}
+          child {node {$302$}
+            child {node {$\{h_2\}$}}}
+          child {node {$301$}
+            child {node {$\{h_1\}$}}}}
+      child[xshift = 20] {node {2}
+          child {node {$301$}
+            child {node {$\{h_2\}$}}}
+          child {node {$300$}
+            child {node {$\{h_1\}$}}}};
+    \node[xshift = 90] {$\eta_y$}
+        child[xshift = -20] {node {1}
+            child {node {$302$}
+                child {node {$\{h_2\}$}}}
+            child {node {$301$}
+                child {node {$\{h_1\}$}}}}
+        child[xshift = 20] {node {2}
+            child {node {$301$}
+                child {node {$\{h_2\}$}}}
+            child {node {$300$}
+                child {node {$\{h_1\}$}}}};
+    \node[xshift=200] {$\mu_y$}
+      child {node[xshift=-20] {1}
+          child {node {$302$}
+            child {node {$\{\{1 \mapsto h_2\}\}$}}}
+          child {node {$301$}
+            child {node {$\{\{1 \mapsto h_1\}\}$}}}}
+      child {node[xshift=20] {2}
+          child {node {$301$}
+            child {node[yshift=-2em] {$\{\{1 \mapsto h_1, 2 \mapsto h_2\}\}$}}}};
+  \end{tikzpicture}
+
+\paragraph{Example 2.}
+We now consider another example with more than one shared variable.
+Let $T_1, T_2$ be two proposition that both depend on the variable $x$ and $y$.
+For simplicity, the rule considred will be $T_1(x,y) \to T_2(x,y) \to R$.
+The hypotheses added to the context will be, in order,
+$h_1 : T_2(a,a)$, $h_2 : T_1(a,b)$ and $h_3 : T_2(a,b)$.
+
+\begin{tikzpicture}[outer sep=auto, level distance=10mm, scale=0.50]
+    \node[xshift = 0] {$\eta_x$}
+      child[xshift = -20] {node {1}
+          child {node {$a$}
+            child {node {$\{ \}$}}}
+          child {node {$b$}
+            child {node {$\{ \}$}}}}
+      child[xshift = 20] {node {2}
+          child {node {$a$}
+            child {node {$\{h_1\}$}}}
+          child {node {$b$}
+            child {node {$\{ \}$}}}};
+    \node[xshift = 90] {$\eta_y$}
+        child[xshift = -20] {node {1}
+            child {node {$a$}
+                child {node {$\{ \}$}}}
+            child {node {$b$}
+                child {node {$\{ \}$}}}}
+        child[xshift = 20] {node {2}
+            child {node {$a$}
+                child {node {$\{h_1\}$}}}
+            child {node {$b$}
+                child {node {$\{ \}$}}}};
+  \end{tikzpicture}
+
+  At this point, there are no matches so there is nothing to extend.
+  We add the next hypothesis in the queue to the variable maps.
+
+  \begin{tikzpicture}[outer sep=auto, level distance=10mm, scale=0.50]
+    \node[xshift = 0] {$\eta_x$}
+      child[xshift = -20] {node {1}
+          child {node {$a$}
+            child {node {$\{h_2\}$}}}
+          child {node {$b$}
+            child {node {$\{ \}$}}}}
+      child[xshift = 20] {node {2}
+          child {node {$a$}
+            child {node {$\{h_1\}$}}}
+          child {node {$b$}
+            child {node {$\{ \}$}}}};
+    \node[xshift = 90] {$\eta_y$}
+        child[xshift = -20] {node {1}
+            child {node {$a$}
+                child {node {$\{ \}$}}}
+            child {node {$b$}
+                child {node {$\{ h_2\}$}}}}
+        child[xshift = 20] {node {2}
+            child {node {$a$}
+                child {node {$\{h_1\}$}}}
+            child {node {$b$}
+                child {node {$\{ \}$}}}};
+  \end{tikzpicture}
+
+  As this hypothesis is in the first slot, we create the corresponding level 1 partial
+  match $m_0 = {(h_2 : T_1(a, b))}$ for which we check for suitable hypotheses that could
+  extend the match.
+  Note that  $s(lvl(m_0) + 1) = s(2)$ which is the intersection of the variables
+  appearing in the slots $i < 2$ with the variables appearing in slot $2$.
+  Thus $s(2) = \{a,b\}$ and we compute :
+  $$\bigcap_{x\in s(lvl(m_0) + 1)} \eta_x \left(i + 1, \sigma_{m_0}(x)\right)
+    = \eta_x(2, a) \cap \eta_y(2, b) = \emptyset $$
+  Since the set is empty, the match can not be currently extended.
+  We turn to the remaining hypothesis in the queue and add it to the maps.
+
+  \begin{tikzpicture}[outer sep=auto, level distance=10mm, scale=0.50]
+    \node[xshift = 0] {$\eta_x$}
+      child[xshift = -20] {node {1}
+          child {node {$a$}
+            child {node {$\{h_2\}$}}}
+          child {node {$b$}
+            child {node {$\{ \}$}}}}
+      child[xshift = 20] {node {2}
+          child {node {$a$}
+            child {node {$\{h_1, h_3\}$}}}
+          child {node {$b$}
+            child {node {$\{ \}$}}}};
+    \node[xshift = 90] {$\eta_y$}
+        child[xshift = -20] {node {1}
+            child {node {$a$}
+                child {node {$\{ \}$}}}
+            child {node {$b$}
+                child {node {$\{h_2\}$}}}}
+        child[xshift = 20] {node {2}
+            child {node {$a$}
+                child {node {$\{h_1\}$}}}
+            child {node {$b$}
+                child {node {$\{h_3\}$}}}};
+  \end{tikzpicture}
+
+  We are now ready to look for potential matches that could be extended by $h_3$.
+  In this particuliar case, since $h_3$ is in slot $2$ we look for $s(2) = \{a,b\}$.
+  Then,
+  $$\bigcap_{x\in s(2)} \mu_x \left(i - 1, \sigma_h(x)\right)
+    = \mu_x(1, a) \cap \mu_y(1, b) = \{m_0\} $$.
+
+  We have found two compatible hypotheses for our rule and we can then add the 
+  result $R$ as hypothesis to the context.
+
+\section{Implementation}
+
+\xcom{Here are some notes about the implementation.}
+The index used to match hypotheses with their matching rules are implemented as
+discrimination trees. A discrimination tree is a trie-like data
+structure that maps expressions to arbitrary data.
+In \cite{}, it is shown to be good solution for the problem of finding
+terms (here the types of hypotheses) that might unify with other terms
+(here the types of the input hypotheses).
+In Lean 4, discrimination trees are also used to index typeclass
+instances and simplifier lemmas.
+
+\xcom{Incrementality?}
+In lean, the type of hypotheses can sometimes change while the hypothesis'
+identifier remains the same. This means that one needs to track the types
+of the hypotheses as well. We call this tracking \textit{goal diffs}, they
+highlight the differences between old and updated goal states.
+The reference to hypotheses which changing types get deleted from the maps
+and the slot. We then add them again as a new hypothesis.
+
+The maps described in the last section are all implemented as persistent
+hash maps \xcom{Why did we not use discrimination trees? $+$
+(pers. important cause we store related forward state for many goal.) }.
+
+One of the more costly operation that we need to run multiple times is the
+unification \xcom{defeq?} between the hypotheses and the types of the
+input hypotheses associated to rules.
+We implement a variant of Lean's unification process as follows :
+The idea is that we wish to normalize the types as much as possible
+
+\xcom{precompilation}
+
+\section{Evaluation}
+
+Analysing the both algorithms of both the stateless and stateful approach, we 
+can draw the following conclusions about their relative performance :
+
+\begin{itemize}
+    \item The stateful algorithm maybe be expected to perform poorly when the
+    context contains many hypotheses that almost complete rules.
+    \item The stateless algorithm should perform decently well on average and 
+    especially when only the last premises are absent; so the rules get never
+    selected by the stateless index
+\end{itemize}
+
+\subsection{Results}
+\xcom{TODO: Add a comment on the effect it had on Mathlib.}
+
+We present here a number of results obtained by running synthetic tests.
+By synthetic we mean that these tests we're constructed solely for the
+purpose of demonstrating the different complexity behavior of both
+algorithms by scaling some parameter. (e.g. the number of rules.)
+The results of the tests are an average of multiple runs.
+
+In \ref{fig:erase}, the step is set up as follows: 
+Given some rule $r$ with $6$ \xcom{motivate 6} premises, 
+we run the procedure in a context with 5 of the 6 premises present.
+The numbers on the $x$ axis represent which hypothesis is missing
+in the order of consideration by the algorithm.
+In a sense, this is how \textit{deep} the algorithm investigate before
+it stops.
+At $0$, the rule are not selected by the index so the procedure stops.
+The stateless algorithm gets progressively slower the further the removed
+premises is located.
+
+The position of the missing premise has little effect on the stateful
+algorithm as the index selects the rule in all case; it only needs one
+of the premise to be present.
+The work the stateful algorithm has to do is thus close to constant
+in this scenario.
+
+\begin{figure}
+% This is the erase without precompilation for 6 premises.
+% Note I have a decent idea that would allow us to compare for multiple
+% number of premises. 
+% (Divide by the average of the stateful time to normalize.)
+\begin{tikzpicture}
+    \begin{axis}[
+        title={Depth Test},
+        xlabel={Depth},
+        ylabel={Time in ms},
+        xmin=0, xmax=6,
+        ymin=50, ymax=65,
+        xtick={0,1,2,3,4,5,6},
+        ytick={50,55,60,65},
+        legend pos=north west,
+        ymajorgrids=true,
+        grid style=dashed,
+    ]
+
+    \addplot[
+        color=orange,
+        mark=square,
+        ]
+        coordinates {
+        (1,50.91)(2,53.14)(3,52.91)(4,51.22)(5,55.14)(6,56.60)
+        };
+        \addlegendentry{Stateless}
+
+    \addplot[
+        color=blue,
+        mark=square,
+        ]
+        coordinates {
+        (1,56.04)(2,56.43)(3,55.58)(4,55.26)(5,54.80)(6,60.61)
+        };
+        \addlegendentry{Stateful}
+
+    \end{axis}
+    \end{tikzpicture}
+\end{figure}
+
+\begin{figure}
+    % This is the erase without precompilation for 6 premises.
+    % Note I have a decent idea that would allow us to compare for multiple
+    % number of premises. 
+    % (Divide by the average of the stateful time to normalize.)
+    \begin{tikzpicture}
+        \begin{axis}[
+            title={Scaled Depth Test},
+            xlabel={Depth},
+            ylabel={Time in ms},
+            xmin=0, xmax=8,
+            ymin=0.5, ymax=2,
+            xtick={0,1,2,3,4,5,6,7,8},
+            ytick={0.75,1,1.25,1.5,1.75},
+            legend pos=north west,
+            ymajorgrids=true,
+            grid style=dashed,
+        ]
+
+        \addplot[color=orange,mark=square,]
+            coordinates {(0, 0.883600)(1, 1.039751)};
+            \addlegendentry{Stateless}
+        \addplot[color=orange,mark=square,]
+            coordinates {(0, 0.804521)(1, 1.017893)(2, 1.022493)};
+        \addplot[color=orange,mark=square,]
+            coordinates {(0, 0.791528)(1, 1.015467)(2, 1.076267)(3, 1.159999)};
+        \addplot[color=orange,mark=square,]
+            coordinates {(0, 0.743256)(1, 0.966436)(2, 1.058876)(3, 1.133514)(4, 1.225561)};
+        \addplot[color=orange,mark=square,]
+            coordinates {(0, 0.770625)(1, 1.003777)(2, 1.109929)(3, 1.215171)(4, 1.305373)(5, 1.390667)};
+        \addplot[color=orange,mark=square,]
+            coordinates {(0, 0.759093)(1, 1.016667)(2, 1.148273)(3, 1.213138)(4, 1.342795)(5, 1.442271)(6, 1.514877)};
+        \addplot[color=orange,mark=square,]
+            coordinates {(0, 0.709965)(1, 0.971294)(2, 1.068745)(3, 1.179333)(4, 1.312798)(5, 1.424054)(6, 1.551690)(7, 1.628344)};
+        \addplot[color=orange,mark=square,]
+            coordinates {(0, 0.728860)(1, 1.001576)(2, 1.097472)(3, 1.219259)(4, 1.366683)(5, 1.444844)(6, 1.576972)(7, 1.668693)(8, 1.815399)};
+
+        \end{axis}
+        \end{tikzpicture}
+    \end{figure}
+
+\begin{figure}
+    \label{fig:erase}
+%erase with precompilation
+    \begin{tikzpicture}
+        \begin{axis}[
+            title={Depth Test},
+            xlabel={Depth},
+            ylabel={Time in ms},
+            xmin=0, xmax=6,
+            ymin=3, ymax=7,
+            xtick={0,1,2,3,4,5,6},
+            ytick={3,4,5,6,7},
+            legend pos=north west,
+            ymajorgrids=true,
+            grid style=dashed,
+        ]
+
+        \addplot[
+            color=blue,
+            mark=square,
+            ]
+            coordinates {
+            (1, 3.715813)(2, 3.893547)(3, 3.856956)(4, 3.722684)(5, 3.714804)(6, 3.900508)
+            };
+            \addlegendentry{Stateful}
+
+        \addplot[
+            color=orange,
+            mark=square,
+            ]
+            coordinates {
+            (1, 2.993757)(2, 3.585232)(3, 3.880661)(4, 4.098640)(5, 4.322570)(6, 5.765420)
+            };
+            \addlegendentry{Stateless}
+
+        \end{axis}
+        \end{tikzpicture}
+    \end{figure}
+
+    \begin{figure}
+        %Indep with precompilation
+            \begin{tikzpicture}
+                \begin{axis}[
+                    title={Independance Test},
+                    xlabel={Number of Rules},
+                    ylabel={Time in ms},
+                    xmin=0, xmax=32,
+                    ymin=0, ymax=1700,
+                    ymode=log,xmode=log,
+                    log basis x=2,
+                    log basis y=2,
+                    legend pos=north west,
+                    ymajorgrids=true,
+                    grid style=dashed,
+                ]
+
+                \addplot[
+                    color=blue,
+                    mark=square,
+                    ]
+                    coordinates {
+                    (1, 0.710963)(2, 1.266787)(4, 2.181111)(8, 4.220279)(16, 8.084744)(32, 15.683840)
+                    };
+                    \addlegendentry{Stateful}
+
+                \addplot[
+                    color=orange,
+                    mark=square,
+                    ]
+                    coordinates {
+                    (1, 0.872730)(2, 2.100940)(4, 7.245598)(8, 35.805959)(16, 222.283526)(32, 1600.978502)
+                    };
+                    \addlegendentry{Stateless}
+
+                \end{axis}
+                \end{tikzpicture}
+            \end{figure}
+
+            \begin{figure}
+                %Casacade with precompilation
+                    \begin{tikzpicture}
+                        \begin{axis}[
+                            title={Casacade Test},
+                            xlabel={Number of Rules},
+                            ylabel={Time in ms},
+                            xmin=0, xmax=32,
+                            ymin=0, ymax=850,
+                            ymode=log,xmode=log,
+                            log basis x=2,
+                            log basis y=2,
+                            legend pos=north west,
+                            ymajorgrids=true,
+                            grid style=dashed,
+                        ]
+
+                        \addplot[
+                            color=blue,
+                            mark=square,
+                            ]
+                            coordinates {
+                            (1, 0.393011)(2, 0.730952)(4, 1.327736)(8, 3.007961)(16, 10.173689)(32, 38.892027)
+                            };
+                            \addlegendentry{Stateful}
+
+                        \addplot[
+                            color=orange,
+                            mark=square,
+                            ]
+                            coordinates {
+                            (1, 0.464973)(2, 0.759516)(4, 1.865191)(8, 7.585588)(16, 64.374740)(32, 813.336527)
+                            };
+                            \addlegendentry{Stateless}
+
+                        \end{axis}
+                        \end{tikzpicture}
+                    \end{figure}
+
+\section{Related Work}
+\section{Conclusion}
+
 \section{First Section}

### 2025-01-31 paper.tex
Remove sample text

@@
 \section{Related Work}
+
+TODO
+
 \section{Conclusion}
 
-\section{First Section}
-\subsection{A Subsection Sample}
-Please note that the first paragraph of a section or subsection is
-not indented. The first paragraph that follows a table, figure,
-equation etc. does not need an indent, either.
+TODO
 
-Subsequent paragraphs, however, are indented.
-
-\subsubsection{Sample Heading (Third Level)} Only two levels of
-headings should be numbered. Lower level headings remain unnumbered;
-they are formatted as run-in headings.
-
-\paragraph{Sample Heading (Fourth Level)}
-The contribution should contain no more than four levels of
-headings. Table~\ref{tab1} gives a summary of all heading levels.
-
-\begin{table}
-\caption{Table captions should be placed above the
-tables.}\label{tab1}
-\begin{tabular}{|l|l|l|}
-\hline
-Heading level &  Example & Font size and style\\
-\hline
-Title (centered) &  {\Large\bfseries Lecture Notes} & 14 point, bold\\
-1st-level heading &  {\large\bfseries 1 Introduction} & 12 point, bold\\
-2nd-level heading & {\bfseries 2.1 Printing Area} & 10 point, bold\\
-3rd-level heading & {\bfseries Run-in Heading in Bold.} Text follows & 10 point, bold\\
-4th-level heading & {\itshape Lowest Level Heading.} Text follows & 10 point, italic\\
-\hline
-\end{tabular}
-\end{table}
-
-\noindent Displayed equations are centered and set on a separate
-line.
-\begin{equation}
-x + y = z
-\end{equation}
-Please try to avoid rasterized images for line-art diagrams and
-schemas. Whenever possible, use vector graphics instead.
-
-\begin{theorem}
-This is a sample theorem. The run-in heading is set in bold, while
-the following text appears in italics. Definitions, lemmas,
-propositions, and corollaries are styled the same way.
-\end{theorem}
-%
-% the environments 'definition', 'lemma', 'proposition', 'corollary',
-% 'remark', and 'example' are defined in the LLNCS documentclass as well.
-%
-\begin{proof}
-Proofs, examples, and remarks have the initial word in italics,
-while the following text appears in normal font.
-\end{proof}
-For citations of references, we prefer the use of square brackets
-and consecutive numbers. Citations using labels or the author/year
-convention are also acceptable. The following bibliography provides
-a sample reference list with entries for journal
-articles~\cite{ref_article1}, an LNCS chapter~\cite{ref_lncs1}, a
-book~\cite{ref_book1}, proceedings without editors~\cite{ref_proc1},
-and a homepage~\cite{ref_url1}. Multiple citations are grouped
-\cite{ref_article1,ref_lncs1,ref_book1},
-\cite{ref_article1,ref_book1,ref_proc1,ref_url1}.
-
-\begin{credits}
-\subsubsection{\ackname} A bold run-in heading in small font size at the end of the paper is
-used for general acknowledgments, for example: This study was funded
-by X (grant number Y).
-
-\subsubsection{\discintname}
-It is now necessary to declare any competing interests or to specifically
-state that the authors have no competing interests. Please place the
-statement with a bold run-in heading in small font size beneath the
-(optional) acknowledgments\footnote{If EquinOCS, our proceedings submission
-system, is used, then the disclaimer can be provided directly in the system.},
-for example: The authors have no competing interests to declare that are
-relevant to the content of this article. Or: Author A has received research
-grants from Company W. Author B has received a speaker honorarium from
-Company X and owns stock in Company Y. Author C is a member of committee Z.
-\end{credits}
-%
-% ---- Bibliography ----
-%
-% BibTeX users should specify bibliography style 'splncs04'.
-% References will then be sorted and formatted in the correct style.
-%
-% \bibliographystyle{splncs04}
-% \bibliography{mybibliography}
-%
-\begin{thebibliography}{8}
-\bibitem{ref_article1}
-Author, F.: Article title. Journal \textbf{2}(5), 99--110 (2016)
-
-\bibitem{ref_lncs1}
-Author, F., Author, S.: Title of a proceedings paper. In: Editor,
-F., Editor, S. (eds.) CONFERENCE 2016, LNCS, vol. 9999, pp. 1--13.
-Springer, Heidelberg (2016). \doi{10.10007/1234567890}
-
-\bibitem{ref_book1}
-Author, F., Author, S., Author, T.: Book title. 2nd edn. Publisher,
-Location (1999)
-
-\bibitem{ref_proc1}
-Author, A.-B.: Contribution title. In: 9th International Proceedings
-on Proceedings, pp. 1--2. Publisher, Location (2010)
-
-\bibitem{ref_url1}
-LNCS Homepage, \url{http://www.springer.com/lncs}, last accessed 2023/10/25
-\end{thebibliography}
 \end{document}

### 2025-01-31 paper.tex
Fix author info
Fix Unicode setup
Use hyperref; remove some comments

@@
-% This is samplepaper.tex, a sample chapter demonstrating the
-% LLNCS macro package for Springer Computer Science proceedings;
-% Version 2.21 of 2022/01/12
-%
 \documentclass[runningheads]{llncs}
-%
+
 \usepackage[T1]{fontenc}
+\usepackage[utf8]{inputenc}
+\usepackage[british]{babel}
+\usepackage{newunicodechar}
 \usepackage{xcolor}
@@
 \usepackage{amsmath,amssymb}
-% T1 fonts will be used to generate the final print and online PDFs,
-% so please use T1 fonts in your manuscript whenever possible.
-% Other font encondings may result in incorrect characters.
-%
+\usepackage{mathtools}
 \usepackage{graphicx}
 \usepackage{pgfplots}
+\usepackage{hyperref}
+
+\renewcommand\UrlFont{\color{blue}\rmfamily}
+\urlstyle{rm}
+
+\newunicodechar{≔}{\ensuremath{\coloneq}}
 
@@
 \newcommand{\jcom}[1]{{\color{orange}{Jannis: #1}} }
-% Used for displaying a sample figure. If possible, figure files should
-% be included in EPS format.
-%
-% If you use the hyperref package, please uncomment the following two lines
-% to display URLs in blue roman font according to Springer's eBook style:
-%\usepackage{color}
-%\renewcommand\UrlFont{\color{blue}\rmfamily}
-%\urlstyle{rm}
-%
+
 \begin{document}
-%
+
 \title{Stateful Forward Reasoning}
-%
+
 %\titlerunning{Abbreviated paper title}
@@
 % an abbreviated paper title here
-%
-\author{First Author\inst{1}\orcidID{0000-1111-2222-3333} \and
-Second Author\inst{2,3}\orcidID{1111-2222-3333-4444} \and
-Third Author\inst{3}\orcidID{2222--3333-4444-5555}}
-%
-\authorrunning{F. Author et al.}
-% First names are abbreviated in the running head.
-% If there are more than two authors, 'et al.' is used.
-%
+
+\author{Jannis Limperg\inst{1}\orcidID{0000-0002-8861-5231} \and
+  Xavier Généreux\inst{1}\orcidID{TODO}}
+
+\authorrunning{J.~Limperg and X.~Généreux}
 \institute{University of Munich (LMU), Munich, Germany\\
-\email{\{xavier.genereux,jannis.limpberg\}@lmu.de}}
-%
-\maketitle              % typeset the header of the contribution
-%
+\email{\{xavier.genereux,jannis.limperg\}@lmu.de}}
+
+\maketitle
+
 \begin{abstract}
@@
 \end{abstract}
-%
-%
-%
 

### 2025-01-31 paper.tex
Remove \inst from authors
Add all the Unicode chars
Add related work

@@
 
+\newunicodechar{₁}{\ensuremath{_{1}}}
+\newunicodechar{₂}{\ensuremath{_{2}}}
+\newunicodechar{₃}{\ensuremath{_{3}}}
+\newunicodechar{₄}{\ensuremath{_{4}}}
+\newunicodechar{₅}{\ensuremath{_{5}}}
+\newunicodechar{₆}{\ensuremath{_{6}}}
+\newunicodechar{₇}{\ensuremath{_{7}}}
+\newunicodechar{₈}{\ensuremath{_{8}}}
+\newunicodechar{₉}{\ensuremath{_{9}}}
+\newunicodechar{ₙ}{\ensuremath{_{n}}}
+\newunicodechar{ₘ}{\ensuremath{_{m}}}
+\newunicodechar{ᵢ}{\ensuremath{_{i}}}
+\newunicodechar{ⱼ}{\ensuremath{_{j}}}
+\newunicodechar{ₗ}{\ensuremath{_{l}}}
+\newunicodechar{ₖ}{\ensuremath{_{k}}}
+\newunicodechar{α}{\ensuremath{\alpha}}
+\newunicodechar{σ}{\ensuremath{\sigma}}
+\newunicodechar{τ}{\ensuremath{\tau}}
+\newunicodechar{λ}{\ensuremath{\lambda}}
+\newunicodechar{Γ}{\ensuremath{\Gamma}}
+\newunicodechar{Δ}{\ensuremath{\Delta}}
+\newunicodechar{Φ}{\ensuremath{\Phi}}
+\newunicodechar{Ψ}{\ensuremath{\Psi}}
+\newunicodechar{Ω}{\ensuremath{\Omega}}
+\newunicodechar{∈}{\ensuremath{\in}}
+\newunicodechar{∉}{\ensuremath{\notin}}
+\newunicodechar{≡}{\ensuremath{\equiv}}
+\newunicodechar{≢}{\ensuremath{\nequiv}}
+\newunicodechar{≠}{\ensuremath{\neq}}
+\newunicodechar{≤}{\ensuremath{\leq}}
 \newunicodechar{≔}{\ensuremath{\coloneq}}
+\newunicodechar{∧}{\ensuremath{\land}}
+\newunicodechar{∨}{\ensuremath{\lor}}
+\newunicodechar{∪}{\ensuremath{\cup}}
+\newunicodechar{∩}{\ensuremath{\cap}}
+\newunicodechar{→}{\ensuremath{\rightarrow}}
+\newunicodechar{↔}{\ensuremath{\leftrightarrow}}
+\newunicodechar{↦}{\ensuremath{\mapsto}}
+\newunicodechar{⊢}{\ensuremath{\vdash}}
+\newunicodechar{∅}{\ensuremath{\emptyset}}
+\newunicodechar{∷}{\ensuremath{\mathbin{::}}}
+\newunicodechar{⇒}{\ensuremath{\Rightarrow}}
+\newunicodechar{ℕ}{\ensuremath{\mathbb{N}}}
+\newunicodechar{⟨}{\ensuremath{\langle}}
+\newunicodechar{⟩}{\ensuremath{\rangle}}
+\newunicodechar{⊤}{\ensuremath{\top}}
+\newunicodechar{⊥}{\ensuremath{\bot}}
 
@@
 
-\author{Jannis Limperg\inst{1}\orcidID{0000-0002-8861-5231} \and
-  Xavier Généreux\inst{1}\orcidID{TODO}}
+\author{Jannis Limperg\orcidID{0000-0002-8861-5231} \and
+  Xavier Généreux\orcidID{TODO}}
 
@@
 
-TODO
+\paragraph{ACL2}
+Among white-box proof automation tactics, only ACL2's \enquote{waterfall} (TODO ref) has special support for forward reasoning.
+(Some such tactics, e.g.\ Coq's \texttt{auto}, can execute arbitrary tactics including forward rules, but make no special provisions for them.)
+Like Aesop, ACL2 works on goals and saturates the local context of each goal, using user-provided forward rules, before using backward reasoning proof methods.
+It uses essentially Aesop's naive algorithm, i.e.\ the local context of each goal is saturated independently without reusing partial results from other goals.
+
+\paragraph{SATCHMO}
+The Prolog-based first-order theorem prover SATCHMO (TODO ref) and its many descendants (TODO refs) uses a tableau calculus (TODO ref) that can be viewed as combined forward and backward reasoning.
+Given a set of implications $A₁ ∧ \dots ∧ Aₙ → B₁ ∨ \dots ∨ Bₘ$ (with special cases $⊤ → B₁ ∨ \dots ∨ Bₘ$ and $A₁ ∧ \dots ∧ Aₙ → ⊥$), SATCHMO attempts to create a model that satisfies all implications.
+To that end, it matches the premises $A₁, \dots, Aₙ$ of each rule against known facts from Prolog's database, which is initially empty.
+If this is successful, SATCHMO first adds $B₁$ to the Prolog database and continues its search; if the continuation fails (i.e., if $⊥$ is derived), SATCHMO backtracks and tries $B₂$, etc.
+
+This search procedure can be viewed as a special case of our technique: the implications are used as forward rules and a single backward rule splits any goal with target $B₁ ∨ \dots ∨ Bₘ$ into $m$ goals with targets $B₁, \dots, Bₘ$.
+Compared to a Prolog-based implementation, our technique has the advantage that at each branching point we remember the partial matches induced by the currently known facts for each implication, rather than matching the implication against the database multiple times (subject to one of Prolog's various indexing schemes).
+Whether this gain outweighs the cost of storing and maintaining the partial match data structures likely depends on the given implications.
+Another advantage is that our technique is agnostic to the search strategy, whereas SATCHMO's Prolog implementation is based on backtracking and hence must use depth-first search.
+
+\paragraph{Knowledge-based reasoning systems}
+Combined forward and backward reasoning is also supported by some knowledge-based reasoning systems, e.g.\ Algernon (TODO ref) and EYE (TODO ref).
+However, to our knowledge none of these systems have backwards rules that can change the facts available to forward rules.
+Hence backward and forward reasoning do not interact and there is no need for our technique.
+Nevertheless, our proposal bears some resemblance to the RETE algorithm (TODO ref) for forward reasoning in large knowledge bases.
+
+\paragraph{Other related work}
+More distantly related are various calculi that combine backward reasoning with some form of incremental forward reasoning or simplification.
+Examples include the DPLL algorithm with its unit resolution rule (TODO ref) or incremental theory reasoners in SMT (TODO ref) and tableaux (TODO refs).
+Focusing (TODO ref) provides a proof-theoretic view on the interplay of forward and backward reasoning.
 

### 2025-01-31 paper.tex
Reformat; semantic linebreaks; some typos

@@
 \begin{abstract}
-The abstract should briefly summarize the contents of the paper in
-150--250 words.
+  TODO
 
@@
 
-An important contextualization is that the forward reasoning procedure in this paper takes place
-within aesop's routine procedure. We will focus on examples that highlight this style of proof
-while noting that in practice, a combination of both forwards and backwards reasoning are used 
+An important contextualization is that the forward reasoning procedure in this paper takes place within Aesop's routine procedure.
+We will focus on examples that highlight this style of proof while noting that in practice, a combination of both forwards and backwards reasoning are used.
+
 \section{Preliminaries}
@@
 
-In what follows, we write $\Gamma \models T$ for a goal where $\Gamma$ represents the context and
-$T$ the target. In a mathematical problem, $\Gamma$ is the list of hypotheses and variable from
-which we wish to show $T$.
+In what follows, we write $\Gamma \models T$ for a goal where $\Gamma$ represents the context and $T$ the target.
+In a mathematical problem, $\Gamma$ is the list of hypotheses and variable from which we wish to show $T$.
 
 \paragraph{Forward Rules.}
-Consider the following lemma :
+Consider the following lemma:
 \begin{lem}[\lean{le_trans}]
-  Given a preorder on some type $\alpha$ and elements $a,b,c$ of type $\alpha$, the following holds :
+  Given a preorder on some type $\alpha$ and elements $a,b,c$ of type $\alpha$, the following holds:
   $$a \leq b \rightarrow b \leq c \rightarrow a \leq c$$
@@
 
-From such lemmas, we will create what we call \textit{forward rules} that represent
-the theorems on which we would like to reason in a forward manner.
-The premises $\alpha$, $ \text{Preorder } \alpha$ , $a$, $b$, $c$, $a \leq b$ and
-$b \leq c$ are separated between \textit{implied hypotheses} $\text{Preorder } \alpha$,
+From such lemmas, we will create what we call \textit{forward rules} that represent the theorems on which we would like to reason in a forward manner.
+The premises $\alpha$, $ \text{Preorder } \alpha$ , $a$, $b$, $c$, $a \leq b$ and $b \leq c$ are separated between \textit{implied hypotheses} $\text{Preorder } \alpha$,
 $a$, $b$ and $c$ and \textit{maximal hypotheses} $a \leq b$ and $b \leq c$.
@@
 
-For the premises $(x_1 : T_1),(x_2 : T_2),\dots,(x_n : T_n)$,
-the input hypotheses consist of the hypotheses for which there are no
-$j > i$ such that $T_j$ depends on $x_i$.
+For the premises $(x_1 : T_1),(x_2 : T_2),\dots,(x_n : T_n)$, the input hypotheses consist of the hypotheses for which there are no $j > i$ such that $T_j$ depends on $x_i$.
 
 \paragraph{Rules.}
-We write $r : \Phi \to \Psi$ for a \textit{rule} $r$ with input
-hypotheses $\Phi$ and output hypothesis $\Psi$.
-For a given rule, we assign each maximal input hypothesis a unique index
-between $1$ and $n$.
-We call theses indices \textit{slots} of $r$ and define $\text{slots}(r)$ as
-the set of slots of $r$.
+We write $r : \Phi \to \Psi$ for a \textit{rule} $r$ with input hypotheses $\Phi$ and output hypothesis $\Psi$.
+For a given rule, we assign each maximal input hypothesis a unique index between $1$ and $n$.
+We call theses indices \textit{slots} of $r$ and define $\text{slots}(r)$ as the set of slots of $r$.
 
-Given a slot $i$ and hypothesis $h : T$ with $T \equiv_{[\sigma]} T_i$
-(i.e. $\sigma$ is the substitution that results from unifying $T$ and $T_i$), 
-we say that $h$ is \textit{suitable for slot $i$} and define $\text{sub}(h) ≔ \sigma$.
+Given a slot $i$ and hypothesis $h : T$ with $T \equiv_{[\sigma]} T_i$ (i.e.
+$\sigma$ is the substitution that results from unifying $T$ and $T_i$), we say that $h$ is \textit{suitable for slot $i$} and define $\text{sub}(h) ≔ \sigma$.
 
-A \textit{shared variable} of $r$ is an input hypothesis $x_i$ such that there are
-at least two input hypothesis types $T_j$, $T_k$ in which $x_i$ occurs.
+A \textit{shared variable} of $r$ is an input hypothesis $x_i$ such that there are at least two input hypothesis types $T_j$, $T_k$ in which $x_i$ occurs.
 A shared variable is necessarily a non-maximal input hypothesis.
-We also write $?x_i$ to emphasise that we interpret $x_i$ as a variable
-(and, in the code, as a metavariable).
-The set $s_r(i)$ of \textit{shared variables of a slot} $i$ of $r$ are those
-shared variables $?x$ that occur in $T_i$ and in at least one slot $j < i$.
+We also write $?x_i$ to emphasise that we interpret $x_i$ as a variable (and, in the code, as a metavariable).
+The set $s_r(i)$ of \textit{shared variables of a slot} $i$ of $r$ are those shared variables $?x$ that occur in $T_i$ and in at least one slot $j < i$.
 If the rule in question is clear from the context, we write $s(i)$.
@@
 
-A \textit{match} $M$ for $r$ is a partial map from slots of $r$ to hypotheses in
-the context.
+A \textit{match} $M$ for $r$ is a partial map from slots of $r$ to hypotheses in the context.
 It must satisfy the following conditions:
 \begin{itemize}
-\item For each slot $i$ of $r$, if $M(i)$ is defined, then
-    $M(i) \equiv_{[\sigma_i]} T_i$ with a substitution $\sigma_i$.
+  \item For each slot $i$ of $r$, if $M(i)$ is defined, then $M(i) \equiv_{[\sigma_i]} T_i$ with a substitution $\sigma_i$.
         The domain of this substitution is a subset of the non-maximal input hypotheses.
-\item For all slots $i$ and $j$ of $r$ such that $M(i)$ and $M(j)$ are defined,
-    $\sigma_i$ and $\sigma_j$ are compatible.
-\item $M$ is downward-closed: if $M(i)$ is defined, then for any slot $j < i$ of
-    $r$, $M(j)$ is also defined.
+  \item For all slots $i$ and $j$ of $r$ such that $M(i)$ and $M(j)$ are defined, $\sigma_i$ and $\sigma_j$ are compatible.
+  \item $M$ is downward-closed: if $M(i)$ is defined, then for any slot $j < i$ of $r$, $M(j)$ is also defined.
+\end{itemize}
 
-\end{itemize}
-A match of $r$ is \textit{complete} if $M(i)$ is defined for every slot $i$ of $r$, and
-\textit{incomplete} otherwise.
-The \textit{level of a match} $M$, $lvl(M)$, is the maximal $i$ such that $M(i)$ 
-is defined.
-The \textit{substitution of a match} $M$, $sub(M)$, is the union of the
-substitutions $\sigma_i$ for each $i$ such that $M_i$ is defined.
+A match of $r$ is \textit{complete} if $M(i)$ is defined for every slot $i$ of $r$, and \textit{incomplete} otherwise.
+The \textit{level of a match} $M$, $lvl(M)$, is the maximal $i$ such that $M(i)$ is defined.
+The \textit{substitution of a match} $M$, $sub(M)$, is the union of the substitutions $\sigma_i$ for each $i$ such that $M_i$ is defined.
 This union is defined since the $\sigma_i$ are pairwise compatible.
 This means that $sub(M)$ contains only one assignement per variable.
-The set of \textit{shared variables of a match} $M$ is $s(lvl(M))$ the
-shared variables of $lvl(M)$.
+The set of \textit{shared variables of a match} $M$ is $s(lvl(M))$ the shared variables of $lvl(M)$.
 
 An output hypothesis is \textit{redundant} if its type is already present in the context.
-A goal is saturated with respect to a set of forward rules if all
-output hypotheses returned by the procedure are redundant.
+A goal is saturated with respect to a set of forward rules if all output hypotheses returned by the procedure are redundant.
 
-The following section introduces an algorithm to construct complete matches $M$ 
-from the context $\Gamma$.
+The following section introduces an algorithm to construct complete matches $M$ from the context $\Gamma$.
 
@@
 
-A straightforward way to solve this problem is to simply test all the hypotheses present in the
-context against the registered rules.
+A straightforward way to solve this problem is to simply test all the hypotheses present in the context against the registered rules.
 
-A few obvious improvements are then available. We can index the rules in terms of one of its
-hypothesis. This index can then be used to guarentee that at least one hypothesis in the context will
-match with a slot of the rule. A heuristic is to take the last slot as it should be more specific
-than the firsts slot of a rule.
+A few obvious improvements are then available.
+We can index the rules in terms of one of its hypothesis.
+This index can then be used to guarentee that at least one hypothesis in the context will match with a slot of the rule.
+A heuristic is to take the last slot as it should be more specific than the firsts slot of a rule.
 
-If the last slot of a rule matches a hypothesis in the context we can then test
-whether other hypotheses of the context can complete that rule.
+If the last slot of a rule matches a hypothesis in the context we can then test whether other hypotheses of the context can complete that rule.
 This is done in two steps, first we check whether the hypotheses in the context match the slot.
-If it does, we check if the substitution given by the unification is coherent with
-the preceding matched hypotheses.
+If it does, we check if the substitution given by the unification is coherent with the preceding matched hypotheses.
 Using the same reasoning as above, we heuristically do this backwards in the slots.
@@
 However, it otherwise needs to test many potential hypothesis candidate which is quite slow.
-Another weakness of this approach is that in the eventuality that a new hypothesis appears in the
-goal, for example as the result of another rule being completed, all previous work done on one rule
-must be repeated with this new hypothesis.
+Another weakness of this approach is that in the eventuality that a new hypothesis appears in the goal, for example as the result of another rule being completed, all previous work done on one rule must be repeated with this new hypothesis.
 
@@
 
-With the weaknesses of the last approach in mind, we now consider a
-\textit{stateful} solution to the problem of finding complete matches.
+With the weaknesses of the last approach in mind, we now consider a \textit{stateful} solution to the problem of finding complete matches.
 The main idea is to save partial progress on complete matches.
 
-This partial progress could be simply saved as trees with root the matched
-hypothesis returned by the index.
+This partial progress could be simply saved as trees with root the matched hypothesis returned by the index.
 
 \paragraph{Indexing Scheme.}
-However, we decided to move towards an index that looks at all the hypotheses of 
-the rule.
-The idea is that we can better track and save information about
-which hypotheses will potentially complete which rules.
-This makes it particularly easy to maintain partial matches which, as we will see,
-are classified in terms of the instantiation of their set of shared variables.
-Another advantage is that we only need to look at each of the context's
-hypotheses once.
+However, we decided to move towards an index that looks at all the hypotheses of the rule.
+The idea is that we can better track and save information about which hypotheses will potentially complete which rules.
+This makes it particularly easy to maintain partial matches which, as we will see, are classified in terms of the instantiation of their set of shared variables.
+Another advantage is that we only need to look at each of the context's hypotheses once.
 This makes the procedure incremental, which is part of the stateful philosophy.
 
-In other words, the indexing is now a discrimination tree that maps 
-types $T$ to sets of pairs $(r,i)$ where $r : (x_1 : T_1) \dots (x_n : T_n) \to \Psi$
-is a rule, $i$ is a slot of $r$ and $T$ is likely unifies with $T_i$.
+In other words, the indexing is now a discrimination tree that maps types $T$ to sets of pairs $(r,i)$ where $r : (x_1 : T_1) \dots (x_n : T_n) \to \Psi$ is a rule, $i$ is a slot of $r$ and $T$ is likely unifies with $T_i$.
 
-Given hypotheses returned by this index, the next challenge comes
-down to finding a complete match of a rule.
+Given hypotheses returned by this index, the next challenge comes down to finding a complete match of a rule.
 This means identifying compatible hypotheses within the hypotheses returned by the index.
@@
 \paragraph{Matches.}
-Compatible hypotheses are hypotheses such that the instantiation of their
-shared variables are the same.
-For example, the hypotheses $h_1 : a \leq b$ and $h_2 : b \leq c$ are compatible
-because they are instantiations of $x \leq y$ and $y \leq z$ where the
-instantiation of their shared variable $\{y\}$, is the same $\{b\}$.
-A complete match is a set of hypotheses such that each of these hypothesis corresponds
-to exactly one slot and they are all compatible.
+Compatible hypotheses are hypotheses such that the instantiation of their shared variables are the same.
+For example, the hypotheses $h_1 : a \leq b$ and $h_2 : b \leq c$ are compatible because they are instantiations of $x \leq y$ and $y \leq z$ where the instantiation of their shared variable $\{y\}$, is the same $\{b\}$.
+A complete match is a set of hypotheses such that each of these hypothesis corresponds to exactly one slot and they are all compatible.
 We proceed in the following way to find a complete match.
 
-First, we order the slots such that each slot has at least one variable in common
-with the previous slots. This can be proven to always be possible. 
+First, we order the slots such that each slot has at least one variable in common with the previous slots.
+This can be proven to always be possible.
 \xcom{We can add the proof if we want to / have space.}
 
-We then aim to construct matches iteratively : 
-We consider each hypotheses one by one, assign them to their
-respective slots and check whether there exist compatible matches
-which they can extend.
-The comptibility check means that all the shared variables of the considered 
-hypotheses need to be compatible.
-It would be costly to do this comparison everytime and indeed
-we choose instead to maintain what we call \textit{variable maps} whose purpose
-is to save the current partial matches.
-Each variable map is associated to one shared variable, these are the
-variables dictate compatibility.
+We then aim to construct matches iteratively: we consider each hypotheses one by one, assign them to their respective slots and check whether there exist compatible matches which they can extend.
+The compatibility check means that all the shared variables of the considered hypotheses need to be compatible.
+It would be costly to do this comparison every time and indeed we choose instead to maintain what we call \textit{variable maps} whose purpose is to save the current partial matches.
+Each variable map is associated to one shared variable, these are the variables that dictate compatibility.
 
 \paragraph{Variable Maps.}
-Let $x$ ba a shared variable, $t$ an expression that represents an
-instantiation of that variable and $i$ an integer associated to a slot.
-We define two variable maps $\mu_x$ and $\eta_x$ that respectively take the 
-pair $(i,t)$ to $\mathbb{M}$ and $\mathbb{H}$ where 
+Let $x$ ba a shared variable, $t$ an expression that represents an instantiation of that variable and $i$ an integer associated to a slot.
+We define two variable maps $\mu_x$ and $\eta_x$ that respectively take the pair $(i,t)$ to $\mathbb{M}$ and $\mathbb{H}$ where
 \begin{itemize}
-    \item $\mathbb{M}$ is the set that contains exactly those
-    partial matches which already contain assignments for all
-    slots up to $i$ and which instantiate $?x$ with $t$.
-    \item $\mathbb{H}$ is the set of hypotheses which match $T_i$ while
-    instantiating $?x$ with $t$.
+  \item $\mathbb{M}$ is the set that contains exactly those partial matches which already contain assignments for all slots up to $i$ and which instantiate $?x$ with $t$.
+  \item $\mathbb{H}$ is the set of hypotheses which match $T_i$ while instantiating $?x$ with $t$.
 \end{itemize}
 
-Given a hypothesis $h : T$ that matches $T_i$ and $\sigma_h$,
-the map sending the variable of $T_i$ to their specific instantiation for $h$.
+Given a hypothesis $h : T$ that matches $T_i$ and $\sigma_h$, the map sending the variable of $T_i$ to their specific instantiation for $h$.
 The compatibility check that looks for suitable partial matches is simply computed by
-$$\bigcap_{s(i)} \mu_x \left(i - 1, \sigma_h(x)\right).$$
+\[
+  \bigcap_{s(i)} \mu_x \left(i - 1, \sigma_h(x)\right)
+\]
 
-Similarly, given a match $m$ and $\sigma_m$ the substitution map of the relevant 
-variable in $m$ then we can check for compatible hypotheses that would extend $m$
-by computing 
-$$\bigcap_{s(lvl(M) + 1)} \eta_x \left(i + 1, \sigma_m(x)\right).$$
+Similarly, given a match $m$ and $\sigma_m$ the substitution map of the relevant  variable in $m$ then we can check for compatible hypotheses that would extend $m$ by computing
+\[
+  \bigcap_{s(lvl(M) + 1)} \eta_x \left(i + 1, \sigma_m(x)\right)
+\]
 
@@
 \xcom{WIP}
-To demonstrate the typical behavior of the algorithm we consider the
-following example which has \textbf{Lemma 1} as a rule and the hypotheses
-$(h_1 : 300 \leq 301)$ and $(h_2 : 301 \leq 302)$. \xcom{format this.}
+To demonstrate the typical behavior of the algorithm we consider the following example which has \textbf{Lemma 1} as a rule and the hypotheses $(h_1 : 300 \leq 301)$ and $(h_2 : 301 \leq 302)$.
+\xcom{format this.}
 We will consider the hypothesis $h_2$ first for demonstration purposes.
 
-At the beginning, for the rule \textit{le\_trans}, we notice that the
-only shared variable is ${y}$. Now the hypothesis $(h_2)$ unifies both
-with slot 1 $x \leq y$ and slot 2 $y \leq z$.
-This updates the map $\eta_y(y \mapsto 302, 1)$ with $\{h_2\}$ and 
-$\eta_y(y \mapsto 301, 2)$ with $\{h_2\}$
+At the beginning, for the rule \textit{le\_trans}, we notice that the only shared variable is ${y}$.
+Now the hypothesis $(h_2)$ unifies both with slot 1 $x \leq y$ and slot 2 $y \leq z$.
+This updates the map $\eta_y(y \mapsto 302, 1)$ with $\{h_2\}$ and $\eta_y(y \mapsto 301, 2)$ with $\{h_2\}$.
 
@@
 Let $T_1, T_2$ be two proposition that both depend on the variable $x$ and $y$.
-For simplicity, the rule considred will be $T_1(x,y) \to T_2(x,y) \to R$.
-The hypotheses added to the context will be, in order,
-$h_1 : T_2(a,a)$, $h_2 : T_1(a,b)$ and $h_3 : T_2(a,b)$.
+For simplicity, the rule considered will be $T_1(x,y) \to T_2(x,y) \to R$.
+The hypotheses added to the context will be, in order, $h_1 : T_2(a,a)$, $h_2 : T_1(a,b)$ and $h_3 : T_2(a,b)$.
 
@@
 
-  As this hypothesis is in the first slot, we create the corresponding level 1 partial
-  match $m_0 = {(h_2 : T_1(a, b))}$ for which we check for suitable hypotheses that could
-  extend the match.
-  Note that  $s(lvl(m_0) + 1) = s(2)$ which is the intersection of the variables
-  appearing in the slots $i < 2$ with the variables appearing in slot $2$.
-  Thus $s(2) = \{a,b\}$ and we compute :
-  $$\bigcap_{x\in s(lvl(m_0) + 1)} \eta_x \left(i + 1, \sigma_{m_0}(x)\right)
-    = \eta_x(2, a) \cap \eta_y(2, b) = \emptyset $$
+As this hypothesis is in the first slot, we create the corresponding level 1 partial match $m_0 = {(h_2 : T_1(a, b))}$ for which we check for suitable hypotheses that could extend the match.
+Note that  $s(lvl(m_0) + 1) = s(2)$ which is the intersection of the variables appearing in the slots $i < 2$ with the variables appearing in slot $2$.
+Thus $s(2) = \{a,b\}$ and we compute:
+\[
+  \bigcap_{x\in s(lvl(m_0) + 1)} \eta_x \left(i + 1, \sigma_{m_0}(x)\right) = \eta_x(2, a) \cap \eta_y(2, b) = \emptyset
+\]
 Since the set is empty, the match can not be currently extended.
@@
 Then,
-  $$\bigcap_{x\in s(2)} \mu_x \left(i - 1, \sigma_h(x)\right)
-    = \mu_x(1, a) \cap \mu_y(1, b) = \{m_0\} $$.
+\[
+\bigcap_{x\in s(2)} \mu_x \left(i - 1, \sigma_h(x)\right) = \mu_x(1, a) \cap \mu_y(1, b) = \{m_0\}
+\]
 
-  We have found two compatible hypotheses for our rule and we can then add the 
-  result $R$ as hypothesis to the context.
+We have found two compatible hypotheses for our rule and we can then add the result $R$ as hypothesis to the context.
 
@@
 \xcom{Here are some notes about the implementation.}
-The index used to match hypotheses with their matching rules are implemented as
-discrimination trees. A discrimination tree is a trie-like data
-structure that maps expressions to arbitrary data.
-In \cite{}, it is shown to be good solution for the problem of finding
-terms (here the types of hypotheses) that might unify with other terms
-(here the types of the input hypotheses).
-In Lean 4, discrimination trees are also used to index typeclass
-instances and simplifier lemmas.
+The index used to match hypotheses with their matching rules are implemented as discrimination trees.
+A discrimination tree is a trie-like data structure that maps expressions to arbitrary data.
+In~\cite{TODO}, it is shown to be good solution for the problem of finding terms (here the types of hypotheses) that might unify with other terms (here the types of the input hypotheses).
+In Lean 4, discrimination trees are also used to index typeclass instances and simplifier lemmas.
 
 \xcom{Incrementality?}
-In lean, the type of hypotheses can sometimes change while the hypothesis'
-identifier remains the same. This means that one needs to track the types
-of the hypotheses as well. We call this tracking \textit{goal diffs}, they
-highlight the differences between old and updated goal states.
-The reference to hypotheses which changing types get deleted from the maps
-and the slot. We then add them again as a new hypothesis.
+In Lean, the type of hypotheses can sometimes change while the hypothesis' identifier remains the same.
+This means that one needs to track the types of the hypotheses as well.
+We call this tracking \textit{goal diffs}, they highlight the differences between old and updated goal states.
+The reference to hypotheses which changing types get deleted from the maps and the slot.
+We then add them again as a new hypothesis.
 
-The maps described in the last section are all implemented as persistent
-hash maps \xcom{Why did we not use discrimination trees? $+$
-(pers. important cause we store related forward state for many goal.) }.
+The maps described in the last section are all implemented as persistent hash maps.
+\xcom{Why did we not use discrimination trees? $+$ (pers. important cause we store related forward state for many goal.)}
 
-One of the more costly operation that we need to run multiple times is the
-unification \xcom{defeq?} between the hypotheses and the types of the
-input hypotheses associated to rules.
-We implement a variant of Lean's unification process as follows :
-The idea is that we wish to normalize the types as much as possible
+One of the more costly operation that we need to run multiple times is the unification \xcom{defeq?} between the hypotheses and the types of the input hypotheses associated to rules.
+We implement a variant of Lean's unification process as follows: the idea is that we wish to normalize the types as much as possible.
 
@@
 
-Analysing the both algorithms of both the stateless and stateful approach, we 
-can draw the following conclusions about their relative performance :
+Comparing the stateless and stateful approach, we can draw the following conclusions about their relative performance:
 
 \begin{itemize}
-    \item The stateful algorithm maybe be expected to perform poorly when the
-    context contains many hypotheses that almost complete rules.
-    \item The stateless algorithm should perform decently well on average and 
-    especially when only the last premises are absent; so the rules get never
-    selected by the stateless index
+    \item The stateful algorithm maybe be expected to perform poorly when the context contains many hypotheses that almost complete rules.
+    \item The stateless algorithm should perform decently well on average and especially when only the last premises are absent, so the rules get never selected by the stateless index.
 \end{itemize}
@@
 We present here a number of results obtained by running synthetic tests.
-By synthetic we mean that these tests we're constructed solely for the
-purpose of demonstrating the different complexity behavior of both
-algorithms by scaling some parameter. (e.g. the number of rules.)
+By synthetic we mean that these tests we're constructed solely for the purpose of demonstrating the different complexity behavior of both algorithms by scaling some parameter (e.g.\ the number of rules).
 The results of the tests are an average of multiple runs.
 
-In \ref{fig:erase}, the step is set up as follows: 
-Given some rule $r$ with $6$ \xcom{motivate 6} premises, 
-we run the procedure in a context with 5 of the 6 premises present.
-The numbers on the $x$ axis represent which hypothesis is missing
-in the order of consideration by the algorithm.
-In a sense, this is how \textit{deep} the algorithm investigate before
-it stops.
+In Fig.~\ref{fig:erase}, the step is set up as follows: given some rule $r$ with $6$ \xcom{motivate 6} premises, we run the procedure in a context with 5 of the 6 premises present.
+The numbers on the $x$ axis represent which hypothesis is missing in the order of consideration by the algorithm.
+In a sense, this is how \textit{deep} the algorithm investigate before it stops.
 At $0$, the rule are not selected by the index so the procedure stops.
-The stateless algorithm gets progressively slower the further the removed
-premises is located.
+The stateless algorithm gets progressively slower the further the removed premises is located.
 
-The position of the missing premise has little effect on the stateful
-algorithm as the index selects the rule in all case; it only needs one
-of the premise to be present.
-The work the stateful algorithm has to do is thus close to constant
-in this scenario.
+The position of the missing premise has little effect on the stateful algorithm as the index selects the rule in all case; it only needs one of the premise to be present.
+The work the stateful algorithm has to do is thus close to constant in this scenario.
 

### 2025-02-03 paper.tex
examples 1 and 2

@@
 From such lemmas, we will create what we call \textit{forward rules} that represent the theorems on which we would like to reason in a forward manner.
-The premises $\alpha$, $ \text{Preorder } \alpha$ , $a$, $b$, $c$, $a \leq b$ and $b \leq c$ are separated between \textit{implied hypotheses} $\text{Preorder } \alpha$,
-$a$, $b$ and $c$ and \textit{maximal hypotheses} $a \leq b$ and $b \leq c$.
-Maximal hypotheses are also called \textit{input hypotheses}.
+The premises $\alpha$, $ \text{Preorder } \alpha$ , $a$, $b$, $c$, $a \leq b$ and $b \leq c$ are separated between \textit{implied premises} $\text{Preorder } \alpha$,
+$a$, $b$ and $c$ and \textit{maximal premises} $a \leq b$ and $b \leq c$.
+Maximal premises are also called \textit{input premises}.
 
-For the premises $(x_1 : T_1),(x_2 : T_2),\dots,(x_n : T_n)$, the input hypotheses consist of the hypotheses for which there are no $j > i$ such that $T_j$ depends on $x_i$.
+For the premises $(x_1 : T_1),(x_2 : T_2),\dots,(x_n : T_n)$, the input premises consist of the premises for which there are no $j > i$ such that $T_j$ depends on $x_i$.
 
 \paragraph{Rules.}
-We write $r : \Phi \to \Psi$ for a \textit{rule} $r$ with input hypotheses $\Phi$ and output hypothesis $\Psi$.
-For a given rule, we assign each maximal input hypothesis a unique index between $1$ and $n$.
+We write $r : \Phi \to \Psi$ for a \textit{rule} $r$ with input premises $\Phi$ and output $\Psi$.
+For a given rule, we assign each maximal input premise a unique index between $1$ and $n$.
 We call theses indices \textit{slots} of $r$ and define $\text{slots}(r)$ as the set of slots of $r$.
 
-Given a slot $i$ and hypothesis $h : T$ with $T \equiv_{[\sigma]} T_i$ (i.e.
+Given a slot $i$ and hypothesis $h : T$ in the context $\Gamma$ with $T \equiv_{[\sigma]} T_i$ (i.e.
 $\sigma$ is the substitution that results from unifying $T$ and $T_i$), we say that $h$ is \textit{suitable for slot $i$} and define $\text{sub}(h) ≔ \sigma$.
 
-A \textit{shared variable} of $r$ is an input hypothesis $x_i$ such that there are at least two input hypothesis types $T_j$, $T_k$ in which $x_i$ occurs.
-A shared variable is necessarily a non-maximal input hypothesis.
+A \textit{shared variable} of $r$ is an input premise $x_i$ such that there are at least two input premise types $T_j$, $T_k$ in which $x_i$ occurs.
+A shared variable is necessarily a non-maximal input premise.
 We also write $?x_i$ to emphasise that we interpret $x_i$ as a variable (and, in the code, as a metavariable).
@@
 %The result is also called the \textit{output hypothesis}.
-%If we find matching hypotheses for all maximal input hypotheses in the local context,
-%we can apply the forward rule, since each non-maximal input hypothesis is uniquely determined
+%If we find matching hypotheses for all maximal input premises in the local context,
+%we can apply the forward rule, since each non-maximal input premise is uniquely determined
 %by maximal hypotheses.
@@
   \item For each slot $i$ of $r$, if $M(i)$ is defined, then $M(i) \equiv_{[\sigma_i]} T_i$ with a substitution $\sigma_i$.
-        The domain of this substitution is a subset of the non-maximal input hypotheses.
+        The domain of this substitution is a subset of the non-maximal input premises.
   \item For all slots $i$ and $j$ of $r$ such that $M(i)$ and $M(j)$ are defined, $\sigma_i$ and $\sigma_j$ are compatible.
@@
 A complete match is a set of hypotheses such that each of these hypothesis corresponds to exactly one slot and they are all compatible.
-We proceed in the following way to find a complete match.
 
+Given a rule, we will partition its premises into subsets for which premises share a dependency.
+To be precise, consider the graph for which the input premises $T_i$ of $r$ are the vertices and whose edges are precisely the pairs of vertices $(T_i, T_j)$ for which $v(T_i) \cap v(T_j) \neq \emptyset$.
+The subsets of premises of $r$ that share a dependency are precisely the connected components of this graph.
+We call these subsets of premises \textit{cluters}.
+
+%I think, if we have space, we could also add the following characterization :
+%Another characterization is the partition with the property that for two parts $C_1$ and $C_2$, $\bigcup_{T_i \in C_1} v(T_i) \cap \bigcup_{T_i \in C_2} v(T_i) = \emptyset$, and for which there do not exists a partition with the same property that has more parts.
+
+One particularity of clusters is that, for two clusters $C_1$ and $C_2$,
+$\bigcup_{T_i \in C_1} v(T_i) \cap \bigcup_{T_i \in C_2} v(T_i) = \emptyset$.
+This means that compatibility need only be considered within clusters as there are no shared variables between distinct clusters.
+In other words, premises in a cluster are independent of the premises in the other clusters.
+Consequently, in what follows, we will assume without loss of generality that there is only one cluster.
+The process we follow would simply be repeated on each cluster.
+
+We proceed in the following way to find a complete match:
 First, we order the slots such that each slot has at least one variable in common with the previous slots.
@@
 \paragraph{Example 1.}
-\xcom{WIP}
-To demonstrate the typical behavior of the algorithm we consider the following example which has \textbf{Lemma 1} as a rule and the hypotheses $(h_1 : 300 \leq 301)$ and $(h_2 : 301 \leq 302)$.
-\xcom{format this.}
-We will consider the hypothesis $h_2$ first for demonstration purposes.
+\xcom{This is quite verbose, maybe I should use a more systematic approach.}
+To demonstrate the typical behavior of the algorithm we consider the following example which has \textbf{Lemma 1} as a rule and for which we add, in order, the hypotheses $h_1 : 301 \leq 302$ and $h_2 : 300 \leq 301$.
+Note that this order is choosed only for demonstration purposes.
 
 At the beginning, for the rule \textit{le\_trans}, we notice that the only shared variable is ${y}$.
-Now the hypothesis $(h_2)$ unifies both with slot 1 $x \leq y$ and slot 2 $y \leq z$.
-This updates the map $\eta_y(y \mapsto 302, 1)$ with $\{h_2\}$ and $\eta_y(y \mapsto 301, 2)$ with $\{h_2\}$.
+Now the hypothesis $h_1 : 301 \leq 302$ unifies both with slot 1: $x \leq y$ and slot 2: $y \leq z$.
+This updates the map $\eta_y(1, y \mapsto 302)$ with $\{h_1\}$.
+An update in slot 1 also marks the begining of a new partial match and we update de match map $\mu_y$ acordingly.
+Note that after the creation of a match of level 1, we look at the hypotheses in slot 2 and check if they can extend the match to level 2.
+However, the instantiation of the current hypotheses in slot 2 are inconsistent with $y \mapsto 302$, so nothing can be extended.
+Finally, as the hypothesis also matches slot 2, we update the map $\eta_y(2, y \mapsto 301)$ with $\{h_1\}$.
+After these updates, the maps have the following state :
 
-\begin{tikzpicture}[outer sep=auto, level distance=10mm, scale=0.50]
+\begin{center}
+\begin{tikzpicture}[outer sep=auto, level distance=12mm, scale=0.50]
   \node[xshift = 0] {$\eta_y$}
@@
       child {node {$302$}
-        child {node {$\{h_2\}$}}}
-      child {node {$301$}
-        child {node {$\{h_1\}$}}}}
+        child {node {$\{h_1\}$}}}
+    %  child {node {$301$}
+    %    child {node {$\{\}$}}}
+        }
     child[xshift = 20] {node {2}
       child {node {$301$}
-        child {node {$\{h_2\}$}}}
-      child {node {$300$}
-        child {node {$\{h_1\}$}}}};
-  \node[xshift = 90] {$\eta_y$}
+        child {node {$\{h_1\}$}}}
+    %  child {node {$300$}
+    %    child {node {$\{\}$}}}
+    };
+  \node[xshift=160] {$\mu_y$}
+    child {node[xshift=-20] {1}
+      child {node {$302$}
+        child {node {$\{[h_1]\}$}}}
+    %  child {node {$301$}
+    %    child {node {$\{\}$}}}
+    }
+    %child {node[xshift=20] {2}
+    %  child {node {$301$}
+    %    child {node[yshift=-2em] {$\{\}$}}}};
+\end{tikzpicture}
+\end{center}
+
+The next step is to add $h_2 : 300 \leq 301$ to the maps.
+Again, it unifies with both slots 1 and 2; we start by updating the map $\eta_y(1, y \mapsto 301)$ with $\{h_2\}$.
+Again, an update in slot 1 causes the creation of a level 1 match.
+We then check whether there are suitable hypotheses in slot 2 with
+with a compatible instantiation of $y$; indeed we find $\eta_y(2, y \mapsto 301) = \{h_1\}$.
+This update the match map with the level 2 match $[h_2,h_1]$.
+This is a complete match.
+The new hypothesis $h_3 : 300 \leq 302$ is added to the context.
+
+The hypothesis $h_2$, also matches with slot 2.
+This causes the update $\eta_y(2, y \mapsto 300)$ with $\{h_2\}$. 
+At this point, we need to check for level 1 that would be compatible, there are none and the work with this hypothesis is done.
+
+\begin{center}
+    \begin{tikzpicture}[outer sep=auto, level distance=12mm, scale=0.50]
+      \node[xshift = 0] {$\eta_y$}
         child[xshift = -20] {node {1}
           child {node {$302$}
+            child {node {$\{h_1\}$}}}
+          child {node {$301$}
             child {node {$\{h_2\}$}}}
-      child {node {$301$}
-        child {node {$\{h_1\}$}}}}
+            }
         child[xshift = 20] {node {2}
           child {node {$301$}
+            child {node {$\{h_1\}$}}}
+          child {node {$300$}
             child {node {$\{h_2\}$}}}
-      child {node {$300$}
-        child {node {$\{h_1\}$}}}};
-  \node[xshift=200] {$\mu_y$}
+        };
+      \node[xshift=160] {$\mu_y$}
         child {node[xshift=-20] {1}
-      child {node {$302$}
-        child {node {$\{\{1 \mapsto h_2\}\}$}}}
-      child {node {$301$}
-        child {node {$\{\{1 \mapsto h_1\}\}$}}}}
+          child[xshift=-10] {node {$302$}
+            child {node {$\{[h_1]\}$}}}
+          child[xshift=10] {node {$301$}
+            child {node {$\{[h_2]\}$}}}
+        }
         child {node[xshift=20] {2}
           child {node {$301$}
-        child {node[yshift=-2em] {$\{\{1 \mapsto h_1, 2 \mapsto h_2\}\}$}}}};
+            child {node {$\{[h_2,h_1]\}$}}}};
     \end{tikzpicture}
+    \end{center}
 
@@
 The hypotheses added to the context will be, in order, $h_1 : T_2(a,a)$, $h_2 : T_1(a,b)$ and $h_3 : T_2(a,b)$.
+First, $h_1 : T_2(a,a)$ is considered :
 
-\begin{tikzpicture}[outer sep=auto, level distance=10mm, scale=0.50]
-  \node[xshift = 0] {$\eta_x$}
-    child[xshift = -20] {node {1}
-      child {node {$a$}
-        child {node {$\{ \}$}}}
-      child {node {$b$}
-        child {node {$\{ \}$}}}}
-    child[xshift = 20] {node {2}
+% NOTE : remove comments to get full trees.
+\begin{tikzpicture}[level distance=12mm, scale=0.50]
+    \node{}
+    \node[xshift = 30] {$\eta_x$}
+        %child[xshift = -20] {node {1}
+        %  child {node {$a$}
+        %    child {node {$\{ \}$}}}
+        %  child {node {$b$}
+        %    child {node {$\{ \}$}}}}
+        child[xshift = 40] {node {2}
         child {node {$a$}
             child {node {$\{h_1\}$}}}
-      child {node {$b$}
-        child {node {$\{ \}$}}}};
-  \node[xshift = 90] {$\eta_y$}
-    child[xshift = -20] {node {1}
-        child {node {$a$}
-            child {node {$\{ \}$}}}
-        child {node {$b$}
-            child {node {$\{ \}$}}}}
-    child[xshift = 20] {node {2}
+        % child {node {$b$}
+        %    child {node {$\{ \}$}}}
+        };
+    \node[xshift = 110] {$\eta_y$}
+        %child[xshift = -20] {node {1}
+        %    child {node {$a$}
+        %        child {node {$\{ \}$}}}
+        %    child {node {$b$}
+        %        child {node {$\{ \}$}}}}
+        child[xshift = 40] {node {2}
             child {node {$a$}
                 child {node {$\{h_1\}$}}}
-        child {node {$b$}
-            child {node {$\{ \}$}}}};
+        %    child {node {$b$}
+        %        child {node {$\{ \}$}}}
+        };
+    \node[xshift = 190, yshift = -25] {$\mu_x$}
+        %child[xshift = -20] {node {1}
+        %    child {node {$a$}
+        %        child {node {$\{ \}$}}}
+        %    child {node {$b$}
+        %        child {node {$\{ \}$}}}}
+        %child[xshift = 20] {node {2}
+        %    child {node {$a$}
+        %        child {node {$\{\}$}}}
+        %    child {node {$b$}
+        %        child {node {$\{ \}$}}}
+        %};
+    \node[xshift = 270, yshift = -25] {$\mu_y$}
+        %child[xshift = -20] {node {1}
+        %    child {node {$a$}
+        %        child {node {$\{ \}$}}}
+        %    child {node {$b$}
+        %        child {node {$\{ \}$}}}}
+        %child[xshift = 20] {node {2}
+        %    child {node {$a$}
+        %        child {node {$\{\}$}}}
+        %    child {node {$b$}
+        %        child {node {$\{ \}$}}}
+        %};
 \end{tikzpicture}
@@
 At this point, there are no matches so there is nothing to extend.
-We add the next hypothesis in the queue to the variable maps.
+We add the next hypothesis in the queue, that is $h_2 : T_1(a,b)$, to the variable maps :
 
-\begin{tikzpicture}[outer sep=auto, level distance=10mm, scale=0.50]
+% NOTE : uncomment to get full trees.
+\begin{tikzpicture}[outer sep=auto, level distance=12mm, scale=0.50]
     \node[xshift = 0] {$\eta_x$}
@@
                 child {node {$\{h_2\}$}}}
-        child {node {$b$}
-        child {node {$\{ \}$}}}}
+            %child {node {$b$}
+            %    child {node {$\{ \}$}}}
+            }
         child[xshift = 20] {node {2}
@@
                 child {node {$\{h_1\}$}}}
-        child {node {$b$}
-        child {node {$\{ \}$}}}};
-\node[xshift = 90] {$\eta_y$}
+            %child {node {$b$}
+            %    child {node {$\{ \}$}}}
+            };
+    \node[xshift = 80] {$\eta_y$}
         child[xshift = -20] {node {1}
-        child {node {$a$}
-            child {node {$\{ \}$}}}
+            %child {node {$a$}
+            %    child {node {$\{ \}$}}}
             child {node {$b$}
@@
                 child {node {$\{h_1\}$}}}
+            %child {node {$b$}
+            %    child {node {$\{ \}$}}}
+                };
+    \node[xshift = 160, yshift = 0] {$\mu_x$}
+        child[xshift = -20] {node {1}
+            child {node {$a$}
+                child {node {$\{[h_2]\}$}}}
+            %child {node {$b$}
+            %   child {node {$\{ \}$}}}
+                }
+        %child[xshift = 20] {node {2}
+        %    child {node {$a$}
+        %        child {node {$\{\}$}}}
+        %    child {node {$b$}
+        %        child {node {$\{ \}$}}}
+        %};
+    \node[xshift = 240, yshift = 0] {$\mu_y$}
+        child[xshift = -20] {node {1}
+        %    child {node {$a$}
+        %        child {node {$\{ \}$}}}
             child {node {$b$}
-            child {node {$\{ \}$}}}};
+                child {node {$\{[h_2]\}$}}}
+        }
+        %child[xshift = 20] {node {2}
+        %    child {node {$a$}
+        %        child {node {$\{\}$}}}
+        %    child {node {$b$}
+        %        child {node {$\{ \}$}}}
+        %};
 \end{tikzpicture}
 
-As this hypothesis is in the first slot, we create the corresponding level 1 partial match $m_0 = {(h_2 : T_1(a, b))}$ for which we check for suitable hypotheses that could extend the match.
+As this hypothesis is in the first slot, we create the corresponding level 1 partial match $m_0 = {(h_2 : T_1(x, y))}$ for which we check for suitable hypotheses that could extend the match.
 Note that  $s(lvl(m_0) + 1) = s(2)$ which is the intersection of the variables appearing in the slots $i < 2$ with the variables appearing in slot $2$.
-Thus $s(2) = \{a,b\}$ and we compute:
+Thus $s(2) = \{x,y\}$ and we compute:
 \[
@@
 Since the set is empty, the match can not be currently extended.
-We turn to the remaining hypothesis in the queue and add it to the maps.
+We turn to the remaining hypothesis $h_3 : T_2(a,b)$ in the queue and add it to the maps:
 
-\begin{tikzpicture}[outer sep=auto, level distance=10mm, scale=0.50]
+\begin{tikzpicture}[outer sep=auto, level distance=12mm, scale=0.50]
 \node[xshift = 0] {$\eta_x$}
@@
             child {node {$\{h_2\}$}}}
-        child {node {$b$}
-        child {node {$\{ \}$}}}}
+        %child {node {$b$}
+        %    child {node {$\{ \}$}}}
+        }
     child[xshift = 20] {node {2}
@@
             child {node {$\{h_1, h_3\}$}}}
-        child {node {$b$}
-        child {node {$\{ \}$}}}};
-\node[xshift = 90] {$\eta_y$}
+        %child {node {$b$}
+        %    child {node {$\{ \}$}}}
+        };
+\node[xshift = 80] {$\eta_y$}
     child[xshift = -20] {node {1}
-        child {node {$a$}
-            child {node {$\{ \}$}}}
+        %child {node {$a$}
+        %    child {node {$\{ \}$}}}
         child {node {$b$}
@@
             child {node {$\{h_3\}$}}}};
+    \node[xshift = 160, yshift = 0] {$\mu_x$}
+        child[xshift = -20] {node {1}
+            child {node {$a$}
+                child {node {$\{[h_2]\}$}}}
+            %child {node {$b$}
+            %   child {node {$\{ \}$}}}
+                }
+        child[xshift = 20] {node {2}
+            child {node {$a$}
+                child {node {$\{[h_2,h_3]\}$}}}
+        %    child {node {$b$}
+        %        child {node {$\{ \}$}}}
+        };
+    \node[xshift = 240, yshift = 0] {$\mu_y$}
+        child[xshift = -20] {node {1}
+        %    child {node {$a$}
+        %        child {node {$\{ \}$}}}
+            child {node {$b$}
+                child {node {$\{[h_2]\}$}}}
+        }
+        child[xshift = 20] {node {2}
+        %    child {node {$a$}
+        %        child {node {$\{\}$}}}
+            child {node {$b$}
+                child {node {$\{[h_2,h_3]\}$}}}
+        };
 \end{tikzpicture}
@@
 We are now ready to look for potential matches that could be extended by $h_3$.
-In this particuliar case, since $h_3$ is in slot $2$ we look for $s(2) = \{a,b\}$.
+In this particuliar case, since $h_3$ is in slot $2$ we look for $s(2) = \{x,y\}$.
 Then,
@@
 A discrimination tree is a trie-like data structure that maps expressions to arbitrary data.
-In~\cite{TODO}, it is shown to be good solution for the problem of finding terms (here the types of hypotheses) that might unify with other terms (here the types of the input hypotheses).
+In~\cite{TODO}, it is shown to be good solution for the problem of finding terms (here the types of hypotheses) that might unify with other terms (here the types of the input premises).
 In Lean 4, discrimination trees are also used to index typeclass instances and simplifier lemmas.
@@
 
-One of the more costly operation that we need to run multiple times is the unification \xcom{defeq?} between the hypotheses and the types of the input hypotheses associated to rules.
+One of the more costly operation that we need to run multiple times is the unification \xcom{defeq?} between the hypotheses and the types of the input premises associated to rules.
 We implement a variant of Lean's unification process as follows: the idea is that we wish to normalize the types as much as possible.

### 2025-02-04 paper.tex
Always end paragraph headings with periods

@@
 
-\paragraph{ACL2}
+\paragraph{ACL2.}
 Among white-box proof automation tactics, only ACL2's \enquote{waterfall} (TODO ref) has special support for forward reasoning.
@@
 
-\paragraph{SATCHMO}
+\paragraph{SATCHMO.}
 The Prolog-based first-order theorem prover SATCHMO (TODO ref) and its many descendants (TODO refs) uses a tableau calculus (TODO ref) that can be viewed as combined forward and backward reasoning.
@@
 
-\paragraph{Knowledge-based reasoning systems}
+\paragraph{Knowledge-based reasoning systems.}
 Combined forward and backward reasoning is also supported by some knowledge-based reasoning systems, e.g.\ Algernon (TODO ref) and EYE (TODO ref).
@@
 
-\paragraph{Other related work}
+\paragraph{Other related work.}
 More distantly related are various calculi that combine backward reasoning with some form of incremental forward reasoning or simplification.

### 2025-02-04 paper.tex
Add missing csquotes package import
Typo
Fix semicolons in tikzpictures
Some implementation content

@@
 \usepackage{hyperref}
+\usepackage{csquotes}
 
@@
 \newunicodechar{λ}{\ensuremath{\lambda}}
+\newunicodechar{μ}{\ensuremath{\mu}}
+\newunicodechar{η}{\ensuremath{\eta}}
 \newunicodechar{Γ}{\ensuremath{\Gamma}}
@@
 \newcommand{\jcom}[1]{{\color{orange}{Jannis: #1}} }
+
+\newcommand{\Lam}[2]{\ensuremath{\lambda\, #1,\; #2}}
+\newcommand{\All}[2]{\ensuremath{\forall\, #1,\; #2}}
+\newcommand{\mvar}[1]{\ensuremath{?#1}}
+\newcommand{\Prop}{\ensuremath{\mathrm{Prop}}}
 
@@
     %    child {node {$\{\}$}}}
-    }
+    };
     %child {node[xshift=20] {2}
@@
 \begin{tikzpicture}[level distance=12mm, scale=0.50]
-    \node{}
     \node[xshift = 30] {$\eta_x$}
@@
         };
-    \node[xshift = 190, yshift = -25] {$\mu_x$}
+    \node[xshift = 190, yshift = -25] {$\mu_x$};
         %child[xshift = -20] {node {1}
@@
         %};
-    \node[xshift = 270, yshift = -25] {$\mu_y$}
+    \node[xshift = 270, yshift = -25] {$\mu_y$};
         %child[xshift = -20] {node {1}
@@
             %   child {node {$\{ \}$}}}
-                }
+                };
         %child[xshift = 20] {node {2}
@@
                 child {node {$\{[h_2]\}$}}}
-        }
+        };
         %child[xshift = 20] {node {2}
@@
 We are now ready to look for potential matches that could be extended by $h_3$.
-In this particuliar case, since $h_3$ is in slot $2$ we look for $s(2) = \{x,y\}$.
+In this particular case, since $h_3$ is in slot $2$ we look for $s(2) = \{x,y\}$.
 Then,
@@
 
+\jcom{TODO: deletion}
+
 \section{Implementation}
+
+Implementing the preceding technique efficiently requires some care, particularly in the context of dependent type theory.
+We briefly discuss the most prominent challenges.
+
+\subsection{Unification}
+
+In Lean's type theory, types contain programs and unification is expected to respect definitional equality, i.e.\ equality up to evaluation.
+For instance, a theorem with premise $P~2$, where $P$ is some predicate, should apply to a hypothesis of type $P~((\Lam{x}{x})~2)$ since the two types are definitionally equal.
+As a result, unifying premises and hypotheses is arbitrarily expensive in theory and a major cost centre in practice.
+Our technique therefore seeks to minimise the number of unifications performed.
+
+To limit the expense of unification, we perform unification at \emph{reducible} transparency.
+Lean has multiple transparency levels that determine which constants are unfolded.
+\enquote{Reducible} is the most restrictive of these, so we unfold few (and, in practice, only non-recursive) definitions, which greatly reduces the cost of unification.
+It also limits the power of forward reasoning since
+
+We also need to account for higher-order unification problems: $Q~\mvar{f}$, where $Q$ is a predicate over functions and $\mvar{f}$ is a function variable, should unify with $Q~(\Lam{x}{x})$.
+Since higher-order unification is undecidable and does not admit most general unifiers (TODO ref), we use instead a unification algorithm provided by Lean that is complete for first-order unification problems and heuristically solves some higher-order problems as well.
+This makes our technique incomplete in the sense that some theorem instantiations will be missed, but since the same unification algorithm is used throughout Lean, users are at least well-acquainted with this source of incompleteness.
+
+\subsection{Data Representation}
+
+Since the maps $μ_{x}$  and $η_{x}$ have the same domain, we may combine them into one map.
+This core data structure is then represented by nested persistent hash maps: a first hash map maps variables to a second hash map, the second hash map maps slots to a third hash map and the third hash map maps instantiations to a set of hypotheses and a set of partial matches.
+
+When we compare instantiations, the comparison should be up to definitional equality.
+For example, when we try to apply a theorem $\All{x}{P~x → Q~x → R~x}$ where the first slot matches a hypothesis of type $P~a$ and the second slot matches a hypothesis of type $P~((\Lam{x}{x})~a)$, the application should succeed even though the instantiations of $\mvar{x}$ are different.
+We therefore do not store $a$ and $(\Lam{x}{x})~a$ in the variable maps as-is.
+
+Instead, we bring all instantiations into a normal form which we call \emph{reducible proof-irrelevant normal form} (RPINF).
+This is the usual normal form of expressions with respect to Lean's notion of computation, except for two modifications:
+\begin{itemize}
+  \item Only \emph{reducible} defined constants are unfolded, i.e.\ replaced with their definiendum.
+        This reflects our choice to perform unification at reducible transparency.
+  \item Parts of the original expression that are proofs (in the sense that their types are members of the universe $\Prop$ of propositions) are not normalised.
+        This reflects the fact that Lean's type theory definitionally equates any two proofs of the same proposition, so there is no need to check whether they are equal.
+        When computing the RPINF, we also mark all proofs occurring in it with some special \enquote{metadata}.
+\end{itemize}
+
+With this setup, two expressions $t$ and $u$ are definitionally equal at reducible transparency if their RPINFs, $t'$ and $u'$, are syntactically equal except for those portions of $t'$ and $u'$ that are proofs, which are always considered equal.
+Since this is an entirely syntactic notion of equality, it can be checked very quickly and we can define an effective hash function that is compatible with it.
+This allows us to use a hash map for the instantiations while still comparing them up to definitional equality.
+
+We believe that this scheme is more efficient than alternative schemes relying on repeated unification.
+For example, we could use Lean's implementation of imperfect discrimination trees (TODO ref), which implement a map from expressions to arbitrary data where lookup respects definitional equality.
+However, since these discrimination trees are imperfect, they may return spurious lookup results, so whether a lookup result is valid must be verified by unifying the query expression with the expression that was originally stored in the discrimination tree.
+Repeatedly performing such unifications may be more expensive than calculation one RPINF per expression (which can also be easily cached).
+
+\subsection{Indexing}
+
+\subsection{Aesop Integration}
 
@@
 The index used to match hypotheses with their matching rules are implemented as discrimination trees.
-A discrimination tree is a trie-like data structure that maps expressions to arbitrary data.
-In~\cite{TODO}, it is shown to be good solution for the problem of finding terms (here the types of hypotheses) that might unify with other terms (here the types of the input premises).
-In Lean 4, discrimination trees are also used to index typeclass instances and simplifier lemmas.
 
@@
 We then add them again as a new hypothesis.
-
-The maps described in the last section are all implemented as persistent hash maps.
-\xcom{Why did we not use discrimination trees? $+$ (pers. important cause we store related forward state for many goal.)}
-
-One of the more costly operation that we need to run multiple times is the unification \xcom{defeq?} between the hypotheses and the types of the input premises associated to rules.
-We implement a variant of Lean's unification process as follows: the idea is that we wish to normalize the types as much as possible.
 

### 2025-02-05 paper.tex
design pseudocode and fixes

@@
 \usepackage{csquotes}
+\usepackage{algorithm2e}
 
@@
 \urlstyle{rm}
+\SetKw{Kwin}{in}
 
@@
 
-\newcommand{\Lam}[2]{\ensuremath{\lambda\, #1,\; #2}}
-\newcommand{\All}[2]{\ensuremath{\forall\, #1,\; #2}}
-\newcommand{\mvar}[1]{\ensuremath{?#1}}
-\newcommand{\Prop}{\ensuremath{\mathrm{Prop}}}
-
 \begin{document}
@@
 
-A \textit{shared variable} of $r$ is an input premise $x_i$ such that there are at least two input premise types $T_j$, $T_k$ in which $x_i$ occurs.
+A \textit{variable} is a premise $x_i$ such that there is at least one premise $T_j$ in which $x_i$ occurs.
+In what follows, $v(T)$ denote the set of variables that appear in $T$.
+
+A \textit{shared variable} of $r$ is a premise $x_i$ such that there are at least two input premise types $T_j$, $T_k$ in which $x_i$ occurs.
 A shared variable is necessarily a non-maximal input premise.
@@
 
+\paragraph{Indexing Scheme.}
 This partial progress could be simply saved as trees with root the matched hypothesis returned by the index.
-
-\paragraph{Indexing Scheme.}
 However, we decided to move towards an index that looks at all the hypotheses of the rule.
@@
 
+Before we look at how to find complete matches, we examine exactly which premises depend on each other and for which we need to insure compatibility.
 Given a rule, we will partition its premises into subsets for which premises share a dependency.
@@
 %Another characterization is the partition with the property that for two parts $C_1$ and $C_2$, $\bigcup_{T_i \in C_1} v(T_i) \cap \bigcup_{T_i \in C_2} v(T_i) = \emptyset$, and for which there do not exists a partition with the same property that has more parts.
+% Do we want to prove this?
 
@@
 This can be proven to always be possible.
+(Again, we assume that there is only one cluster.)
 \xcom{We can add the proof if we want to / have space.}
@@
 We then aim to construct matches iteratively: we consider each hypotheses one by one, assign them to their respective slots and check whether there exist compatible matches which they can extend.
+Whenever we find an hypothesis that fits with the premise in slot one, we create a new (level one) match.
+This process, that we will introduce in more details shortly, will be refered to as Algorithm \ref{alg:addhyp}.
+On the other hand, whenever a match is updated (or created), we also look through the hypotheses we have already came accross, this is Algorithm \ref{alg:updatematch}.
+
 The compatibility check means that all the shared variables of the considered hypotheses need to be compatible.
-It would be costly to do this comparison every time and indeed we choose instead to maintain what we call \textit{variable maps} whose purpose is to save the current partial matches.
+It would be costly to do this comparison every time and indeed we choose instead to maintain what we call \textit{variable maps} whose purpose is to save the current partial matches in terms of their instantiation.
 Each variable map is associated to one shared variable, these are the variables that dictate compatibility.
@@
 \[
-  \bigcap_{s(i)} \mu_x \left(i - 1, \sigma_h(x)\right)
+  \bigcap_{x \in s(i)} \mu_x \left(i - 1, \sigma_h(x)\right).
 \]
 
-Similarly, given a match $m$ and $\sigma_m$ the substitution map of the relevant  variable in $m$ then we can check for compatible hypotheses that would extend $m$ by computing
+Similarly, given a match $m$ of level $i$ and $\sigma_m$ the substitution map of the relevant  variable in $m$ then we can check for compatible hypotheses that would extend $m$ by computing
 \[
-  \bigcap_{s(lvl(M) + 1)} \eta_x \left(i + 1, \sigma_m(x)\right)
+  \bigcap_{x \in s(i + 1)} \eta_x \left(i + 1, \sigma_m(x)\right).
 \]
 
+With this in mind, Algorithm \ref{alg:addhyp} and \ref{alg:updatematch} simply maintain these maps up to date and performs compatibility checks by computing the above intersection of sets.
+
+\begin{algorithm}
+    \caption{New hypothesis}\label{alg:addhyp}
+    \KwData{A hypothesis $h : T$ that matches with $T_i$.}
+    \KwResult{The maps $\eta$ and $\mu$ are updated.}
+        \eIf{$i = 1$}{
+            \For{$x$ \Kwin $s(i)$}{
+                Update $\mu_x$ with $(1,\sigma_m(x)) \mapsto \mu_x(1,\sigma_m(x))\cup\{[h]\}$ (A level 1 match).\\
+            }
+            Extend match $[h]$.
+            }
+        {
+        \For{$x$ \Kwin $s(i)$}{
+            Update $\eta_x$ with $(i,\sigma_h(x)) \mapsto \eta_x(i,\sigma_h(x))\cup\{h\}$.\\        
+        }    
+        $ms$ \gets \ $\bigcap_{x \in s(i)} \mu_x \left(i - 1, \sigma_h(x)\right)$\;
+        \For{$m$ \Kwin $ms$}{
+            $m'$ \gets m.\text{append } h\\
+            \For{$x$ \Kwin $s(i)$}{
+            Update $\mu_x$ with $(i,\sigma_m(x)) \mapsto \mu_x(i,\sigma_m(x))\cup\{m'\}$.}\\
+            Extend match $m'$.
+        }}
+\end{algorithm}
+
+\begin{algorithm}
+    \caption{Extend match}\label{alg:updatematch}
+    \KwData{A partial match $m$ of level $i$.}
+    \KwResult{The map $\mu$ is updated.}  
+        $hs$ \gets \ $\bigcap_{x \in s(i + 1)} \eta_x \left(i + 1, \sigma_m(x)\right)$\\
+        \For{$h$ \Kwin $hs$}{
+            $m'$ \gets m.\text{append } h\\
+            \For{$x$ \Kwin $s(i + 1)$}{
+            Update $\mu_x$ with $(i + 1,\sigma_{m'}(x)) \mapsto \mu_x(i + 1,\sigma_{m'}(x))\cup\{m'\}$.}\\
+            Extend match $m'$.
+        }
+\end{algorithm}
+
+In sum, given a hypothesis $h :T$ we
+\begin{enumerate}
+\item Save it in the variable maps $\eta_x$, for the relevant shared variables $x$.
+\item If it matches slot 1, we create a level one match.
+\item Look if we can extend existing partial matches using $h$.
+\item For any modified partial match, we look whether there exists suitable hypotheses that can extend them and repeat in case of success.
+\end{enumerate}
+
 \paragraph{Example 1.}
-\xcom{This is quite verbose, maybe I should use a more systematic approach.}
+
+\xcom{Technically, it is useless to save level 1 hyps in $\eta_x$. Should we remove them from the examples?.}
 To demonstrate the typical behavior of the algorithm we consider the following example which has \textbf{Lemma 1} as a rule and for which we add, in order, the hypotheses $h_1 : 301 \leq 302$ and $h_2 : 300 \leq 301$.
@@
 
-At the beginning, for the rule \textit{le\_trans}, we notice that the only shared variable is ${y}$.
+At the beginning, we feed $h_1$ to Algorithm \ref{alg:addhyp}.
+Furthmore, for the rule \textit{le\_trans}, we notice that the only shared variable is ${y}$.
 Now the hypothesis $h_1 : 301 \leq 302$ unifies both with slot 1: $x \leq y$ and slot 2: $y \leq z$.
-This updates the map $\eta_y(1, y \mapsto 302)$ with $\{h_1\}$.
+Starting with slot 1, this updates the map $\eta_y(1, y \mapsto 302)$ with $\{h_1\}$.
 An update in slot 1 also marks the begining of a new partial match and we update de match map $\mu_y$ acordingly.
-Note that after the creation of a match of level 1, we look at the hypotheses in slot 2 and check if they can extend the match to level 2.
+Note that after the creation of a match of level 1, we look at the hypotheses in slot 2 and check if they can extend the match to level 2 (Algorithm \ref{alg:updatematch}).
 However, the instantiation of the current hypotheses in slot 2 are inconsistent with $y \mapsto 302$, so nothing can be extended.
+Indeed,
+\[
+  \bigcap_{x\in s(lvl(m_0) + 1)} \eta_x \left(i + 1, \sigma_{m_0}(x)\right) = \eta_y(2, 302) = \emptyset.
+\]
 Finally, as the hypothesis also matches slot 2, we update the map $\eta_y(2, y \mapsto 301)$ with $\{h_1\}$.
+At this point the intersection
+\[
+  \bigcap_{x\in s(2)} \eta_x \left(1, \sigma_{h}(x)\right) = \mu_y(1, 301) = \emptyset
+\]
+and at this point the algorithm is done with $h_1$ and stops.
 After these updates, the maps have the following state :

### 2025-02-05 paper.tex
Restore some newcommands that were accidentally deleted
This partially reverts commit 62d7792a5e283487598f09894f862964333ca330.
Fix algorithms
Introduction first draft

@@
 
+\newcommand{\Lam}[2]{\ensuremath{\lambda\, #1,\; #2}}
+\newcommand{\All}[2]{\ensuremath{\forall\, #1,\; #2}}
+\newcommand{\mvar}[1]{\ensuremath{?#1}}
+\newcommand{\Prop}{\ensuremath{\mathrm{Prop}}}
+
 \begin{document}
@@
 \section{Introduction}
-\xcom{TODO}
 
-\xcom{ Specify the problem (user defined rules)
- + deal with changes with local context}
-\xcom{ Talk about aesop}
-\xcom{ Do we outline the solution?}
+In many proof assistants, simple automation tactics based on tree search play a substantial role in practice: Isabelle's (TODO ref) and Rocq's (TODO ref) auto, ACL2's (TODO ref) waterfall, PVS's (TODO ref) grind, Lean's (TODO ref) Aesop (TODO ref), etc.
+These tactics operate on \emph{goals} $Γ ⊢ A$, where the \emph{local context} $Γ$ is a list of assumptions and $A$ is the proposition to be proved.
+Faced with an initial goal, the tactics try various user-specified \emph{rules}, which correspond to admissible inferences in the prover's logic.
+A successful rule reduces the current goal to zero or more subgoals, which are solved recursively until either a full proof is found or no rule applies to any open goal or the search times out.
 
-An important contextualization is that the forward reasoning procedure in this paper takes place within Aesop's routine procedure.
-We will focus on examples that highlight this style of proof while noting that in practice, a combination of both forwards and backwards reasoning are used.
+Typical rules would be the $∧$-introduction lemma $A → B → A ∧ B$, which, when applied backwards, reduces a goal $Γ ⊢ A ∧ B$ to subgoals $Γ ⊢ A$ and $Γ ⊢ B$; or left $∨$-introduction, which reduces $Γ ⊢ A ∨ B$ to $Γ ⊢ A$.
+Such \emph{backward rules} are the most common rule type, but most search tactics allow users to use any tactic as a rule.
+These rule tactics can output arbitrary subgoals, subject only to the condition that proofs of the returned subgoals can be assembled into a proof of the goal on which the rule was run.
+
+ACL2 and Aesop additionally provide special support for \emph{forward rules}.
+Given, for example, the transitivity lemma $\All{x~y~z}{x < y → y < z → x < z}$, a forward rule run on the goal $a < b,\, b < c ⊢ A$ produces the single subgoal $a < b,\, b < c,\, a < c ⊢ A$.
+Hence forward rules can be used to establish additional facts that may be relevant for other rules.
+For example, the Aeneas project (TODO ref) uses Aesop's forward rules \jcom{(ensure that this is accurate)} in its proof automation to establish facts that are later used by an arithmetic decision procedure.
+
+However, Aesop's (and ACL2's) support for forward rules was, until recently, quite basic.
+In particular, each forward rule was tried on each goal separately (subject to pre-filtering by a discrimination tree (TODO ref) index).
+This is wasteful since in practice, many rules produce subgoals with a similar---or even the same---local context as the initial goal.
+For example, the ∧-introduction rule affects only the goal's target, not its context.
+Still, all forward rules applied to the initial $Γ ⊢ A ∧ B$ would be tried again on the subgoals $Γ ⊢ A$ and $Γ ⊢ B$, with predictably little effect.
+
+In this paper, we present an alternative way to apply forward rules that addresses this issue.
+The central idea is to remember, for each goal $Γ ⊢ A$, which forward rules can be partially applied to which sets of hypotheses.
+When a non-forward rule then generates subgoals $Δ₁ ⊢ A₁$, \dots, $Δₙ ⊢ Aₙ$, the new local contexts $Δᵢ$ are likely similar to $Γ$.
+Hence it is generally cheaper (and often much cheaper) to start with the partial matches for $Γ$ and update them for each of the $Δᵢ$.
+The update first deletes partial matches that include hypotheses no longer present in $Δᵢ$.
+Afterwards it adds new partial matches for the hypotheses that are new in $Δᵢ$, extending the existing partial matches for a rule if the new hypothesis is compatible with them.
+Whenever this process produces a full match---i.e., all premises of a forward rule are satisfied by a set of hypothesis in $Δᵢ$---the rule is applied and adds a corresponding new hypothesis to $Δᵢ$.
+Our main contribution is a data structure, the \emph{partial match state}, that organises the partial matches in such a way that the update can be done efficiently.
+
+In addition to supporting such \emph{incremental} forward reasoning, the partial match state satisfies two other desiderata:
+\begin{itemize}
+  \item It is \emph{persistent}, i.e.\ partial match states for parent and child goals share much of their structure.
+        This is desirable since Aesop supports search strategies other than depth-first search and must therefore keep all goals encountered during the search, along with their partial match states, in memory.
+  \item It reduces the number of \emph{unifications} performed when matching hypotheses against rule premises.
+        In dependent type theory, unification happens up to definitional equality, which makes it quite expensive.
+        We therefore want to avoid redundant unifications.
+\end{itemize}
+
+Our evaluation (Sec.~\ref{sec:evaluation}) shows that this new implementation of forward reasoning in Aesop substantially outperforms the previous, naive implementation in some (though not all) cases.
+The implementation, the benchmarks we used and the full evaluation data are available in a supplement to this paper (TODO ref).
+Mainline Aesop already uses the new implementation by default.
 
@@
       \For{$x$ \Kwin $s(i)$}{
-                Update $\mu_x$ with $(1,\sigma_m(x)) \mapsto \mu_x(1,\sigma_m(x))\cup\{[h]\}$ (A level 1 match).\\
+        Update $\mu_x$ with $(1,\sigma_m(x)) \mapsto \mu_x(1,\sigma_m(x))\cup\{[h]\}$ (a level 1 match)
       }
-            Extend match $[h]$.
+      Extend match $[h]$
+    }{
+      \For{$x$ \Kwin $s(i)$}{
+        Update $\eta_x$ with $(i,\sigma_h(x)) \mapsto \eta_x(i,\sigma_h(x))\cup\{h\}$
       }
-        {
+      $ms \gets \bigcap_{x \in s(i)} \mu_x \left(i - 1, \sigma_h(x)\right)$ \\
+      \For{$m$ \Kwin $ms$}{
+        $m' \gets m.\text{append } h$ \\
         \For{$x$ \Kwin $s(i)$}{
-            Update $\eta_x$ with $(i,\sigma_h(x)) \mapsto \eta_x(i,\sigma_h(x))\cup\{h\}$.\\        
+          Update $\mu_x$ with $(i,\sigma_m(x)) \mapsto \mu_x(i,\sigma_m(x))\cup\{m'\}$
         }
-        $ms$ \gets \ $\bigcap_{x \in s(i)} \mu_x \left(i - 1, \sigma_h(x)\right)$\;
-        \For{$m$ \Kwin $ms$}{
-            $m'$ \gets m.\text{append } h\\
-            \For{$x$ \Kwin $s(i)$}{
-            Update $\mu_x$ with $(i,\sigma_m(x)) \mapsto \mu_x(i,\sigma_m(x))\cup\{m'\}$.}\\
-            Extend match $m'$.
-        }}
+        Extend match $m'$
+      }
+    }
 \end{algorithm}
@@
     \KwResult{The map $\mu$ is updated.}  
-        $hs$ \gets \ $\bigcap_{x \in s(i + 1)} \eta_x \left(i + 1, \sigma_m(x)\right)$\\
+        $hs \gets \bigcap_{x \in s(i + 1)} \eta_x \left(i + 1, \sigma_m(x)\right)$ \\
         \For{$h$ \Kwin $hs$}{
-            $m'$ \gets m.\text{append } h\\
-            \For{$x$ \Kwin $s(i + 1)$}{
-            Update $\mu_x$ with $(i + 1,\sigma_{m'}(x)) \mapsto \mu_x(i + 1,\sigma_{m'}(x))\cup\{m'\}$.}\\
+            $m' \gets m.\text{append } h$\\
+            \For{$x$ \Kwin $s(i + 1)$}{%
+              Update $\mu_x$ with $(i + 1,\sigma_{m'}(x)) \mapsto \mu_x(i + 1,\sigma_{m'}(x))\cup\{m'\}$.%
+            }
             Extend match $m'$.

### 2025-02-07 paper.tex
Edit preliminaries

@@
 \newunicodechar{∩}{\ensuremath{\cap}}
+\newunicodechar{⊆}{\ensuremath{\subseteq}}
+\newunicodechar{⋃}{\ensuremath{\bigcup}}
 \newunicodechar{→}{\ensuremath{\rightarrow}}
@@
 \newcommand{\Prop}{\ensuremath{\mathrm{Prop}}}
+\newcommand{\vars}{\ensuremath{\mathrm{vars}}}
+\newcommand{\dom}{\ensuremath{\mathrm{dom}}}
+\newcommand{\sub}{\ensuremath{\mathrm{sub}}}
+\newcommand{\lvl}{\ensuremath{\mathrm{lvl}}}
 
@@
 The implementation, the benchmarks we used and the full evaluation data are available in a supplement to this paper (TODO ref).
-Mainline Aesop already uses the new implementation by default.
+Mainline Aesop already uses the new implementation by default (TODO true?).
 
@@
 
-\xcom{Adapt for CADE.}
+We first introduce some terminology for matches and for forward rules (which we identify with their underlying lemmas).
+Our foundation in this paper is dependent type theory, but our technique can be adapted to other logics.
 
-In what follows, we write $\Gamma \models T$ for a goal where $\Gamma$ represents the context and $T$ the target.
-In a mathematical problem, $\Gamma$ is the list of hypotheses and variable from which we wish to show $T$.
+Given a lemma $\All{(x₁ : A₁) \dots (xₙ : Aₙ)}{B}$, the $xᵢ$ are \emph{premises} and $B$ is the lemma's \emph{conclusion}.
+The type $Aᵢ$ of the $i$-th premise may depend on (i.e., refer to) any premise $xⱼ$ with $j < i$.
+Implication $A → B$ is an abbreviation for $\All{x : A}{B}$ where $B$ does not depend on $x$.
+Propositions are (certain) types and we do not distinguish between the two concepts.
 
-\paragraph{Forward Rules.}
-Consider the following lemma:
-\begin{lem}[\lean{le_trans}]
-  Given a preorder on some type $\alpha$ and elements $a,b,c$ of type $\alpha$, the following holds:
-  $$a \leq b \rightarrow b \leq c \rightarrow a \leq c$$
-\end{lem}
+Premises on which a later premise depends are \emph{variables}.
+All other premises are \emph{input premises}.
+For example, the lemma $\All{x~y~z}{x < y → y < z → x < z}$ has variables $x, y, z$ and (unnamed) input premises of types $x < y$ and $y < z$.
+If a variable occurs in at least two premises, like $y$ above, it is \emph{shared}.
 
-From such lemmas, we will create what we call \textit{forward rules} that represent the theorems on which we would like to reason in a forward manner.
-The premises $\alpha$, $ \text{Preorder } \alpha$ , $a$, $b$, $c$, $a \leq b$ and $b \leq c$ are separated between \textit{implied premises} $\text{Preorder } \alpha$,
-$a$, $b$ and $c$ and \textit{maximal premises} $a \leq b$ and $b \leq c$.
-Maximal premises are also called \textit{input premises}.
+For a lemma with $n$ input premises, we assign to each input premise a unique number between $1$ and $n$.
+This is the premise's \emph{slot}, which we often identify with the premise itself.
+For now, we assume that the input premises are numbered consecutively, so in the above example slot 1 has type $x < y$ and slot 2 has type $y < z$.
+The \emph{shared variables of slot $i$}, $\vars(i)$ are those variables that appear in the type of $i$ and also in at least one type of a slot $j < i$.
 
-For the premises $(x_1 : T_1),(x_2 : T_2),\dots,(x_n : T_n)$, the input premises consist of the premises for which there are no $j > i$ such that $T_j$ depends on $x_i$.
+If $h : A$ is a hypothesis in the local context of a goal and $A$ unifies with the type $B$ of slot $i$, we say that $h$ \emph{matches} $i$.
+Unification induces a substitution $σ$ (with $\dom(σ) ⊆ \vars(i)$) such that $A[σ]$ is definitionally equal to $B[σ]$; we call this substitution $\sub_{i}(h)$.
+For now we do not specify exactly what \enquote{unifies} means; see Sec.~\ref{sec:unification}.
 
-\paragraph{Rules.}
-We write $r : \Phi \to \Psi$ for a \textit{rule} $r$ with input premises $\Phi$ and output $\Psi$.
-For a given rule, we assign each maximal input premise a unique index between $1$ and $n$.
-We call theses indices \textit{slots} of $r$ and define $\text{slots}(r)$ as the set of slots of $r$.
+A \textit{match} $M$ for a given rule $r$ is a partial map from slots of $r$ to hypotheses.
+The map must satisfy the following conditions:
+\begin{enumerate}
+  \item\label{cond:match:matches} For each slot $i$, if $M(i)$ is defined, then $M(i)$ matches $i$.
+  \item\label{cond:match:compat} For all slots $i$ and $j$ of $r$ such that $M(i)$ and $M(j)$ are defined, the substitutions $\sub_{i}(M(i))$ and $\sub_{j}(M(j))$ are \emph{compatible}, i.e.\ they agree on all variables that occur in both $i$ and $j$.
+  \item\label{cond:match:closed} $M$ is downward-closed: if $M(i)$ is defined, then for any slot $j < i$ of $r$, $M(j)$ is also defined.
+\end{enumerate}
+The \emph{substitution of a match $M$}, $\sub(M)$, is the union of the substitutions induced by the hypotheses in $M$, i.e. $\sub(M) ≔ ⋃_{i ∈ \dom(M)}\sub_{i}(M(i))$.
+This is well-defined by requirement~\ref{cond:match:compat}.
+The \textit{level of a match $M$}, $\lvl(M)$, is the maximal $i$ such that $i ∈ \dom(M)$.
+A match for rule $r$ is \textit{complete} if $i ∈ \dom(M)$ for every slot $i$ of $r$, and \emph{partial} otherwise.
 
-Given a slot $i$ and hypothesis $h : T$ in the context $\Gamma$ with $T \equiv_{[\sigma]} T_i$ (i.e.
-$\sigma$ is the substitution that results from unifying $T$ and $T_i$), we say that $h$ is \textit{suitable for slot $i$} and define $\text{sub}(h) ≔ \sigma$.
-
-A \textit{variable} is a premise $x_i$ such that there is at least one premise $T_j$ in which $x_i$ occurs.
-In what follows, $v(T)$ denote the set of variables that appear in $T$.
-
-A \textit{shared variable} of $r$ is a premise $x_i$ such that there are at least two input premise types $T_j$, $T_k$ in which $x_i$ occurs.
-A shared variable is necessarily a non-maximal input premise.
-We also write $?x_i$ to emphasise that we interpret $x_i$ as a variable (and, in the code, as a metavariable).
-The set $s_r(i)$ of \textit{shared variables of a slot} $i$ of $r$ are those shared variables $?x$ that occur in $T_i$ and in at least one slot $j < i$.
-If the rule in question is clear from the context, we write $s(i)$.
-
-%The typical behavior is that running the procedure in a context containing suitable
-%candidates for the maximal hypotheses will add the \textit{result} of the forward
-%rule $a < c$ to the context.
-%The result is also called the \textit{output hypothesis}.
-%If we find matching hypotheses for all maximal input premises in the local context,
-%we can apply the forward rule, since each non-maximal input premise is uniquely determined
-%by maximal hypotheses.
-
-A \textit{match} $M$ for $r$ is a partial map from slots of $r$ to hypotheses in the context.
-It must satisfy the following conditions:
-\begin{itemize}
-  \item For each slot $i$ of $r$, if $M(i)$ is defined, then $M(i) \equiv_{[\sigma_i]} T_i$ with a substitution $\sigma_i$.
-        The domain of this substitution is a subset of the non-maximal input premises.
-  \item For all slots $i$ and $j$ of $r$ such that $M(i)$ and $M(j)$ are defined, $\sigma_i$ and $\sigma_j$ are compatible.
-  \item $M$ is downward-closed: if $M(i)$ is defined, then for any slot $j < i$ of $r$, $M(j)$ is also defined.
-\end{itemize}
-
-A match of $r$ is \textit{complete} if $M(i)$ is defined for every slot $i$ of $r$, and \textit{incomplete} otherwise.
-The \textit{level of a match} $M$, $lvl(M)$, is the maximal $i$ such that $M(i)$ is defined.
-The \textit{substitution of a match} $M$, $sub(M)$, is the union of the substitutions $\sigma_i$ for each $i$ such that $M_i$ is defined.
-This union is defined since the $\sigma_i$ are pairwise compatible.
-This means that $sub(M)$ contains only one assignement per variable.
-The set of \textit{shared variables of a match} $M$ is $s(lvl(M))$ the shared variables of $lvl(M)$.
-
-An output hypothesis is \textit{redundant} if its type is already present in the context.
-A goal is saturated with respect to a set of forward rules if all output hypotheses returned by the procedure are redundant.
-
-The following section introduces an algorithm to construct complete matches $M$ from the context $\Gamma$.
+A partial match $M$ for rule $r$ corresponds to a partial application of the rule to the terms given in $\sub(M)$.
+The application is type-correct by requirements~\ref{cond:match:matches} and~\ref{cond:match:compat}.
+If $M$ is complete, then $\sub(M)$ contains a term for every slot and every variable---hence for every premise---of $r$ and we have a full application.
 
@@
 
-\subsection{Unification}
+\subsection{Unification}%
+\label{sec:unification}
 
@@
 However, since these discrimination trees are imperfect, they may return spurious lookup results, so whether a lookup result is valid must be verified by unifying the query expression with the expression that was originally stored in the discrimination tree.
-Repeatedly performing such unifications may be more expensive than calculation one RPINF per expression (which can also be easily cached).
-
-\subsection{Indexing}
+We believe that repeatedly performing such unifications would be, on average, more expensive than calculating one RPINF (which can be cached) for each type in a goal.
 
 \subsection{Aesop Integration}
-
-\xcom{Here are some notes about the implementation.}
-The index used to match hypotheses with their matching rules are implemented as discrimination trees.
 
@@
 
+\subsection{Indexing}
+
+\subsection{Deduplication}
+
 \xcom{precompilation}
 
-\section{Evaluation}
+\section{Evaluation}%
+\label{sec:evaluation}
 
@@
 
+\jcom{auto2}
+
 \section{Conclusion}

### 2025-02-07 paper.tex
Fix order of email addresses

@@
 \institute{University of Munich (LMU), Munich, Germany\\
-\email{\{xavier.genereux,jannis.limperg\}@lmu.de}}
+\email{\{jannis.limperg,xavier.genereux\}@lmu.de}}
 
@@
 
+These tests were performed on a Macbook Pro with the Apple M2 Pro chip equipped with 32 Go of RAM.
+The tests showcased in this section are available here \footnote{\xcom{TODO : Link Github repo with tests.}}
 \subsection{Results}
@@
 
-In Fig.~\ref{fig:erase}, the step is set up as follows: given some rule $r$ with $6$ \xcom{motivate 6} premises, we run the procedure in a context with 5 of the 6 premises present.
-The numbers on the $x$ axis represent which hypothesis is missing in the order of consideration by the algorithm.
-In a sense, this is how \textit{deep} the algorithm investigate before it stops.
-At $0$, the rule are not selected by the index so the procedure stops.
-The stateless algorithm gets progressively slower the further the removed premises is located.
-
-The position of the missing premise has little effect on the stateful algorithm as the index selects the rule in all case; it only needs one of the premise to be present.
-The work the stateful algorithm has to do is thus close to constant in this scenario.
+\paragraph{Transitivity.}
 
 \begin{figure}
-% This is the erase without precompilation for 6 premises.
-% Note I have a decent idea that would allow us to compare for multiple
-% number of premises. 
-% (Divide by the average of the stateful time to normalize.)
+    \label{fig:trans}
+    % Transitivity with precompilation.
+    % 
     \begin{tikzpicture}
         \begin{axis}[
-        title={Depth Test},
-        xlabel={Depth},
+            title={Transitivity rule},
+            xlabel={Number of hypotheses},
             ylabel={Time in ms},
-        xmin=0, xmax=6,
-        ymin=50, ymax=65,
-        xtick={0,1,2,3,4,5,6},
-        ytick={50,55,60,65},
+            xmin=2, xmax=6,
+            ymin=2^3, ymax=2^13,
+            xtick={2,3,4,5,6},
+            ytick={2^3,2^5,2^7,2^9,2^11,2^13},
+            ymode=log,
+            %xmode=log,
+            log basis y=2,
             legend pos=north west,
@@
             coordinates {
-        (1,50.91)(2,53.14)(3,52.91)(4,51.22)(5,55.14)(6,56.60)
+                (2, 74.566944) (3, 444.481222) (4, 811.149916) (5, 2955.663569) (6, 4703.832541) 
             };
             \addlegendentry{Stateless}
+        \addplot[
+            color=blue,
+            mark=square,
+            ]
+            coordinates {
+                (2, 13.089680) (3, 12.556333) (4, 14.278986) (5, 17.385902) (6, 20.682083)
+            };
+            \addlegendentry{Stateful}
+        \end{axis}
+        \end{tikzpicture}
+    \end{figure}
+
+As a first example, which we believe highlights well the advantages of the stateful approach, we consider a scaled up version of a transitivity problem.
+As variables we choose a custom version of natural numbers.
+The reason being that Lean has special optimization procedures for its integer and thus makes evaluation quite hard on problems that use them; the unification is too quickly resolved, making algorithmic complexity difficultly distinguishable from noise.
+We aim to emulate a more general problem where the objects unified do not benefit of some special support.
+
+Given some predicate $P(x, y)$, we register the rule $r : P(x, y) \to P(y, z) \to P(x, z)$ to both algorithms.
+Then, in Figure \ref{fig:trans}, we compare them on a context containing $n$ hypotheses of the form $(h_1 : P(a, n + 1)), \dots, (h_n : P(a + n - 1, a + n))$.
+Saturating this goal adds a total of $n(n-1)/2$ hypotheses to the context.
+
+The results in Figure \ref{fig:trans} have nothing surprising : The stateless algorithm needs to go over the whole context, which contains initially $n$ hypotheses and compares them with these same $n$ hypotheses.
+This is repeated approximately for each of the $n(n-1)/2$ added hypotheses giving the expected complexity of $O(n^4)$.
+
+The stateful algorithm considers each of the problem's hypothesis, both the initial and added ones, only once.
+As the hypotheses get matched to both slots, it then compares it with at most $n$ hypotheses for each slot.
+Indeed, by the end of the saturation procedure, for a value $k$ fixed, the first slot $\eta_y(1,k)$, contains at most $n$ hypotheses of the form $h : P(m,k)$ and symmetrically for the second slot.
+This leaves an expected complexity of $O(n^2)$.
+
+There are of course obvious inefficiencies, even with the stateful algorithm.
+For this specific kind of problem, the hypotheses need only be used once and this leads to much repeated work.
+For these scenarios, users can register the rule as a \textit{destruct} rule, eliminating hypotheses after their first use.
+In this case, the complexity is linear in $n$.
+For more edge case examples like this one, consult the project's repository.
+
+\paragraph{Independent rules.}
+
+\begin{figure}
+    %Indep with precompilation
+        \begin{tikzpicture}
+            \begin{axis}[
+                title={Independance Test},
+                xlabel={Number of Rules},
+                ylabel={Time in ms},
+                xmin=0, xmax=32,
+                ymin=0, ymax=1700,
+                ymode=log,xmode=log,
+                log basis x=2,
+                log basis y=2,
+                legend pos=north west,
+                ymajorgrids=true,
+                grid style=dashed,
+            ]
 
@@
                 coordinates {
-        (1,56.04)(2,56.43)(3,55.58)(4,55.26)(5,54.80)(6,60.61)
+                (1, 0.710963)(2, 1.266787)(4, 2.181111)(8, 4.220279)(16, 8.084744)(32, 15.683840)
                 };
                 \addlegendentry{Stateful}
-
+            \addplot[
+                color=orange,
+                mark=square,
+                ]
+                coordinates {
+                (1, 0.872730)(2, 2.100940)(4, 7.245598)(8, 35.805959)(16, 222.283526)(32, 1600.978502)
+                };
+                \addlegendentry{Stateless}
             \end{axis}
@@
 
-\begin{figure}
-    % This is the erase without precompilation for 6 premises.
-    % Note I have a decent idea that would allow us to compare for multiple
-    % number of premises. 
-    % (Divide by the average of the stateful time to normalize.)
-    \begin{tikzpicture}
-        \begin{axis}[
-            title={Scaled Depth Test},
-            xlabel={Depth},
-            ylabel={Time in ms},
-            xmin=0, xmax=8,
-            ymin=0.5, ymax=2,
-            xtick={0,1,2,3,4,5,6,7,8},
-            ytick={0.75,1,1.25,1.5,1.75},
-            legend pos=north west,
-            ymajorgrids=true,
-            grid style=dashed,
-        ]
+In the last example, part of the stateful algorithm's success stems from the fact that it could reuse a lot of its previous work.
+An obvious question is then how well does the stateful algorithm work when there is no work to be reused.
 
-        \addplot[color=orange,mark=square,]
-            coordinates {(0, 0.883600)(1, 1.039751)};
-            \addlegendentry{Stateless}
-        \addplot[color=orange,mark=square,]
-            coordinates {(0, 0.804521)(1, 1.017893)(2, 1.022493)};
-        \addplot[color=orange,mark=square,]
-            coordinates {(0, 0.791528)(1, 1.015467)(2, 1.076267)(3, 1.159999)};
-        \addplot[color=orange,mark=square,]
-            coordinates {(0, 0.743256)(1, 0.966436)(2, 1.058876)(3, 1.133514)(4, 1.225561)};
-        \addplot[color=orange,mark=square,]
-            coordinates {(0, 0.770625)(1, 1.003777)(2, 1.109929)(3, 1.215171)(4, 1.305373)(5, 1.390667)};
-        \addplot[color=orange,mark=square,]
-            coordinates {(0, 0.759093)(1, 1.016667)(2, 1.148273)(3, 1.213138)(4, 1.342795)(5, 1.442271)(6, 1.514877)};
-        \addplot[color=orange,mark=square,]
-            coordinates {(0, 0.709965)(1, 0.971294)(2, 1.068745)(3, 1.179333)(4, 1.312798)(5, 1.424054)(6, 1.551690)(7, 1.628344)};
-        \addplot[color=orange,mark=square,]
-            coordinates {(0, 0.728860)(1, 1.001576)(2, 1.097472)(3, 1.219259)(4, 1.366683)(5, 1.444844)(6, 1.576972)(7, 1.668693)(8, 1.815399)};
+Given a set of propositions $\mathcal{P} := \{P_{i,j} | 1 \leq i \leq n \text{ and } 1 \leq j \leq 6\}$ and $\mathcal{Q} :=\{Q_{i} | 1 \leq i \leq n \}$, in this test, we consider the $n$ following rules $r_i : P_{i,1}\to \dots P_{i,6} \to Q_i$.
+The procedures are then run on a context containing exactly $\mathcal{P}$.
 
-        \end{axis}
-        \end{tikzpicture}
-    \end{figure}
+In this case, we sketch the behavior of the stateless algorithm:
+It pick a first hypothesis, $h_{i,j} : P_{i,j}$.
+If $j < 6$, the index returns nothing and it selects another hypothesis.
+For $h_{i,6} : P_{i,6}$, the rule $r_i$ is selected and it then needs to try and fill its remaining five slot with the context's $6n$ hypotheses which it tries all.
+This gives a complexity of about $O(n^2)$.
+
+For the stateful algorithm, the hypotheses are considered one by one and match directly to the rule and the slot.
+The expected complexity is then linear in the number of rules.
+
+\paragraph{Depth of considered rules.}
 
@@
             \addlegendentry{Stateful}
-
         \addplot[
@@
             \addlegendentry{Stateless}
-
         \end{axis}
@@
 
-    \begin{figure}
-        %Indep with precompilation
-            \begin{tikzpicture}
-                \begin{axis}[
-                    title={Independance Test},
-                    xlabel={Number of Rules},
-                    ylabel={Time in ms},
-                    xmin=0, xmax=32,
-                    ymin=0, ymax=1700,
-                    ymode=log,xmode=log,
-                    log basis x=2,
-                    log basis y=2,
-                    legend pos=north west,
-                    ymajorgrids=true,
-                    grid style=dashed,
-                ]
+In Fig.~\ref{fig:erase}, the step is set up as follows: given some rule $r$ with $6$ \xcom{motivate 6} premises, we run the procedure in a context with 5 of the 6 premises present.
+The numbers on the $x$ axis represent which hypothesis is missing in the order of consideration by the algorithm.
+In a sense, this is how \textit{deep} the algorithm investigate before it stops.
+At $0$, the rule are not selected by the index so the procedure stops.
+The stateless algorithm gets progressively slower the further the removed premises is located.
 
-                \addplot[
-                    color=blue,
-                    mark=square,
-                    ]
-                    coordinates {
-                    (1, 0.710963)(2, 1.266787)(4, 2.181111)(8, 4.220279)(16, 8.084744)(32, 15.683840)
-                    };
-                    \addlegendentry{Stateful}
-
-                \addplot[
-                    color=orange,
-                    mark=square,
-                    ]
-                    coordinates {
-                    (1, 0.872730)(2, 2.100940)(4, 7.245598)(8, 35.805959)(16, 222.283526)(32, 1600.978502)
-                    };
-                    \addlegendentry{Stateless}
-
-                \end{axis}
-                \end{tikzpicture}
-            \end{figure}
-
-            \begin{figure}
-                %Casacade with precompilation
-                    \begin{tikzpicture}
-                        \begin{axis}[
-                            title={Casacade Test},
-                            xlabel={Number of Rules},
-                            ylabel={Time in ms},
-                            xmin=0, xmax=32,
-                            ymin=0, ymax=850,
-                            ymode=log,xmode=log,
-                            log basis x=2,
-                            log basis y=2,
-                            legend pos=north west,
-                            ymajorgrids=true,
-                            grid style=dashed,
-                        ]
-
-                        \addplot[
-                            color=blue,
-                            mark=square,
-                            ]
-                            coordinates {
-                            (1, 0.393011)(2, 0.730952)(4, 1.327736)(8, 3.007961)(16, 10.173689)(32, 38.892027)
-                            };
-                            \addlegendentry{Stateful}
-
-                        \addplot[
-                            color=orange,
-                            mark=square,
-                            ]
-                            coordinates {
-                            (1, 0.464973)(2, 0.759516)(4, 1.865191)(8, 7.585588)(16, 64.374740)(32, 813.336527)
-                            };
-                            \addlegendentry{Stateless}
-
-                        \end{axis}
-                        \end{tikzpicture}
-                    \end{figure}
+The position of the missing premise has little effect on the stateful algorithm as the index selects the rule in all case; it only needs one of the premise to be present.
+The work the stateful algorithm has to do is thus close to constant in this scenario.
 
@@
 \end{document}
+
+% \begin{figure}
+%     % This is the erase without precompilation for 6 premises.
+%     % Note I have a decent idea that would allow us to compare for multiple
+%     % number of premises. 
+%     % (Divide by the average of the stateful time to normalize.)
+%     \begin{tikzpicture}
+%         \begin{axis}[
+%             title={Depth Test},
+%             xlabel={Depth},
+%             ylabel={Time in ms},
+%             xmin=0, xmax=6,
+%             ymin=50, ymax=65,
+%             xtick={0,1,2,3,4,5,6},
+%             ytick={50,55,60,65},
+%             legend pos=north west,
+%             ymajorgrids=true,
+%             grid style=dashed,
+%         ]
+
+%         \addplot[
+%             color=orange,
+%             mark=square,
+%             ]
+%             coordinates {
+%             (1,50.91)(2,53.14)(3,52.91)(4,51.22)(5,55.14)(6,56.60)
+%             };
+%             \addlegendentry{Stateless}
+
+%         \addplot[
+%             color=blue,
+%             mark=square,
+%             ]
+%             coordinates {
+%             (1,56.04)(2,56.43)(3,55.58)(4,55.26)(5,54.80)(6,60.61)
+%             };
+%             \addlegendentry{Stateful}
+
+%         \end{axis}
+%         \end{tikzpicture}
+%     \end{figure}
+
+%     \begin{figure}
+%         % This is the erase without precompilation for 6 premises.
+%         % Note I have a decent idea that would allow us to compare for multiple
+%         % number of premises. 
+%         % (Divide by the average of the stateful time to normalize.)
+%         \begin{tikzpicture}
+%             \begin{axis}[
+%                 title={Scaled Depth Test},
+%                 xlabel={Depth},
+%                 ylabel={Time in ms},
+%                 xmin=0, xmax=8,
+%                 ymin=0.5, ymax=2,
+%                 xtick={0,1,2,3,4,5,6,7,8},
+%                 ytick={0.75,1,1.25,1.5,1.75},
+%                 legend pos=north west,
+%                 ymajorgrids=true,
+%                 grid style=dashed,
+%             ]
+
+%             \addplot[color=orange,mark=square,]
+%                 coordinates {(0, 0.883600)(1, 1.039751)};
+%                 \addlegendentry{Stateless}
+%             \addplot[color=orange,mark=square,]
+%                 coordinates {(0, 0.804521)(1, 1.017893)(2, 1.022493)};
+%             \addplot[color=orange,mark=square,]
+%                 coordinates {(0, 0.791528)(1, 1.015467)(2, 1.076267)(3, 1.159999)};
+%             \addplot[color=orange,mark=square,]
+%                 coordinates {(0, 0.743256)(1, 0.966436)(2, 1.058876)(3, 1.133514)(4, 1.225561)};
+%             \addplot[color=orange,mark=square,]
+%                 coordinates {(0, 0.770625)(1, 1.003777)(2, 1.109929)(3, 1.215171)(4, 1.305373)(5, 1.390667)};
+%             \addplot[color=orange,mark=square,]
+%                 coordinates {(0, 0.759093)(1, 1.016667)(2, 1.148273)(3, 1.213138)(4, 1.342795)(5, 1.442271)(6, 1.514877)};
+%             \addplot[color=orange,mark=square,]
+%                 coordinates {(0, 0.709965)(1, 0.971294)(2, 1.068745)(3, 1.179333)(4, 1.312798)(5, 1.424054)(6, 1.551690)(7, 1.628344)};
+%             \addplot[color=orange,mark=square,]
+%                 coordinates {(0, 0.728860)(1, 1.001576)(2, 1.097472)(3, 1.219259)(4, 1.366683)(5, 1.444844)(6, 1.576972)(7, 1.668693)(8, 1.815399)};
+
+%             \end{axis}
+%             \end{tikzpicture}
+%         \end{figure}
+
+%                 \begin{figure}
+%                     %Casacade with precompilation
+%                         \begin{tikzpicture}
+%                             \begin{axis}[
+%                                 title={Casacade Test},
+%                                 xlabel={Number of Rules},
+%                                 ylabel={Time in ms},
+%                                 xmin=0, xmax=32,
+%                                 ymin=0, ymax=850,
+%                                 ymode=log,xmode=log,
+%                                 log basis x=2,
+%                                 log basis y=2,
+%                                 legend pos=north west,
+%                                 ymajorgrids=true,
+%                                 grid style=dashed,
+%                             ]
+
+%                             \addplot[
+%                                 color=blue,
+%                                 mark=square,
+%                                 ]
+%                                 coordinates {
+%                                 (1, 0.393011)(2, 0.730952)(4, 1.327736)(8, 3.007961)(16, 10.173689)(32, 38.892027)
+%                                 };
+%                                 \addlegendentry{Stateful}
+
+%                             \addplot[
+%                                 color=orange,
+%                                 mark=square,
+%                                 ]
+%                                 coordinates {
+%                                 (1, 0.464973)(2, 0.759516)(4, 1.865191)(8, 7.585588)(16, 64.374740)(32, 813.336527)
+%                                 };
+%                                 \addlegendentry{Stateless}
+
+%                             \end{axis}
+%                             \end{tikzpicture}
+%                         \end{figure}

### 2025-02-10 paper.tex
Minor edits in preliminaries

@@
 \begin{enumerate}
-  \item\label{cond:match:matches} For each slot $i$, if $M(i)$ is defined, then $M(i)$ matches $i$.
-  \item\label{cond:match:compat} For all slots $i$ and $j$ of $r$ such that $M(i)$ and $M(j)$ are defined, the substitutions $\sub_{i}(M(i))$ and $\sub_{j}(M(j))$ are \emph{compatible}, i.e.\ they agree on all variables that occur in both $i$ and $j$.
-  \item\label{cond:match:closed} $M$ is downward-closed: if $M(i)$ is defined, then for any slot $j < i$ of $r$, $M(j)$ is also defined.
+  \item\label{cond:match:matches} For each slot $i ∈ \dom(M)$, the hypothesis $M(i)$ matches $i$.
+  \item\label{cond:match:compat} For all slots $i,j ∈ \dom(M)$, the substitutions $\sub_{i}(M(i))$ and $\sub_{j}(M(j))$ are \emph{compatible}.
+        This means that for each variable $x ∈ \dom(\sub_{i}(M(i))) ∩ \dom(\sub_{j}(M(j)))$ we have $\sub_{i}(M(i))(x) = \sub_{j}(M(j))(x)$, i.e.\ the substitutions agree on all common variables.
+  \item\label{cond:match:closed} $M$ is \emph{downward-closed}: if $i ∈ \dom(M)$, then for any slot $j < i$ of $r$ we also have $j ∈ \dom(M)$.
 \end{enumerate}

### 2025-02-10 paper.tex
Edit naive forward reasoning section

@@
 \newunicodechar{⊥}{\ensuremath{\bot}}
+\newunicodechar{×}{\ensuremath{\times}}
 
@@
 \newcommand{\lvl}{\ensuremath{\mathrm{lvl}}}
+\newcommand{\Rules}{\ensuremath{\mathcal{R}}}
 
@@
 
-\section{Design}
-
-We now focus on the problem of constructing complete matches.
-
-\subsection{A naive design}
-
-A straightforward way to solve this problem is to simply test all the hypotheses present in the context against the registered rules.
-
-A few obvious improvements are then available.
-We can index the rules in terms of one of its hypothesis.
-This index can then be used to guarentee that at least one hypothesis in the context will match with a slot of the rule.
-A heuristic is to take the last slot as it should be more specific than the firsts slot of a rule.
-
-If the last slot of a rule matches a hypothesis in the context we can then test whether other hypotheses of the context can complete that rule.
-This is done in two steps, first we check whether the hypotheses in the context match the slot.
-If it does, we check if the substitution given by the unification is coherent with the preceding matched hypotheses.
-Using the same reasoning as above, we heuristically do this backwards in the slots.
-
-This particular design is fast when the final hypothesis of a rule is not present in the context. 
-However, it otherwise needs to test many potential hypothesis candidate which is quite slow.
-Another weakness of this approach is that in the eventuality that a new hypothesis appears in the goal, for example as the result of another rule being completed, all previous work done on one rule must be repeated with this new hypothesis.
-
-We call stateless this kind of approach to the problem as no previous work is saved.
-
-\subsection{An improved, stateful design}
+\section{Naive Forward Reasoning}
+
+Prior to this work, Aesop used what we call naive forward reasoning (TODO ref).
+This approach treats a forward rule essentially like any other Aesop rule.
+Hence, given a goal $h₁ : T₁,\, \dots, hₙ : Tₙ ⊢ U$, we first query an index to determine which rules might apply to the goal.
+For forward rules, this index stores the type of the last premise (since later premises tend to be more specific).
+The index is queried for each $Tᵢ$, yielding a set of rules that have a premise which likely unifies with at least one hypothesis.
+Of these rules and any non-forward rules selected in a similar way, we select the highest-priority one.
+
+If the selected rule is a forward rule $r : \All{(x₁ : A₁) \dots (xₘ : Aₘ)}{B}$, we first unify $Aₘ$ with one of the candidate hypotheses determined by the index.
+If successful, this results in a substitution $σ$ whose domain contains $xₘ$ and any variables occurring in $Aₘ$.
+We then proceed to unify $A_{m-1}[σ]$ with the type of each hypothesis in the context, etc., until we have found a complete match for $r$ (skipping premises already determined by $σ$).
+If a premise type $A_{i}[σ]$ does not unify with any hypothesis type, we backtrack and consider other potential matching hypotheses for $A_{i+1}$.
+This may lead to a different substitution $τ$ and therefore to matching hypotheses for $A_{i}[τ]$.
+
+If this process yields one or more full applications $r~a₁ \dots aₘ$, we add $B[x₁ ↦ a₁, \dots, xₘ ↦ aₘ]$ as a new hypothesis unless a hypothesis of this type is already present in the context.
+
+\jcom{Add example?}
+
+This approach is efficient when the last premise of a rule is not present in the context since the rule is then never run.
+If the last premise is present, however, we perform a large number of unifications between premises and hypotheses.
+Furthermore, these unifications are repeated when we proceed to another goal with a similar context.
+In particular, the contexts of subgoals produced by whatever rule is eventually applied to the current goal are likely to be substantially similar, if not identical, to the current context, but these similarities are not exploited.
+
+\jcom{Mention fast subsumption paper?}
+
+\section{Incremental Forward Reasoning}
 
@@
 
-\subsection{Deduplication}
-
-\xcom{precompilation}
+\subsection{Match Redundancy}
+
+\subsection{Hypothesis Redundancy}
+
+\jcom{Mention here that we also use this for the naive algorithm.}
+
+\subsection{Precompilation}
 

### 2025-02-11 paper.tex
wip eval
Intro to incremental algorithm and plan for reorg

@@
 \newcommand{\Rules}{\ensuremath{\mathcal{R}}}
+\newcommand{\Hyps}{\ensuremath{\mathcal{H}}}
+\newcommand{\States}{\ensuremath{\mathcal{S}}}
+\newcommand{\CMatches}{\ensuremath{\mathcal{C}}}
+\newcommand{\addHyp}{\ensuremath{\mathrm{addHyp}}}
+\newcommand{\delHyp}{\ensuremath{\mathrm{delHyp}}}
+\newcommand{\powerset}{\ensuremath{\mathbb{P}}}
 
@@
 Furthermore, these unifications are repeated when we proceed to another goal with a similar context.
-In particular, the contexts of subgoals produced by whatever rule is eventually applied to the current goal are likely to be substantially similar, if not identical, to the current context, but these similarities are not exploited.
+In particular, the contexts of subgoals produced by whatever rule is eventually applied to the current goal are likely to be substantially similar, if not identical, to the current context, but these similarities are not exploited since the naive algorithm is stateless.
 
@@
 
-With the weaknesses of the last approach in mind, we now consider a \textit{stateful} solution to the problem of finding complete matches.
-The main idea is to save partial progress on complete matches.
+With the weaknesses of the naive approach in mind, we introduce an incremental, stateful solution to the problem of finding complete matches.
+The main idea is to store, for each goal $Γ ⊢ A$, the partial matches implied by the hypotheses in $Γ$, and to reuse these partial matches for subgoals $Δ ⊢ B$.
+This rests on the assumption that $Δ$ is likely similar to $Γ$ in practice, and that it is therefore cheaper to update the partial matches for $Γ$ to obtain those of $Δ$ than to compute those of $Δ$ from scratch.
+Our main contribution is a data structure, the \emph{partial match state}, that allows us to perform these updates efficiently.
+We first discuss the interface of this data structure and its integration into Aesop, then its construction.
+
+\subsection{Interface}
+
+We associate to each goal in Aesop's search tree a partial match state that reflects the goal's context.
+Let $\States$ be the set of forward match states, $\Hyps$ the set of hypotheses, $\Rules$ the set of rules and $\CMatches$ the set of complete matches.
+The forward match state's interface then consists of two functions:
+\begin{align*}
+  \addHyp &: \States → \Hyps → \States × \powerset(\Rules × \CMatches) \\
+  \delHyp &: \States → \Hyps → \States
+\end{align*}
+The $\addHyp$ function adds a hypothesis to the partial match state.
+Hence, if $s$ is the partial match state corresponding to context $Γ$ and $\addHyp(s, h : T) = (s', M)$, then $s'$ is the partial match state corresponding to $Γ,\, h : T$.
+Similarly, if $s$ is the partial match state corresponding to $Γ, h : T$, then $\delHyp(s, h : T)$ is the partial match state corresponding to $Γ$.
+The second output of $\addHyp$, $M$, is a set of pairs $(r, m)$ where $r$ is a rule and $m$ is a complete match for $r$.
+These are the matches that were completed by adding $h : T$.
+
+When Aesop is run on an initial goal with context $h₁ : T₁, \dots, hₙ : Tₙ$, the partial match state for it is built by starting with an empty state and applying $\addHyp$ to each hypothesis $hᵢ$.
+Afterwards, whenever an Aesop rule is run on a goal with context $Γ$, producing subgoals with contexts $Δ₁, \dots, Δₙ$, we derive the partial match state for each subgoal context $Δⱼ$ from the partial match state for $Γ$, in two steps:
+\begin{enumerate}
+  \item Apply $\delHyp$ once for each hypothesis that was deleted, i.e.\ that appears in $Γ$ but not in $Δⱼ$.
+  \item Apply $\addHyp$ once for each hypothesis that was added, i.e.\ that appears in $Δⱼ$ but not in $Γ$.
+\end{enumerate}
+Changes to a hypothesis (e.g.\ a change of its type) are treated as a deletion followed by an addition.
+
+\subsection{Partial Match State}%
+\label{sec:state}
+
+\jcom{partial match state is composed of rule states}
+
+\jcom{rule states contain variable maps}
+
+\subsection{Variable Clusters}%
+\label{sec:clusters}
+
+\jcom{=== Initial draft follows}
 
@@
 The subsets of premises of $r$ that share a dependency are precisely the connected components of this graph.
-We call these subsets of premises \textit{cluters}.
+We call these subsets of premises \textit{clusters}.
 
@@
 
-\subsection{Aesop Integration}
-
-\xcom{Incrementality?}
+\subsection{Goal Diffs}
+
+\jcom{Remove this whole section as too low-level?}
 In Lean, the type of hypotheses can sometimes change while the hypothesis' identifier remains the same.

### 2025-02-11 paper.tex
rephrase

@@
 The procedures are then run on a context containing exactly $\mathcal{P}$.
-
 In this case, we sketch the behavior of the stateless algorithm:
@@
 If $j < 6$, the index returns nothing and it selects another hypothesis.
-For $h_{i,6} : P_{i,6}$, the rule $r_i$ is selected and it then needs to try and fill its remaining five slot with the context's $6n$ hypotheses which it tries all.
+For $h_{i,6} : P_{i,6}$, the rule $r_i$ is selected, and it then needs to try and fill its remaining five slot with the context's $6n$ hypotheses which it tries all.
 This gives a complexity of about $O(n^2)$.

### 2025-02-12 paper.tex
add possible todo
New title
xavier orcidID
Add abstract, modify keywords

@@
 
-\title{Stateful Forward Reasoning}
+\title{Incremental Forward Reasoning for White-Box Proof Search}
 
@@
 \author{Jannis Limperg\orcidID{0000-0002-8861-5231} \and
-  Xavier Généreux\orcidID{TODO}}
+  Xavier Généreux\orcidID{0000-0003-4952-9557}}
 
@@
 \begin{abstract}
-  TODO
-
-  \keywords{Forward Reasoning \and Proof Automation \and Theorem Proving.}
+  Many proof assistants provide automation tactics based on simple, tableau-style tree search, such as Isabelle's and Rocq's auto or Lean's Aesop.
+  These tactics solve a given goal by recursively applying \emph{rules}, which return zero or more subgoals.
+  In most cases, the rules themselves can be arbitrary tactics, so they can return any set of subgoals corresponding to an admissible inference in the prover's logic.
+
+  In this setting, we consider \emph{forward rules}, which apply a given theorem, say $A → B → C$, to any goal containing hypotheses $A$ and $B$, adding $C$ as an additional hypothesis.
+  When treated naively, such rules are tried on every goal encountered during the search, leading to repeated unifications of premises $A$ and $B$ with the hypotheses of each goal.
+  Much of this work is redundant since goals often have similar hypotheses.
+
+  We present an implementation of forward rules that takes advantage of these similarities.
+  For each goal, we cache certain partial applications of forward rules in a custom data structure, the \emph{partial match state}.
+  This enables efficient updates, so if goals have similar hypotheses, we can quickly update the partial match state rather than re-running all forward rules.
+
+  Our technique is compatible with any search strategy and most logics.
+  It has been implemented in the Aesop proof search tactic for Lean.
+
+  \keywords{Forward Reasoning \and Forward Chaining \and Tactics \and Interactive Theorem Proving.}
 \end{abstract}
@@
 
-In many proof assistants, simple automation tactics based on tree search play a substantial role in practice: Isabelle's (TODO ref) and Rocq's (TODO ref) auto, ACL2's (TODO ref) waterfall, PVS's (TODO ref) grind, Lean's (TODO ref) Aesop (TODO ref), etc.
+In many proof assistants, simple automation tactics based on tableau-style tree search play a substantial role in practice: Isabelle's (TODO ref) and Rocq's (TODO ref) auto, ACL2's (TODO ref) waterfall, PVS's (TODO ref) grind, Lean's (TODO ref) Aesop (TODO ref), etc.
 These tactics operate on \emph{goals} $Γ ⊢ A$, where the \emph{local context} $Γ$ is a list of assumptions and $A$ is the proposition to be proved.
@@
 
+\xcom{Do we want to add potential improvements? Like an index closer to the naive algorithm.}
+
 TODO

### 2025-02-12 paper.tex
Explain powerset notation

@@
 
-\title{Incremental Forward Reasoning for White-Box Proof Search}
+\title{Incremental Forward Reasoning for White-Box~Proof~Search}
 
@@
 \authorrunning{J.~Limperg and X.~Généreux}
-\institute{University of Munich (LMU), Munich, Germany\\
+\institute{Ludwig-Maximilians-Universität München, Munich, Germany\\
 \email{\{jannis.limperg,xavier.genereux\}@lmu.de}}
@@
 \begin{abstract}
-  Many proof assistants provide automation tactics based on simple, tableau-style tree search, such as Isabelle's and Rocq's auto or Lean's Aesop.
-  These tactics solve a given goal by recursively applying \emph{rules}, which return zero or more subgoals.
-  In most cases, the rules themselves can be arbitrary tactics, so they can return any set of subgoals corresponding to an admissible inference in the prover's logic.
-
-  In this setting, we consider \emph{forward rules}, which apply a given theorem, say $A → B → C$, to any goal containing hypotheses $A$ and $B$, adding $C$ as an additional hypothesis.
+  Several proof assistants provide automation tactics based on tableau-style tree search, such as Isabelle's and Rocq's auto and Lean's Aesop.
+  We consider \emph{forward rules}, which apply a given theorem, say, $A → B → C$, to any goal containing hypotheses $A$ and $B$, adding $C$ as a hypothesis.
   When treated naively, such rules are tried on every goal encountered during the search, leading to repeated unifications of premises $A$ and $B$ with the hypotheses of each goal.
-  Much of this work is redundant since goals often have similar hypotheses.
-
-  We present an implementation of forward rules that takes advantage of these similarities.
-  For each goal, we cache certain partial applications of forward rules in a custom data structure, the \emph{partial match state}.
-  This enables efficient updates, so if goals have similar hypotheses, we can quickly update the partial match state rather than re-running all forward rules.
-
+  We present an approach to forward rules that takes advantage of similarities between successive goals.
+  For each goal, we cache partial applications of forward rules in a custom data structure, the \emph{partial match state}, which enables efficient updates.
   Our technique is compatible with any search strategy and most logics.
-  It has been implemented in the Aesop proof search tactic for Lean.
+  It has been implemented in Aesop.
 
@@
 \end{abstract}
+
+% \begin{abstract}
+%   Many proof assistants provide automation tactics based on simple, tableau-style tree search, such as Isabelle's and Rocq's auto or Lean's Aesop.
+%   These tactics solve a given goal by recursively applying \emph{rules}, which return zero or more subgoals.
+%   In most cases, the rules themselves can be arbitrary tactics, so they can return any set of subgoals corresponding to an admissible inference in the prover's logic.
+
+%   In this setting, we consider \emph{forward rules}, which apply a given theorem, say $A → B → C$, to any goal containing hypotheses $A$ and $B$, adding $C$ as an additional hypothesis.
+%   When treated naively, such rules are tried on every goal encountered during the search, leading to repeated unifications of premises $A$ and $B$ with the hypotheses of each goal.
+%   Much of this work is redundant since goals often have similar hypotheses.
+
+%   We present an implementation of forward rules that takes advantage of these similarities.
+%   For each goal, we cache certain partial applications of forward rules in a custom data structure, the \emph{partial match state}.
+%   This enables efficient updates, so if goals have similar hypotheses, we can quickly update the partial match state rather than re-running all forward rules.
+
+%   Our technique is compatible with any search strategy and most logics.
+%   It has been implemented in the Aesop proof search tactic for Lean.
+
+%   \keywords{Forward Reasoning \and Forward Chaining \and Tactics \and Interactive Theorem Proving.}
+% \end{abstract}
 
@@
 We associate to each goal in Aesop's search tree a partial match state that reflects the goal's context.
-Let $\States$ be the set of forward match states, $\Hyps$ the set of hypotheses, $\Rules$ the set of rules and $\CMatches$ the set of complete matches.
-The forward match state's interface then consists of two functions:
+Let $\States$ be the set of partial match states, $\Hyps$ the set of hypotheses, $\Rules$ the set of rules, $\CMatches$ the set of complete matches and $\powerset(X)$ the powerset of $X$.
+The partial match state's interface then consists of two functions:
 \begin{align*}

### 2025-02-13 paper.tex
Abstract proposal

@@
 We associate to each goal in Aesop's search tree a partial match state that reflects the goal's context.
-Let $\States$ be the set of partial match states, $\Hyps$ the set of hypotheses, $\Rules$ the set of rules, $\CMatches$ the set of complete matches and $\powerset(X)$ the powerset of $X$.
-The partial match state's interface then consists of two functions:
+Let $\States$ be the set of forward match states, $\Hyps$ the set of hypotheses, $\Rules$ the set of rules and $\CMatches$ the set of complete matches.
+The forward match state's interface then consists of two functions:
 \begin{align*}

### 2025-02-13 paper.tex
Some clarifications in abstract

--- 2025-02-12 paper.tex
@@
   Several proof assistants provide automation tactics based on tableau-style tree search, such as Isabelle's and Rocq's auto and Lean's Aesop.
-  We consider \emph{forward rules}, which apply a given theorem, say, $A → B → C$, to any goal containing hypotheses $A$ and $B$, adding $C$ as a hypothesis.
+  In this setting we consider \emph{forward rules}, which apply a given theorem, say, $A → B → C$, to any goal containing hypotheses $A$ and $B$, adding $C$ as a new hypothesis.
   When treated naively, such rules are tried on every goal encountered during the search, leading to repeated unifications of premises $A$ and $B$ with the hypotheses of each goal.
-  We present an approach to forward rules that takes advantage of similarities between successive goals.
+  We present an approach to forward rules that avoids some of this repeated work by taking advantage of similarities between successive goals.
   For each goal, we cache partial applications of forward rules in a custom data structure, the \emph{partial match state}, which enables efficient updates.
@@
 \end{abstract}
-
-% \begin{abstract}
-%   Many proof assistants provide automation tactics based on simple, tableau-style tree search, such as Isabelle's and Rocq's auto or Lean's Aesop.
-%   These tactics solve a given goal by recursively applying \emph{rules}, which return zero or more subgoals.
-%   In most cases, the rules themselves can be arbitrary tactics, so they can return any set of subgoals corresponding to an admissible inference in the prover's logic.
-
-%   In this setting, we consider \emph{forward rules}, which apply a given theorem, say $A → B → C$, to any goal containing hypotheses $A$ and $B$, adding $C$ as an additional hypothesis.
-%   When treated naively, such rules are tried on every goal encountered during the search, leading to repeated unifications of premises $A$ and $B$ with the hypotheses of each goal.
-%   Much of this work is redundant since goals often have similar hypotheses.
-
-%   We present an implementation of forward rules that takes advantage of these similarities.
-%   For each goal, we cache certain partial applications of forward rules in a custom data structure, the \emph{partial match state}.
-%   This enables efficient updates, so if goals have similar hypotheses, we can quickly update the partial match state rather than re-running all forward rules.
-
-%   Our technique is compatible with any search strategy and most logics.
-%   It has been implemented in the Aesop proof search tactic for Lean.
-
-%   \keywords{Forward Reasoning \and Forward Chaining \and Tactics \and Interactive Theorem Proving.}
-% \end{abstract}
 

### 2025-02-14 paper.tex
change orcid format
Edit incremental forward reasoning section

@@
 \documentclass[runningheads]{llncs}
+
+% TODO: standardise indices vs indexes
+% TODO: standardise dots after equations
+% TODO: rename variable maps to variable indices everywhere
+% TODO: standardise Rocq vs Coq
 
@@
 \usepackage{csquotes}
-\usepackage{algorithm2e}
+\usepackage[noend]{algorithm2e}
 
@@
 \urlstyle{rm}
-\SetKw{Kwin}{in}
-
+\def\orcid#1{{\href{http://orcid.org/#1}{\protect\raisebox{-1.25pt}{\protect\includegraphics{orcid.pdf}}}}}
+\SetKw{KwIn}{in}
+\SetKw{KwLet}{let}
+\SetKw{KwYield}{yield}
+\SetKwInOut{Input}{Input}
+\SetKwInOut{Output}{Output}
+\SetKwProg{Procedure}{Procedure}{}{}
+
+\newunicodechar{₀}{\ensuremath{_{0}}}
 \newunicodechar{₁}{\ensuremath{_{1}}}
@@
 \newcommand{\delHyp}{\ensuremath{\mathrm{delHyp}}}
+\newcommand{\addMatch}{\ensuremath{\mathrm{addMatch}}}
 \newcommand{\powerset}{\ensuremath{\mathbb{P}}}
@@
 
-\author{Jannis Limperg\orcidID{0000-0002-8861-5231} \and
-  Xavier Généreux\orcidID{0000-0003-4952-9557}}
+\author{Jannis Limperg\orcid{0000-0002-8861-5231} \and
+  Xavier Généreux\orcid{0000-0003-4952-9557}}
 
@@
 If $h : A$ is a hypothesis in the local context of a goal and $A$ unifies with the type $B$ of slot $i$, we say that $h$ \emph{matches} $i$.
-Unification induces a substitution $σ$ (with $\dom(σ) ⊆ \vars(i)$) such that $A[σ]$ is definitionally equal to $B[σ]$; we call this substitution $\sub_{i}(h)$.
+Unification induces a substitution $σ$ such that $A[σ]$ is definitionally equal to $B[σ]$; we call this substitution $\sub_{i}(h)$.
 For now we do not specify exactly what \enquote{unifies} means; see Sec.~\ref{sec:unification}.
@@
 Our main contribution is a data structure, the \emph{partial match state}, that allows us to perform these updates efficiently.
-We first discuss the interface of this data structure and its integration into Aesop, then its construction.
-
-\subsection{Interface}
-
+
+\paragraph{Partial Match State.}
 We associate to each goal in Aesop's search tree a partial match state that reflects the goal's context.
@@
 
-\subsection{Partial Match State}%
-\label{sec:state}
-
-\jcom{partial match state is composed of rule states}
-
-\jcom{rule states contain variable maps}
-
-\subsection{Variable Clusters}%
-\label{sec:clusters}
-
-\jcom{=== Initial draft follows}
-
-\paragraph{Indexing Scheme.}
-This partial progress could be simply saved as trees with root the matched hypothesis returned by the index.
-However, we decided to move towards an index that looks at all the hypotheses of the rule.
-The idea is that we can better track and save information about which hypotheses will potentially complete which rules.
-This makes it particularly easy to maintain partial matches which, as we will see, are classified in terms of the instantiation of their set of shared variables.
-Another advantage is that we only need to look at each of the context's hypotheses once.
-This makes the procedure incremental, which is part of the stateful philosophy.
-
-In other words, the indexing is now a discrimination tree that maps types $T$ to sets of pairs $(r,i)$ where $r : (x_1 : T_1) \dots (x_n : T_n) \to \Psi$ is a rule, $i$ is a slot of $r$ and $T$ is likely unifies with $T_i$.
-
-Given hypotheses returned by this index, the next challenge comes down to finding a complete match of a rule.
-This means identifying compatible hypotheses within the hypotheses returned by the index.
-
-\paragraph{Matches.}
-Compatible hypotheses are hypotheses such that the instantiation of their shared variables are the same.
-For example, the hypotheses $h_1 : a \leq b$ and $h_2 : b \leq c$ are compatible because they are instantiations of $x \leq y$ and $y \leq z$ where the instantiation of their shared variable $\{y\}$, is the same $\{b\}$.
-A complete match is a set of hypotheses such that each of these hypothesis corresponds to exactly one slot and they are all compatible.
-
-Before we look at how to find complete matches, we examine exactly which premises depend on each other and for which we need to insure compatibility.
-Given a rule, we will partition its premises into subsets for which premises share a dependency.
-To be precise, consider the graph for which the input premises $T_i$ of $r$ are the vertices and whose edges are precisely the pairs of vertices $(T_i, T_j)$ for which $v(T_i) \cap v(T_j) \neq \emptyset$.
-The subsets of premises of $r$ that share a dependency are precisely the connected components of this graph.
-We call these subsets of premises \textit{clusters}.
-
-%I think, if we have space, we could also add the following characterization :
-%Another characterization is the partition with the property that for two parts $C_1$ and $C_2$, $\bigcup_{T_i \in C_1} v(T_i) \cap \bigcup_{T_i \in C_2} v(T_i) = \emptyset$, and for which there do not exists a partition with the same property that has more parts.
-% Do we want to prove this?
-
-One particularity of clusters is that, for two clusters $C_1$ and $C_2$,
-$\bigcup_{T_i \in C_1} v(T_i) \cap \bigcup_{T_i \in C_2} v(T_i) = \emptyset$.
-This means that compatibility need only be considered within clusters as there are no shared variables between distinct clusters.
-In other words, premises in a cluster are independent of the premises in the other clusters.
-Consequently, in what follows, we will assume without loss of generality that there is only one cluster.
-The process we follow would simply be repeated on each cluster.
-
-We proceed in the following way to find a complete match:
-First, we order the slots such that each slot has at least one variable in common with the previous slots.
-This can be proven to always be possible.
-(Again, we assume that there is only one cluster.)
-\xcom{We can add the proof if we want to / have space.}
-
-We then aim to construct matches iteratively: we consider each hypotheses one by one, assign them to their respective slots and check whether there exist compatible matches which they can extend.
-Whenever we find an hypothesis that fits with the premise in slot one, we create a new (level one) match.
-This process, that we will introduce in more details shortly, will be refered to as Algorithm \ref{alg:addhyp}.
-On the other hand, whenever a match is updated (or created), we also look through the hypotheses we have already came accross, this is Algorithm \ref{alg:updatematch}.
-
-The compatibility check means that all the shared variables of the considered hypotheses need to be compatible.
-It would be costly to do this comparison every time and indeed we choose instead to maintain what we call \textit{variable maps} whose purpose is to save the current partial matches in terms of their instantiation.
-Each variable map is associated to one shared variable, these are the variables that dictate compatibility.
-
-\paragraph{Variable Maps.}
-Let $x$ ba a shared variable, $t$ an expression that represents an instantiation of that variable and $i$ an integer associated to a slot.
-We define two variable maps $\mu_x$ and $\eta_x$ that respectively take the pair $(i,t)$ to $\mathbb{M}$ and $\mathbb{H}$ where
-\begin{itemize}
-  \item $\mathbb{M}$ is the set that contains exactly those partial matches which already contain assignments for all slots up to $i$ and which instantiate $?x$ with $t$.
-  \item $\mathbb{H}$ is the set of hypotheses which match $T_i$ while instantiating $?x$ with $t$.
-\end{itemize}
-
-Given a hypothesis $h : T$ that matches $T_i$ and $\sigma_h$, the map sending the variable of $T_i$ to their specific instantiation for $h$.
-The compatibility check that looks for suitable partial matches is simply computed by
+\paragraph{Rule State.}
+Partial matches are specific to each rule, so the partial match state naturally decomposes into a set of \emph{rule states}, one per rule.
+Given a goal $Γ ⊢ A$ and a rule $r$, the rule state for $r$ contains exactly the partial matches of $r$ in context $Γ$.
+A rule state supports $\addHyp$ and $\delHyp$ methods analogous to those on partial match states.
+Adding a hypothesis to a partial match state then amounts to adding it to each rule state (which may yield new complete matches for the rule); deleting a hypothesis amounts to deleting it from each rule state.
+We initially focus on how to add hypotheses, since this question motivates much of the design of the rule state data structure.
+
+When a hypothesis is added to a rule state, two things can happen.
+First, the hypothesis may allow us to extend partial matches already present in the state.
+For instance, consider the rule $r : \All{x}{P~x → Q~x → R~x → S~x}$ and a rule state containing the partial match $\{1 ↦ h₁\}$, with $1$ referring to the slot of type $P~x$ and $h₁ : P~a$ for some $a$.
+Adding a new hypothesis $h₂ : Q~a$ to this rule state means that we should add to the state the extended match $\{1 ↦ h₁, 2 ↦ h₂\}$.
+
+Second, adding a match may trigger the insertion of further matches.
+Suppose that before adding $h₁$ and $h₂$, we had already added a hypothesis $h₀ : R~a$ to the rule state.
+Once the partial match $\{1 ↦ h₁, 2 ↦ h₂\}$ arrives, we should recognise that it can be extended with $h₀$ to obtain the (complete) match $\{1 ↦ h₁, 2 ↦ h₂, 3 ↦ h₀\}$.
+
+\paragraph{Variable Indices.}
+Extending the matches above was possible only because all hypotheses are compatible on the shared variable $x$, i.e.\ they all instantiate $x$ with the same term $a$.
+In general, to extend a match $M$ with a hypothesis $h : T$ at slot $i$, we need to ensure that the substitution of $M$, $\sub(M)$, is compatible with the substitution of $h$, $\sub_{i}(h)$.
+
+The rule state therefore uses two \emph{variable indices} $H$ and $M$ to index hypotheses and matches by their instantiations for shared variables, allowing us to quickly determine compatible pairs of matches and hypotheses.
+We could use a substitution tree (TODO ref, confirm) for this, but our custom data structure is simpler and perhaps more efficient.
+The \emph{hypothesis index} $H$ maps each triple $(x, i, t)$ (where $i > 1$ is a slot, $x ∈ \vars(i)$ a variable and $t$ a term) to the set of hypotheses $h$ added to the rule state such that $h$ matches $i$ and $\sub_{i}(h)(x) = t$.
+In other words, $H$ indexes the known hypotheses by the slots they match and the instantiations of shared variables.
+Note that hypotheses matching slot $1$ are not stored since they can never extend a match.
+
+Similarly, the \emph{match index} $M$ maps each triple $(x, i, t)$ to the set of partial matches $m$ (that can be derived from the hypotheses added to the rule state) such that $i = \lvl(m) $, $x ∈ \vars(i + 1)$ and $\sub(m)(x) = t$.
+Matches for the maximal slot are not stored because, being complete, they never need to be extended.
+
+Given a hypothesis $h : T$ that matches slot $i > 1$ with substitution $σ = \sub_{i}(h)$, the partial matches that can be extended by $h$ are those in slot $i - 1$ whose substitutions are compatible with $σ$ on the variables $\vars(i)$ (i.e., the variables shared between $i$ and any earlier slot).
+Assuming for the moment that $\vars(i)$ is nonempty for all slots $i > 1$, these matches are
 \[
-  \bigcap_{x \in s(i)} \mu_x \left(i - 1, \sigma_h(x)\right).
+  \bigcap_{x \in \vars(i)} M(x, i-1, σ(x))
 \]
-
-Similarly, given a match $m$ of level $i$ and $\sigma_m$ the substitution map of the relevant  variable in $m$ then we can check for compatible hypotheses that would extend $m$ by computing
+Similarly, given a partial match $m$ at level $i$ with substitution $σ = \sub(m)$, the hypotheses that could extend $m$ are
 \[
-  \bigcap_{x \in s(i + 1)} \eta_x \left(i + 1, \sigma_m(x)\right).
+  \bigcap_{x \in \vars(i+1)} H(x, i+1, σ(x))
 \]
-
-With this in mind, Algorithm \ref{alg:addhyp} and \ref{alg:updatematch} simply maintain these maps up to date and performs compatibility checks by computing the above intersection of sets.
+The number of shared variables tends to be small in practice, so these intersections can be computed efficiently.
+
+From these observations we can directly derive two procedures for adding hypotheses and matches  to the rule state.
+The procedures maintain the invariants of the variable indices while reporting any complete matches discovered during this process.
 
 \begin{algorithm}
-    \caption{New hypothesis}\label{alg:addhyp}
-    \KwData{A hypothesis $h : T$ that matches with $T_i$.}
-    \KwResult{The maps $\eta$ and $\mu$ are updated.}
+  \SetKwInOut{Precondition}{Precondition}
+  \SetKwInOut{Postcondition}{Postcondition}
+  \Precondition{$H$ and $M$ contain the hypotheses and matches corresponding to a context $Γ$}
+  \Input{A hypothesis $h : T$ matching slot $i$ of $r$ with substitution $σ$}
+  \Output{Complete matches of $r$ in context $Γ,\, h : T$ that contain $h$}
+  \Postcondition{$H$ and $M$ contain the hypotheses and matches corresponding to $Γ,\, h : T$}
+  \BlankLine
+  \Procedure{$\mathrm{addHyp}(h : T)$}{
     \eIf{$i = 1$}{
-      \For{$x$ \Kwin $s(i)$}{
-        Update $\mu_x$ with $(1,\sigma_m(x)) \mapsto \mu_x(1,\sigma_m(x))\cup\{[h]\}$ (a level 1 match)
-      }
-      Extend match $[h]$
+      $\mathrm{addMatch}(\{1 ↦ h\})$
     }{
-      \For{$x$ \Kwin $s(i)$}{
-        Update $\eta_x$ with $(i,\sigma_h(x)) \mapsto \eta_x(i,\sigma_h(x))\cup\{h\}$
-      }
-      $ms \gets \bigcap_{x \in s(i)} \mu_x \left(i - 1, \sigma_h(x)\right)$ \\
-      \For{$m$ \Kwin $ms$}{
-        $m' \gets m.\text{append } h$ \\
-        \For{$x$ \Kwin $s(i)$}{
-          Update $\mu_x$ with $(i,\sigma_m(x)) \mapsto \mu_x(i,\sigma_m(x))\cup\{m'\}$
-        }
-        Extend match $m'$
-      }
-    }
+      \For{$x$ \KwIn $\vars(i)$}{
+        Insert $h$ into the set $H(x, i, σ(h))$
+      }
+      \KwLet $\mathit{ms} ≔ \bigcap_{x \in \vars(i)} M(x, i-1, σ(x))$ \\
+      $\mathit{completeMs} \gets ∅$ \\
+      \For{$m$ \KwIn $\mathit{ms}$}{
+        $\mathit{completeMs} \gets \mathit{completeMs} ∪ \mathrm{addMatch}(m ∪ \{i ↦ h\})$
+      }
+      \Return{$\mathit{completeMs}$}
+    }
+  }
+  \BlankLine % without these, the caption is not offset properly
+  \BlankLine
+  \caption{Adding a hypothesis to the rule state for a rule $r$}%
+  \label{alg:addhyp}
 \end{algorithm}
 
+The rule state's $\addHyp$ method (Algorithm~\ref{alg:addhyp}) adds a hypothesis $h : T$ matching slot $i$ to the rule state.
+(If a hypothesis matches multiple slots, $\addHyp$ is called multiple times.)
+For $i = 1$, we add the singleton match $\{1 ↦ h\}$ to the rule state.
+For $i > 1$, we first insert $h$ into the hypothesis index $H$.
+Then we query the match index $M$ for matches that can be extended by $h$ and add the extended matches to the rule state.
+While adding matches, we may discover new complete matches, which are collected and returned.
+
+% TODO pre- and postcondition?
 \begin{algorithm}
-    \caption{Extend match}\label{alg:updatematch}
-    \KwData{A partial match $m$ of level $i$.}
-    \KwResult{The map $\mu$ is updated.}  
-        $hs \gets \bigcap_{x \in s(i + 1)} \eta_x \left(i + 1, \sigma_m(x)\right)$ \\
-        \For{$h$ \Kwin $hs$}{
-            $m' \gets m.\text{append } h$\\
-            \For{$x$ \Kwin $s(i + 1)$}{%
-              Update $\mu_x$ with $(i + 1,\sigma_{m'}(x)) \mapsto \mu_x(i + 1,\sigma_{m'}(x))\cup\{m'\}$.%
-            }
-            Extend match $m'$.
-        }
+  \Input{A partial match $m$ of $r$ at level $i$ with substitution $σ$}
+  \Output{A set of complete matches of $r$}
+  \BlankLine
+  \Procedure{$\mathrm{addMatch}(h : T)$}{
+    \eIf{$i$ is complete}{
+      \Return{$\{m\}$}
+    }{
+      \For{$x$ \KwIn $\vars(i + 1)$}{
+        Insert $m$ into the set $M(x, i, σ(x))$
+      }
+      \KwLet $\mathit{hs} ≔ \bigcap_{x \in \vars(i + 1)} H(x, i + 1, σ(x))$ \\
+      $\mathit{completeMs} \gets ∅$ \\
+      \For{$h$ \KwIn $\mathit{hs}$}{
+        $\mathit{completeMs} \gets \mathit{completeMs} ∪ \mathrm{addMatch}(m ∪ \{i + 1 ↦ h\})$
+      }
+      \Return{$\mathit{completeMs}$}
+    }
+  }
+  \BlankLine % without these, the caption is not offset properly
+  \BlankLine
+  \caption{Adding a match to the rule state for a rule $r$}%
+  \label{alg:addmatch}
 \end{algorithm}
 
-In sum, given a hypothesis $h :T$ we
-\begin{enumerate}
-\item Save it in the variable maps $\eta_x$, for the relevant shared variables $x$.
-\item If it matches slot 1, we create a level one match.
-\item Look if we can extend existing partial matches using $h$.
-\item For any modified partial match, we look whether there exists suitable hypotheses that can extend them and repeat in case of success.
-\end{enumerate}
+The rule state's $\addMatch$ method (Algorithm~\ref{alg:addmatch}) adds a match $m$ with level $i = \lvl(m)$ to the rule state.
+If $m$ is already complete, it is returned.
+Otherwise $m$ is inserted into $M$ and we query $H$ for hypotheses that can extend $m$.
+Any extended matches thus produced are added to the rule state recursively.
+
+\paragraph{Variable Clusters.}
+We have so far assumed that for each slot $i > 1$, the set of shared variables $\vars(i)$ is nonempty, i.e.\ each slot shares at least one variable with at least one of the previous slots.
+This assumption is violated for rules such as $\All{x~y}{P~x → Q~y → R~x~y}$: slot 2 ($Q~y$) shares no variables with slot 1 ($P~x$).
+As a result, our above algorithms fail.
+Suppose we are adding the hypothesis $h₂ : Q~b$, having already added the hypothesis $h₁ : P~a$ and hence the partial match $\{1 ↦ h₁\}$.
+The set of matches compatible with $h₂$ is supposedly $\bigcap_{x \in \vars(2)} M(x, i-1, σ(x)) = ∅$, but in fact the match $\{1 ↦ h₁\}$ can be extended with $h₂$.
+In fact, any match at level 1 can be extended with any hypothesis matching slot 2 since the two slots are independent.
+
+To account for this fact, we partition the set of premises of each rule into maximal subsets of premises connected by shared variables, using a construction from (TODO ref).
+These subsets, called \emph{variable clusters}, share no variables with each other, so each can be treated as a separate rule with its own rule state.
+Finding a complete match for a rule amounts to finding one complete match for each of its clusters.
+The slots within each cluster can be ordered to ensure that each slot shares at least one variable with the previous slots, satisfying our assumption.
 
 \paragraph{Example 1.}
-
 \xcom{Technically, it is useless to save level 1 hyps in $\eta_x$. Should we remove them from the examples?.}
@@
 An update in slot 1 also marks the begining of a new partial match and we update de match map $\mu_y$ acordingly.
-Note that after the creation of a match of level 1, we look at the hypotheses in slot 2 and check if they can extend the match to level 2 (Algorithm \ref{alg:updatematch}).
+Note that after the creation of a match of level 1, we look at the hypotheses in slot 2 and check if they can extend the match to level 2 (Algorithm \ref{alg:addmatch}).
 However, the instantiation of the current hypotheses in slot 2 are inconsistent with $y \mapsto 302$, so nothing can be extended.
 Indeed,
-\[
+\[ %\textstyle - \nolimits 
   \bigcap_{x\in s(lvl(m_0) + 1)} \eta_x \left(i + 1, \sigma_{m_0}(x)\right) = \eta_y(2, 302) = \emptyset.
@@
 
-\section{Implementation}
+\section{Implementation}%
+\label{sec:implementation}
 
@@
 This makes our technique incomplete in the sense that some theorem instantiations will be missed, but since the same unification algorithm is used throughout Lean, users are at least well-acquainted with this source of incompleteness.
+
+\jcom{TODO: mvars in hyps are treated as rigid}
 
@@
 \subsection{Indexing}
+
+\jcom{TODO adjust}
+
+This partial progress could be simply saved as trees with root the matched hypothesis returned by the index.
+However, we decided to move towards an index that looks at all the hypotheses of the rule.
+The idea is that we can better track and save information about which hypotheses will potentially complete which rules.
+This makes it particularly easy to maintain partial matches which, as we will see, are classified in terms of the instantiation of their set of shared variables.
+Another advantage is that we only need to look at each of the context's hypotheses once.
+This makes the procedure incremental, which is part of the stateful philosophy.
+
+In other words, the indexing is now a discrimination tree that maps types $T$ to sets of pairs $(r,i)$ where $r : (x_1 : T_1) \dots (x_n : T_n) \to \Psi$ is a rule, $i$ is a slot of $r$ and $T$ is likely unifies with $T_i$.
 
@@
 \jcom{auto2}
+\jcom{Mention ACL2 forward rule triggers}
 

### 2025-02-14 paper.tex
Add a paragraph about deletion

@@
 \newunicodechar{≤}{\ensuremath{\leq}}
+\newunicodechar{≥}{\ensuremath{\geq}}
 \newunicodechar{≔}{\ensuremath{\coloneq}}
@@
 \paragraph{Example 1.}
-\xcom{Technically, it is useless to save level 1 hyps in $\eta_x$. Should we remove them from the examples?.}
+\xcom{Technically, it is useless to save level 1 hyps in $\eta_x$. Should we remove them from the examples?.} \jcom{Yes, since the previous section now contains this optimisation.}
 To demonstrate the typical behavior of the algorithm we consider the following example which has \textbf{Lemma 1} as a rule and for which we add, in order, the hypotheses $h_1 : 301 \leq 302$ and $h_2 : 300 \leq 301$.
@@
 
-\jcom{TODO: deletion}
+\paragraph{Deletion.}
+When a hypothesis $h$ is removed from a goal's context, we must update the partial match state accordingly.
+This means deleting $h$ from any rule state to which it was previously added, i.e.\ from any rule state for a rule $r$ with a slot $i$ matching $h$.
+To facilitate deletion, we cache this information when $h$ is added to the rule state.
+When $h$ is deleted, we can then efficiently determine the relevant rule states and delete $h$ from the hypothesis index $H$ at slot $i$ and from the match index $M$ at all slots $j ≥ i$.
 

### 2025-02-14 paper.tex
Editing in implementation; mention rigid mvars in hyps

@@
 Unification induces a substitution $σ$ such that $A[σ]$ is definitionally equal to $B[σ]$; we call this substitution $\sub_{i}(h)$.
-For now we do not specify exactly what \enquote{unifies} means; see Sec.~\ref{sec:unification}.
+For now we do not specify exactly what \enquote{unifies} means; see Sec.~\ref{sec:implementation}.
 
@@
 
-\subsection{Unification}%
-\label{sec:unification}
-
+\paragraph{Unification.}
 In Lean's type theory, types contain programs and unification is expected to respect definitional equality, i.e.\ equality up to evaluation.
@@
 \enquote{Reducible} is the most restrictive of these, so we unfold few (and, in practice, only non-recursive) definitions, which greatly reduces the cost of unification.
-It also limits the power of forward reasoning since
+However, this limits the power of forward reasoning since fewer terms can be unified.
 
 We also need to account for higher-order unification problems: $Q~\mvar{f}$, where $Q$ is a predicate over functions and $\mvar{f}$ is a function variable, should unify with $Q~(\Lam{x}{x})$.
-Since higher-order unification is undecidable and does not admit most general unifiers (TODO ref), we use instead a unification algorithm provided by Lean that is complete for first-order unification problems and heuristically solves some higher-order problems as well.
-This makes our technique incomplete in the sense that some theorem instantiations will be missed, but since the same unification algorithm is used throughout Lean, users are at least well-acquainted with this source of incompleteness.
-
-\jcom{TODO: mvars in hyps are treated as rigid}
-
-\subsection{Data Representation}
-
-Since the maps $μ_{x}$  and $η_{x}$ have the same domain, we may combine them into one map.
-This core data structure is then represented by nested persistent hash maps: a first hash map maps variables to a second hash map, the second hash map maps slots to a third hash map and the third hash map maps instantiations to a set of hypotheses and a set of partial matches.
+Since higher-order unification is undecidable and does not admit most general unifiers (TODO ref), we instead use a unification algorithm provided by Lean that is complete for first-order unification problems and heuristically solves some higher-order problems as well.
+This makes our technique incomplete in the sense that some theorem instantiations will be missed, but since the same unification algorithm is used throughout Lean, users are well acquainted with this form of incompleteness.
+
+During unification, we treat metavariables in hypotheses as constants, i.e.\ we never assign them.
+This reflects the expected behaviour of forward rules: given a hypothesis $\mvar{x} < \mvar{y}$, we would not expect a forward rule to commit to the choice $\mvar{x} ≔ 1$ and $\mvar{y} ≔ 0$ just because this allows the rule to derive a contradiction.
+
+\paragraph{Data Representation.}
+Since the hypothesis index $H$ and the variable index $M$ of a rule state have the same domain, we combine them into one map.
+This core data structure is then represented by nested persistent hash maps: a first hash map maps variables to a second hash map; the second hash map maps slots to a third hash map; and the third hash map maps instantiations to sets of hypotheses and sets of partial matches.
 
@@
 Instead, we bring all instantiations into a normal form which we call \emph{reducible proof-irrelevant normal form} (RPINF).
-This is the usual normal form of expressions with respect to Lean's notion of computation, except for two modifications:
-\begin{itemize}
-  \item Only \emph{reducible} defined constants are unfolded, i.e.\ replaced with their definiendum.
+This is the usual normal form of expressions with respect to Lean's notion of computation, with two modifications.
+First, only \emph{reducible} defined constants are unfolded, i.e.\ replaced with their definiendum.
 This reflects our choice to perform unification at reducible transparency.
-  \item Parts of the original expression that are proofs (in the sense that their types are members of the universe $\Prop$ of propositions) are not normalised.
+Second, parts of the original expression that are proofs (in the sense that their types are members of the universe $\Prop$ of propositions) are not normalised.
 This reflects the fact that Lean's type theory definitionally equates any two proofs of the same proposition, so there is no need to check whether they are equal.
 When computing the RPINF, we also mark all proofs occurring in it with some special \enquote{metadata}.
-\end{itemize}
-
-With this setup, two expressions $t$ and $u$ are definitionally equal at reducible transparency if their RPINFs, $t'$ and $u'$, are syntactically equal except for those portions of $t'$ and $u'$ that are proofs, which are always considered equal.
-Since this is an entirely syntactic notion of equality, it can be checked very quickly and we can define an effective hash function that is compatible with it.
+
+With this setup, two expressions $t$ and $u$ are definitionally equal at reducible transparency if and only if their RPINFs, $t'$ and $u'$, are syntactically equal except for those portions of $t'$ and $u'$ that are proofs, which are always considered equal.
+Since this is an entirely syntactic notion of equality, it can be checked quickly and we can define an effective hash function that is compatible with it.
 This allows us to use a hash map for the instantiations while still comparing them up to definitional equality.
@@
 However, since these discrimination trees are imperfect, they may return spurious lookup results, so whether a lookup result is valid must be verified by unifying the query expression with the expression that was originally stored in the discrimination tree.
-We believe that repeatedly performing such unifications would be, on average, more expensive than calculating one RPINF (which can be cached) for each type in a goal.
-
-\subsection{Goal Diffs}
-
-\jcom{Remove this whole section as too low-level?}
-In Lean, the type of hypotheses can sometimes change while the hypothesis' identifier remains the same.
-This means that one needs to track the types of the hypotheses as well.
-We call this tracking \textit{goal diffs}, they highlight the differences between old and updated goal states.
-The reference to hypotheses which changing types get deleted from the maps and the slot.
-We then add them again as a new hypothesis.
-
-\subsection{Indexing}
-
+We believe that repeatedly performing such unifications would generally be more expensive than calculating the RPINF of each hypothesis type (which can be cached).
+
+\paragraph{Indexing.}
 \jcom{TODO adjust}
-
 This partial progress could be simply saved as trees with root the matched hypothesis returned by the index.
@@
 
-In other words, the indexing is now a discrimination tree that maps types $T$ to sets of pairs $(r,i)$ where $r : (x_1 : T_1) \dots (x_n : T_n) \to \Psi$ is a rule, $i$ is a slot of $r$ and $T$ is likely unifies with $T_i$.
-
-\subsection{Match Redundancy}
-
-\subsection{Hypothesis Redundancy}
-
+In other words, the indexing is now a discrimination tree that maps types $T$ to sets of pairs $(r,i)$ where $r : (x_1 : T_1) \dots (x_n : T_n) \to \Psi$ is a rule, $i$ is a slot of $r$ and $T$ likely unifies with $T_i$.
+
+\paragraph{Match Redundancy.}
+TODO
+
+\paragraph{Hypothesis Redundancy.}
 \jcom{Mention here that we also use this for the naive algorithm.}
-
-\subsection{Precompilation}
 

### 2025-02-14 paper.tex
Fix comment commands
Finish implementation draft

@@
 
-\newcommand{\xcom}[1]{{\color{cyan}{Xavier: #1}} }
-\newcommand{\jcom}[1]{{\color{orange}{Jannis: #1}} }
+\newcommand{\xcom}[1]{{\textcolor{cyan}{Xavier: #1}} }
+\newcommand{\jcom}[1]{{\textcolor{orange}{Jannis: #1}} }
 
@@
 \paragraph{Indexing.}
-\jcom{TODO adjust}
-This partial progress could be simply saved as trees with root the matched hypothesis returned by the index.
-However, we decided to move towards an index that looks at all the hypotheses of the rule.
-The idea is that we can better track and save information about which hypotheses will potentially complete which rules.
-This makes it particularly easy to maintain partial matches which, as we will see, are classified in terms of the instantiation of their set of shared variables.
-Another advantage is that we only need to look at each of the context's hypotheses once.
-This makes the procedure incremental, which is part of the stateful philosophy.
-
-In other words, the indexing is now a discrimination tree that maps types $T$ to sets of pairs $(r,i)$ where $r : (x_1 : T_1) \dots (x_n : T_n) \to \Psi$ is a rule, $i$ is a slot of $r$ and $T$ likely unifies with $T_i$.
-
-\paragraph{Match Redundancy.}
-TODO
-
-\paragraph{Hypothesis Redundancy.}
-\jcom{Mention here that we also use this for the naive algorithm.}
+When adding a hypothesis $h : T$ to the partial match state, we add it to the rule state of each rule $r$ such that $h$ matches a slot of $r$.
+To optimise this operation, we use an imperfect discrimination tree index that, for each rule $r$ with slot $i$ corresponding to a premise of type $Tᵢ$, maps the type $Tᵢ$ to the pair $(r, i)$.
+Querying this index for $T$ quickly determines all rules and slots for which $h$ is likely relevant.
+
+\paragraph{Match Equivalence.}
+Naively, two matches $\{1 ↦ h₁, \dots, n ↦ hₙ\}$ and $\{1 ↦ h₁', \dots, n ↦ hₙ'\}$ would be considered equal if they are structurally equal, i.e.\ $h₁ = h₁'$, \dots, $hₙ = hₙ'$.
+However, consider the rule $r : \All{x~y}{P~x~y → Q~x~z → R~x}$ and the hypotheses $h₁ : P~a~b$, $h₂ : P~a~c$, $h₃ : Q~a~d$ and $h₄ : Q~a~e$.
+The complete matches $m₁ ≔ \{1 ↦ h₁, 2 ↦ h₃\}$ and $m₁' ≔ \{1 ↦ h₁, 2 ↦ h₄\}$ are structurally different, but the difference does not matter because the applications $r~h₁~h₃$ and $r~h₁~h₄$ both have type $R~a$.
+Generating both $m₁$ and $m₁'$ would therefore be redundant.
+Similarly, the partial matches $m₂ ≔ \{1 ↦ h₁\}$ and $m₂' ≔ \{1 ↦ h₂\}$ are redundant because the choice between $h₁$ and $h₂$ affects neither the type of any eventual application derived from $m₂$ and $m₂'$ nor the hypotheses that can be used to complete $m₂$ and $m₂'$.
+Hence we should store only one of $m₂$ and $m₂'$ in the rule state of $r$.
+Based on these observations, we consider two matches $m$ and $m'$ of a rule $r$ with conclusion $A$ equal if $\lvl(m) = \lvl(m')$ and for each variable $x$ that appears either in the conclusion of $r$ or in the type of a slot $i$ of $r$ with $i > \lvl(m)$ we have $\sub(m)(x) = \sub(m')(x)$.
+To ensure that we respect definitional equality, the instantiations in the substitutions are kept in RPINF.
+
+\paragraph{Redundant Hypotheses.}
+When we find a complete match, we add a new hypothesis of the corresponding type $T$ to the goal, but only if there is no other hypothesis of type $T$ already present.
+To efficiently implement this check, we store in the forward match state a set containing the RPINFs of all hypotheses that have been added (and not subsequently removed) from the state.
+We modified the naive forward reasoning implementation to use the same mechanism.
 

### 2025-02-14 paper.tex
update graphs

@@
     \label{fig:trans}
-    % Transitivity with precompilation.
+    % Transitivity without precompilation.
     % 
@@
             ylabel={Time in ms},
-            xmin=2, xmax=6,
+            xmin=1, xmax=16,
             ymin=2^3, ymax=2^13,
-            xtick={2,3,4,5,6},
+            xtick={1,2,4,8,16},
             ytick={2^3,2^5,2^7,2^9,2^11,2^13},
+            xmode=log,
             ymode=log,
-            %xmode=log,
+            log basis x=2,
             log basis y=2,
@@
             grid style=dashed,
+            legend image post style={mark=},
         ]
@@
             coordinates {
-                (2, 74.566944) (3, 444.481222) (4, 811.149916) (5, 2955.663569) (6, 4703.832541) 
+              (1, 7.663472) (2, 9.354750) (4, 15.768194) (8, 104.921791) (16, 1882.683111)
             };
@@
             coordinates {
-                (2, 13.089680) (3, 12.556333) (4, 14.278986) (5, 17.385902) (6, 20.682083)
+              (1, 10.308541) (2, 10.704638) (4, 16.274305) (8, 38.800833) (16, 179.697681) 
             };
             \addlegendentry{Stateful}
+            \addplot[
+              color=orange,
+              mark=o,
+              ]
+              coordinates {
+                (1, 10.591138) (2, 23.881333) (4, 142.013389) (8, 1676.148375) (16, 24327.359417)
+              };
+          \addplot[
+              color=blue,
+              mark=o,
+              ]
+              coordinates {
+                (1, 13.974694) (2, 26.216833) (4, 68.953444) (8, 245.601028) (16, 1028.670680) 
+              };
         \end{axis}
@@
 \begin{figure}
-    %Indep with precompilation
+    %Indep without precompilation
         \begin{tikzpicture}
@@
                 ylabel={Time in ms},
-                xmin=0, xmax=32,
-                ymin=0, ymax=1700,
+                xmin=1, xmax=32,
+                ymin=2^3, ymax=1700,
                 ymode=log,xmode=log,
@@
                 grid style=dashed,
-            ]
-
+                legend image post style={mark=},
+            ]
+
+            \addplot[
+              color=orange,
+              mark=square,
+              ]
+              coordinates {
+                (1, 10.399958) (2, 16.426541) (4, 43.461403) (8, 205.595541) (16, 1238.090889) (32, 8927.857819) 
+              };
+              \addlegendentry{Stateless}
             \addplot[
@@
                 coordinates {
-                (1, 0.710963)(2, 1.266787)(4, 2.181111)(8, 4.220279)(16, 8.084744)(32, 15.683840)
+                  (1, 11.231875) (2, 12.879500) (4, 16.851930) (8, 24.666485) (16, 42.054055) (32, 76.369778) 
                 };
@@
               color=orange,
-                mark=square,
+              mark=o,
               ]
               coordinates {
-                (1, 0.872730)(2, 2.100940)(4, 7.245598)(8, 35.805959)(16, 222.283526)(32, 1600.978502)
-                };
-                \addlegendentry{Stateless}
+                (1, 19.405222) (2, 33.487472) (4, 80.107597) (8, 284.954791) (16, 1536.089624) (32, 10721.729861)
+              };
+            \addplot[
+                color=blue,
+                mark=o,
+                ]
+                coordinates {
+                  (1, 25.892069) (2, 40.031930) (4, 69.446902) (8, 130.640986) (16, 260.653917) (32, 481.827902)
+                };
             \end{axis}
@@
     \label{fig:erase}
-%erase with precompilation
+%erase without precompilation
     \begin{tikzpicture}
@@
             ylabel={Time in ms},
-            xmin=0, xmax=6,
-            ymin=3, ymax=7,
-            xtick={0,1,2,3,4,5,6},
-            ytick={3,4,5,6,7},
+            xmin=1, xmax=6,
+            ymin=6, ymax=20,
+            xtick={1,2,3,4,5,6},
+            ytick={10,15},
             legend pos=north west,
@@
             grid style=dashed,
-        ]
-
+            legend image post style={mark=},
+        ]
+        \addplot[
+            color=orange,
+            mark=square,
+            ]
+            coordinates {
+              (1, 6.407973) (2, 9.831676) (3, 10.301718) (4, 10.626144) (5, 11.261754) (6, 11.795240) 
+            };
+            \addlegendentry{Stateless}
         \addplot[
@@
           coordinates {
-            (1, 3.715813)(2, 3.893547)(3, 3.856956)(4, 3.722684)(5, 3.714804)(6, 3.900508)
+            (1, 15.835558) (2, 16.961994) (3, 16.424259) (4, 17.605449) (5, 15.691536) (6, 14.982820) 
           };
           \addlegendentry{Stateful}
-        \addplot[
-            color=orange,
-            mark=square,
-            ]
-            coordinates {
-            (1, 2.993757)(2, 3.585232)(3, 3.880661)(4, 4.098640)(5, 4.322570)(6, 5.765420)
-            };
-            \addlegendentry{Stateless}
         \end{axis}

### 2025-02-14 paper.tex
Add a note about term orders

@@
 
+Requirement~\ref{cond:match:closed} implies that we will be adding hypotheses to matches in the order in which the slots appear.
+Hence for a rule $r : A → B → C → D$, we effectively consider partial applications $r~a$ and $r~a~b$ for some $a : A$ and $b : B$, but not $r~\text{--}~b$.
+This prevents us from constructing $r~a~b$ in two different but ultimately redundant ways.
+The term orders used in various saturation calculi serve a similar function.
+
 \section{Naive Forward Reasoning}

### 2025-02-14 paper.tex
Fix in implementation

@@
 When we find a complete match, we add a new hypothesis of the corresponding type $T$ to the goal, but only if there is no other hypothesis of type $T$ already present.
-To efficiently implement this check, we store in the forward match state a set containing the RPINFs of all hypotheses that have been added (and not subsequently removed) from the state.
+To efficiently implement this check, we store in the forward match state a set containing the RPINFs of all hypotheses that have been added to (and not subsequently removed from) the state.
 We modified the naive forward reasoning implementation to use the same mechanism.

### 2025-02-14 paper.tex
Save a line
Move deletion before examples
Edit first example

@@
 Let $\States$ be the set of partial match states, $\Hyps$ the set of hypotheses, $\Rules$ the set of rules, $\CMatches$ the set of complete matches and $\powerset(X)$ the powerset of $X$.
-The partial match state's interface then consists of two functions:
-\begin{align*}
-  \addHyp &: \States → \Hyps → \States × \powerset(\Rules × \CMatches) \\
-  \delHyp &: \States → \Hyps → \States
-\end{align*}
+The partial match state's interface then consists of two functions
+\begin{equation*}
+  \addHyp : \States → \Hyps → \States × \powerset(\Rules × \CMatches) \qquad \delHyp : \States → \Hyps → \States
+\end{equation*}
 The $\addHyp$ function adds a hypothesis to the partial match state.
@@
 
+\paragraph{Deletion.}
+When a hypothesis $h$ is removed from a goal's context, we must update the partial match state accordingly.
+This means deleting $h$ from any rule state to which it was previously added, i.e.\ from any rule state for a rule $r$ with a slot $i$ matching $h$.
+To facilitate deletion, we cache this information when $h$ is added to the rule state.
+When $h$ is deleted, we can then efficiently determine the relevant rule states and delete $h$ from the hypothesis index $H$ at slot $i$ and from the match index $M$ at all slots $j ≥ i$.
+
 \paragraph{Example 1.}
-\xcom{Technically, it is useless to save level 1 hyps in $\eta_x$. Should we remove them from the examples?.} \jcom{Yes, since the previous section now contains this optimisation.}
-To demonstrate the typical behavior of the algorithm we consider the following example which has \textbf{Lemma 1} as a rule and for which we add, in order, the hypotheses $h_1 : 301 \leq 302$ and $h_2 : 300 \leq 301$.
-Note that this order is choosed only for demonstration purposes.
-
-At the beginning, we feed $h_1$ to Algorithm \ref{alg:addhyp}.
-Furthmore, for the rule \textit{le\_trans}, we notice that the only shared variable is ${y}$.
-Now the hypothesis $h_1 : 301 \leq 302$ unifies both with slot 1: $x \leq y$ and slot 2: $y \leq z$.
-Starting with slot 1, this updates the map $\eta_y(1, y \mapsto 302)$ with $\{h_1\}$.
-An update in slot 1 also marks the begining of a new partial match and we update de match map $\mu_y$ acordingly.
-Note that after the creation of a match of level 1, we look at the hypotheses in slot 2 and check if they can extend the match to level 2 (Algorithm \ref{alg:addmatch}).
-However, the instantiation of the current hypotheses in slot 2 are inconsistent with $y \mapsto 302$, so nothing can be extended.
-Indeed,
-\[ %\textstyle - \nolimits 
-  \bigcap_{x\in s(lvl(m_0) + 1)} \eta_x \left(i + 1, \sigma_{m_0}(x)\right) = \eta_y(2, 302) = \emptyset.
+As a first example, consider a rule state for the transitivity rule $r : \All{x~y~z}{x ≤ y → y ≤ z → x ≤ z}$ to which we add, in order, the hypotheses
+$h₁ : 301 ≤ 302$ and $h₂ : 300 ≤ 301$.
+The rule has two slots of types $x ≤ y$ and $y ≤ z$ which share the variable $y$.
+
+The hypothesis $h₁ : 301 ≤ 302$ unifies with both slots, so we run Algorithm~\ref{alg:addhyp} twice.
+For slot 1, unification yields the substitution $\{x ↦ 301, y ↦ 302\}$.
+Since this is the first slot, we start a new partial match $\{1 ↦ h₁\}$, which Algorithm~\ref{alg:addmatch} inserts into the set $M(y, 1, 302)$.
+This match cannot be extended since the hypothesis index $H$ does not contain a hypothesis for slot 2.
+
+For slot 2, unification yields the substitution $σ ≔ \{y ↦ 301, z ↦ 302\}$, so we insert $h₁$ into $H(2, y, 301)$.
+No matches at level 1 can be extended by $h₁$ since
+\[
+  \bigcap_{x \in \vars(2)} M(x, 1, σ(x)) = M(y, 1, 301) = \emptyset
 \]
-Finally, as the hypothesis also matches slot 2, we update the map $\eta_y(2, y \mapsto 301)$ with $\{h_1\}$.
-At this point the intersection
-\[
-  \bigcap_{x\in s(2)} \eta_x \left(1, \sigma_{h}(x)\right) = \mu_y(1, 301) = \emptyset
-\]
-and at this point the algorithm is done with $h_1$ and stops.
-After these updates, the maps have the following state :
-
+At this point, $H$ and $M$ are in the following state.
+For brevity, we write matches $\{1 ↦ h₁, \dots, n ↦ hₙ\}$ as $[h₁, \dots, hₙ]$.
+We also write $H_{y}$ instead of $H(y, \text{--}, \text{--})$, and similar for $M$.
 \begin{center}
 \begin{tikzpicture}[outer sep=auto, level distance=12mm, scale=0.50]
-  \node[xshift = 0] {$\eta_y$}
-    child[xshift = -20] {node {1}
-      child {node {$302$}
-        child {node {$\{h_1\}$}}}
-    %  child {node {$301$}
-    %    child {node {$\{\}$}}}
-        }
-    child[xshift = 20] {node {2}
+  \node[xshift = 0] {$H_{y}$}
+    child {node {2}
       child {node {$301$}
-        child {node {$\{h_1\}$}}}
+        child {node {$\{h₁\}$}}}
     %  child {node {$300$}
@@
     };
-  \node[xshift=160] {$\mu_y$}
-    child {node[xshift=-20] {1}
+  \node[xshift=160] {$M_y$}
+    child {node {1}
       child {node {$302$}
-        child {node {$\{[h_1]\}$}}}
+        child {node {$\{[h₁]\}$}}}
     %  child {node {$301$}
@@
 
-The next step is to add $h_2 : 300 \leq 301$ to the maps.
-Again, it unifies with both slots 1 and 2; we start by updating the map $\eta_y(1, y \mapsto 301)$ with $\{h_2\}$.
-Again, an update in slot 1 causes the creation of a level 1 match.
-We then check whether there are suitable hypotheses in slot 2 with
-with a compatible instantiation of $y$; indeed we find $\eta_y(2, y \mapsto 301) = \{h_1\}$.
-This update the match map with the level 2 match $[h_2,h_1]$.
-This is a complete match.
-The new hypothesis $h_3 : 300 \leq 302$ is added to the context.
-
-The hypothesis $h_2$, also matches with slot 2.
-This causes the update $\eta_y(2, y \mapsto 300)$ with $\{h_2\}$. 
-At this point, we need to check for level 1 that would be compatible, there are none and the work with this hypothesis is done.
-
+Now we add $h₂ : 300 ≤ 301$, which again unifies with both slots.
+For slot 1, $\sub₁(h₂)(y) = 301$, so we add the match $\{1 ↦ h₂\}$ to $M(y, 1, 301)$.
+We then check for hypotheses in slot 2 that also instantiate $y$ with $301$, and indeed we find $M(2, y, 301) = \{h₁\}$.
+Hence we return the complete match $\{1 ↦ h₂, 2 ↦ h₁\}$, which induces a new hypothesis $h₃ : 300 ≤ 302$ to be added later.
+
+For slot 2, we add $h₂$ to $H(y, 2, 300)$.
+There are no compatible matches at level 1, so we are done with $h₂$, yielding the following final rule state.
 \begin{center}
     \begin{tikzpicture}[outer sep=auto, level distance=12mm, scale=0.50]
-      \node[xshift = 0] {$\eta_y$}
-        child[xshift = -20] {node {1}
-          child {node {$302$}
-            child {node {$\{h_1\}$}}}
+      \node[xshift = 0] {$H_y$}
+        child {node {2}
           child {node {$301$}
-            child {node {$\{h_2\}$}}}
-            }
-        child[xshift = 20] {node {2}
-          child {node {$301$}
-            child {node {$\{h_1\}$}}}
+            child {node {$\{h₁\}$}}}
           child {node {$300$}
-            child {node {$\{h_2\}$}}}
-        };
-      \node[xshift=160] {$\mu_y$}
-        child {node[xshift=-20] {1}
-          child[xshift=-10] {node {$302$}
-            child {node {$\{[h_1]\}$}}}
-          child[xshift=10] {node {$301$}
-            child {node {$\{[h_2]\}$}}}
-        }
-        child {node[xshift=20] {2}
-          child {node {$301$}
-            child {node {$\{[h_2,h_1]\}$}}}};
+            child {node {$\{h₂\}$}}}
+        };
+      \node[xshift=160] {$M_y$}
+        child {node {1}
+          child[xshift=-20] {node {$302$}
+            child {node {$\{[h₁]\}$}}}
+          child[xshift=20] {node {$301$}
+            child {node {$\{[h₂]\}$}}}
+        };
     \end{tikzpicture}
@@
 We have found two compatible hypotheses for our rule and we can then add the result $R$ as hypothesis to the context.
-
-\paragraph{Deletion.}
-When a hypothesis $h$ is removed from a goal's context, we must update the partial match state accordingly.
-This means deleting $h$ from any rule state to which it was previously added, i.e.\ from any rule state for a rule $r$ with a slot $i$ matching $h$.
-To facilitate deletion, we cache this information when $h$ is added to the rule state.
-When $h$ is deleted, we can then efficiently determine the relevant rule states and delete $h$ from the hypothesis index $H$ at slot $i$ and from the match index $M$ at all slots $j ≥ i$.
 

### 2025-02-14 paper.tex
Edit example 2

@@
 \paragraph{Example 2.}
-We now consider another example with more than one shared variable.
-Let $T_1, T_2$ be two proposition that both depend on the variable $x$ and $y$.
-For simplicity, the rule considered will be $T_1(x,y) \to T_2(x,y) \to R$.
-The hypotheses added to the context will be, in order, $h_1 : T_2(a,a)$, $h_2 : T_1(a,b)$ and $h_3 : T_2(a,b)$.
-First, $h_1 : T_2(a,a)$ is considered :
-
+We now consider a rule $s : \All{x~y}{A~x~y → B~x~y → C~x~y}$ with multiple shared variables.
+We add hypotheses $h₁ : B~a~a$, $h₂ : A~a~b$ and $h₃ : B~a~b$ to the rule state of $s$, starting with $h₁$:
 % NOTE : remove comments to get full trees.
+\begin{center}
   \begin{tikzpicture}[level distance=12mm, scale=0.50]
-    \node[xshift = 30] {$\eta_x$}
+    \node[xshift = 30] {$H_x$}
         %child[xshift = -20] {node {1}
@@
         %    child {node {$\{ \}$}}}}
-        child[xshift = 40] {node {2}
+        child {node {2}
         child {node {$a$}
@@
         };
-    \node[xshift = 110] {$\eta_y$}
+    \node[xshift = 110] {$H_y$}
         %child[xshift = -20] {node {1}
@@
         %        child {node {$\{ \}$}}}}
-        child[xshift = 40] {node {2}
+        child {node {2}
             child {node {$a$}
@@
         };
-    \node[xshift = 190, yshift = -25] {$\mu_x$};
+    \node[xshift = 190] {$M_x$};
         %child[xshift = -20] {node {1}
@@
         %};
-    \node[xshift = 270, yshift = -25] {$\mu_y$};
+    \node[xshift = 270] {$M_y$};
         %child[xshift = -20] {node {1}
@@
   \end{tikzpicture}
-
-At this point, there are no matches so there is nothing to extend.
-We add the next hypothesis in the queue, that is $h_2 : T_1(a,b)$, to the variable maps :
-
+\end{center}
+At this point there are no matches, so there is nothing to extend.
+We now add $h_2 : A~a~b$, which initially generates a match $m = \{1 ↦ h₂\}$ that is added to $M$:
 % NOTE : uncomment to get full trees.
+\begin{center}
   \begin{tikzpicture}[outer sep=auto, level distance=12mm, scale=0.50]
-    \node[xshift = 0] {$\eta_x$}
-        child[xshift = -20] {node {1}
-            child {node {$a$}
-                child {node {$\{h_2\}$}}}
-            %child {node {$b$}
-            %    child {node {$\{ \}$}}}
-            }
-        child[xshift = 20] {node {2}
+    \node[xshift = 0] {$H_x$}
+        child {node {2}
             child {node {$a$}
@@
             };
-    \node[xshift = 80] {$\eta_y$}
-        child[xshift = -20] {node {1}
-            %child {node {$a$}
-            %    child {node {$\{ \}$}}}
-            child {node {$b$}
-                child {node {$\{ h_2\}$}}}}
-        child[xshift = 20] {node {2}
+    \node[xshift = 80] {$H_y$}
+        child {node {2}
             child {node {$a$}
@@
                 };
-    \node[xshift = 160, yshift = 0] {$\mu_x$}
-        child[xshift = -20] {node {1}
+    \node[xshift = 160, yshift = 0] {$M_x$}
+        child {node {1}
             child {node {$a$}
@@
         %};
-    \node[xshift = 240, yshift = 0] {$\mu_y$}
-        child[xshift = -20] {node {1}
+    \node[xshift = 240, yshift = 0] {$M_y$}
+        child {node {1}
         %    child {node {$a$}
@@
   \end{tikzpicture}
-
-As this hypothesis is in the first slot, we create the corresponding level 1 partial match $m_0 = {(h_2 : T_1(x, y))}$ for which we check for suitable hypotheses that could extend the match.
-Note that  $s(lvl(m_0) + 1) = s(2)$ which is the intersection of the variables appearing in the slots $i < 2$ with the variables appearing in slot $2$.
-Thus $s(2) = \{x,y\}$ and we compute:
+\end{center}
+To find hypotheses compatible with $m$, we consider the set of variables $\vars(\lvl(m) + 1)$.
+These are the variables shared between slot $\lvl(m) + 1 = 2$ and all previous slots, i.e.\ $x$ and $y$.
+Hence the compatible hypotheses are
 \[
-  \bigcap_{x\in s(lvl(m_0) + 1)} \eta_x \left(i + 1, \sigma_{m_0}(x)\right) = \eta_x(2, a) \cap \eta_y(2, b) = \emptyset
+  \bigcap_{x \in \vars(\lvl(m) + 1)} H(x, i + 1, \sub(m)(x)) = H(x, 2, a) \cap H(y, 2, b) = \emptyset
 \]
-Since the set is empty, the match can not be currently extended.
-We turn to the remaining hypothesis $h_3 : T_2(a,b)$ in the queue and add it to the maps:
-
+This means the match cannot currently be extended.
+We now turn to the remaining hypothesis $h_3 : B~a~b$, adding it first to $H$:
+\begin{center}
   \begin{tikzpicture}[outer sep=auto, level distance=12mm, scale=0.50]
-\node[xshift = 0] {$\eta_x$}
-    child[xshift = -20] {node {1}
-        child {node {$a$}
-            child {node {$\{h_2\}$}}}
-        %child {node {$b$}
-        %    child {node {$\{ \}$}}}
-        }
-    child[xshift = 20] {node {2}
+    \node[xshift = 0] {$H_x$}
+      child {node {2}
           child {node {$a$}
@@
           };
-\node[xshift = 80] {$\eta_y$}
-    child[xshift = -20] {node {1}
-        %child {node {$a$}
-        %    child {node {$\{ \}$}}}
-        child {node {$b$}
-            child {node {$\{h_2\}$}}}}
-    child[xshift = 20] {node {2}
+    \node[xshift = 80] {$H_y$}
+      child {node {2}
           child {node {$a$}
@@
               child {node {$\{h_3\}$}}}};
-    \node[xshift = 160, yshift = 0] {$\mu_x$}
-        child[xshift = -20] {node {1}
+    \node[xshift = 160, yshift = 0] {$M_x$}
+      child {node {1}
           child {node {$a$}
@@
           %   child {node {$\{ \}$}}}
-                }
-        child[xshift = 20] {node {2}
-            child {node {$a$}
-                child {node {$\{[h_2,h_3]\}$}}}
-        %    child {node {$b$}
-        %        child {node {$\{ \}$}}}
-        };
-    \node[xshift = 240, yshift = 0] {$\mu_y$}
-        child[xshift = -20] {node {1}
+              };
+    \node[xshift = 240, yshift = 0] {$M_y$}
+      child {node {1}
       %    child {node {$a$}
@@
               child {node {$\{[h_2]\}$}}}
-        }
-        child[xshift = 20] {node {2}
-        %    child {node {$a$}
-        %        child {node {$\{\}$}}}
-            child {node {$b$}
-                child {node {$\{[h_2,h_3]\}$}}}
       };
   \end{tikzpicture}
-
-We are now ready to look for potential matches that could be extended by $h_3$.
-In this particular case, since $h_3$ is in slot $2$ we look for $s(2) = \{x,y\}$.
-Then,
+\end{center}
+Since $h₃$ matches slot $i = 2$, the matches that can be extended by $h₃$ are
 \[
-\bigcap_{x\in s(2)} \mu_x \left(i - 1, \sigma_h(x)\right) = \mu_x(1, a) \cap \mu_y(1, b) = \{m_0\}
+  \bigcap_{x \in \vars(i)} M(x, i - 1, \sub_{i}(x)) = M(x, 1, a) \cap M(y, 1, b) = \{[h₂]\}
 \]
-
-We have found two compatible hypotheses for our rule and we can then add the result $R$ as hypothesis to the context.
+We have found the complete match $[h₂, h₃]$ and can add a new hypothesis of type $C~a~b$.
 

### 2025-02-14 paper.tex
Compress related work

@@
 Among white-box proof automation tactics, only ACL2's \enquote{waterfall} (TODO ref) has special support for forward reasoning.
-(Some such tactics, e.g.\ Coq's \texttt{auto}, can execute arbitrary tactics including forward rules, but make no special provisions for them.)
-Like Aesop, ACL2 works on goals and saturates the local context of each goal, using user-provided forward rules, before using backward reasoning proof methods.
+(Some such tactics, e.g.\ Coq's auto, can execute arbitrary tactics including forward rules, but make no special provisions for them.)
 It uses essentially Aesop's naive algorithm, i.e.\ the local context of each goal is saturated independently without reusing partial results from other goals.
+ACL2 additionally supports \emph{trigger terms}, which allow forward rules to be selected based on the presence of a particular term anywhere in the goal, rather than based on particular hypotheses.
+Aesop implements a similar mechanism, \emph{rule patterns}, not described here due to space constraints.
 
@@
 The Prolog-based first-order theorem prover SATCHMO (TODO ref) and its many descendants (TODO refs) uses a tableau calculus (TODO ref) that can be viewed as combined forward and backward reasoning.
-Given a set of implications $A₁ ∧ \dots ∧ Aₙ → B₁ ∨ \dots ∨ Bₘ$ (with special cases $⊤ → B₁ ∨ \dots ∨ Bₘ$ and $A₁ ∧ \dots ∧ Aₙ → ⊥$), SATCHMO attempts to create a model that satisfies all implications.
-To that end, it matches the premises $A₁, \dots, Aₙ$ of each rule against known facts from Prolog's database, which is initially empty.
-If this is successful, SATCHMO first adds $B₁$ to the Prolog database and continues its search; if the continuation fails (i.e., if $⊥$ is derived), SATCHMO backtracks and tries $B₂$, etc.
-
-This search procedure can be viewed as a special case of our technique: the implications are used as forward rules and a single backward rule splits any goal with target $B₁ ∨ \dots ∨ Bₘ$ into $m$ goals with targets $B₁, \dots, Bₘ$.
-Compared to a Prolog-based implementation, our technique has the advantage that at each branching point we remember the partial matches induced by the currently known facts for each implication, rather than matching the implication against the database multiple times (subject to one of Prolog's various indexing schemes).
-Whether this gain outweighs the cost of storing and maintaining the partial match data structures likely depends on the given implications.
-Another advantage is that our technique is agnostic to the search strategy, whereas SATCHMO's Prolog implementation is based on backtracking and hence must use depth-first search.
+Given a set of implications $A₁ ∧ \dots ∧ Aₙ → B₁ ∨ \dots ∨ Bₘ$, SATCHMO attempts to create a model that satisfies all implications.
+Its search procedure is a special case of our technique: the implications are used as forward rules and a single backward rule splits any goal with target $B₁ ∨ \dots ∨ Bₘ$ into goals with targets $B₁, \dots, Bₘ$.
+This is a simpler setting than ours since the backward rule does not change the facts available to the forward rules.
+Still, our technique has the advantage that at each branching point we remember the partial matches induced by the currently known facts for each implication, rather than matching the implication against the database multiple times.
 
@@
 Combined forward and backward reasoning is also supported by some knowledge-based reasoning systems, e.g.\ Algernon (TODO ref) and EYE (TODO ref).
-However, to our knowledge none of these systems have backwards rules that can change the facts available to forward rules.
-Hence backward and forward reasoning do not interact and there is no need for our technique.
+To our knowledge, none of these systems have backwards rules that can change the facts available to forward rules.
 Nevertheless, our proposal bears some resemblance to the RETE algorithm (TODO ref) for forward reasoning in large knowledge bases.
 
-\paragraph{Other related work.}
-More distantly related are various calculi that combine backward reasoning with some form of incremental forward reasoning or simplification.
-Examples include the DPLL algorithm with its unit resolution rule (TODO ref) or incremental theory reasoners in SMT (TODO ref) and tableaux (TODO refs).
-Focusing (TODO ref) provides a proof-theoretic view on the interplay of forward and backward reasoning.
-
-\jcom{auto2}
-\jcom{Mention ACL2 forward rule triggers}
+\jcom{auto2?}
 

### 2025-02-14 paper.tex
First draft of conclusion

@@
 
-\xcom{Do we want to add potential improvements? Like an index closer to the naive algorithm.}
-
-TODO
+We have presented a technique that integrates forward reasoning into a tree-based proof search procedure supporting arbitrary proof rules.
+We avoid redundant reasoning steps by caching partial forward rule applications in a data structure, the \emph{forward match state}, that enables efficient updates.
+The technique is independent of the search strategy and logic used.
+Our evaluation shows a substantial speedup on some synthetic benchmarks, relative to Aesop's previous implementation.
 

### 2025-02-16 paper.tex
Remove a confusing formulation
Minor edits

@@
 In many proof assistants, simple automation tactics based on tableau-style tree search play a substantial role in practice: Isabelle's (TODO ref) and Rocq's (TODO ref) auto, ACL2's (TODO ref) waterfall, PVS's (TODO ref) grind, Lean's (TODO ref) Aesop (TODO ref), etc.
-These tactics operate on \emph{goals} $Γ ⊢ A$, where the \emph{local context} $Γ$ is a list of assumptions and $A$ is the proposition to be proved.
+These tactics operate on \emph{goals} $Γ ⊢ A$, where the \emph{context} $Γ$ is a list of assumptions and $A$ is the proposition to be proved.
 Faced with an initial goal, the tactics try various user-specified \emph{rules}, which correspond to admissible inferences in the prover's logic.
@@
 In particular, each forward rule was tried on each goal separately (subject to pre-filtering by a discrimination tree (TODO ref) index).
-This is wasteful since in practice, many rules produce subgoals with a similar---or even the same---local context as the initial goal.
+This is wasteful since in practice, many rules produce subgoals with a similar---or even the same---context as the initial goal.
 For example, the ∧-introduction rule affects only the goal's target, not its context.
@@
 The central idea is to remember, for each goal $Γ ⊢ A$, which forward rules can be partially applied to which sets of hypotheses.
-When a non-forward rule then generates subgoals $Δ₁ ⊢ A₁$, \dots, $Δₙ ⊢ Aₙ$, the new local contexts $Δᵢ$ are likely similar to $Γ$.
+When a non-forward rule then generates subgoals $Δ₁ ⊢ A₁$, \dots, $Δₙ ⊢ Aₙ$, the new contexts $Δᵢ$ are likely similar to $Γ$.
 Hence it is generally cheaper (and often much cheaper) to start with the partial matches for $Γ$ and update them for each of the $Δᵢ$.
@@
 
-In addition to supporting such \emph{incremental} forward reasoning, the partial match state satisfies two other desiderata:
+In addition to this \emph{incrementality}, the partial match state satisfies two other desiderata:
 \begin{itemize}
@@
 
-Our evaluation (Sec.~\ref{sec:evaluation}) shows that this new implementation of forward reasoning in Aesop substantially outperforms the previous, naive implementation in some (though not all) cases.
-The implementation, the benchmarks we used and the full evaluation data are available in a supplement to this paper (TODO ref).
+Our evaluation (Sec.~\ref{sec:evaluation}) shows that this new implementation of forward reasoning in Aesop substantially outperforms the previous, naive implementation on some (though not all) synthetic benchmarks.
+The implementation, benchmarks and evaluation data are available in a supplement to this paper (TODO ref).
 Mainline Aesop already uses the new implementation by default (TODO true?).
@@
 
-If $h : A$ is a hypothesis in the local context of a goal and $A$ unifies with the type $B$ of slot $i$, we say that $h$ \emph{matches} $i$.
+If $h : A$ is a hypothesis in the context of a goal and $A$ unifies with the type $B$ of slot $i$, we say that $h$ \emph{matches} $i$.
 Unification induces a substitution $σ$ such that $A[σ]$ is definitionally equal to $B[σ]$; we call this substitution $\sub_{i}(h)$.
@@
 This is the usual normal form of expressions with respect to Lean's notion of computation, with two modifications.
-First, only \emph{reducible} defined constants are unfolded, i.e.\ replaced with their definiendum.
+First, only \emph{reducible} defined constants are unfolded.
 This reflects our choice to perform unification at reducible transparency.
@@
 (Some such tactics, e.g.\ Coq's auto, can execute arbitrary tactics including forward rules, but make no special provisions for them.)
-It uses essentially Aesop's naive algorithm, i.e.\ the local context of each goal is saturated independently without reusing partial results from other goals.
+It uses essentially Aesop's naive algorithm, i.e.\ the context of each goal is saturated independently without reusing partial results from other goals.
 ACL2 additionally supports \emph{trigger terms}, which allow forward rules to be selected based on the presence of a particular term anywhere in the goal, rather than based on particular hypotheses.
@@
 Our evaluation shows a substantial speedup on some synthetic benchmarks, relative to Aesop's previous implementation.
-
 \end{document}

### 2025-02-16 paper.tex
Edit evaluation

@@
 \usepackage[noend]{algorithm2e}
+\usepackage{siunitx}
 
@@
 \newunicodechar{≥}{\ensuremath{\geq}}
+\newunicodechar{≺}{\ensuremath{\prec}}
 \newunicodechar{≔}{\ensuremath{\coloneq}}
@@
 
-\section{Incremental Forward Reasoning}
+\section{Incremental Forward Reasoning}%
+\label{sec:incremental}
 
@@
 
-Comparing the stateless and stateful approach, we can draw the following conclusions about their relative performance:
-
-\begin{itemize}
-    \item The stateful algorithm maybe be expected to perform poorly when the context contains many hypotheses that almost complete rules.
-    \item The stateless algorithm should perform decently well on average and especially when only the last premises are absent, so the rules get never selected by the stateless index.
-\end{itemize}
-
-These tests were performed on a Macbook Pro with the Apple M2 Pro chip equipped with 32 Go of RAM.
-The tests showcased in this section are available here \footnote{\xcom{TODO : Link Github repo with tests.}}
-\subsection{Results}
-\xcom{TODO: Add a comment on the effect it had on Mathlib.}
-
-We present here a number of results obtained by running synthetic tests.
-By synthetic we mean that these tests we're constructed solely for the purpose of demonstrating the different complexity behavior of both algorithms by scaling some parameter (e.g.\ the number of rules).
-The results of the tests are an average of multiple runs.
+There are unfortunately no standard benchmarks for Aesop-like tactics, and since Aesop is not expected to solve most goals without an extensive, manually curated set of rules, benchmarks for push-button theorem provers such as TPTP (TODO ref) are not very meaningful.
+At the same time, forward rules are currently little used by Aesop's main client, the Mathlib mathematics library (TODO ref).
+We therefore evaluate our technique against the previous, naive Aesop implementation on several synthetic benchmarks that showcase the performance characteristics of incremental forward reasoning.
+
+For each benchmark, we set up certain forward rules and a goal with certain hypotheses.
+We then run the \texttt{saturate} tactic, which uses either naive or incremental forward reasoning to exhaustively apply the forward rules.
+Performing only forward reasoning is the best-case scenario for the incremental algorithm since previous results can be fully reused.
+When forward reasoning is combined with other rules, we would expect smaller gains, depending on how much reuse the non-forward rules allow.
+
+Benchmark code and data are available in our supplement (TODO ref).
+The results shown below were obtain on a Macbook Pro with an Apple~M2 Pro processor and \SI{32}{GB} of RAM, averaging over \textcolor{red}{N} (TODO) runs per benchmark.
 
 \paragraph{Transitivity.}
+Our first benchmark is a scaled-up variant of the transitivity example from Sec.~\ref{sec:incremental}.
+Given some relation $≺$ on natural numbers, we register the lemma $\All{x~y}{x ≺ y → y ≺ z → x ≺ z}$ as the sole forward rule and set up a goal with hypotheses $h₁ : a ≺ a + 1, \dots, hₙ : a + n - 1 ≺ a + n$ (TODO what is $a$?).
+Saturating this goal adds $n(n-1)/2$ hypotheses to the context.
+
+Instead of the natural number type of Lean's standard library, which has special support in the compiler and kernel, we use a custom type of Peano naturals.
+Unifying two concrete Peano natural numbers $n$ and $m$ involves comparing $\min(n, m)$ constructors.
+We can therefore use moderately large numbers to simulate the bigger unification problems that Aesop encounters, for example, in Mathlib's category theory development.
 
 \begin{figure}
-    \label{fig:trans}
-    % Transitivity without precompilation.
-    % 
+  \begin{center}
     \begin{tikzpicture}
       \begin{axis}[
-            title={Transitivity rule},
         xlabel={Number of hypotheses},
@@
         ]
-
         \addplot[
@@
           };
-            \addlegendentry{Stateless}
+        \addlegendentry{naive}
         \addplot[
@@
           };
-            \addlegendentry{Stateful}
+        \addlegendentry{incremental}
         \addplot[
@@
     \end{tikzpicture}
+  \end{center}
+  \caption{Results of the transitivity benchmark. TODO detailed description}%
+  \label{fig:trans}
 \end{figure}
 
-As a first example, which we believe highlights well the advantages of the stateful approach, we consider a scaled up version of a transitivity problem.
-As variables we choose a custom version of natural numbers.
-The reason being that Lean has special optimization procedures for its integer and thus makes evaluation quite hard on problems that use them; the unification is too quickly resolved, making algorithmic complexity difficultly distinguishable from noise.
-We aim to emulate a more general problem where the objects unified do not benefit of some special support.
-
-Given some predicate $P(x, y)$, we register the rule $r : P(x, y) \to P(y, z) \to P(x, z)$ to both algorithms.
-Then, in Figure \ref{fig:trans}, we compare them on a context containing $n$ hypotheses of the form $(h_1 : P(a, n + 1)), \dots, (h_n : P(a + n - 1, a + n))$.
-Saturating this goal adds a total of $n(n-1)/2$ hypotheses to the context.
-
-The results in Figure \ref{fig:trans} have nothing surprising : The stateless algorithm needs to go over the whole context, which contains initially $n$ hypotheses and compares them with these same $n$ hypotheses.
-This is repeated approximately for each of the $n(n-1)/2$ added hypotheses giving the expected complexity of $O(n^4)$.
-
-The stateful algorithm considers each of the problem's hypothesis, both the initial and added ones, only once.
-As the hypotheses get matched to both slots, it then compares it with at most $n$ hypotheses for each slot.
-Indeed, by the end of the saturation procedure, for a value $k$ fixed, the first slot $\eta_y(1,k)$, contains at most $n$ hypotheses of the form $h : P(m,k)$ and symmetrically for the second slot.
-This leaves an expected complexity of $O(n^2)$.
-
-There are of course obvious inefficiencies, even with the stateful algorithm.
-For this specific kind of problem, the hypotheses need only be used once and this leads to much repeated work.
-For these scenarios, users can register the rule as a \textit{destruct} rule, eliminating hypotheses after their first use.
-In this case, the complexity is linear in $n$.
-For more edge case examples like this one, consult the project's repository.
-
-\paragraph{Independent rules.}
+The results in Fig.~\ref{fig:trans} are unsurprising.
+The naive algorithm goes over the context, containing initially $n$ hypotheses, and compares all hypotheses pairwise.
+This is repeated approximately for each of the $n(n-1)/2$ added hypotheses, so the overall complexity is $O(n^4)$.
+The incremental algorithm considers each hypothesis only once.
+As the hypotheses get matched to both slots, it then compares each hypothesis with at most $n$ other hypotheses.
+This yields a complexity of $O(n^2)$.
+
+Either algorithm generates all possible transitivity chains.
+If only the longest such chain, from $a$ to $a + n$, is needed, the transitivity rule can be registered as a \emph{destruct} rule, a special sort of forward rule that, whenever it is applied to certain hypotheses, removes the hypotheses from the context.
+This reduces the complexity of the incremental algorithm to $O(n)$.
+
+\paragraph{Independent Rules.}
+In the last example, part of the stateful algorithm's success stems from the fact that it could reuse a lot of its previous work.
+An obvious question is then how well does the stateful algorithm work when there is no work to be reused.
+
+Given a set of propositions $\mathcal{P} := \{P_{i,j} | 1 \leq i \leq n \text{ and } 1 \leq j \leq 6\}$ and $\mathcal{Q} :=\{Q_{i} | 1 \leq i \leq n \}$, in this test, we consider the $n$ following rules $r_i : P_{i,1}\to \dots P_{i,6} \to Q_i$.
+The procedures are then run on a context containing exactly $\mathcal{P}$.
+In this case, we sketch the behavior of the stateless algorithm:
+It pick a first hypothesis, $h_{i,j} : P_{i,j}$.
+If $j < 6$, the index returns nothing and it selects another hypothesis.
+For $h_{i,6} : P_{i,6}$, the rule $r_i$ is selected, and it then needs to try and fill its remaining five slot with the context's $6n$ hypotheses which it tries all.
+This gives a complexity of about $O(n^2)$.
+
+For the stateful algorithm, the hypotheses are considered one by one and match directly to the rule and the slot.
+The expected complexity is then linear in the number of rules.
 
 \begin{figure}
-    %Indep without precompilation
+  \begin{center}
     \begin{tikzpicture}
       \begin{axis}[
-                title={Independance Test},
         xlabel={Number of Rules},
@@
           };
-              \addlegendentry{Stateless}
+        \addlegendentry{naive}
         \addplot[
@@
           };
-                \addlegendentry{Stateful}
+        \addlegendentry{incremental}
         \addplot[
@@
     \end{tikzpicture}
+  \end{center}
+  \caption{Results of the independence benchmark. TODO detailed description}%
+  \label{fig:indep}
 \end{figure}
 
-In the last example, part of the stateful algorithm's success stems from the fact that it could reuse a lot of its previous work.
-An obvious question is then how well does the stateful algorithm work when there is no work to be reused.
-
-Given a set of propositions $\mathcal{P} := \{P_{i,j} | 1 \leq i \leq n \text{ and } 1 \leq j \leq 6\}$ and $\mathcal{Q} :=\{Q_{i} | 1 \leq i \leq n \}$, in this test, we consider the $n$ following rules $r_i : P_{i,1}\to \dots P_{i,6} \to Q_i$.
-The procedures are then run on a context containing exactly $\mathcal{P}$.
-In this case, we sketch the behavior of the stateless algorithm:
-It pick a first hypothesis, $h_{i,j} : P_{i,j}$.
-If $j < 6$, the index returns nothing and it selects another hypothesis.
-For $h_{i,6} : P_{i,6}$, the rule $r_i$ is selected, and it then needs to try and fill its remaining five slot with the context's $6n$ hypotheses which it tries all.
-This gives a complexity of about $O(n^2)$.
-
-For the stateful algorithm, the hypotheses are considered one by one and match directly to the rule and the slot.
-The expected complexity is then linear in the number of rules.
-
-\paragraph{Depth of considered rules.}
+\paragraph{Rules of Varying Depth.}
+In Fig.~\ref{fig:erase}, the step is set up as follows: given some rule $r$ with $6$ \xcom{motivate 6} premises, we run the procedure in a context with 5 of the 6 premises present.
+The numbers on the $x$ axis represent which hypothesis is missing in the order of consideration by the algorithm.
+In a sense, this is how \textit{deep} the algorithm investigate before it stops.
+At $0$, the rule are not selected by the index so the procedure stops.
+The stateless algorithm gets progressively slower the further the removed premises is located.
 
 \begin{figure}
-    \label{fig:erase}
-%erase without precompilation
+  \begin{center}
     \begin{tikzpicture}
@@
           };
-          \addlegendentry{Stateful}
+        \addlegendentry{incremental}
       \end{axis}
     \end{tikzpicture}
+  \end{center}
+  \caption{Results of the depth benchmark. TODO detailed description}%
+  \label{fig:erase}
 \end{figure}
-
-In Fig.~\ref{fig:erase}, the step is set up as follows: given some rule $r$ with $6$ \xcom{motivate 6} premises, we run the procedure in a context with 5 of the 6 premises present.
-The numbers on the $x$ axis represent which hypothesis is missing in the order of consideration by the algorithm.
-In a sense, this is how \textit{deep} the algorithm investigate before it stops.
-At $0$, the rule are not selected by the index so the procedure stops.
-The stateless algorithm gets progressively slower the further the removed premises is located.
 
@@
 \paragraph{SATCHMO.}
-The Prolog-based first-order theorem prover SATCHMO (TODO ref) and its many descendants (TODO refs) uses a tableau calculus (TODO ref) that can be viewed as combined forward and backward reasoning.
+The Prolog-based first-order theorem prover SATCHMO (TODO ref) and its many descendants (TODO refs) use a tableau calculus (TODO ref) that can be viewed as combined forward and backward reasoning.
 Given a set of implications $A₁ ∧ \dots ∧ Aₙ → B₁ ∨ \dots ∨ Bₘ$, SATCHMO attempts to create a model that satisfies all implications.

### 2025-02-17 paper.tex
Remove commented-out figures
update graphs

@@
 \usepackage{siunitx}
+\usepackage{caption}
+\usepackage{subcaption}
 
@@
 \begin{figure}
-  \begin{center}
-    \begin{tikzpicture}
+  \centering
+  \begin{subfigure}{.5\textwidth}
+    \centering
+    \begin{tikzpicture}[scale=0.75]
       \begin{axis}[
@@
     \end{tikzpicture}
-  \end{center}
-  \caption{Results of the transitivity benchmark. TODO detailed description}%
+    \caption{Transitivity benchmark}
     \label{fig:trans}
-\end{figure}
-
-The results in Fig.~\ref{fig:trans} are unsurprising.
-The naive algorithm goes over the context, containing initially $n$ hypotheses, and compares all hypotheses pairwise.
-This is repeated approximately for each of the $n(n-1)/2$ added hypotheses, so the overall complexity is $O(n^4)$.
-The incremental algorithm considers each hypothesis only once.
-As the hypotheses get matched to both slots, it then compares each hypothesis with at most $n$ other hypotheses.
-This yields a complexity of $O(n^2)$.
-
-Either algorithm generates all possible transitivity chains.
-If only the longest such chain, from $a$ to $a + n$, is needed, the transitivity rule can be registered as a \emph{destruct} rule, a special sort of forward rule that, whenever it is applied to certain hypotheses, removes the hypotheses from the context.
-This reduces the complexity of the incremental algorithm to $O(n)$.
-
-\paragraph{Independent Rules.}
-In the last example, part of the stateful algorithm's success stems from the fact that it could reuse a lot of its previous work.
-An obvious question is then how well does the stateful algorithm work when there is no work to be reused.
-
-Given a set of propositions $\mathcal{P} := \{P_{i,j} | 1 \leq i \leq n \text{ and } 1 \leq j \leq 6\}$ and $\mathcal{Q} :=\{Q_{i} | 1 \leq i \leq n \}$, in this test, we consider the $n$ following rules $r_i : P_{i,1}\to \dots P_{i,6} \to Q_i$.
-The procedures are then run on a context containing exactly $\mathcal{P}$.
-In this case, we sketch the behavior of the stateless algorithm:
-It pick a first hypothesis, $h_{i,j} : P_{i,j}$.
-If $j < 6$, the index returns nothing and it selects another hypothesis.
-For $h_{i,6} : P_{i,6}$, the rule $r_i$ is selected, and it then needs to try and fill its remaining five slot with the context's $6n$ hypotheses which it tries all.
-This gives a complexity of about $O(n^2)$.
-
-For the stateful algorithm, the hypotheses are considered one by one and match directly to the rule and the slot.
-The expected complexity is then linear in the number of rules.
-
-\begin{figure}
-  \begin{center}
-    \begin{tikzpicture}
+  \end{subfigure}%
+  \begin{subfigure}{.5\textwidth}
+    \centering
+    \begin{tikzpicture}[scale=0.75]
       \begin{axis}[
         xlabel={Number of Rules},
-        ylabel={Time in ms},
+        %ylabel={Time in ms},
         xmin=1, xmax=32,
@@
     \end{tikzpicture}
-  \end{center}
-  \caption{Results of the independence benchmark. TODO detailed description}%
+    % graph \includegraphics[width=.4\linewidth]{image1}
+    \caption{Independence benchmark}
     \label{fig:indep}
+  \end{subfigure}
+  \caption{The above results depict the behavior of both algorithms in a context where there is a lot of work that can be reused (Transitivity) and no work to be reused (Independence).
+The lines with square entry points represent problems where unification is easy.
+The lines with circle entry points are the same tests with hard unification problems.}
+  \label{fig:benchmark}
   \end{figure}
+
+The results in Fig.~\ref{fig:trans} are unsurprising.
+The naive algorithm goes over the context, containing initially $n$ hypotheses, and compares all hypotheses pairwise.
+This is repeated approximately for each of the $n(n-1)/2$ added hypotheses, so the overall complexity is $O(n^4)$.
+The incremental algorithm considers each hypothesis only once.
+As the hypotheses get matched to both slots, it then compares each hypothesis with at most $n$ other hypotheses.
+This yields a complexity of $O(n^2)$.
+
+Either algorithm generates all possible transitivity chains.
+If only the longest such chain, from $a$ to $a + n$, is needed, the transitivity rule can be registered as a \emph{destruct} rule, a special sort of forward rule that, whenever it is applied to certain hypotheses, removes the hypotheses from the context.
+This reduces the complexity of the incremental algorithm to $O(n)$.
+
+\paragraph{Independent Rules.}
+In the last example, part of the stateful algorithm's success stems from the fact that it could reuse a lot of its previous work.
+An obvious question is then how well does the stateful algorithm work when there is no work to be reused.
+
+Given a set of propositions $\mathcal{P} := \{P_{i,j} | 1 \leq i \leq n \text{ and } 1 \leq j \leq 6\}$ and $\mathcal{Q} :=\{Q_{i} | 1 \leq i \leq n \}$, in this test, we consider the $n$ following rules $r_i : P_{i,1}\to \dots P_{i,6} \to Q_i$.
+The procedures are then run on a context containing exactly $\mathcal{P}$.
+In this case, we sketch the behavior of the stateless algorithm:
+It pick a first hypothesis, $h_{i,j} : P_{i,j}$.
+If $j < 6$, the index returns nothing and it selects another hypothesis.
+For $h_{i,6} : P_{i,6}$, the rule $r_i$ is selected, and it then needs to try and fill its remaining five slot with the context's $6n$ hypotheses which it tries all.
+This gives a complexity of about $O(n^2)$.
+
+For the stateful algorithm, the hypotheses are considered one by one and match directly to the rule and the slot.
+The expected complexity is then linear in the number of rules.
 
@@
         xmin=1, xmax=6,
-        ymin=6, ymax=20,
+        ymin=0, ymax=650,
         xtick={1,2,3,4,5,6},
-        ytick={10,15},
+        ytick={2^2,2^4,2^6,2^8},
+        ymode=log,
+        log basis y=2,
         legend pos=north west,
@@
           coordinates {
-            (1, 6.407973) (2, 9.831676) (3, 10.301718) (4, 10.626144) (5, 11.261754) (6, 11.795240)
+            (1, 32.580083) (2, 33.212805) (3, 40.239833) (4, 47.670277) (5, 54.647458) (6, 61.044958) 
           };
@@
           coordinates {
-            (1, 15.835558) (2, 16.961994) (3, 16.424259) (4, 17.605449) (5, 15.691536) (6, 14.982820)
+            (1, 127.925541) (2, 140.497597) (3, 123.719264) (4, 130.819361) (5, 121.047667) (6, 123.789500)
           };
         \addlegendentry{incremental}
+        \addplot[
+          color=orange,
+          mark=o,
+          ]
+          coordinates {
+            (1, 108.120444) (2, 107.352694) (3, 116.182833) (4, 125.995639) (5, 137.815527) (6, 142.310597) 
+          };
+        \addplot[
+          color=blue,
+          mark=o,
+          ]
+          coordinates {
+            (1, 604.595791) (2, 624.541194) (3, 608.671111) (4, 614.285014) (5, 588.636819) (6, 595.276806) 
+          };
       \end{axis}
     \end{tikzpicture}
-  \end{center}
-  \caption{Results of the depth benchmark. TODO detailed description}%
+    \caption{Depth benchmark no-precompilation}
+    \label{fig:erase-noprecomp}
+  \end{subfigure}%
+  \begin{subfigure}{.5\textwidth}
+    \centering
+    \begin{tikzpicture}[scale=0.75]
+      \begin{axis}[
+        title={Depth Test},
+        xlabel={Depth},
+        %ylabel={Time in ms},
+        xmin=1, xmax=6,
+        ymin=0, ymax=128,
+        xtick={1,2,3,4,5,6},
+        ytick={2^1,2^3,2^5,2^7},
+        ymode=log,
+        log basis y=2,
+        legend pos=north west,
+        ymajorgrids=true,
+        grid style=dashed,
+        legend image post style={mark=},
+        ]
+        \addplot[
+          color=orange,
+          mark=square,
+          ]
+          coordinates {
+            (1, 12.820514) (2, 12.645666) (3, 17.582138) (4, 22.730528) (5, 27.280097) (6, 32.578722) 
+          };
+        \addlegendentry{naive}
+        \addplot[
+          color=blue,
+          mark=square,
+          ]
+          coordinates {
+            (1, 13.406416) (2, 13.747139) (3, 12.503764) (4, 12.950486) (5, 12.282083) (6, 12.151930) 
+          };
+        \addlegendentry{incremental}
+        \addplot[
+          color=orange,
+          mark=o,
+          ]
+          coordinates {
+            (1, 83.032819) (2, 79.396889) (3, 85.939180) (4, 92.495542) (5, 96.461291) (6, 105.517458)
+          };
+        \addplot[
+          color=blue,
+          mark=o,
+          ]
+          coordinates {
+            (1, 25.681583) (2, 25.322444) (3, 27.672833) (4, 26.999055) (5, 25.690319) (6, 25.257291)
+          };
+      \end{axis}
+    \end{tikzpicture}
+    % graph \includegraphics[width=.4\linewidth]{image1}
+    \caption{Depth benchmark with precompilation}
+    \label{fig:erase-precomp}
+  \end{subfigure}
+  \caption{The above results depict the behavior of both algorithms in depending on the \text{depth} of potential partial matches.
+The lines with square entry points represent problems where unification is easy.
+The lines with circle entry points are the same tests with hard unification problems.}
   \label{fig:erase}
@@
 \end{document}
-
-% \begin{figure}
-%     % This is the erase without precompilation for 6 premises.
-%     % Note I have a decent idea that would allow us to compare for multiple
-%     % number of premises. 
-%     % (Divide by the average of the stateful time to normalize.)
-%     \begin{tikzpicture}
-%         \begin{axis}[
-%             title={Depth Test},
-%             xlabel={Depth},
-%             ylabel={Time in ms},
-%             xmin=0, xmax=6,
-%             ymin=50, ymax=65,
-%             xtick={0,1,2,3,4,5,6},
-%             ytick={50,55,60,65},
-%             legend pos=north west,
-%             ymajorgrids=true,
-%             grid style=dashed,
-%         ]
-
-%         \addplot[
-%             color=orange,
-%             mark=square,
-%             ]
-%             coordinates {
-%             (1,50.91)(2,53.14)(3,52.91)(4,51.22)(5,55.14)(6,56.60)
-%             };
-%             \addlegendentry{Stateless}
-
-%         \addplot[
-%             color=blue,
-%             mark=square,
-%             ]
-%             coordinates {
-%             (1,56.04)(2,56.43)(3,55.58)(4,55.26)(5,54.80)(6,60.61)
-%             };
-%             \addlegendentry{Stateful}
-
-%         \end{axis}
-%         \end{tikzpicture}
-%     \end{figure}
-
-%     \begin{figure}
-%         % This is the erase without precompilation for 6 premises.
-%         % Note I have a decent idea that would allow us to compare for multiple
-%         % number of premises. 
-%         % (Divide by the average of the stateful time to normalize.)
-%         \begin{tikzpicture}
-%             \begin{axis}[
-%                 title={Scaled Depth Test},
-%                 xlabel={Depth},
-%                 ylabel={Time in ms},
-%                 xmin=0, xmax=8,
-%                 ymin=0.5, ymax=2,
-%                 xtick={0,1,2,3,4,5,6,7,8},
-%                 ytick={0.75,1,1.25,1.5,1.75},
-%                 legend pos=north west,
-%                 ymajorgrids=true,
-%                 grid style=dashed,
-%             ]
-
-%             \addplot[color=orange,mark=square,]
-%                 coordinates {(0, 0.883600)(1, 1.039751)};
-%                 \addlegendentry{Stateless}
-%             \addplot[color=orange,mark=square,]
-%                 coordinates {(0, 0.804521)(1, 1.017893)(2, 1.022493)};
-%             \addplot[color=orange,mark=square,]
-%                 coordinates {(0, 0.791528)(1, 1.015467)(2, 1.076267)(3, 1.159999)};
-%             \addplot[color=orange,mark=square,]
-%                 coordinates {(0, 0.743256)(1, 0.966436)(2, 1.058876)(3, 1.133514)(4, 1.225561)};
-%             \addplot[color=orange,mark=square,]
-%                 coordinates {(0, 0.770625)(1, 1.003777)(2, 1.109929)(3, 1.215171)(4, 1.305373)(5, 1.390667)};
-%             \addplot[color=orange,mark=square,]
-%                 coordinates {(0, 0.759093)(1, 1.016667)(2, 1.148273)(3, 1.213138)(4, 1.342795)(5, 1.442271)(6, 1.514877)};
-%             \addplot[color=orange,mark=square,]
-%                 coordinates {(0, 0.709965)(1, 0.971294)(2, 1.068745)(3, 1.179333)(4, 1.312798)(5, 1.424054)(6, 1.551690)(7, 1.628344)};
-%             \addplot[color=orange,mark=square,]
-%                 coordinates {(0, 0.728860)(1, 1.001576)(2, 1.097472)(3, 1.219259)(4, 1.366683)(5, 1.444844)(6, 1.576972)(7, 1.668693)(8, 1.815399)};
-
-%             \end{axis}
-%             \end{tikzpicture}
-%         \end{figure}
-
-%                 \begin{figure}
-%                     %Casacade with precompilation
-%                         \begin{tikzpicture}
-%                             \begin{axis}[
-%                                 title={Casacade Test},
-%                                 xlabel={Number of Rules},
-%                                 ylabel={Time in ms},
-%                                 xmin=0, xmax=32,
-%                                 ymin=0, ymax=850,
-%                                 ymode=log,xmode=log,
-%                                 log basis x=2,
-%                                 log basis y=2,
-%                                 legend pos=north west,
-%                                 ymajorgrids=true,
-%                                 grid style=dashed,
-%                             ]
-
-%                             \addplot[
-%                                 color=blue,
-%                                 mark=square,
-%                                 ]
-%                                 coordinates {
-%                                 (1, 0.393011)(2, 0.730952)(4, 1.327736)(8, 3.007961)(16, 10.173689)(32, 38.892027)
-%                                 };
-%                                 \addlegendentry{Stateful}
-
-%                             \addplot[
-%                                 color=orange,
-%                                 mark=square,
-%                                 ]
-%                                 coordinates {
-%                                 (1, 0.464973)(2, 0.759516)(4, 1.865191)(8, 7.585588)(16, 64.374740)(32, 813.336527)
-%                                 };
-%                                 \addlegendentry{Stateless}
-
-%                             \end{axis}
-%                             \end{tikzpicture}
-%                         \end{figure}

### 2025-02-17 paper.tex
Correct and expand PVS info

@@
 % TODO: rename variable maps to variable indices everywhere
-% TODO: standardise Rocq vs Coq
+% TODO: Title Case for paragraphs
 
@@
 
-In many proof assistants, simple automation tactics based on tableau-style tree search play a substantial role in practice: Isabelle's (TODO ref) and Rocq's (TODO ref) auto, ACL2's (TODO ref) waterfall, PVS's (TODO ref) grind, Lean's (TODO ref) Aesop (TODO ref), etc.
+In many proof assistants, simple \enquote*{white-box} automation tactics based on tableau-style tree search play a substantial role in practice: Isabelle's (TODO ref) and Rocq's (TODO ref) auto, ACL2's (TODO ref) waterfall, Lean's (TODO ref) Aesop (TODO ref), etc.
 These tactics operate on \emph{goals} $Γ ⊢ A$, where the \emph{context} $Γ$ is a list of assumptions and $A$ is the proposition to be proved.
@@
 
-\paragraph{ACL2.}
-Among white-box proof automation tactics, only ACL2's \enquote{waterfall} (TODO ref) has special support for forward reasoning.
-(Some such tactics, e.g.\ Coq's auto, can execute arbitrary tactics including forward rules, but make no special provisions for them.)
-It uses essentially Aesop's naive algorithm, i.e.\ the context of each goal is saturated independently without reusing partial results from other goals.
+\paragraph{ACL2 and PVS.}
+Among white-box proof automation tactics, to our knowledge only ACL2's \enquote{waterfall} (TODO ref) and PVS's various proof search strategies (TODO ref) have special support for forward reasoning.
+Both use essentially Aesop's naive algorithm, i.e.\ the context of each goal is saturated independently without reusing partial results from other goals.
 ACL2 additionally supports \emph{trigger terms}, which allow forward rules to be selected based on the presence of a particular term anywhere in the goal, rather than based on particular hypotheses.
 Aesop implements a similar mechanism, \emph{rule patterns}, not described here due to space constraints.
+Some other white-box tactics, e.g.\ Rocq's auto, can execute arbitrary tactics including forward rules, but make no special provisions for them.
 
@@
 Our evaluation shows a substantial speedup on some synthetic benchmarks, relative to Aesop's previous implementation.
+
 \end{document}

### 2025-02-17 paper.tex
Fix SATCHMO related work

@@
 
-\paragraph{SATCHMO.}
-The Prolog-based first-order theorem prover SATCHMO (TODO ref) and its many descendants (TODO refs) use a tableau calculus (TODO ref) that can be viewed as combined forward and backward reasoning.
-Given a set of implications $A₁ ∧ \dots ∧ Aₙ → B₁ ∨ \dots ∨ Bₘ$, SATCHMO attempts to create a model that satisfies all implications.
-Its search procedure is a special case of our technique: the implications are used as forward rules and a single backward rule splits any goal with target $B₁ ∨ \dots ∨ Bₘ$ into goals with targets $B₁, \dots, Bₘ$.
-This is a simpler setting than ours since the backward rule does not change the facts available to the forward rules.
-Still, our technique has the advantage that at each branching point we remember the partial matches induced by the currently known facts for each implication, rather than matching the implication against the database multiple times.
-
-\paragraph{Knowledge-based reasoning systems.}
-Combined forward and backward reasoning is also supported by some knowledge-based reasoning systems, e.g.\ Algernon (TODO ref) and EYE (TODO ref).
-To our knowledge, none of these systems have backwards rules that can change the facts available to forward rules.
-Nevertheless, our proposal bears some resemblance to the RETE algorithm (TODO ref) for forward reasoning in large knowledge bases.
+\paragraph{Mixed-Mode Reasoning.}
+Some knowledge-based reasoning systems, e.g.\ Algernon (TODO ref) and EYE (TODO ref), support combined forward and backward (\enquote*{mixed-mode}) reasoning.
+However, to our knowledge, rules in these systems cannot change the assumptions available to forward rules, so they do not require incrementality.
+Nevertheless, our technique bears some resemblance to the RETE algorithm (TODO ref) for forward reasoning in large knowledge bases.
+
+The Prolog-based first-order theorem prover SATCHMO (TODO ref) and its many descendants also use mixed-mode reasoning.
+SATCHMO's search procedure can be viewed as a special case of our technique: a given set of implications are used as forward rules and a single case-splitting rule splits any goal with hypothesis $h : B₁ ∨ \dots ∨ Bₘ$ into $m$ goals with hypotheses $h₁ : B₁, \dots, hₘ : Bₘ$.
+A SATCHMO-style prover may benefit from our incremental approach if the set of implications is large, though in a setting where unification is cheap it is not clear whether avoiding repeated unifications outweighs the cost of maintaining the partial match state.
+SATCHMO also employs an optimisation that \enquote*{focuses} forward reasoning on the hypothesis established by the last case split, which provides a form of incrementality.
 

### 2025-02-17 paper.tex
Add lazy insertion optimisation in implementation and remove indexing

@@
 
-\paragraph{Indexing.}
-When adding a hypothesis $h : T$ to the partial match state, we add it to the rule state of each rule $r$ such that $h$ matches a slot of $r$.
-To optimise this operation, we use an imperfect discrimination tree index that, for each rule $r$ with slot $i$ corresponding to a premise of type $Tᵢ$, maps the type $Tᵢ$ to the pair $(r, i)$.
-Querying this index for $T$ quickly determines all rules and slots for which $h$ is likely relevant.
+\paragraph{Lazy Insertion.}
+Adding a hypothesis to a rule state is fairly cheap, but not free: it involves one unification for each premise that the hypothesis likely matches (according to a discrimination tree index), followed by several operations on the variable indices.
+We therefore delay insertions until the index has selected at least one hypothesis for each slot.
+This helps, for example, with the definitional rule for involutive functions: if $f : α → α$ is involutive, then for any $a : α$ we have $f (f a) = a$.
+When used as a forward rule, the premise $a$ matches any non-propositional hypothesis, but this is only relevant if there is, in fact, an involutive function is the context.
 

### 2025-02-17 paper.tex
minor fixes

@@
 In the last example, part of the stateful algorithm's success stems from the fact that it could reuse a lot of its previous work.
-An obvious question is then how well does the stateful algorithm work when there is no work to be reused.
+An obvious question is then how well does the stateful algorithm behave when there is no work to be reused.
 
@@
         xmin=1, xmax=6,
-        ymin=0, ymax=650,
+        ymin=0, ymax=2^10,
         xtick={1,2,3,4,5,6},
-        ytick={2^2,2^4,2^6,2^8},
+        ytick={4,16,64,256,1024},
         ymode=log,
@@
         xmin=1, xmax=6,
-        ymin=0, ymax=128,
+        ymin=0, ymax=170,
         xtick={1,2,3,4,5,6},

### 2025-02-17 paper.tex
Edit involutive example

@@
 We therefore delay insertions until the index has selected at least one hypothesis for each slot.
-This helps, for example, with the definitional rule for involutive functions: if $f : α → α$ is involutive, then for any $a : α$ we have $f (f a) = a$.
-When used as a forward rule, the premise $a$ matches any non-propositional hypothesis, but this is only relevant if there is, in fact, an involutive function is the context.
+This helps, for example, with the defining rule for involutive functions: if $f : α → α$ is involutive, then for any $a : α$ we have $f (f a) = a$.
+When this rule is used as a forward rule, the premise $a$ matches any non-propositional hypothesis, but with lazy insertion we only process $a$ if there is, in fact, an involutive function is the context.
 

### 2025-02-17 paper.tex
Mention PVS also in intro

@@
 
-In many proof assistants, simple \enquote*{white-box} automation tactics based on tableau-style tree search play a substantial role in practice: Isabelle's (TODO ref) and Rocq's (TODO ref) auto, ACL2's (TODO ref) waterfall, Lean's (TODO ref) Aesop (TODO ref), etc.
+In many proof assistants, simple \enquote*{white-box} automation tactics based on tableau-style tree search play a substantial role in practice: Isabelle's (TODO ref) and Rocq's (TODO ref) auto, ACL2's (TODO ref) waterfall, various PVS (TODO ref) tactics, Lean's (TODO ref) Aesop (TODO ref), etc.
 These tactics operate on \emph{goals} $Γ ⊢ A$, where the \emph{context} $Γ$ is a list of assumptions and $A$ is the proposition to be proved.
@@
 
-ACL2 and Aesop additionally provide special support for \emph{forward rules}.
-Given, for example, the transitivity lemma $\All{x~y~z}{x < y → y < z → x < z}$, a forward rule run on the goal $a < b,\, b < c ⊢ A$ produces the single subgoal $a < b,\, b < c,\, a < c ⊢ A$.
+ACL2, PVS and Aesop provide special support for \emph{forward rules}.
+Given, for example, the transitivity lemma $\All{x~y~z}{x < y → y < z → x < z}$, a forward rule run on the goal $h₁ : a < b,\, h₂ : b < c ⊢ A$ produces the single subgoal $h₁ : a < b,\, h₂ : b < c,\, h₃ : a < c ⊢ A$.
 Hence forward rules can be used to establish additional facts that may be relevant for other rules.
@@
 
-However, Aesop's (and ACL2's) support for forward rules was, until recently, quite basic.
+However, Aesop's support for forward rules was, until recently, quite basic.
 In particular, each forward rule was tried on each goal separately (subject to pre-filtering by a discrimination tree (TODO ref) index).

### 2025-02-17 paper.tex
Add some intro TODOs

@@
 These rule tactics can output arbitrary subgoals, subject only to the condition that proofs of the returned subgoals can be assembled into a proof of the goal on which the rule was run.
+\jcom{The framing here is wrong: we want to talk about non-forward rules that mess with the context, e.g. case splitting.}
 
@@
 Hence forward rules can be used to establish additional facts that may be relevant for other rules.
-For example, the Aeneas project (TODO ref) uses Aesop's forward rules \jcom{(ensure that this is accurate)} in its proof automation to establish facts that are later used by an arithmetic decision procedure.
+For example, the Aeneas project (TODO ref) uses Aesop's forward rules \jcom{ensure that this is accurate} in its proof automation to establish facts that are later used by an arithmetic decision procedure.
 

### 2025-02-17 paper.tex
Clarify and compress naive algorithm

@@
 Prior to this work, Aesop used what we call naive forward reasoning (TODO ref).
-This approach treats a forward rule essentially like any other Aesop rule.
-Hence, given a goal $h₁ : T₁,\, \dots, hₙ : Tₙ ⊢ U$, we first query an index to determine which rules might apply to the goal.
-For forward rules, this index stores the type of the last premise (since later premises tend to be more specific).
-The index is queried for each $Tᵢ$, yielding a set of rules that have a premise which likely unifies with at least one hypothesis.
-Of these rules and any non-forward rules selected in a similar way, we select the highest-priority one.
-
-If the selected rule is a forward rule $r : \All{(x₁ : A₁) \dots (xₘ : Aₘ)}{B}$, we first unify $Aₘ$ with one of the candidate hypotheses determined by the index.
-If successful, this results in a substitution $σ$ whose domain contains $xₘ$ and any variables occurring in $Aₘ$.
-We then proceed to unify $A_{m-1}[σ]$ with the type of each hypothesis in the context, etc., until we have found a complete match for $r$ (skipping premises already determined by $σ$).
-If a premise type $A_{i}[σ]$ does not unify with any hypothesis type, we backtrack and consider other potential matching hypotheses for $A_{i+1}$.
-This may lead to a different substitution $τ$ and therefore to matching hypotheses for $A_{i}[τ]$.
-
-If this process yields one or more full applications $r~a₁ \dots aₘ$, we add $B[x₁ ↦ a₁, \dots, xₘ ↦ aₘ]$ as a new hypothesis unless a hypothesis of this type is already present in the context.
-
-\jcom{Add example?}
-
-This approach is efficient when the last premise of a rule is not present in the context since the rule is then never run.
-If the last premise is present, however, we perform a large number of unifications between premises and hypotheses.
-Furthermore, these unifications are repeated when we proceed to another goal with a similar context.
-In particular, the contexts of subgoals produced by whatever rule is eventually applied to the current goal are likely to be substantially similar, if not identical, to the current context, but these similarities are not exploited since the naive algorithm is stateless.
-
-\jcom{Mention fast subsumption paper?}
+This approach treats forward rules essentially like any other Aesop rule.
+When Aesop selects a forward rule $r : \All{(x₁ : A₁) \dots (xₘ : Aₘ)}{B}$ as the highest-priority rule currently applicable to a goal $Γ ⊢ T$, we first unify the last premise $Aₘ$ with each hypothesis $h : C$ such that $C$ and $Aₘ$ likely unify (as determined by a discrimination tree (TODO ref) index).
+If successful, this results in a substitution $σ₁$ containing $xₘ$ and any variables occurring in $Aₘ$.
+We then unify $A_{m-1}[σ₁]$ with the type of each hypothesis in the context, yielding a substitution $σ₂$, and proceed with $A_{m-2}[σ₂]$, etc., until we obtain a complete match for $r$ (skipping premises $A_{m-i}$ if they are already determined by $σᵢ$).
+If a premise $A_{m-i}[σᵢ]$ does not unify with any hypothesis type, we backtrack and consider other potential matching hypotheses for $A_{i+1}$.
+These may lead to a different substitution $τ$ and therefore to matching hypotheses for $A_{i}[τ]$.
+If this process yields at least one full application $r~a₁ \dots aₘ$, we can add a new hypothesis of type $T = B[x₁ ↦ a₁, \dots, xₘ ↦ aₘ]$, meaning we generate the subgoal $Γ, h : T$ (unless $Γ$ already contains a hypothesis of type $T$).
+
+This approach is efficient when the last premise of a rule is not present in the context since the rule is then excluded by indexing and never run.
+If the last premise is present, however, we perform a large number of unifications between premises and hypotheses (though some of these could perhaps be avoided heuristically (TODO ref fast subsumption)).
+Furthermore, the whole process is repeated when we proceed to another goal with a similar context.
+In particular, the contexts of subgoals produced by whatever rule is eventually applied to the current goal are likely to be substantially similar, if not identical, to the current context, but these similarities are not exploited by the naive algorithm.
 

### 2025-02-17 paper.tex
Minor edit

@@
 The main idea is to store, for each goal $Γ ⊢ A$, the partial matches implied by the hypotheses in $Γ$, and to reuse these partial matches for subgoals $Δ ⊢ B$.
-This rests on the assumption that $Δ$ is likely similar to $Γ$ in practice, and that it is therefore cheaper to update the partial matches for $Γ$ to obtain those of $Δ$ than to compute those of $Δ$ from scratch.
+This rests on the assumption that $Δ$ is likely similar to $Γ$ in practice, and that it is therefore cheaper to update the partial matches for $Γ$ than to compute those of $Δ$ from scratch.
 Our main contribution is a data structure, the \emph{partial match state}, that allows us to perform these updates efficiently.

### 2025-02-17 paper.tex
Rename partial match state => forward state

@@
   We present an approach to forward rules that avoids some of this repeated work by taking advantage of similarities between successive goals.
-  For each goal, we cache partial applications of forward rules in a custom data structure, the \emph{partial match state}, which enables efficient updates.
+  For each goal, we cache partial applications of forward rules in a custom data structure, the \emph{forward state}, which enables efficient updates.
   Our technique is compatible with any search strategy and most logics.
@@
 Whenever this process produces a full match---i.e., all premises of a forward rule are satisfied by a set of hypothesis in $Δᵢ$---the rule is applied and adds a corresponding new hypothesis to $Δᵢ$.
-Our main contribution is a data structure, the \emph{partial match state}, that organises the partial matches in such a way that the update can be done efficiently.
-
-In addition to this \emph{incrementality}, the partial match state satisfies two other desiderata:
+Our main contribution is a data structure, the \emph{forward state}, that organises the partial matches in such a way that the update can be done efficiently.
+
+In addition to this \emph{incrementality}, the forward state satisfies two other desiderata:
 \begin{itemize}
-  \item It is \emph{persistent}, i.e.\ partial match states for parent and child goals share much of their structure.
-        This is desirable since Aesop supports search strategies other than depth-first search and must therefore keep all goals encountered during the search, along with their partial match states, in memory.
+  \item It is \emph{persistent}, i.e.\ forward states for parent and child goals share much of their structure.
+        This is desirable since Aesop supports search strategies other than depth-first search and must therefore keep all goals encountered during the search, along with their forward states, in memory.
   \item It reduces the number of \emph{unifications} performed when matching hypotheses against rule premises.
@@
 This rests on the assumption that $Δ$ is likely similar to $Γ$ in practice, and that it is therefore cheaper to update the partial matches for $Γ$ than to compute those of $Δ$ from scratch.
-Our main contribution is a data structure, the \emph{partial match state}, that allows us to perform these updates efficiently.
-
-\paragraph{Partial Match State.}
-We associate to each goal in Aesop's search tree a partial match state that reflects the goal's context.
-Let $\States$ be the set of partial match states, $\Hyps$ the set of hypotheses, $\Rules$ the set of rules, $\CMatches$ the set of complete matches and $\powerset(X)$ the powerset of $X$.
-The partial match state's interface then consists of two functions
+Our main contribution is a data structure, the \emph{forward state}, that allows us to perform these updates efficiently.
+
+\paragraph{Forward State.}
+We associate to each goal in Aesop's search tree a forward state that reflects the goal's context.
+Let $\States$ be the set of forward states, $\Hyps$ the set of hypotheses, $\Rules$ the set of rules, $\CMatches$ the set of complete matches and $\powerset(X)$ the powerset of $X$.
+The forward state's interface then consists of two functions
 \begin{equation*}
@@
 \end{equation*}
-The $\addHyp$ function adds a hypothesis to the partial match state.
-Hence, if $s$ is the partial match state corresponding to context $Γ$ and $\addHyp(s, h : T) = (s', M)$, then $s'$ is the partial match state corresponding to $Γ,\, h : T$.
-Similarly, if $s$ is the partial match state corresponding to $Γ, h : T$, then $\delHyp(s, h : T)$ is the partial match state corresponding to $Γ$.
+The $\addHyp$ function adds a hypothesis to the forward state.
+Hence, if $s$ is the forward state corresponding to context $Γ$ and $\addHyp(s, h : T) = (s', M)$, then $s'$ is the forward state corresponding to $Γ,\, h : T$.
+Similarly, if $s$ is the forward state corresponding to $Γ, h : T$, then $\delHyp(s, h : T)$ is the forward state corresponding to $Γ$.
 The second output of $\addHyp$, $M$, is a set of pairs $(r, m)$ where $r$ is a rule and $m$ is a complete match for $r$.
@@
 
-When Aesop is run on an initial goal with context $h₁ : T₁, \dots, hₙ : Tₙ$, the partial match state for it is built by starting with an empty state and applying $\addHyp$ to each hypothesis $hᵢ$.
-Afterwards, whenever an Aesop rule is run on a goal with context $Γ$, producing subgoals with contexts $Δ₁, \dots, Δₙ$, we derive the partial match state for each subgoal context $Δⱼ$ from the partial match state for $Γ$, in two steps:
+When Aesop is run on an initial goal with context $h₁ : T₁, \dots, hₙ : Tₙ$, the forward state for it is built by starting with an empty state and applying $\addHyp$ to each hypothesis $hᵢ$.
+Afterwards, whenever an Aesop rule is run on a goal with context $Γ$, producing subgoals with contexts $Δ₁, \dots, Δₙ$, we derive the forward state for each subgoal context $Δⱼ$ from the forward state for $Γ$, in two steps:
 \begin{enumerate}
@@
 \paragraph{Rule State.}
-Partial matches are specific to each rule, so the partial match state naturally decomposes into a set of \emph{rule states}, one per rule.
+Partial matches are specific to each rule, so the forward state naturally decomposes into a set of \emph{rule states}, one per rule.
 Given a goal $Γ ⊢ A$ and a rule $r$, the rule state for $r$ contains exactly the partial matches of $r$ in context $Γ$.
-A rule state supports $\addHyp$ and $\delHyp$ methods analogous to those on partial match states.
-Adding a hypothesis to a partial match state then amounts to adding it to each rule state (which may yield new complete matches for the rule); deleting a hypothesis amounts to deleting it from each rule state.
+A rule state supports $\addHyp$ and $\delHyp$ methods analogous to those on forward states.
+Adding a hypothesis to a forward state then amounts to adding it to each rule state (which may yield new complete matches for the rule); deleting a hypothesis amounts to deleting it from each rule state.
 We initially focus on how to add hypotheses, since this question motivates much of the design of the rule state data structure.
@@
 \paragraph{Deletion.}
-When a hypothesis $h$ is removed from a goal's context, we must update the partial match state accordingly.
+When a hypothesis $h$ is removed from a goal's context, we must update the forward state accordingly.
 This means deleting $h$ from any rule state to which it was previously added, i.e.\ from any rule state for a rule $r$ with a slot $i$ matching $h$.
@@
 SATCHMO's search procedure can be viewed as a special case of our technique: a given set of implications are used as forward rules and a single case-splitting rule splits any goal with hypothesis $h : B₁ ∨ \dots ∨ Bₘ$ into $m$ goals with hypotheses $h₁ : B₁, \dots, hₘ : Bₘ$.
-A SATCHMO-style prover may benefit from our incremental approach if the set of implications is large, though in a setting where unification is cheap it is not clear whether avoiding repeated unifications outweighs the cost of maintaining the partial match state.
+A SATCHMO-style prover may benefit from our incremental approach if the set of implications is large, though in a setting where unification is cheap it is not clear whether avoiding repeated unifications outweighs the cost of maintaining the forward state.
 SATCHMO also employs an optimisation that \enquote*{focuses} forward reasoning on the hypothesis established by the last case split, which provides a form of incrementality.

### 2025-02-17 paper.tex
Mathcal powerset
Write H(x, i, t) as H_x(i, t) and similar for M

@@
 \newcommand{\addMatch}{\ensuremath{\mathrm{addMatch}}}
-\newcommand{\powerset}{\ensuremath{\mathbb{P}}}
+\newcommand{\powerset}{\ensuremath{\mathcal{P}}}
 
@@
 Matches for the maximal slot are not stored because, being complete, they never need to be extended.
+We write $M_{x}(i, t)$ instead of $M(x, i, t)$ and $H_{x}(i, t)$ instead of $H(x, i, t)$.
 
@@
 \[
-  \bigcap_{x \in \vars(i)} M(x, i-1, σ(x))
+  \bigcap_{x \in \vars(i)} M_{x}(i-1, σ(x))
 \]
@@
 \[
-  \bigcap_{x \in \vars(i+1)} H(x, i+1, σ(x))
+  \bigcap_{x \in \vars(i+1)} H_{x}(i+1, σ(x))
 \]
@@
       \For{$x$ \KwIn $\vars(i)$}{
-        Insert $h$ into the set $H(x, i, σ(h))$
+        Insert $h$ into the set $H_{x}(i, σ(h))$
       }
-      \KwLet $\mathit{ms} ≔ \bigcap_{x \in \vars(i)} M(x, i-1, σ(x))$ \\
+      \KwLet $\mathit{ms} ≔ \bigcap_{x \in \vars(i)} M_{x}(i-1, σ(x))$ \\
       $\mathit{completeMs} \gets ∅$ \\
@@
       \For{$x$ \KwIn $\vars(i + 1)$}{
-        Insert $m$ into the set $M(x, i, σ(x))$
+        Insert $m$ into the set $M_{x}(i, σ(x))$
       }
-      \KwLet $\mathit{hs} ≔ \bigcap_{x \in \vars(i + 1)} H(x, i + 1, σ(x))$ \\
+      \KwLet $\mathit{hs} ≔ \bigcap_{x \in \vars(i + 1)} H_{x}(i + 1, σ(x))$ \\
       $\mathit{completeMs} \gets ∅$ \\
@@
 Suppose we are adding the hypothesis $h₂ : Q~b$, having already added the hypothesis $h₁ : P~a$ and hence the partial match $\{1 ↦ h₁\}$.
-The set of matches compatible with $h₂$ is supposedly $\bigcap_{x \in \vars(2)} M(x, i-1, σ(x)) = ∅$, but in fact the match $\{1 ↦ h₁\}$ can be extended with $h₂$.
+The set of matches compatible with $h₂$ is supposedly $\bigcap_{x \in \vars(2)} M_{x}(i-1, σ(x)) = ∅$, but the match $\{1 ↦ h₁\}$ can be extended with $h₂$.
 In fact, any match at level 1 can be extended with any hypothesis matching slot 2 since the two slots are independent.
 
-To account for this fact, we partition the set of premises of each rule into maximal subsets of premises connected by shared variables, using a construction from (TODO ref).
+To account for this fact, we partition the set of premises of each rule into maximal subsets of premises connected by shared variables, using a construction from (TODO ref Aesop Sec. 4.2).
 These subsets, called \emph{variable clusters}, share no variables with each other, so each can be treated as a separate rule with its own rule state.
@@
 For slot 1, unification yields the substitution $\{x ↦ 301, y ↦ 302\}$.
-Since this is the first slot, we start a new partial match $\{1 ↦ h₁\}$, which Algorithm~\ref{alg:addmatch} inserts into the set $M(y, 1, 302)$.
+Since this is the first slot, we start a new partial match $\{1 ↦ h₁\}$, which Algorithm~\ref{alg:addmatch} inserts into the set $M_{y}(1, 302)$.
 This match cannot be extended since the hypothesis index $H$ does not contain a hypothesis for slot 2.
 
-For slot 2, unification yields the substitution $σ ≔ \{y ↦ 301, z ↦ 302\}$, so we insert $h₁$ into $H(2, y, 301)$.
+For slot 2, unification yields the substitution $σ ≔ \{y ↦ 301, z ↦ 302\}$, so we insert $h₁$ into $H_{y}(2, 301)$.
 No matches at level 1 can be extended by $h₁$ since
 \[
-  \bigcap_{x \in \vars(2)} M(x, 1, σ(x)) = M(y, 1, 301) = \emptyset
+  \bigcap_{x \in \vars(2)} M_{x}(1, σ(x)) = M_{y}(1, 301) = \emptyset
 \]
@@
 For brevity, we write matches $\{1 ↦ h₁, \dots, n ↦ hₙ\}$ as $[h₁, \dots, hₙ]$.
-We also write $H_{y}$ instead of $H(y, \text{--}, \text{--})$, and similar for $M$.
 \begin{center}
@@
 Now we add $h₂ : 300 ≤ 301$, which again unifies with both slots.
-For slot 1, $\sub₁(h₂)(y) = 301$, so we add the match $\{1 ↦ h₂\}$ to $M(y, 1, 301)$.
-We then check for hypotheses in slot 2 that also instantiate $y$ with $301$, and indeed we find $M(2, y, 301) = \{h₁\}$.
+For slot 1, $\sub₁(h₂)(y) = 301$, so we add the match $\{1 ↦ h₂\}$ to $M_{y}(1, 301)$.
+We then check for hypotheses in slot 2 that also instantiate $y$ with $301$, and indeed we find $M_{y}(2, 301) = \{h₁\}$.
 Hence we return the complete match $\{1 ↦ h₂, 2 ↦ h₁\}$, which induces a new hypothesis $h₃ : 300 ≤ 302$ to be added later.
 
-For slot 2, we add $h₂$ to $H(y, 2, 300)$.
+For slot 2, we add $h₂$ to $H_{y}(2, 300)$.
 There are no compatible matches at level 1, so we are done with $h₂$, yielding the following final rule state.
@@
 \[
-  \bigcap_{x \in \vars(\lvl(m) + 1)} H(x, i + 1, \sub(m)(x)) = H(x, 2, a) \cap H(y, 2, b) = \emptyset
+  \bigcap_{x \in \vars(\lvl(m) + 1)} H_{x}(i + 1, \sub(m)(x)) = H_{x}(2, a) \cap H_{y}(2, b) = \emptyset
 \]
@@
 \[
-  \bigcap_{x \in \vars(i)} M(x, i - 1, \sub_{i}(x)) = M(x, 1, a) \cap M(y, 1, b) = \{[h₂]\}
+  \bigcap_{x \in \vars(i)} M_{x}(i - 1, \sub_{i}(x)) = M_{x}(1, a) \cap M_{y}(1, b) = \{[h₂]\}
 \]
@@
 There are unfortunately no standard benchmarks for Aesop-like tactics, and since Aesop is not expected to solve most goals without an extensive, manually curated set of rules, benchmarks for push-button theorem provers such as TPTP (TODO ref) are not very meaningful.
-At the same time, forward rules are currently little used by Aesop's main client, the Mathlib mathematics library (TODO ref).
-We therefore evaluate our technique against the previous, naive Aesop implementation on several synthetic benchmarks that showcase the performance characteristics of incremental forward reasoning.
+Additionally, forward rules are currently little used by Aesop's main client, the Mathlib mathematics library (TODO ref).
+We therefore only evaluate our technique against the previous, naive Aesop implementation on several synthetic benchmarks that showcase the performance characteristics of incremental forward reasoning.
 

### 2025-02-17 paper.tex
Define matches as lists

@@
 
-A \textit{match} $M$ for a given rule $r$ is a partial map from slots of $r$ to hypotheses.
-The map must satisfy the following conditions:
+A \textit{match} $m$ for a give rule $r$ with $n$ slots is a list containing at least one and at most $n$ hypotheses.
+We interpret $m$ as a partial map from slots of $r$ to hypotheses, so $\dom(m) = \{i \mid 1 ≤ i ≤ |m| \}$ and for each slot $i ∈ \dom(m)$, $m(i)$ is the hypothesis assigned to $i$.
+Each match $m$ must satisfy the following requirements:
 \begin{enumerate}
-  \item\label{cond:match:matches} For each slot $i ∈ \dom(M)$, the hypothesis $M(i)$ matches $i$.
-  \item\label{cond:match:compat} For all slots $i,j ∈ \dom(M)$, the substitutions $\sub_{i}(M(i))$ and $\sub_{j}(M(j))$ are \emph{compatible}.
-        This means that for each variable $x ∈ \dom(\sub_{i}(M(i))) ∩ \dom(\sub_{j}(M(j)))$ we have $\sub_{i}(M(i))(x) = \sub_{j}(M(j))(x)$, i.e.\ the substitutions agree on all common variables.
-  \item\label{cond:match:closed} $M$ is \emph{downward-closed}: if $i ∈ \dom(M)$, then for any slot $j < i$ of $r$ we also have $j ∈ \dom(M)$.
+  \item\label{cond:match:matches} For each slot $i ∈ \dom(m)$, the hypothesis $m(i)$ matches $i$.
+  \item\label{cond:match:compat} For all slots $i,j ∈ \dom(m)$, the substitutions $\sub_{i}(m(i))$ and $\sub_{j}(m(j))$ are \emph{compatible}.
+        This means that the substitutions agree on all common variables, i.e.\ for each variable $x ∈ \dom(\sub_{i}(m(i))) ∩ \dom(\sub_{j}(m(j)))$ we have $\sub_{i}(m(i))(x) = \sub_{j}(m(j))(x)$.
 \end{enumerate}
-The \emph{substitution of a match $M$}, $\sub(M)$, is the union of the substitutions induced by the hypotheses in $M$, i.e. $\sub(M) ≔ ⋃_{i ∈ \dom(M)}\sub_{i}(M(i))$.
+The \emph{substitution of a match $m$}, $\sub(m)$, is the union of the substitutions induced by the hypotheses in $m$, i.e. $\sub(m) ≔ ⋃_{i ∈ \dom(m)}\sub_{i}(m(i))$.
 This is well-defined by requirement~\ref{cond:match:compat}.
-The \textit{level of a match $M$}, $\lvl(M)$, is the maximal $i$ such that $i ∈ \dom(M)$.
+The \textit{level of a match $m$}, $\lvl(m)$, is the length of $m$.
 A match for rule $r$ is \textit{complete} if $i ∈ \dom(M)$ for every slot $i$ of $r$, and \emph{partial} otherwise.
 
-A partial match $M$ for rule $r$ corresponds to a partial application of the rule to the terms given in $\sub(M)$.
+A partial match $m$ for rule $r$ corresponds to a partial application of the rule to the terms given in $\sub(M)$.
 The application is type-correct by requirements~\ref{cond:match:matches} and~\ref{cond:match:compat}.
-If $M$ is complete, then $\sub(M)$ contains a term for every slot and every variable---hence for every premise---of $r$ and we have a full application.
-
-Requirement~\ref{cond:match:closed} implies that we will be adding hypotheses to matches in the order in which the slots appear.
-Hence for a rule $r : A → B → C → D$, we effectively consider partial applications $r~a$ and $r~a~b$ for some $a : A$ and $b : B$, but not $r~\text{--}~b$.
-This prevents us from constructing $r~a~b$ in two different but ultimately redundant ways.
+If $m$ is complete, then $\sub(m)$ contains a term for every slot and every variable---hence for every premise---of $r$ and we have a full application.
+
+Note that matches represent only those partial applications where arguments are added without gaps.
+Hence for a rule $r : A → B → C → D$, the partial applications $r~a$ and $r~a~b$ for some $a : A$ and $b : B$ correspond to matches, but not $r~\text{--}~b$.
+This will prevent us from constructing $r~a~b$ in two different but ultimately redundant ways.
 The term orders used in various saturation calculi serve a similar function.
@@
 
+\jcom{Formulate this in terms of matches.}
 Prior to this work, Aesop used what we call naive forward reasoning (TODO ref).
@@
 The $\addHyp$ function adds a hypothesis to the forward state.
-Hence, if $s$ is the forward state corresponding to context $Γ$ and $\addHyp(s, h : T) = (s', M)$, then $s'$ is the forward state corresponding to $Γ,\, h : T$.
+Hence, if $s$ is the forward state corresponding to context $Γ$ and $\addHyp(s, h : T) = (s', C)$, then $s'$ is the forward state corresponding to $Γ,\, h : T$.
 Similarly, if $s$ is the forward state corresponding to $Γ, h : T$, then $\delHyp(s, h : T)$ is the forward state corresponding to $Γ$.
-The second output of $\addHyp$, $M$, is a set of pairs $(r, m)$ where $r$ is a rule and $m$ is a complete match for $r$.
+The second output of $\addHyp$, $C$, is a set of pairs $(r, m)$ where $r$ is a rule and $m$ is a complete match for $r$.
 These are the matches that were completed by adding $h : T$.
@@
 First, the hypothesis may allow us to extend partial matches already present in the state.
-For instance, consider the rule $r : \All{x}{P~x → Q~x → R~x → S~x}$ and a rule state containing the partial match $\{1 ↦ h₁\}$, with $1$ referring to the slot of type $P~x$ and $h₁ : P~a$ for some $a$.
-Adding a new hypothesis $h₂ : Q~a$ to this rule state means that we should add to the state the extended match $\{1 ↦ h₁, 2 ↦ h₂\}$.
+For instance, consider the rule $r : \All{x}{P~x → Q~x → R~x → S~x}$ and a rule state containing the partial match $[h₁]$, which assigns some hypothesis $h₁ : P~a$ to the first slot of type $P~x$.
+Adding a new hypothesis $h₂ : Q~a$ to this rule state means that we should add to the state the extended match $[h₁, h₂]$.
 
@@
 Suppose that before adding $h₁$ and $h₂$, we had already added a hypothesis $h₀ : R~a$ to the rule state.
-Once the partial match $\{1 ↦ h₁, 2 ↦ h₂\}$ arrives, we should recognise that it can be extended with $h₀$ to obtain the (complete) match $\{1 ↦ h₁, 2 ↦ h₂, 3 ↦ h₀\}$.
+Once the partial match $[h₁, h₂]$ arrives, we should recognise that it can be extended with $h₀$ to obtain the (complete) match $[h₁, h₂, h₀]$.
 
@@
 Extending the matches above was possible only because all hypotheses are compatible on the shared variable $x$, i.e.\ they all instantiate $x$ with the same term $a$.
-In general, to extend a match $M$ with a hypothesis $h : T$ at slot $i$, we need to ensure that the substitution of $M$, $\sub(M)$, is compatible with the substitution of $h$, $\sub_{i}(h)$.
+In general, to extend a match $m$ with a hypothesis $h : T$ at slot $i$, we need to ensure that the substitution of $m$, $\sub(m)$, is compatible with the substitution of $h$, $\sub_{i}(h)$.
 
@@
     \eIf{$i = 1$}{
-      $\mathrm{addMatch}(\{1 ↦ h\})$
+      $\mathrm{addMatch}([h])$
     }{
@@
       \For{$m$ \KwIn $\mathit{ms}$}{
-        $\mathit{completeMs} \gets \mathit{completeMs} ∪ \mathrm{addMatch}(m ∪ \{i ↦ h\})$
+        $\mathit{completeMs} \gets \mathit{completeMs} ∪ \mathrm{addMatch}(m ++ [h])$
       }
@@
 (If a hypothesis matches multiple slots, $\addHyp$ is called multiple times.)
-For $i = 1$, we add the singleton match $\{1 ↦ h\}$ to the rule state.
+For $i = 1$, we add the singleton match $[h]$ to the rule state.
 For $i > 1$, we first insert $h$ into the hypothesis index $H$.
@@
       \For{$h$ \KwIn $\mathit{hs}$}{
-        $\mathit{completeMs} \gets \mathit{completeMs} ∪ \mathrm{addMatch}(m ∪ \{i + 1 ↦ h\})$
+        $\mathit{completeMs} \gets \mathit{completeMs} ∪ \mathrm{addMatch}(m ++ [h])$
       }
@@
 As a result, our above algorithms fail.
-Suppose we are adding the hypothesis $h₂ : Q~b$, having already added the hypothesis $h₁ : P~a$ and hence the partial match $\{1 ↦ h₁\}$.
-The set of matches compatible with $h₂$ is supposedly $\bigcap_{x \in \vars(2)} M_{x}(i-1, σ(x)) = ∅$, but the match $\{1 ↦ h₁\}$ can be extended with $h₂$.
+Suppose we are adding the hypothesis $h₂ : Q~b$, having already added the hypothesis $h₁ : P~a$ and hence the partial match $[h₁]$.
+The set of matches compatible with $h₂$ is supposedly $\bigcap_{x \in \vars(2)} M_{x}(i-1, σ(x)) = ∅$, but the match $[h₁]$ can be extended with $h₂$.
 In fact, any match at level 1 can be extended with any hypothesis matching slot 2 since the two slots are independent.
@@
 \paragraph{Example 1.}
-As a first example, consider a rule state for the transitivity rule $r : \All{x~y~z}{x ≤ y → y ≤ z → x ≤ z}$ to which we add, in order, the hypotheses
+As a first example, consider a rule state for the transitivity rule $r₁ : \All{x~y~z}{x ≤ y → y ≤ z → x ≤ z}$ to which we add, in order, the hypotheses
 $h₁ : 301 ≤ 302$ and $h₂ : 300 ≤ 301$.
@@
 For slot 1, unification yields the substitution $\{x ↦ 301, y ↦ 302\}$.
-Since this is the first slot, we start a new partial match $\{1 ↦ h₁\}$, which Algorithm~\ref{alg:addmatch} inserts into the set $M_{y}(1, 302)$.
+Since this is the first slot, we start a new partial match $[h₁]$, which Algorithm~\ref{alg:addmatch} inserts into the set $M_{y}(1, 302)$.
 This match cannot be extended since the hypothesis index $H$ does not contain a hypothesis for slot 2.
@@
 At this point, $H$ and $M$ are in the following state.
-For brevity, we write matches $\{1 ↦ h₁, \dots, n ↦ hₙ\}$ as $[h₁, \dots, hₙ]$.
 \begin{center}
@@
 Now we add $h₂ : 300 ≤ 301$, which again unifies with both slots.
-For slot 1, $\sub₁(h₂)(y) = 301$, so we add the match $\{1 ↦ h₂\}$ to $M_{y}(1, 301)$.
+For slot 1, $\sub₁(h₂)(y) = 301$, so we add the match $[h₂]$ to $M_{y}(1, 301)$.
 We then check for hypotheses in slot 2 that also instantiate $y$ with $301$, and indeed we find $M_{y}(2, 301) = \{h₁\}$.
-Hence we return the complete match $\{1 ↦ h₂, 2 ↦ h₁\}$, which induces a new hypothesis $h₃ : 300 ≤ 302$ to be added later.
+Hence we return the complete match $[h₂, h₁]$, which induces a new hypothesis $h₃ : 300 ≤ 302$ to be added later.
 
@@
 \paragraph{Example 2.}
-We now consider a rule $s : \All{x~y}{A~x~y → B~x~y → C~x~y}$ with multiple shared variables.
-We add hypotheses $h₁ : B~a~a$, $h₂ : A~a~b$ and $h₃ : B~a~b$ to the rule state of $s$, starting with $h₁$:
+We now consider a rule $r₂ : \All{x~y}{A~x~y → B~x~y → C~x~y}$ with multiple shared variables.
+We add hypotheses $h₁ : B~a~a$, $h₂ : A~a~b$ and $h₃ : B~a~b$ to the rule state of $r₂$, starting with $h₁$:
 % NOTE : remove comments to get full trees.
@@
 At this point there are no matches, so there is nothing to extend.
-We now add $h_2 : A~a~b$, which initially generates a match $m = \{1 ↦ h₂\}$ that is added to $M$:
+We now add $h_2 : A~a~b$, which initially generates a match $m = [h₂]$ that is added to $M$:
 % NOTE : uncomment to get full trees.
@@
 \paragraph{Match Equivalence.}
-Naively, two matches $\{1 ↦ h₁, \dots, n ↦ hₙ\}$ and $\{1 ↦ h₁', \dots, n ↦ hₙ'\}$ would be considered equal if they are structurally equal, i.e.\ $h₁ = h₁'$, \dots, $hₙ = hₙ'$.
+Naively, two matches $[h₁, \dots, hₙ]$ and $[h₁', \dots, hₙ']$ would be considered equal if they are structurally equal, i.e.\ $h₁ = h₁'$, \dots, $hₙ = hₙ'$.
 However, consider the rule $r : \All{x~y}{P~x~y → Q~x~z → R~x}$ and the hypotheses $h₁ : P~a~b$, $h₂ : P~a~c$, $h₃ : Q~a~d$ and $h₄ : Q~a~e$.
-The complete matches $m₁ ≔ \{1 ↦ h₁, 2 ↦ h₃\}$ and $m₁' ≔ \{1 ↦ h₁, 2 ↦ h₄\}$ are structurally different, but the difference does not matter because the applications $r~h₁~h₃$ and $r~h₁~h₄$ both have type $R~a$.
+The complete matches $m₁ ≔ [h₁, h₃]$ and $m₁' ≔ [h₁, h₄]$ are structurally different, but the difference does not matter because the applications $r~h₁~h₃$ and $r~h₁~h₄$ both have type $R~a$.
 Generating both $m₁$ and $m₁'$ would therefore be redundant.
-Similarly, the partial matches $m₂ ≔ \{1 ↦ h₁\}$ and $m₂' ≔ \{1 ↦ h₂\}$ are redundant because the choice between $h₁$ and $h₂$ affects neither the type of any eventual application derived from $m₂$ and $m₂'$ nor the hypotheses that can be used to complete $m₂$ and $m₂'$.
+Similarly, the partial matches $m₂ ≔ [h₁]$ and $m₂' ≔ [h₂]$ are redundant because the choice between $h₁$ and $h₂$ affects neither the type of any eventual application derived from $m₂$ and $m₂'$ nor the hypotheses that can be used to complete $m₂$ and $m₂'$.
 Hence we should store only one of $m₂$ and $m₂'$ in the rule state of $r$.
@@
 The procedures are then run on a context containing exactly $\mathcal{P}$.
-In this case, we sketch the behavior of the stateless algorithm:
+In this case, we sketch the behaviour of the stateless algorithm:
 It pick a first hypothesis, $h_{i,j} : P_{i,j}$.

### 2025-02-17 paper.tex
Minor edits

@@
 \paragraph{Lazy Insertion.}
-Adding a hypothesis to a rule state is fairly cheap, but not free: it involves one unification for each premise that the hypothesis likely matches (according to a discrimination tree index), followed by several operations on the variable indices.
+Adding a hypothesis to a rule state is fairly cheap, but not free: it involves one unification for each slot that the hypothesis likely matches (according to a discrimination tree index), followed by several operations on the variable indices.
 We therefore delay insertions until the index has selected at least one hypothesis for each slot.
@@
     \end{tikzpicture}
-    \caption{Transitivity benchmark}
+    \caption{Transitivity benchmark}%
     \label{fig:trans}
@@
     % graph \includegraphics[width=.4\linewidth]{image1}
-    \caption{Independence benchmark}
+    \caption{Independence benchmark}%
     \label{fig:indep}
@@
 The lines with square entry points represent problems where unification is easy.
-The lines with circle entry points are the same tests with hard unification problems.}
+The lines with circle entry points are the same tests with hard unification problems.}%
   \label{fig:benchmark}
@@
     \end{tikzpicture}
-    \caption{Depth benchmark no-precompilation}
+    \caption{Depth benchmark no-precompilation}%
     \label{fig:erase-noprecomp}
@@
     % graph \includegraphics[width=.4\linewidth]{image1}
-    \caption{Depth benchmark with precompilation}
+    \caption{Depth benchmark with precompilation}%
     \label{fig:erase-precomp}
@@
 The lines with square entry points represent problems where unification is easy.
-The lines with circle entry points are the same tests with hard unification problems.}
+The lines with circle entry points are the same tests with hard unification problems.}%
   \label{fig:erase}

### 2025-02-18 paper.tex
Move example rules states into a figure

@@
 \]
-At this point, $H$ and $M$ are in the following state.
-\begin{center}
+At this point, $H$ and $M$ are in the state shown in Fig.~\ref{fig:example1:state1}.
+
+\begin{figure}
+  \centering
+  \begin{minipage}{\textwidth}
+    \begin{subcaptiongroup}
+      \subcaptionlistentry{}
+      \begin{minipage}{0.5\textwidth}
         \begin{tikzpicture}[outer sep=auto, level distance=12mm, scale=0.50]
+          \node at (-2,0) {\captiontext*{}\label{fig:example1:state1}};
           \node[xshift = 0] {$H_{y}$}
@@
             };
-  \node[xshift=160] {$M_y$}
+          \node[xshift=80] {$M_y$}
             child {node {1}
@@
         \end{tikzpicture}
-\end{center}
-
-Now we add $h₂ : 300 ≤ 301$, which again unifies with both slots.
-For slot 1, $\sub₁(h₂)(y) = 301$, so we add the match $[h₂]$ to $M_{y}(1, 301)$.
-We then check for hypotheses in slot 2 that also instantiate $y$ with $301$, and indeed we find $M_{y}(2, 301) = \{h₁\}$.
-Hence we return the complete match $[h₂, h₁]$, which induces a new hypothesis $h₃ : 300 ≤ 302$ to be added later.
-
-For slot 2, we add $h₂$ to $H_{y}(2, 300)$.
-There are no compatible matches at level 1, so we are done with $h₂$, yielding the following final rule state.
-\begin{center}
+      \end{minipage}
+    \end{subcaptiongroup}
+    \begin{subcaptiongroup}
+      \subcaptionlistentry{}
+      \begin{minipage}{0.5\textwidth}
         \begin{tikzpicture}[outer sep=auto, level distance=12mm, scale=0.50]
+          \node at (-2,0) {\captiontext*{}\label{fig:example1:state2}};
           \node[xshift = 0] {$H_y$}
@@
             };
-      \node[xshift=160] {$M_y$}
+          \node[xshift=80] {$M_y$}
             child {node {1}
@@
         \end{tikzpicture}
-    \end{center}
-
-\paragraph{Example 2.}
-We now consider a rule $r₂ : \All{x~y}{A~x~y → B~x~y → C~x~y}$ with multiple shared variables.
-We add hypotheses $h₁ : B~a~a$, $h₂ : A~a~b$ and $h₃ : B~a~b$ to the rule state of $r₂$, starting with $h₁$:
-% NOTE : remove comments to get full trees.
-\begin{center}
-  \begin{tikzpicture}[level distance=12mm, scale=0.50]
-    \node[xshift = 30] {$H_x$}
-        %child[xshift = -20] {node {1}
-        %  child {node {$a$}
-        %    child {node {$\{ \}$}}}
-        %  child {node {$b$}
-        %    child {node {$\{ \}$}}}}
-        child {node {2}
-        child {node {$a$}
-            child {node {$\{h_1\}$}}}
-        % child {node {$b$}
-        %    child {node {$\{ \}$}}}
-        };
-    \node[xshift = 110] {$H_y$}
-        %child[xshift = -20] {node {1}
-        %    child {node {$a$}
-        %        child {node {$\{ \}$}}}
-        %    child {node {$b$}
-        %        child {node {$\{ \}$}}}}
-        child {node {2}
-            child {node {$a$}
-                child {node {$\{h_1\}$}}}
-        %    child {node {$b$}
-        %        child {node {$\{ \}$}}}
-        };
-    \node[xshift = 190] {$M_x$};
-        %child[xshift = -20] {node {1}
-        %    child {node {$a$}
-        %        child {node {$\{ \}$}}}
-        %    child {node {$b$}
-        %        child {node {$\{ \}$}}}}
-        %child[xshift = 20] {node {2}
-        %    child {node {$a$}
-        %        child {node {$\{\}$}}}
-        %    child {node {$b$}
-        %        child {node {$\{ \}$}}}
-        %};
-    \node[xshift = 270] {$M_y$};
-        %child[xshift = -20] {node {1}
-        %    child {node {$a$}
-        %        child {node {$\{ \}$}}}
-        %    child {node {$b$}
-        %        child {node {$\{ \}$}}}}
-        %child[xshift = 20] {node {2}
-        %    child {node {$a$}
-        %        child {node {$\{\}$}}}
-        %    child {node {$b$}
-        %        child {node {$\{ \}$}}}
-        %};
-  \end{tikzpicture}
-\end{center}
-At this point there are no matches, so there is nothing to extend.
-We now add $h_2 : A~a~b$, which initially generates a match $m = [h₂]$ that is added to $M$:
-% NOTE : uncomment to get full trees.
-\begin{center}
+      \end{minipage}
+    \end{subcaptiongroup}
+  \end{minipage}
+  \\[1em]
+  \begin{subcaptiongroup}
+    \subcaptionlistentry{}
+    \begin{minipage}{\textwidth}
       \begin{tikzpicture}[outer sep=auto, level distance=12mm, scale=0.50]
+        \node at (-2,0) {\captiontext*{}\label{fig:example2:state1}};
         \node[xshift = 0] {$H_x$}
@@
       \end{tikzpicture}
-\end{center}
-To find hypotheses compatible with $m$, we consider the set of variables $\vars(\lvl(m) + 1)$.
-These are the variables shared between slot $\lvl(m) + 1 = 2$ and all previous slots, i.e.\ $x$ and $y$.
-Hence the compatible hypotheses are
-\[
-  \bigcap_{x \in \vars(\lvl(m) + 1)} H_{x}(i + 1, \sub(m)(x)) = H_{x}(2, a) \cap H_{y}(2, b) = \emptyset
-\]
-This means the match cannot currently be extended.
-We now turn to the remaining hypothesis $h_3 : B~a~b$, adding it first to $H$:
-\begin{center}
+    \end{minipage}
+  \end{subcaptiongroup}
+  \\[1em]
+  \begin{subcaptiongroup}
+    \subcaptionlistentry{}
+    \begin{minipage}{\textwidth}
       \begin{tikzpicture}[outer sep=auto, level distance=12mm, scale=0.50]
+        \node at (-2,0) {\captiontext*{}\label{fig:example2:state2}};
         \node[xshift = 0] {$H_x$}
@@
       \end{tikzpicture}
-\end{center}
+    \end{minipage}
+  \end{subcaptiongroup}
+  \caption{Rule states encountered during Examples~1 and~2}
+\end{figure}
+
+Now we add $h₂ : 300 ≤ 301$, which again unifies with both slots.
+For slot 1, $\sub₁(h₂)(y) = 301$, so we add the match $[h₂]$ to $M_{y}(1, 301)$.
+We then check for hypotheses in slot 2 that also instantiate $y$ with $301$, and indeed we find $M_{y}(2, 301) = \{h₁\}$.
+Hence we return the complete match $[h₂, h₁]$, which induces a new hypothesis $h₃ : 300 ≤ 302$ to be added later.
+
+For slot 2, we add $h₂$ to $H_{y}(2, 300)$.
+There are no compatible matches at level 1, so we are done with $h₂$, yielding the final rule state in Fig.~\ref{fig:example1:state2}.
+
+\paragraph{Example 2.}
+We now consider a rule $r₂ : \All{x~y}{A~x~y → B~x~y → C~x~y}$ with multiple shared variables.
+We add hypotheses $h₁ : B~a~a$, $h₂ : A~a~b$ and $h₃ : B~a~b$ to the rule state of $r₂$, starting with $h₁$, which is added to $H$ with slot 2.
+There are no matches yet, so there is nothing to extend.
+We then add $h_2 : A~a~b$, which initially generates a match $m = [h₂]$ that is added to $M$, yielding the rule state in Fig.~\ref{fig:example2:state1}.
+% NOTE : uncomment to get full trees.
+To find hypotheses compatible with $m$, we consider the set of variables $\vars(\lvl(m) + 1)$.
+These are the variables shared between slot $\lvl(m) + 1 = 2$ and all previous slots, i.e.\ $x$ and $y$.
+Hence the compatible hypotheses are
+\[
+  \bigcap_{x \in \vars(\lvl(m) + 1)} H_{x}(i + 1, \sub(m)(x)) = H_{x}(2, a) \cap H_{y}(2, b) = \emptyset
+\]
+This means the match cannot currently be extended.
+We now turn to the remaining hypothesis $h_3 : B~a~b$, adding it first to $H$.
+This yields the rule state in Fig.~\ref{fig:example2:state2}.
 Since $h₃$ matches slot $i = 2$, the matches that can be extended by $h₃$ are
@@
 \]
-We have found the complete match $[h₂, h₃]$ and can add a new hypothesis of type $C~a~b$.
+This produces the complete match $[h₂, h₃]$, so we can add a new hypothesis of type $C~a~b$.
 

### 2025-02-18 paper.tex
partial eval changes

@@
 \def\orcid#1{{\href{http://orcid.org/#1}{\protect\raisebox{-1.25pt}{\protect\includegraphics{orcid.pdf}}}}}
+\def\showpgfcircle{\tikz[baseline=-0.7ex]\node[mark size=0.7ex]
+{\pgfuseplotmark{o}};}
+\def\showpgfsquare{\tikz[baseline=-0.7ex]\node[mark size=0.7ex]
+{\pgfuseplotmark{square}};}
 \SetKw{KwIn}{in}
@@
 
-For each benchmark, we set up certain forward rules and a goal with certain hypotheses.
+For each benchmark, we set up certain forward rules and a goal with handpicked hypotheses.
 We then run the \texttt{saturate} tactic, which uses either naive or incremental forward reasoning to exhaustively apply the forward rules.
@@
 
+In what follows, instead of the natural number type of Lean's standard library, which has special support in the compiler and kernel, we use a custom type of Peano naturals.
+Unifying two concrete Peano natural numbers $n$ and $m$ involves comparing $\min(n, m)$ constructors.
+We can therefore use moderately large numbers to simulate the bigger unification problems that Aesop encounters, for example, in Mathlib's category theory development.
+
 Benchmark code and data are available in our supplement (TODO ref).
-The results shown below were obtain on a Macbook Pro with an Apple~M2 Pro processor and \SI{32}{GB} of RAM, averaging over \textcolor{red}{N} (TODO) runs per benchmark.
+The results shown below were obtained on a MacBook Pro with an Apple~M2 Pro processor and \SI{32}{GB} of RAM.
+The results are an average of \textcolor{red}{N} (TODO) runs per benchmark.
 
@@
 Our first benchmark is a scaled-up variant of the transitivity example from Sec.~\ref{sec:incremental}.
-Given some relation $≺$ on natural numbers, we register the lemma $\All{x~y}{x ≺ y → y ≺ z → x ≺ z}$ as the sole forward rule and set up a goal with hypotheses $h₁ : a ≺ a + 1, \dots, hₙ : a + n - 1 ≺ a + n$ (TODO what is $a$?).
+Given some relation $≺$ on natural numbers, we register the lemma $\All{x~y}{x ≺ y → y ≺ z → x ≺ z}$ as the sole forward rule and set up a goal with hypotheses $h₁ : a ≺ a + 1, \dots, hₙ : a + n - 1 ≺ a + n$.
+The test is run for $a = 0$, represented with a square \protect\showpgfsquare\, and $a = 100$, represented with a circle \protect\showpgfcircle\.
 Saturating this goal adds $n(n-1)/2$ hypotheses to the context.
-
-Instead of the natural number type of Lean's standard library, which has special support in the compiler and kernel, we use a custom type of Peano naturals.
-Unifying two concrete Peano natural numbers $n$ and $m$ involves comparing $\min(n, m)$ constructors.
-We can therefore use moderately large numbers to simulate the bigger unification problems that Aesop encounters, for example, in Mathlib's category theory development.
 
@@
         xmin=1, xmax=16,
-        ymin=2^3, ymax=2^13,
-        xtick={1,2,4,8,16},
-        ytick={2^3,2^5,2^7,2^9,2^11,2^13},
+        ymin=4, ymax=2^15,
+        %xtick={1,2,4,8,16},
+        %ytick={2^3,2^5,2^7,2^9,2^11,2^13},
         xmode=log,
@@
           coordinates {
-            (1, 7.663472) (2, 9.354750) (4, 15.768194) (8, 104.921791) (16, 1882.683111)
+            (1, 7.663472) (2, 9.354750) (4, 15.768194) (8, 104.921791) (16, 1882.683111) % (32, 1175.888750) 
           };
@@
           coordinates {
-            (1, 10.308541) (2, 10.704638) (4, 16.274305) (8, 38.800833) (16, 179.697681)
+            (1, 10.308541) (2, 10.704638) (4, 16.274305) (8, 38.800833) (16, 179.697681) % (32, 45220.919667) 
           };
@@
           coordinates {
-            (1, 10.591138) (2, 23.881333) (4, 142.013389) (8, 1676.148375) (16, 24327.359417)
+            (1, 10.591138) (2, 23.881333) (4, 142.013389) (8, 1676.148375) (16, 24327.359417) %(32, 392049.849708) 
           };
@@
           coordinates {
-            (1, 13.974694) (2, 26.216833) (4, 68.953444) (8, 245.601028) (16, 1028.670680)
+            (1, 13.974694) (2, 26.216833) (4, 68.953444) (8, 245.601028) (16, 1028.670680) %(32, 4388.917500) 
           };
@@
         xmin=1, xmax=32,
-        ymin=2^3, ymax=1700,
+        ymin=4, ymax=2^15,
         ymode=log,xmode=log,
@@
   \end{subfigure}
-  \caption{The above results depict the behavior of both algorithms in a context where there is a lot of work that can be reused (Transitivity) and no work to be reused (Independence).
-The lines with square entry points represent problems where unification is easy.
-The lines with circle entry points are the same tests with hard unification problems.}%
+  \caption{The above results depict the behavior of both algorithms in a context where a lot of work can be reused (Transitivity) and in another where no work can be reused (Independence). 
+  In both benchmarks, results denoted with a square \protect\showpgfsquare\ come from a context where the argument of the predicates is small ($a = 0$), leading to easy unification problems; results denoted with a circle \protect\showpgfcircle\ come from a context where the unification problem is hard ($a = 100$).}%
   \label{fig:benchmark}
@@
 
-The results in Fig.~\ref{fig:trans} are unsurprising.
+We see in Fig.~\ref{fig:trans} that the incremental algorithm has a sizable advantage over the naive algorithm.
+An important detail is that the scales are logarithmic
+These results are unsurprising:
 The naive algorithm goes over the context, containing initially $n$ hypotheses, and compares all hypotheses pairwise.
@@
 
-Given a set of propositions $\mathcal{P} := \{P_{i,j} | 1 \leq i \leq n \text{ and } 1 \leq j \leq 6\}$ and $\mathcal{Q} :=\{Q_{i} | 1 \leq i \leq n \}$, in this test, we consider the $n$ following rules $r_i : P_{i,1}\to \dots P_{i,6} \to Q_i$.
+Given a set of predicates $\mathcal{P} := \{P_{i,j} (x) | 1 \leq i \leq n \text{ and } 1 \leq j \leq 6\}$ and $\mathcal{Q} :=\{Q_{i}(x) | 1 \leq i \leq n \}$, in this test, we consider the $n$ following rules $r_i : \forall x, P_{i,1}(x)\to \dots P_{i,6}(x) \to Q_i(x)$.
 The procedures are then run on a context containing exactly $\mathcal{P}$.
 In this case, we sketch the behaviour of the stateless algorithm:
-It pick a first hypothesis, $h_{i,j} : P_{i,j}$.
+It pick a first hypothesis, $h_{i,j} : P_{i,j}(a)$.
 If $j < 6$, the index returns nothing and it selects another hypothesis.
-For $h_{i,6} : P_{i,6}$, the rule $r_i$ is selected, and it then needs to try and fill its remaining five slot with the context's $6n$ hypotheses which it tries all.
+For $h_{i,6} : P_{i,6}(a)$, the rule $r_i$ is selected, and it then needs to try and fill its remaining five slot with the context's $6n$ hypotheses which it tries all.
 This gives a complexity of about $O(n^2)$.

### 2025-02-18 paper.tex
Further compress rule state figure

@@
       \begin{minipage}{0.5\textwidth}
-        \begin{tikzpicture}[outer sep=auto, level distance=12mm, scale=0.50]
+        \begin{tikzpicture}[outer sep=auto, level distance=4em, scale=0.50]
           \node at (-2,0) {\captiontext*{}\label{fig:example1:state1}};
-          \node[xshift = 0] {$H_{y}$}
-            child {node {2}
+          \node {$H_{y}\colon$};
+          \node at (1.5,0) {2}
             child {node {$301$}
-                child {node {$\{h₁\}$}}}
-            %  child {node {$300$}
-            %    child {node {$\{\}$}}}
-            };
-          \node[xshift=80] {$M_y$}
-            child {node {1}
+              child {node {$\{h₁\}$}}};
+          \node at (5, 0) {$M_y\colon$};
+          \node at (6.5, 0) {1}
             child {node {$302$}
-                child {node {$\{[h₁]\}$}}}
-            %  child {node {$301$}
-            %    child {node {$\{\}$}}}
-            };
-            %child {node[xshift=20] {2}
-            %  child {node {$301$}
-            %    child {node[yshift=-2em] {$\{\}$}}}};
+              child {node {$\{[h₁]\}$}}};
         \end{tikzpicture}
@@
       \begin{minipage}{0.5\textwidth}
-        \begin{tikzpicture}[outer sep=auto, level distance=12mm, scale=0.50]
+        \begin{tikzpicture}[outer sep=auto, level distance=4em, scale=0.50]
           \node at (-2,0) {\captiontext*{}\label{fig:example1:state2}};
-          \node[xshift = 0] {$H_y$}
-            child {node {2}
-              child {node {$301$}
+          \node {$H_y\colon$};
+          \node at (1.5, 0) {2}
+            child {node[xshift=-2] {$301$}
               child {node {$\{h₁\}$}}}
-              child {node {$300$}
-                child {node {$\{h₂\}$}}}
-            };
-          \node[xshift=80] {$M_y$}
-            child {node {1}
-              child[xshift=-20] {node {$302$}
+            child {node[xshift=2] {$300$}
+              child {node {$\{h₂\}$}}};
+          \node at (5, 0) {$M_y\colon$};
+          \node at (6.5, 0) {1}
+            child {node[xshift=-2] {$302$}
               child {node {$\{[h₁]\}$}}}
-              child[xshift=20] {node {$301$}
-                child {node {$\{[h₂]\}$}}}
-            };
+            child {node[xshift=2] {$301$}
+              child {node {$\{[h₂]\}$}}};
         \end{tikzpicture}
@@
     \begin{minipage}{\textwidth}
-      \begin{tikzpicture}[outer sep=auto, level distance=12mm, scale=0.50]
+      \begin{tikzpicture}[outer sep=auto, level distance=4em, scale=0.50]
         \node at (-2,0) {\captiontext*{}\label{fig:example2:state1}};
-        \node[xshift = 0] {$H_x$}
-            child {node {2}
+        \node {$H_x\colon$};
+        \node at (1.5, 0) {2}
           child {node {$a$}
-                    child {node {$\{h_1\}$}}}
-                %child {node {$b$}
-                %    child {node {$\{ \}$}}}
-                };
-        \node[xshift = 80] {$H_y$}
-            child {node {2}
+            child {node {$\{h_1\}$}}};
+        \node at (5, 0) {$H_y\colon$};
+        \node at (6.5, 0) {2}
           child {node {$a$}
-                    child {node {$\{h_1\}$}}}
-                %child {node {$b$}
-                %    child {node {$\{ \}$}}}
-                    };
-        \node[xshift = 160, yshift = 0] {$M_x$}
-            child {node {1}
+            child {node {$\{h_1\}$}}};
+        \node at (10, 0) {$M_x\colon$};
+        \node at (11.5, 0) {1}
           child {node {$a$}
-                    child {node {$\{[h_2]\}$}}}
-                %child {node {$b$}
-                %   child {node {$\{ \}$}}}
-                    };
-            %child[xshift = 20] {node {2}
-            %    child {node {$a$}
-            %        child {node {$\{\}$}}}
-            %    child {node {$b$}
-            %        child {node {$\{ \}$}}}
-            %};
-        \node[xshift = 240, yshift = 0] {$M_y$}
-            child {node {1}
-            %    child {node {$a$}
-            %        child {node {$\{ \}$}}}
+            child {node {$\{[h_2]\}$}}};
+        \node at (15, 0) {$M_y\colon$};
+        \node at (16.5, 0) {1}
           child {node {$b$}
-                    child {node {$\{[h_2]\}$}}}
-            };
-            %child[xshift = 20] {node {2}
-            %    child {node {$a$}
-            %        child {node {$\{\}$}}}
-            %    child {node {$b$}
-            %        child {node {$\{ \}$}}}
-            %};
+            child {node {$\{[h_2]\}$}}};
       \end{tikzpicture}
@@
         \node at (-2,0) {\captiontext*{}\label{fig:example2:state2}};
-        \node[xshift = 0] {$H_x$}
-          child {node {2}
+        \node {$H_{x}\colon$};
+        \node at (1.5, 0) {2}
           child {node {$a$}
-                  child {node {$\{h_1, h_3\}$}}}
-              %child {node {$b$}
-              %    child {node {$\{ \}$}}}
-              };
-        \node[xshift = 80] {$H_y$}
-          child {node {2}
+            child {node {$\{h_1, h_3\}$}}};
+        \node at (5, 0) {$H_y\colon$};
+        \node at (6.5, 0) {2}
           child {node {$a$}
@@
           child {node {$b$}
-                  child {node {$\{h_3\}$}}}};
-        \node[xshift = 160, yshift = 0] {$M_x$}
-          child {node {1}
+            child {node {$\{h_3\}$}}};
+        \node at (10, 0) {$M_x\colon$};
+        \node at (11.5, 0) {1}
           child {node {$a$}
-                  child {node {$\{[h_2]\}$}}}
-              %child {node {$b$}
-              %   child {node {$\{ \}$}}}
-                  };
-        \node[xshift = 240, yshift = 0] {$M_y$}
-          child {node {1}
-          %    child {node {$a$}
-          %        child {node {$\{ \}$}}}
+            child {node {$\{[h_2]\}$}}};
+        \node at (15, 0) {$M_y\colon$};
+        \node at (16.5, 0) {1}
           child {node {$b$}
-                  child {node {$\{[h_2]\}$}}}
-          };
+            child {node {$\{[h_2]\}$}}};
       \end{tikzpicture}

### 2025-02-18 paper.tex
More detailed description of rule state figure
Use a, b, c instead of 300, 301, 301 in Ex. 1; add non-breakable spaces

@@
 This is the premise's \emph{slot}, which we often identify with the premise itself.
-For now, we assume that the input premises are numbered consecutively, so in the above example slot 1 has type $x < y$ and slot 2 has type $y < z$.
+For now, we assume that the input premises are numbered consecutively, so in the above example slot~1 has type $x < y$ and slot~2 has type $y < z$.
 The \emph{shared variables of slot $i$}, $\vars(i)$ are those variables that appear in the type of $i$ and also in at least one type of a slot $j < i$.
@@
 We have so far assumed that for each slot $i > 1$, the set of shared variables $\vars(i)$ is nonempty, i.e.\ each slot shares at least one variable with at least one of the previous slots.
-This assumption is violated for rules such as $\All{x~y}{P~x → Q~y → R~x~y}$: slot 2 ($Q~y$) shares no variables with slot 1 ($P~x$).
+This assumption is violated for rules such as $\All{x~y}{P~x → Q~y → R~x~y}$: slot~2 ($Q~y$) shares no variables with slot~1 ($P~x$).
 As a result, our above algorithms fail.
@@
 The set of matches compatible with $h₂$ is supposedly $\bigcap_{x \in \vars(2)} M_{x}(i-1, σ(x)) = ∅$, but the match $[h₁]$ can be extended with $h₂$.
-In fact, any match at level 1 can be extended with any hypothesis matching slot 2 since the two slots are independent.
+In fact, any match at level~1 can be extended with any hypothesis matching slot~2 since the two slots are independent.
 
@@
 As a first example, consider a rule state for the transitivity rule $r₁ : \All{x~y~z}{x ≤ y → y ≤ z → x ≤ z}$ to which we add, in order, the hypotheses
-$h₁ : 301 ≤ 302$ and $h₂ : 300 ≤ 301$.
+$h₁ : b ≤ c$ and $h₂ : a ≤ b$.
 The rule has two slots of types $x ≤ y$ and $y ≤ z$ which share the variable $y$.
 
-The hypothesis $h₁ : 301 ≤ 302$ unifies with both slots, so we run Algorithm~\ref{alg:addhyp} twice.
-For slot 1, unification yields the substitution $\{x ↦ 301, y ↦ 302\}$.
-Since this is the first slot, we start a new partial match $[h₁]$, which Algorithm~\ref{alg:addmatch} inserts into the set $M_{y}(1, 302)$.
+The hypothesis $h₁ : b ≤ c$ unifies with both slots, so we run Algorithm~\ref{alg:addhyp} twice.
+For slot 1, unification yields the substitution $\{x ↦ b, y ↦ c\}$.
+Since this is the first slot, Algorithm~\ref{alg:addmatch} inserts the new partial match $[h₁]$ into the set $M_{y}(1, c)$.
 This match cannot be extended since the hypothesis index $H$ does not contain a hypothesis for slot 2.
 
-For slot 2, unification yields the substitution $σ ≔ \{y ↦ 301, z ↦ 302\}$, so we insert $h₁$ into $H_{y}(2, 301)$.
-No matches at level 1 can be extended by $h₁$ since
+For slot 2, unification yields the substitution $σ ≔ \{y ↦ b, z ↦ c\}$, so we insert $h₁$ into $H_{y}(2, b)$.
+No matches at level~1 can be extended by $h₁$ since
 \[
-  \bigcap_{x \in \vars(2)} M_{x}(1, σ(x)) = M_{y}(1, 301) = \emptyset
+  \bigcap_{x \in \vars(2)} M_{x}(1, σ(x)) = M_{y}(1, b) = \emptyset
 \]
@@
           \node at (1.5,0) {2}
-            child {node {$301$}
+            child {node {$b$}
               child {node {$\{h₁\}$}}};
@@
           \node at (6.5, 0) {1}
-            child {node {$302$}
+            child {node {$c$}
               child {node {$\{[h₁]\}$}}};
@@
           \node at (1.5, 0) {2}
-            child {node[xshift=-2] {$301$}
+            child {node[xshift=-2] {$b$}
               child {node {$\{h₁\}$}}}
-            child {node[xshift=2] {$300$}
+            child {node[xshift=2] {$a$}
               child {node {$\{h₂\}$}}};
@@
           \node at (6.5, 0) {1}
-            child {node[xshift=-2] {$302$}
+            child {node[xshift=-2] {$c$}
               child {node {$\{[h₁]\}$}}}
-            child {node[xshift=2] {$301$}
+            child {node[xshift=2] {$b$}
               child {node {$\{[h₂]\}$}}};
@@
     \begin{minipage}{\textwidth}
-      \begin{tikzpicture}[outer sep=auto, level distance=12mm, scale=0.50]
+      \begin{tikzpicture}[outer sep=auto, level distance=4em, scale=0.50]
         \node at (-2,0) {\captiontext*{}\label{fig:example2:state2}};
@@
   \end{subcaptiongroup}
-  \caption{Rule states encountered during Examples~1 and~2}
+  \caption{%
+    Rule states encountered during Examples~1 and~2.
+    The variable indices are represented as trees, with nodes in the first row denoting slots and nodes in the second row denoting variable instantiations.%
+  }
 \end{figure}
 
-Now we add $h₂ : 300 ≤ 301$, which again unifies with both slots.
-For slot 1, $\sub₁(h₂)(y) = 301$, so we add the match $[h₂]$ to $M_{y}(1, 301)$.
-We then check for hypotheses in slot 2 that also instantiate $y$ with $301$, and indeed we find $M_{y}(2, 301) = \{h₁\}$.
-Hence we return the complete match $[h₂, h₁]$, which induces a new hypothesis $h₃ : 300 ≤ 302$ to be added later.
+Now we add $h₂ : a ≤ b$, which again unifies with both slots.
+For slot~1, $\sub₁(h₂)(y) = b$, so we add the match $[h₂]$ to $M_{y}(1, b)$.
+We then check for hypotheses in slot~2 that also instantiate $y$ with $b$, and indeed we find $M_{y}(2, b) = \{h₁\}$.
+Hence we return the complete match $[h₂, h₁]$, which induces a new hypothesis $h₃ : a ≤ c$ to be added later.
 
-For slot 2, we add $h₂$ to $H_{y}(2, 300)$.
-There are no compatible matches at level 1, so we are done with $h₂$, yielding the final rule state in Fig.~\ref{fig:example1:state2}.
+For slot~2, we add $h₂$ to $H_{y}(2, a)$.
+There are no compatible matches at level~1 so we are done with $h₂$, yielding the final rule state in Fig.~\ref{fig:example1:state2}.
 
@@
 We now consider a rule $r₂ : \All{x~y}{A~x~y → B~x~y → C~x~y}$ with multiple shared variables.
-We add hypotheses $h₁ : B~a~a$, $h₂ : A~a~b$ and $h₃ : B~a~b$ to the rule state of $r₂$, starting with $h₁$, which is added to $H$ with slot 2.
+We add hypotheses $h₁ : B~a~a$, $h₂ : A~a~b$ and $h₃ : B~a~b$ to the rule state of $r₂$, starting with $h₁$, which is added to $H$ with slot~2.
 There are no matches yet, so there is nothing to extend.
@@
 To find hypotheses compatible with $m$, we consider the set of variables $\vars(\lvl(m) + 1)$.
-These are the variables shared between slot $\lvl(m) + 1 = 2$ and all previous slots, i.e.\ $x$ and $y$.
+These are the variables shared between slot~$\lvl(m) + 1 = 2$ and all previous slots, i.e.\ $x$ and $y$.
 Hence the compatible hypotheses are
@@
 Additionally, forward rules are currently little used by Aesop's main client, the Mathlib mathematics library (TODO ref).
-We therefore only evaluate our technique against the previous, naive Aesop implementation on several synthetic benchmarks that showcase the performance characteristics of incremental forward reasoning.
+We therefore only evaluate our technique against the previous, naive Aesop implementation on several synthetic benchmarks that demonstrate the performance characteristics of incremental forward reasoning.
 

### 2025-02-18 paper.tex
Minor edit in related work

@@
 The Prolog-based first-order theorem prover SATCHMO (TODO ref) and its many descendants also use mixed-mode reasoning.
-SATCHMO's search procedure can be viewed as a special case of our technique: a given set of implications are used as forward rules and a single case-splitting rule splits any goal with hypothesis $h : B₁ ∨ \dots ∨ Bₘ$ into $m$ goals with hypotheses $h₁ : B₁, \dots, hₘ : Bₘ$.
+SATCHMO's search procedure can be viewed as a special case of our technique: a given set of implications are used as forward rules and a single non-forward rule performs case analysis on disjunctive hypotheses.
 A SATCHMO-style prover may benefit from our incremental approach if the set of implications is large, though in a setting where unification is cheap it is not clear whether avoiding repeated unifications outweighs the cost of maintaining the forward state.

### 2025-02-18 paper.tex
Split SATCHMO and KBS related work; rewrite SATCHMO

@@
 
-\paragraph{Mixed-Mode Reasoning.}
-Some knowledge-based reasoning systems, e.g.\ Algernon (TODO ref) and EYE (TODO ref), support combined forward and backward (\enquote*{mixed-mode}) reasoning.
-However, to our knowledge, rules in these systems cannot change the assumptions available to forward rules, so they do not require incrementality.
+\paragraph{SATCHMO.}
+The SATCHMO first-order theorem prover (TODO ref) uses a search procedure that treats a given set of implications as forward rules and uses a single non-forward rule to perform case analysis on disjunctive hypotheses.
+SATCHMO-style provers may benefit from our incremental approach if the set of implications is large.
+However, since unification is comparatively cheap in first-order logic, it is not clear whether avoiding repeated unifications outweighs the cost of maintaining the forward state.
+SATCHMO also employs an optimisation that \enquote*{focuses} forward reasoning on the hypothesis established by the last case split, providing a different form of incrementality.
+
+\paragraph{Knowledge-Based Systems.}
+Some knowledge-based reasoning systems, e.g.\ Algernon (TODO ref) and EYE (TODO ref), support combined forward and backward reasoning.
+However, to our knowledge, no such system allows non-forward rules to change the assumptions available to forward rules, so they do not require incrementality.
 Nevertheless, our technique bears some resemblance to the RETE algorithm (TODO ref) for forward reasoning in large knowledge bases.
-
-The Prolog-based first-order theorem prover SATCHMO (TODO ref) and its many descendants also use mixed-mode reasoning.
-SATCHMO's search procedure can be viewed as a special case of our technique: a given set of implications are used as forward rules and a single non-forward rule performs case analysis on disjunctive hypotheses.
-A SATCHMO-style prover may benefit from our incremental approach if the set of implications is large, though in a setting where unification is cheap it is not clear whether avoiding repeated unifications outweighs the cost of maintaining the forward state.
-SATCHMO also employs an optimisation that \enquote*{focuses} forward reasoning on the hypothesis established by the last case split, which provides a form of incrementality.
 

### 2025-02-18 paper.tex
Minor
long version eval

@@
 \section{Naive Forward Reasoning}
+\label{sec:naive}
 
@@
 Given some relation $≺$ on natural numbers, we register the lemma $\All{x~y}{x ≺ y → y ≺ z → x ≺ z}$ as the sole forward rule and set up a goal with hypotheses $h₁ : a ≺ a + 1, \dots, hₙ : a + n - 1 ≺ a + n$.
-The test is run for $a = 0$, represented with a square \protect\showpgfsquare\, and $a = 100$, represented with a circle \protect\showpgfcircle\.
+The test is run for $a = 0$, represented with a square \protect\showpgfsquare\, and $a = 100$, represented with a circle \protect\showpgfcircle\ .
 Saturating this goal adds $n(n-1)/2$ hypotheses to the context.
@@
         ylabel={Time in ms},
-        xmin=1, xmax=16,
-        ymin=4, ymax=2^15,
+        xmin=1, xmax=32,
+        ymin=4, ymax=2^16,
         %xtick={1,2,4,8,16},
@@
           coordinates {
-            (1, 7.663472) (2, 9.354750) (4, 15.768194) (8, 104.921791) (16, 1882.683111) % (32, 1175.888750) 
+            (1, 7.663472) (2, 9.354750) (4, 15.768194) (8, 104.921791) (16, 1882.683111) (32, 45220.919667)  
           };
@@
           coordinates {
-            (1, 10.308541) (2, 10.704638) (4, 16.274305) (8, 38.800833) (16, 179.697681) % (32, 45220.919667) 
+            (1, 10.308541) (2, 10.704638) (4, 16.274305) (8, 38.800833) (16, 179.697681) (32, 1175.888750)
           };
@@
           coordinates {
-            (1, 10.591138) (2, 23.881333) (4, 142.013389) (8, 1676.148375) (16, 24327.359417) %(32, 392049.849708) 
+            (1, 10.591138) (2, 23.881333) (4, 142.013389) (8, 1676.148375) (16, 24327.359417) (32, 392049.849708) 
           };
@@
           coordinates {
-            (1, 13.974694) (2, 26.216833) (4, 68.953444) (8, 245.601028) (16, 1028.670680) %(32, 4388.917500) 
+            (1, 13.974694) (2, 26.216833) (4, 68.953444) (8, 245.601028) (16, 1028.670680) (32, 4388.917500) 
           };
@@
         xmin=1, xmax=32,
-        ymin=4, ymax=2^15,
+        ymin=4, ymax=2^16,
         ymode=log,xmode=log,
@@
 We see in Fig.~\ref{fig:trans} that the incremental algorithm has a sizable advantage over the naive algorithm.
-An important detail is that the scales are logarithmic
+An important detail is that the scales are logarithmic; the dotted horizontal lines represent a relative $2^3 = 8$ fold increase.
+To highlight this, we consider the time taken for $2^5 = 32$ hypotheses and $a = 0$, these are the square marks on the rightmost entries: The incremental algorithm dominates the naive algorithm by a relative factor of more than 38.
+
 These results are unsurprising:
@@
 The incremental algorithm considers each hypothesis only once.
-As the hypotheses get matched to both slots, it then compares each hypothesis with at most $n$ other hypotheses.
-This yields a complexity of $O(n^2)$.
+The hypotheses get matched to both slots and are then compared with at most $n$ other hypotheses.
+This yields an expected complexity of $O(n^2)$.
 
@@
 In the last example, part of the stateful algorithm's success stems from the fact that it could reuse a lot of its previous work.
-An obvious question is then how well does the stateful algorithm behave when there is no work to be reused.
+An obvious question is then how well does the stateful algorithm behave when there is les work to be reused.
 
@@
 The procedures are then run on a context containing exactly $\mathcal{P}$.
-In this case, we sketch the behaviour of the stateless algorithm:
-It pick a first hypothesis, $h_{i,j} : P_{i,j}(a)$.
-If $j < 6$, the index returns nothing and it selects another hypothesis.
-For $h_{i,6} : P_{i,6}(a)$, the rule $r_i$ is selected, and it then needs to try and fill its remaining five slot with the context's $6n$ hypotheses which it tries all.
-This gives a complexity of about $O(n^2)$.
+Here, as the rules are independent, the interference between them is minimal and there is less work to be reused.
 
-For the stateful algorithm, the hypotheses are considered one by one and match directly to the rule and the slot.
-The expected complexity is then linear in the number of rules.
+This doesn't mean, however, that no work is reused: A successful rule application creates a subgoal with the newly added fact.
+This means that the naive algorithm treats this subgoal as a completely new goal whereas the incremental algorithm can reuse all the work already done on the other hypotheses.
+A similar analysis to the one done in the first benchmark suggests a cubic complexity in the number of rules for the naive algorithm against a linear one for the incremental algorithm.
+The advantage is clearly seen in Fig.~\ref{fig:indep}.
+%In this case, we sketch the behaviour of the stateless algorithm:
+%It pick a first hypothesis, $h_{i,j} : P_{i,j}(a)$.
+%If $j < 6$, the index returns nothing and it selects another hypothesis.
+%For $h_{i,6} : P_{i,6}(a)$, the rule $r_i$ is selected, and it then needs to try and fill its remaining five slot with the context's $6n$ hypotheses which it tries all.
+%This gives a complexity of about $O(n^2)$.
+
+%For the stateful algorithm, the hypotheses are considered one by one and match directly to the rule and the slot.
+%The expected complexity is then linear in the number of rules.
 
 \paragraph{Rules of Varying Depth.}
-In Fig.~\ref{fig:erase}, the step is set up as follows: given some rule $r$ with $6$ \xcom{motivate 6} premises, we run the procedure in a context with 5 of the 6 premises present.
-The numbers on the $x$ axis represent which hypothesis is missing in the order of consideration by the algorithm.
-In a sense, this is how \textit{deep} the algorithm investigate before it stops.
-At $0$, the rule are not selected by the index so the procedure stops.
-The stateless algorithm gets progressively slower the further the removed premises is located.
+As a last test case, we consider a scenario for which the naive algorithm should have a clear advantage over the incremental algorithm.
+The main idea is the following: The naive algorithm sometimes does a better job at recognising that a rule is unprovable.
+For example, when the context contains all but the last premise of a rule; the naive algorithm's index doesn't even select the rule.
+(See Sec.
+\ref{sec:naive} for the complete description of the naive algorithm.)
+This leads to a much faster treatment of this rule than the incremental algorithm that will still process the hypotheses in the context that match the other premises.
+
+Inspired by this, we find it interesting to consider a closely related test: A context with all the necessary hypotheses matching the premises however with one of them being incompatible with the others.
+The reason why we introduce an incomptatible hypothesis, instead of having one absent hypothesis in the context, is because even thought the naive algorithm is sure to be fast, the incremental algorithm also handles this specific situations well.
+See \textbf{Lazy Insertion} in Sec. \ref{sec:implementation}.
+
+With all of this in mind, in Fig.~\ref{fig:erase}, the test is set up as follows: given some rule $r$ with $n$ premises $P_i(x)$, $1 \leq i \leq n$, we run the procedure in a context where $n-1$ premises are instanciated with $a$ and one of them with $a + 1$.
+The numbers on the $x$ axis represent which hypothesis has the incompatible instantiation in the order of consideration by the algorithm.
+In a sense, this is how \textit{deep} the naive algorithm investigate before it stops.
+The reason we see similar times for the naive algorithm at $1$ and $2$ is because for $1$, it tries to build the theorem using $P_n(a + 1)$ and immediately fails because $P_{n - 1}(a)$ is incompatible.
+For $2$, the same thing happens but the instantiations are swapped.
+For the next depths, the stateless algorithm gets progressively slower the further the incompatible hypothesis is located.
+The position of the incompatible hypothesis has little effect on the stateful algorithm as the index selects the rule in all case.
+The work the stateful algorithm has to do is thus close to constant in this scenario.
+The Figure \ref{fig:erase-noprecomp} showcases that the naive algorithm has a sizable advantage in this test, especially when unification is difficult.
+
+\paragraph{Precompilation.}
+
+Some of the modules implemented in Lean can be precompiled.
+This makes a lot of sense for our approach since it is inevitably more complex that the naive approach.
+It thus benefits from getting precompiled.
+This translate into even more substantial gains in the first two tests and in the third test, the incremental algorithm again takes the advantage over the naive algorithm. See \ref{fig:erase-precomp}.
+%Precompilation is by default turned off for Mathlib4 as it amplifies its already long compilation time.
+%Users working on projects that uses a lot of forward reasoning can however chose to pay this compilation time price once and then benefit from fast forward reasoning.
 
@@
 
-The position of the missing premise has little effect on the stateful algorithm as the index selects the rule in all case; it only needs one of the premise to be present.
-The work the stateful algorithm has to do is thus close to constant in this scenario.
-
 \section{Related Work}
@@
 The technique is independent of the search strategy and logic used.
-Our evaluation shows a substantial speedup on some synthetic benchmarks, relative to Aesop's previous implementation.
+Our evaluation shows substantial speedups on some synthetic benchmarks, relative to Aesop's previous implementation.
 

### 2025-02-18 paper.tex
add comment

@@
 In a sense, this is how \textit{deep} the naive algorithm investigate before it stops.
+\xcom{I would remove either 1 or 2 from the graph (they have the same depth and then remove this explanation. I can then either modify the test accordingly or just move the explanation to the test.)}
 The reason we see similar times for the naive algorithm at $1$ and $2$ is because for $1$, it tries to build the theorem using $P_n(a + 1)$ and immediately fails because $P_{n - 1}(a)$ is incompatible.

### 2025-02-18 paper.tex
fix depth

--- 2025-02-18 paper.tex
@@
 In a sense, this is how \textit{deep} the naive algorithm investigate before it stops.
-The reason we see similar times for the naive algorithm at $1$ and $2$ is because for $1$, it tries to build the theorem using $P_n(a + 1)$ and immediately fails because $P_{n - 1}(a)$ is incompatible.
-For $2$, the same thing happens but the instantiations are swapped.
-For the next depths, the stateless algorithm gets progressively slower the further the incompatible hypothesis is located.
 The position of the incompatible hypothesis has little effect on the stateful algorithm as the index selects the rule in all case.
@@
         ylabel={Time in ms},
-        xmin=1, xmax=6,
-        ymin=0, ymax=2^10,
-        xtick={1,2,3,4,5,6},
-        ytick={4,16,64,256,1024},
+        xmin=1, xmax=5,
+        ymin=8, ymax=2^10,
+        xtick={1,2,3,4,5},
         ymode=log,
@@
           coordinates {
-            (1, 32.580083) (2, 33.212805) (3, 40.239833) (4, 47.670277) (5, 54.647458) (6, 61.044958) 
+            (1, 35.466292) (2, 40.918638) (3, 46.936125) (4, 54.933124) (5, 61.269861)
           };
@@
           coordinates {
-            (1, 127.925541) (2, 140.497597) (3, 123.719264) (4, 130.819361) (5, 121.047667) (6, 123.789500)
+            (1, 135.967736) (2, 127.324444) (3, 136.891250) (4, 118.720444) (5, 116.293222) 
           };
@@
           coordinates {
-            (1, 108.120444) (2, 107.352694) (3, 116.182833) (4, 125.995639) (5, 137.815527) (6, 142.310597) 
+            (1, 107.978069) (2, 115.900527) (3, 122.454958) (4, 137.732764) (5, 144.037681) 
           };
@@
           coordinates {
-            (1, 604.595791) (2, 624.541194) (3, 608.671111) (4, 614.285014) (5, 588.636819) (6, 595.276806) 
+            (1, 619.932611) (2, 601.456805) (3, 600.926110) (4, 589.144389) (5, 595.091180) 
           };
@@
         %ylabel={Time in ms},
-        xmin=1, xmax=6,
-        ymin=0, ymax=170,
-        xtick={1,2,3,4,5,6},
-        ytick={2^1,2^3,2^5,2^7},
+        xmin=1, xmax=5,
+        ymin=8, ymax=1024,
+        xtick={1,2,3,4,5},
         ymode=log,
@@
           coordinates {
-            (1, 12.820514) (2, 12.645666) (3, 17.582138) (4, 22.730528) (5, 27.280097) (6, 32.578722) 
+            (1, 13.728805) (2, 17.909194) (3, 23.324666) (4, 28.030819) (5, 33.279611) 
           };
@@
           coordinates {
-            (1, 13.406416) (2, 13.747139) (3, 12.503764) (4, 12.950486) (5, 12.282083) (6, 12.151930) 
+            (1, 13.841305) (2, 12.801791) (3, 13.358389) (4, 13.200611) (5, 13.251902)
           };
@@
           coordinates {
-            (1, 83.032819) (2, 79.396889) (3, 85.939180) (4, 92.495542) (5, 96.461291) (6, 105.517458)
+            (1, 81.546486) (2, 85.345444) (3, 90.532347) (4, 98.247764) (5, 105.423875) 
           };
@@
           coordinates {
-            (1, 25.681583) (2, 25.322444) (3, 27.672833) (4, 26.999055) (5, 25.690319) (6, 25.257291)
+            (1, 25.539097) (2, 24.432485) (3, 25.629666) (4, 23.935250) (5, 23.911139)
           };

### 2025-02-18 paper.tex
Add citations

@@
 \usepackage{subcaption}
+
+\bibliographystyle{splncs04}
 
@@
 
-In many proof assistants, simple \enquote*{white-box} automation tactics based on tableau-style tree search play a substantial role in practice: Isabelle's (TODO ref) and Rocq's (TODO ref) auto, ACL2's (TODO ref) waterfall, various PVS (TODO ref) tactics, Lean's (TODO ref) Aesop (TODO ref), etc.
+In many proof assistants, simple \enquote*{white-box} automation tactics based on tableau-style tree search play a substantial role in practice: Isabelle's~\cite{Isabelle} and Rocq's~\cite{Coq} auto~\cite{IsabelleAuto}, ACL2's \enquote{waterfall}~\cite{ACL2}, various PVS~\cite{PVS,PVS-tutorial} tactics, Lean's~\cite{Lean4} Aesop~\cite{Aesop}, etc.
 These tactics operate on \emph{goals} $Γ ⊢ A$, where the \emph{context} $Γ$ is a list of assumptions and $A$ is the proposition to be proved.
@@
 Hence forward rules can be used to establish additional facts that may be relevant for other rules.
-For example, the Aeneas project (TODO ref) uses Aesop's forward rules \jcom{ensure that this is accurate} in its proof automation to establish facts that are later used by an arithmetic decision procedure.
+For example, the Aeneas project~\cite{Aeneas} uses Aesop's forward rules \jcom{ensure that this is accurate} in its proof automation to establish facts that are later used by an arithmetic decision procedure.
 
 However, Aesop's support for forward rules was, until recently, quite basic.
-In particular, each forward rule was tried on each goal separately (subject to pre-filtering by a discrimination tree (TODO ref) index).
+In particular, each forward rule was tried on each goal separately (subject to pre-filtering by a discrimination tree~\cite{DiscriminationTrees} index).
 This is wasteful since in practice, many rules produce subgoals with a similar---or even the same---context as the initial goal.
@@
 Our evaluation (Sec.~\ref{sec:evaluation}) shows that this new implementation of forward reasoning in Aesop substantially outperforms the previous, naive implementation on some (though not all) synthetic benchmarks.
-The implementation, benchmarks and evaluation data are available in a supplement to this paper (TODO ref).
+The implementation, benchmarks and evaluation data are available in a supplement to this paper~\cite{supplement}.
 Mainline Aesop already uses the new implementation by default (TODO true?).
@@
 \jcom{Formulate this in terms of matches.}
-Prior to this work, Aesop used what we call naive forward reasoning (TODO ref).
+Prior to this work, Aesop used what we call naive forward reasoning~\cite{Aesop}.
 This approach treats forward rules essentially like any other Aesop rule.
-When Aesop selects a forward rule $r : \All{(x₁ : A₁) \dots (xₘ : Aₘ)}{B}$ as the highest-priority rule currently applicable to a goal $Γ ⊢ T$, we first unify the last premise $Aₘ$ with each hypothesis $h : C$ such that $C$ and $Aₘ$ likely unify (as determined by a discrimination tree (TODO ref) index).
+When Aesop selects a forward rule $r : \All{(x₁ : A₁) \dots (xₘ : Aₘ)}{B}$ as the highest-priority rule currently applicable to a goal $Γ ⊢ T$, we first unify the last premise $Aₘ$ with each hypothesis $h : C$ such that $C$ and $Aₘ$ likely unify (as determined by a discrimination tree index).
 If successful, this results in a substitution $σ₁$ containing $xₘ$ and any variables occurring in $Aₘ$.
@@
 This approach is efficient when the last premise of a rule is not present in the context since the rule is then excluded by indexing and never run.
-If the last premise is present, however, we perform a large number of unifications between premises and hypotheses (though some of these could perhaps be avoided heuristically (TODO ref fast subsumption)).
+If the last premise is present, however, we perform a large number of unifications between premises and hypotheses (though some of these could perhaps be avoided heuristically~\cite{TammetSubsumption}).
 Furthermore, the whole process is repeated when we proceed to another goal with a similar context.
@@
 The rule state therefore uses two \emph{variable indices} $H$ and $M$ to index hypotheses and matches by their instantiations for shared variables, allowing us to quickly determine compatible pairs of matches and hypotheses.
-We could use a substitution tree (TODO ref, confirm) for this, but our custom data structure is simpler and perhaps more efficient.
+We could use a substitution tree~\cite{SubstitutionTrees}, but our custom data structure is simpler and perhaps more efficient.
 The \emph{hypothesis index} $H$ maps each triple $(x, i, t)$ (where $i > 1$ is a slot, $x ∈ \vars(i)$ a variable and $t$ a term) to the set of hypotheses $h$ added to the rule state such that $h$ matches $i$ and $\sub_{i}(h)(x) = t$.
@@
 
-To account for this fact, we partition the set of premises of each rule into maximal subsets of premises connected by shared variables, using a construction from (TODO ref Aesop Sec. 4.2).
-These subsets, called \emph{variable clusters}, share no variables with each other, so each can be treated as a separate rule with its own rule state.
+To account for this fact, we partition the set of premises of each rule into maximal subsets of premises connected by shared variables, called \emph{variable clusters}, using the same construction as Aesop's \emph{metavariable clusters}~\cite{Aesop}.
+The variable clusters share no variables with each other, so each can be treated as a separate rule with its own rule state.
 Finding a complete match for a rule amounts to finding one complete match for each of its clusters.
@@
 We also need to account for higher-order unification problems: $Q~\mvar{f}$, where $Q$ is a predicate over functions and $\mvar{f}$ is a function variable, should unify with $Q~(\Lam{x}{x})$.
-Since higher-order unification is undecidable and does not admit most general unifiers (TODO ref), we instead use a unification algorithm provided by Lean that is complete for first-order unification problems and heuristically solves some higher-order problems as well.
+Since higher-order unification is undecidable~\cite{ThirdOrderUnificationUndecidableLucchesi,ThirdOrderUnificationUndecidableHuet,SecondOrderUnificationUndecidable}, we instead use a unification algorithm provided by Lean that is complete for first-order unification problems and heuristically solves some higher-order problems as well.
 This makes our technique incomplete in the sense that some theorem instantiations will be missed, but since the same unification algorithm is used throughout Lean, users are well acquainted with this form of incompleteness.
@@
 We believe that this scheme is more efficient than alternative schemes relying on repeated unification.
-For example, we could use Lean's implementation of imperfect discrimination trees (TODO ref), which implement a map from expressions to arbitrary data where lookup respects definitional equality.
+For example, we could use Lean's implementation of imperfect discrimination trees, which implement a map from expressions to arbitrary data where lookup respects definitional equality.
 However, since these discrimination trees are imperfect, they may return spurious lookup results, so whether a lookup result is valid must be verified by unifying the query expression with the expression that was originally stored in the discrimination tree.
@@
 
-There are unfortunately no standard benchmarks for Aesop-like tactics, and since Aesop is not expected to solve most goals without an extensive, manually curated set of rules, benchmarks for push-button theorem provers such as TPTP (TODO ref) are not very meaningful.
-Additionally, forward rules are currently little used by Aesop's main client, the Mathlib mathematics library (TODO ref).
+There are unfortunately no standard benchmarks for Aesop-like tactics, and since Aesop is not expected to solve most goals without an extensive, manually curated set of rules, benchmarks for push-button theorem provers such as TPTP~\cite{TPTP} are not very meaningful.
+Additionally, forward rules are currently little used by Aesop's main client, the Mathlib mathematics library~\cite{Mathlib}.
 We therefore only evaluate our technique against the previous, naive Aesop implementation on several synthetic benchmarks that demonstrate the performance characteristics of incremental forward reasoning.
@@
 
-Benchmark code and data are available in our supplement (TODO ref).
+Benchmark code and data are available in our supplement~\cite{supplement}.
 The results shown below were obtained on a MacBook Pro with an Apple~M2 Pro processor and \SI{32}{GB} of RAM.
@@
 \paragraph{ACL2 and PVS.}
-Among white-box proof automation tactics, to our knowledge only ACL2's \enquote{waterfall} (TODO ref) and PVS's various proof search strategies (TODO ref) have special support for forward reasoning.
+Among white-box proof automation tactics, to our knowledge only ACL2's \enquote{waterfall}~\cite{ACL2} and PVS's various proof search strategies~\cite{ACL2,PVS} have special support for forward reasoning.
 Both use essentially Aesop's naive algorithm, i.e.\ the context of each goal is saturated independently without reusing partial results from other goals.
@@
 \paragraph{SATCHMO.}
-The SATCHMO first-order theorem prover (TODO ref) uses a search procedure that treats a given set of implications as forward rules and uses a single non-forward rule to perform case analysis on disjunctive hypotheses.
+The SATCHMO first-order theorem prover~\cite{SATCHMO} uses a search procedure that treats a given set of implications as forward rules and uses a single non-forward rule to perform case analysis on disjunctive hypotheses.
 SATCHMO-style provers may benefit from our incremental approach if the set of implications is large.
@@
 \paragraph{Knowledge-Based Systems.}
-Some knowledge-based reasoning systems, e.g.\ Algernon (TODO ref) and EYE (TODO ref), support combined forward and backward reasoning.
+Some knowledge-based reasoning systems, e.g.\ Algernon~\cite{Algernon} and EYE~\cite{EYE}, support combined forward and backward reasoning.
 However, to our knowledge, no such system allows non-forward rules to change the assumptions available to forward rules, so they do not require incrementality.
-Nevertheless, our technique bears some resemblance to the RETE algorithm (TODO ref) for forward reasoning in large knowledge bases.
+Nevertheless, our technique bears some resemblance to the RETE algorithm~\cite{RETE} for forward reasoning in large knowledge bases.
 
@@
 
+\bibliography{lit}
+
 \end{document}

### 2025-02-18 paper.tex
No dot in keywords
fix axis label

@@
 
-  \keywords{Forward Reasoning \and Forward Chaining \and Tactics \and Interactive Theorem Proving.}
+  \keywords{Forward Reasoning \and Forward Chaining \and Tactics \and Interactive Theorem Proving}
 \end{abstract}
@@
 \begin{figure}
+  Depth Test
   \centering
@@
       \begin{axis}[
-        title={Depth Test},
+        %title={Depth Test},
         xlabel={Depth},
@@
         log basis y=2,
+        yticklabels={},
         legend pos=north west,

### 2025-02-18 paper.tex
Edit evaluation

@@
 \usepackage{pgfplots}
-\usepackage{hyperref}
+\usepackage[colorlinks]{hyperref}
 \usepackage{csquotes}
@@
 \def\orcid#1{{\href{http://orcid.org/#1}{\protect\raisebox{-1.25pt}{\protect\includegraphics{orcid.pdf}}}}}
+
+\pgfplotsset{compat=1.18, width=8cm}
 \def\showpgfcircle{\tikz[baseline=-0.7ex]\node[mark size=0.7ex]
@@
 {\pgfuseplotmark{square}};}
+
 \SetKw{KwIn}{in}
@@
 
-\section{Naive Forward Reasoning}
+\section{Naive Forward Reasoning}%
 \label{sec:naive}
@@
 
-For slot 2, unification yields the substitution $σ ≔ \{y ↦ b, z ↦ c\}$, so we insert $h₁$ into $H_{y}(2, b)$.
+For slot~2, unification yields the substitution $σ ≔ \{y ↦ b, z ↦ c\}$, so we insert $h₁$ into $H_{y}(2, b)$.
 No matches at level~1 can be extended by $h₁$ since
@@
 Hence we return the complete match $[h₂, h₁]$, which induces a new hypothesis $h₃ : a ≤ c$ to be added later.
-
 For slot~2, we add $h₂$ to $H_{y}(2, a)$.
@@
 
-In what follows, instead of the natural number type of Lean's standard library, which has special support in the compiler and kernel, we use a custom type of Peano naturals.
+Instead of the natural number type of Lean's standard library, which has special support in the compiler and kernel, we use a custom type of Peano naturals for the benchmarks.
 Unifying two concrete Peano natural numbers $n$ and $m$ involves comparing $\min(n, m)$ constructors.
@@
 
-\paragraph{Transitivity.}
+\paragraph{Transitivity Benchmark.}
 Our first benchmark is a scaled-up variant of the transitivity example from Sec.~\ref{sec:incremental}.
-Given some relation $≺$ on natural numbers, we register the lemma $\All{x~y}{x ≺ y → y ≺ z → x ≺ z}$ as the sole forward rule and set up a goal with hypotheses $h₁ : a ≺ a + 1, \dots, hₙ : a + n - 1 ≺ a + n$.
-The test is run for $a = 0$, represented with a square \protect\showpgfsquare\, and $a = 100$, represented with a circle \protect\showpgfcircle\ .
+Given some relation $≺$ on natural numbers, we register the lemma $\All{x~y}{x ≺ y → y ≺ z → x ≺ z}$ as the sole forward rule and set up a goal with hypotheses $h₁ : a ≺ a + 1, \dots, hₙ : a + n - 1 ≺ a + n$, where either $a = 0$ (yielding easy unification problems) or $a = 100$.
 Saturating this goal adds $n(n-1)/2$ hypotheses to the context.
@@
   \centering
-  \begin{subfigure}{.5\textwidth}
+  \begin{subfigure}[b]{.5\textwidth}
     \centering
@@
         grid style=dashed,
-        legend image post style={mark=},
+        legend image post style={mark=}
         ]
@@
           mark=square,
+          style=densely dashed,
+          mark options={style={solid}}
           ]
@@
           mark=o,
+          style=densely dashed,
+          mark options={style={solid}}
           ]
@@
   \end{subfigure}%
-  \begin{subfigure}{.5\textwidth}
+  \begin{subfigure}[b]{.5\textwidth}
     \centering
@@
       \begin{axis}[
-        xlabel={Number of Rules},
+        xlabel={Number of rules},
         %ylabel={Time in ms},
@@
           };
-        \addlegendentry{naive}
         \addplot[
@@
           };
-        \addlegendentry{incremental}
         \addplot[
@@
     \end{tikzpicture}
-    % graph \includegraphics[width=.4\linewidth]{image1}
     \caption{Independence benchmark}%
@@
   \end{subfigure}
-  \caption{The above results depict the behavior of both algorithms in a context where a lot of work can be reused (Transitivity) and in another where no work can be reused (Independence). 
-  In both benchmarks, results denoted with a square \protect\showpgfsquare\ come from a context where the argument of the predicates is small ($a = 0$), leading to easy unification problems; results denoted with a circle \protect\showpgfcircle\ come from a context where the unification problem is hard ($a = 100$).}%
+  \caption{Results of the transitivity (\subref{fig:trans}) and independence (\subref{fig:indep}) benchmarks for $a = 0$ (\protect\showpgfsquare) and $a = 100$ (\protect\showpgfcircle).}%
   \label{fig:benchmark}
@@
 
-We see in Fig.~\ref{fig:trans} that the incremental algorithm has a sizable advantage over the naive algorithm.
-An important detail is that the scales are logarithmic; the dotted horizontal lines represent a relative $2^3 = 8$ fold increase.
-To highlight this, we consider the time taken for $2^5 = 32$ hypotheses and $a = 0$, these are the square marks on the rightmost entries: The incremental algorithm dominates the naive algorithm by a relative factor of more than 38.
+Fig.~\ref{fig:trans} shows that the incremental algorithm has a sizeable advantage over the naive algorithm.
+The scales are logarithmic, with each dotted horizontal line representing an 8-fold increase.
+Hence, for $2^{5} = 32$ hypotheses and $a = 0$, incremental forward reasoning is over 38 times faster than naive forward reasoning.
+The difference for $a = 100$ is even larger.
 
-These results are unsurprising:
+These results are unsurprising, given the complexity of both algorithms on this benchmark.
 The naive algorithm goes over the context, containing initially $n$ hypotheses, and compares all hypotheses pairwise.
 This is repeated approximately for each of the $n(n-1)/2$ added hypotheses, so the overall complexity is $O(n^4)$.
-The incremental algorithm considers each hypothesis only once.
-The hypotheses get matched to both slots and are then compared with at most $n$ other hypotheses.
-This yields an expected complexity of $O(n^2)$.
+By contrast, the incremental algorithm considers each hypothesis only once, matching it against both slots and then comparing it with at most $n$ other hypotheses, so the complexity is $O(n^{2})$.
 
-Either algorithm generates all possible transitivity chains.
-If only the longest such chain, from $a$ to $a + n$, is needed, the transitivity rule can be registered as a \emph{destruct} rule, a special sort of forward rule that, whenever it is applied to certain hypotheses, removes the hypotheses from the context.
+Both algorithms generate all possible transitivity chains.
+If only the longest such chain, from $a$ to $a + n$, is desired, the transitivity rule can be registered as a \emph{destruct} rule, a special sort of forward rule that removes from the context any hypotheses it is applied to.
 This reduces the complexity of the incremental algorithm to $O(n)$.
 
-\paragraph{Independent Rules.}
-In the last example, part of the stateful algorithm's success stems from the fact that it could reuse a lot of its previous work.
-An obvious question is then how well does the stateful algorithm behave when there is les work to be reused.
+\paragraph{Independence Benchmark.}
+In the last example, the incremental algorithm performs well partly because it can reuse much of its work.
+We now consider a test with almost no potential for reuse.
 
-Given a set of predicates $\mathcal{P} := \{P_{i,j} (x) | 1 \leq i \leq n \text{ and } 1 \leq j \leq 6\}$ and $\mathcal{Q} :=\{Q_{i}(x) | 1 \leq i \leq n \}$, in this test, we consider the $n$ following rules $r_i : \forall x, P_{i,1}(x)\to \dots P_{i,6}(x) \to Q_i(x)$.
-The procedures are then run on a context containing exactly $\mathcal{P}$.
-Here, as the rules are independent, the interference between them is minimal and there is less work to be reused.
+Given $n ≥ 1$, we use $n$ forward rules of the form $\All{x}{P_{1}~x → \dots → P_{6}~x}$, where $P_{1}, \dots, P_{6}$ are arbitrary predicates that are distinct from each other and from the predicates of all other rules.
+The initial goal contains one hypothesis of type $P_{i}~a$ for each predicate $P_{i}$ that appears in a rule, where $a$ is again either 0 or 100.
+Hence, each rule can be applied exactly once and each hypothesis matches exactly one slot.
 
-This doesn't mean, however, that no work is reused: A successful rule application creates a subgoal with the newly added fact.
-This means that the naive algorithm treats this subgoal as a completely new goal whereas the incremental algorithm can reuse all the work already done on the other hypotheses.
-A similar analysis to the one done in the first benchmark suggests a cubic complexity in the number of rules for the naive algorithm against a linear one for the incremental algorithm.
-The advantage is clearly seen in Fig.~\ref{fig:indep}.
-%In this case, we sketch the behaviour of the stateless algorithm:
-%It pick a first hypothesis, $h_{i,j} : P_{i,j}(a)$.
-%If $j < 6$, the index returns nothing and it selects another hypothesis.
-%For $h_{i,6} : P_{i,6}(a)$, the rule $r_i$ is selected, and it then needs to try and fill its remaining five slot with the context's $6n$ hypotheses which it tries all.
-%This gives a complexity of about $O(n^2)$.
-
-%For the stateful algorithm, the hypotheses are considered one by one and match directly to the rule and the slot.
-%The expected complexity is then linear in the number of rules.
-
-\paragraph{Rules of Varying Depth.}
-As a last test case, we consider a scenario for which the naive algorithm should have a clear advantage over the incremental algorithm.
-The main idea is the following: The naive algorithm sometimes does a better job at recognising that a rule is unprovable.
-For example, when the context contains all but the last premise of a rule; the naive algorithm's index doesn't even select the rule.
-(See Sec.
-\ref{sec:naive} for the complete description of the naive algorithm.)
-This leads to a much faster treatment of this rule than the incremental algorithm that will still process the hypotheses in the context that match the other premises.
-
-Inspired by this, we find it interesting to consider a closely related test: A context with all the necessary hypotheses matching the premises however with one of them being incompatible with the others.
-The reason why we introduce an incomptatible hypothesis, instead of having one absent hypothesis in the context, is because even thought the naive algorithm is sure to be fast, the incremental algorithm also handles this specific situations well.
-See \textbf{Lazy Insertion} in Sec. \ref{sec:implementation}.
-
-With all of this in mind, in Fig.~\ref{fig:erase}, the test is set up as follows: given some rule $r$ with $n$ premises $P_i(x)$, $1 \leq i \leq n$, we run the procedure in a context where $n-1$ premises are instanciated with $a$ and one of them with $a + 1$.
-The numbers on the $x$ axis represent which hypothesis has the incompatible instantiation in the order of consideration by the algorithm.
-In a sense, this is how \textit{deep} the naive algorithm investigate before it stops.
-The position of the incompatible hypothesis has little effect on the stateful algorithm as the index selects the rule in all case.
-The work the stateful algorithm has to do is thus close to constant in this scenario.
-The Figure \ref{fig:erase-noprecomp} showcases that the naive algorithm has a sizable advantage in this test, especially when unification is difficult.
-
-\paragraph{Precompilation.}
-
-Some of the modules implemented in Lean can be precompiled.
-This makes a lot of sense for our approach since it is inevitably more complex that the naive approach.
-It thus benefits from getting precompiled.
-This translate into even more substantial gains in the first two tests and in the third test, the incremental algorithm again takes the advantage over the naive algorithm. See \ref{fig:erase-precomp}.
-%Precompilation is by default turned off for Mathlib4 as it amplifies its already long compilation time.
-%Users working on projects that uses a lot of forward reasoning can however chose to pay this compilation time price once and then benefit from fast forward reasoning.
+Still, Fig.~\ref{fig:indep} shows that the incremental approach outperforms the naive one for $n ≥ 4$.
+The is because the naive algorithm fails to take much advantage of the rules' independence:
+whenever a hypothesis is added (and thus a new subgoal generated), it re-runs all the rules it had already run on the previous subgoal, realising only after the fact that the rules were already applied.
+This yields a complexity of $O(n^{3})$ for the naive algorithm versus $O(n)$ for the incremental algorithm.
 
 \begin{figure}
-  Depth Test
   \centering
@@
       \begin{axis}[
-        %title={Depth Test},
         xlabel={Depth},
@@
         log basis y=2,
-        legend pos=north west,
         ymajorgrids=true,
-        grid style=dashed,
-        legend image post style={mark=},
+        grid style=dashed
         ]
@@
     \end{tikzpicture}
-    \caption{Depth benchmark no-precompilation}%
+    \caption{Without precompilation}%
     \label{fig:erase-noprecomp}
@@
       \begin{axis}[
-        %title={Depth Test},
         xlabel={Depth},
-        %ylabel={Time in ms},
         xmin=1, xmax=5,
@@
     \end{tikzpicture}
-    % graph \includegraphics[width=.4\linewidth]{image1}
-    \caption{Depth benchmark with precompilation}%
+    \caption{With precompilation}%
     \label{fig:erase-precomp}
   \end{subfigure}
-  \caption{The above results depict the behavior of both algorithms in depending on the \text{depth} of potential partial matches.
-The lines with square entry points represent problems where unification is easy.
-The lines with circle entry points are the same tests with hard unification problems.}%
+  \caption{Results of the depth benchmark with precompilation disabled (\subref{fig:erase-noprecomp}) or enabled (\subref{fig:erase-precomp}) and with $a = 0$ (\protect\showpgfsquare) or $a = 100$ (\protect\showpgfcircle).}%
   \label{fig:erase}
 \end{figure}
+
+\paragraph{Depth Benchmark.}
+Our final benchmark concerns situations where a rule matches almost, but not quite.
+We use a single rule $r : \All{x}{P_{1}~x → \dots → P_{n}~x → A}$, where the predicates $P_{i}$ are all distinct and $A$ is an arbitrary proposition.
+\jcom{What is $n$?}
+Given a \emph{depth} $k ≤ n$, the initial goal contains hypotheses of types $P_{i}~a$ for $1 ≤ i ≤ n$ and $i ≠ n - k$ as well as one hypothesis $P_{n-k}~(a + 1)$ (with $a = 0$ or $a = 100$ as before).
+In other words, the hypotheses match all premises with compatible substitutions, except the $k$-th premise from the back.
+Since the naive algorithm processes the premises back-to-front, $k$ corresponds to the number of premises that it matches before it becomes clear that the rule cannot be applied.
+
+This setup is challenging for the incremental approach because (a) no rule is ever applied, so incrementality does not help; and (b) the incremental algorithm expends a non-negligible amount of effort for each of the $n$ hypothesis, whereas the naive algorithm can, at least for small $k$, decide fairly quickly that the rule is not applicable.
+And indeed, Fig.~\ref{fig:erase-noprecomp} shows that the naive approach scales with $k$ while the incremental one does not, but the constant factors are so high that the incremental approach remains slower for $k$ up to 5.
+
+Fig.~\ref{fig:erase-precomp}, however, shows a different picture, with the incremental approach leading in most cases.
+The figure shows the results of running the same depth benchmark with \emph{precompilation} enabled.
+This is a feature of Lean that compiles tactics ahead of time (whereas they are normally interpreted).
+Our technique appears to benefit from this much more than the naive approach, perhaps because the RPINF calculation incurs a large interpretation overhead.
+%Precompilation is by default turned off for Mathlib4 as it amplifies its already long compilation time.
+%Users working on projects that uses a lot of forward reasoning can however chose to pay this compilation time price once and then benefit from fast forward reasoning.
 

### 2025-02-19 paper.tex
Edit evaluation again

@@
 \paragraph{Independence Benchmark.}
-In the last example, the incremental algorithm performs well partly because it can reuse much of its work.
-We now consider a test with almost no potential for reuse.
+In the last example, the incremental algorithm performs well partly because each match at level~1 can be combined with many hypotheses at level~2 without processing the level~1 hypothesis repeatedly.
+We now consider a benchmark with many independent rules where this kind of reuse does not occur.
 
-Given $n ≥ 1$, we use $n$ forward rules of the form $\All{x}{P_{1}~x → \dots → P_{6}~x}$, where $P_{1}, \dots, P_{6}$ are arbitrary predicates that are distinct from each other and from the predicates of all other rules.
-The initial goal contains one hypothesis of type $P_{i}~a$ for each predicate $P_{i}$ that appears in a rule, where $a$ is again either 0 or 100.
+Given $n ≥ 1$, we use $n$ forward rules of the form $\All{x}{P_{1}~x → \dots → P_{6}~x → Q~x}$, where $P_{1}, \dots, P_{6}$ and $Q$ are arbitrary predicates that are distinct from each other and from the predicates of all other rules.
+The initial goal contains one hypothesis of type $P_{i}~a$ for each predicate $P_{i}$ that appears in a rule slot, where $a$ is again either 0 or 100.
 Hence, each rule can be applied exactly once and each hypothesis matches exactly one slot.
@@
 Still, Fig.~\ref{fig:indep} shows that the incremental approach outperforms the naive one for $n ≥ 4$.
-The is because the naive algorithm fails to take much advantage of the rules' independence:
-whenever a hypothesis is added (and thus a new subgoal generated), it re-runs all the rules it had already run on the previous subgoal, realising only after the fact that the rules were already applied.
-This yields a complexity of $O(n^{3})$ for the naive algorithm versus $O(n)$ for the incremental algorithm.
+This is because incrementality helps even in this scenario.
+Whenever a hypothesis is added (and hence a new subgoal created), the naive algorithm re-runs all the rules it had already run on the previous subgoal, realising only after the fact that the rules were already applied.
+The incremental algorithm avoids this source of redundancy.
+Accordingly, the naive algorithm has complexity $O(n^{3})$ on this benchmark, versus $O(n)$ for the incremental algorithm.
 
@@
 \paragraph{Depth Benchmark.}
-Our final benchmark concerns situations where a rule matches almost, but not quite.
-We use a single rule $r : \All{x}{P_{1}~x → \dots → P_{n}~x → A}$, where the predicates $P_{i}$ are all distinct and $A$ is an arbitrary proposition.
+Our final benchmark concerns situations where many rules match almost, but not quite.
+We use 100 copies of a rule $r : \All{x}{P_{1}~x → \dots → P_{n}~x → A}$, where the predicates $P_{i}$ are all distinct and $A$ is an arbitrary proposition.
 \jcom{What is $n$?}
-Given a \emph{depth} $k ≤ n$, the initial goal contains hypotheses of types $P_{i}~a$ for $1 ≤ i ≤ n$ and $i ≠ n - k$ as well as one hypothesis $P_{n-k}~(a + 1)$ (with $a = 0$ or $a = 100$ as before).
-In other words, the hypotheses match all premises with compatible substitutions, except the $k$-th premise from the back.
-Since the naive algorithm processes the premises back-to-front, $k$ corresponds to the number of premises that it matches before it becomes clear that the rule cannot be applied.
+Given a \emph{depth} $k ≤ n$, the initial goal contains hypotheses of types $P_{i}~a$ for each $i ≠ n - k$ with $1 ≤ i ≤ n$, as well as one hypothesis $P_{n-k}~(a + 1)$ (with $a = 0$ or $a = 100$).
+In other words, the hypotheses match all premises with compatible substitutions, except the $k^{\text{th}}$ premise from the back.
+Since the naive algorithm processes the premises back-to-front, $k$ determines how many premises the algorithm matches before it notices that the rule cannot be applied.
 
-This setup is challenging for the incremental approach because (a) no rule is ever applied, so incrementality does not help; and (b) the incremental algorithm expends a non-negligible amount of effort for each of the $n$ hypothesis, whereas the naive algorithm can, at least for small $k$, decide fairly quickly that the rule is not applicable.
-And indeed, Fig.~\ref{fig:erase-noprecomp} shows that the naive approach scales with $k$ while the incremental one does not, but the constant factors are so high that the incremental approach remains slower for $k$ up to 5.
+This setup is challenging for the incremental approach because (1) no rule is ever applied, so incrementality does not help; and (2) the incremental algorithm expends a non-negligible amount of effort for each of the $n$ hypotheses, whereas the naive algorithm can, at least for small $k$, decide fairly quickly that the rule is not applicable.
+And indeed, Fig.~\ref{fig:erase-noprecomp} shows that while the naive approach slows down for greater $k$, big constant factors cause the incremental approach to remain slower for $k$ up to 5.
 
-Fig.~\ref{fig:erase-precomp}, however, shows a different picture, with the incremental approach leading in most cases.
-The figure shows the results of running the same depth benchmark with \emph{precompilation} enabled.
-This is a feature of Lean that compiles tactics ahead of time (whereas they are normally interpreted).
+Fig.~\ref{fig:erase-precomp}, however, shows a different picture, with the incremental approach leading for all $k > 1$.
+The figure visualises the results of the depth benchmark with \emph{precompilation} enabled.
+This is a feature of Lean that compiles tactics ahead of time instead of interpreting them.
 Our technique appears to benefit from this much more than the naive approach, perhaps because the RPINF calculation incurs a large interpretation overhead.
-%Precompilation is by default turned off for Mathlib4 as it amplifies its already long compilation time.
-%Users working on projects that uses a lot of forward reasoning can however chose to pay this compilation time price once and then benefit from fast forward reasoning.
+We observe the same effect when enabling precompilation for the other benchmarks.
+Unfortunately, precompilation also substantially increases Aesop's build time and slows down Lean's build tool, Lake, so we disable it by default, but users of Aesop can choose to enable it for faster forward reasoning.
 

### 2025-02-19 paper.tex
Clarify addHyp

@@
 The procedures maintain the invariants of the variable indices while reporting any complete matches discovered during this process.
+Specifically, if a rule state for a rule $r$ represents the context $Γ$---i.e., it contains the hypotheses of $Γ$ and all partial matches of $r$ that can be derived from them---then the rule state obtained from $\addHyp(h : T)$ represents $Γ,\, h : T$.
 
 \begin{algorithm}
-  \SetKwInOut{Precondition}{Precondition}
-  \SetKwInOut{Postcondition}{Postcondition}
-  \Precondition{$H$ and $M$ contain the hypotheses and matches corresponding to a context $Γ$}
-  \Input{A hypothesis $h : T$ matching slot $i$ of $r$ with substitution $σ$}
-  \Output{Complete matches of $r$ in context $Γ,\, h : T$ that contain $h$}
-  \Postcondition{$H$ and $M$ contain the hypotheses and matches corresponding to $Γ,\, h : T$}
+  \Input{A hypothesis $h : T$}
+  \Output{All complete matches of $r$ in context $Γ,\, h : T$ that contain $h$}
   \BlankLine
   \Procedure{$\mathrm{addHyp}(h : T)$}{
+    \For{all slots $i$ of $r$ such that $h$ matches $i$}{
       \eIf{$i = 1$}{
@@
       }{
+        \KwLet $σ ≔ \sub_{i}(h)$ \\
         \For{$x$ \KwIn $\vars(i)$}{
@@
         \For{$m$ \KwIn $\mathit{ms}$}{
-        $\mathit{completeMs} \gets \mathit{completeMs} ∪ \mathrm{addMatch}(m ++ [h])$
+          $\mathit{completeMs} \gets \mathit{completeMs} ∪ \mathrm{addMatch}(m \mathbin{++} [h])$
         }
@@
     }
+  }
   \BlankLine % without these, the caption is not offset properly
   \BlankLine
-  \caption{Adding a hypothesis to the rule state for a rule $r$}%
+  \caption{Add a hypothesis to the rule state for a rule $r$ and context $Γ$}%
   \label{alg:addhyp}
@@
       \For{$h$ \KwIn $\mathit{hs}$}{
-        $\mathit{completeMs} \gets \mathit{completeMs} ∪ \mathrm{addMatch}(m ++ [h])$
+        $\mathit{completeMs} \gets \mathit{completeMs} ∪ \mathrm{addMatch}(m \mathbin{++} [h])$
       }
@@
   \BlankLine
-  \caption{Adding a match to the rule state for a rule $r$}%
+  \caption{Add a match to the rule state for a rule $r$}%
   \label{alg:addmatch}

### 2025-02-19 paper.tex
Fix signature of addMatch
Edit implementation

@@
   \BlankLine
-  \Procedure{$\mathrm{addMatch}(h : T)$}{
+  \Procedure{$\mathrm{addMatch}(m)$}{
     \eIf{$i$ is complete}{
@@
 Implementing the preceding technique efficiently requires some care, particularly in the context of dependent type theory.
-We briefly discuss the most prominent challenges.
 
@@
 Lean has multiple transparency levels that determine which constants are unfolded.
-\enquote{Reducible} is the most restrictive of these, so we unfold few (and, in practice, only non-recursive) definitions, which greatly reduces the cost of unification.
-However, this limits the power of forward reasoning since fewer terms can be unified.
-
-We also need to account for higher-order unification problems: $Q~\mvar{f}$, where $Q$ is a predicate over functions and $\mvar{f}$ is a function variable, should unify with $Q~(\Lam{x}{x})$.
-Since higher-order unification is undecidable~\cite{ThirdOrderUnificationUndecidableLucchesi,ThirdOrderUnificationUndecidableHuet,SecondOrderUnificationUndecidable}, we instead use a unification algorithm provided by Lean that is complete for first-order unification problems and heuristically solves some higher-order problems as well.
-This makes our technique incomplete in the sense that some theorem instantiations will be missed, but since the same unification algorithm is used throughout Lean, users are well acquainted with this form of incompleteness.
+\enquote{Reducible} is the most restrictive of these, so we unfold few (and, in practice, only non-recursive) definitions.
+Additionally, we use Lean's usual approximation of higher-order unification, which is complete for first-order unification problems and heuristically solves some higher-order problems.
+Either of these limitations render forward reasoning incomplete, but since the same unification is used throughout Lean, users are used to this.
 
 During unification, we treat metavariables in hypotheses as constants, i.e.\ we never assign them.
-This reflects the expected behaviour of forward rules: given a hypothesis $\mvar{x} < \mvar{y}$, we would not expect a forward rule to commit to the choice $\mvar{x} ≔ 1$ and $\mvar{y} ≔ 0$ just because this allows the rule to derive a contradiction.
+This reflects the expected behaviour of forward rules: given a hypothesis $h : \mvar{x} < \mvar{y}$ and rule $1 < 0 → ⊥$, applying the rule to $h$ would replace $\mvar{x}$ with 1 and $\mvar{y}$ with 0 in all goals where $\mvar{x}$ and $\mvar{y}$ appear, even though this may render some goals unprovable.
 
-\paragraph{Data Representation.}
-Since the hypothesis index $H$ and the variable index $M$ of a rule state have the same domain, we combine them into one map.
-This core data structure is then represented by nested persistent hash maps: a first hash map maps variables to a second hash map; the second hash map maps slots to a third hash map; and the third hash map maps instantiations to sets of hypotheses and sets of partial matches.
+\paragraph{RPINF.}
+Since the hypothesis index $H$ and the match index $M$ of a rule state have the same domain, we combine them into one map.
+This map, whose domain is a set of triples $(x, i, t)$, is represented by three nested persistent hash maps.
 
-When we compare instantiations, the comparison should be up to definitional equality.
-For example, when we try to apply a theorem $\All{x}{P~x → Q~x → R~x}$ where the first slot matches a hypothesis of type $P~a$ and the second slot matches a hypothesis of type $P~((\Lam{x}{x})~a)$, the application should succeed even though the instantiations of $\mvar{x}$ are different.
-We therefore do not store $a$ and $(\Lam{x}{x})~a$ in the variable maps as-is.
-
-Instead, we bring all instantiations into a normal form which we call \emph{reducible proof-irrelevant normal form} (RPINF).
+When looking up matching instantiations in the combined map, the instantiations should be compared up to definitional equality.
+For example, the application of a theorem $\All{x}{P~x → Q~x → R~x}$ to hypotheses of type $P~a$ and $Q~((\Lam{x}{x})~a)$ should succeed even though $x$ is instantiated with different terms.
+We therefore do not store $a$ and $(\Lam{x}{x})~a$ in the variable indices as-is.
+Instead, we bring all instantiations into \emph{reducible proof-irrelevant normal form} (RPINF).
 This is the usual normal form of expressions with respect to Lean's notion of computation, with two modifications.
-First, only \emph{reducible} defined constants are unfolded.
-This reflects our choice to perform unification at reducible transparency.
-Second, parts of the original expression that are proofs (in the sense that their types are members of the universe $\Prop$ of propositions) are not normalised.
-This reflects the fact that Lean's type theory definitionally equates any two proofs of the same proposition, so there is no need to check whether they are equal.
+First, only \emph{reducible} defined constants are unfolded, reflecting our choice to perform unification at reducible transparency.
+Second, parts of the original expression that are proofs (in the sense that their types are in the universe of propositions) are not normalised.
+This is justified because Lean's type theory definitionally equates any two proofs of the same proposition, so there is no need to check whether they are equal.
 When computing the RPINF, we also mark all proofs occurring in it with some special \enquote{metadata}.
 
-With this setup, two expressions $t$ and $u$ are definitionally equal at reducible transparency if and only if their RPINFs, $t'$ and $u'$, are syntactically equal except for those portions of $t'$ and $u'$ that are proofs, which are always considered equal.
-Since this is an entirely syntactic notion of equality, it can be checked quickly and we can define an effective hash function that is compatible with it.
+With this setup, two expressions $t$ and $u$ are definitionally equal at reducible transparency if and only if their RPINFs, $t'$ and $u'$, are syntactically equal except for those subexpressions marked as proofs.
+This notion of equality is syntactic, so it can be checked quickly and we can define an effective hash function that is compatible with it.
 This allows us to use a hash map for the instantiations while still comparing them up to definitional equality.
 
-We believe that this scheme is more efficient than alternative schemes relying on repeated unification.
-For example, we could use Lean's implementation of imperfect discrimination trees, which implement a map from expressions to arbitrary data where lookup respects definitional equality.
-However, since these discrimination trees are imperfect, they may return spurious lookup results, so whether a lookup result is valid must be verified by unifying the query expression with the expression that was originally stored in the discrimination tree.
-We believe that repeatedly performing such unifications would generally be more expensive than calculating the RPINF of each hypothesis type (which can be cached).
-
-\paragraph{Lazy Insertion.}
-Adding a hypothesis to a rule state is fairly cheap, but not free: it involves one unification for each slot that the hypothesis likely matches (according to a discrimination tree index), followed by several operations on the variable indices.
-We therefore delay insertions until the index has selected at least one hypothesis for each slot.
-This helps, for example, with the defining rule for involutive functions: if $f : α → α$ is involutive, then for any $a : α$ we have $f (f a) = a$.
-When this rule is used as a forward rule, the premise $a$ matches any non-propositional hypothesis, but with lazy insertion we only process $a$ if there is, in fact, an involutive function is the context.
+We believe that this scheme is more efficient than alternative schemes relying on repeated unifications.
+For example, we could use Lean's discrimination trees, which implement a map from expressions to arbitrary data where lookup respects definitional equality.
+However, these discrimination trees are imperfect, so the validity of a lookup result must be verified by unifying the query expression with the expression that was originally stored in the discrimination tree.
+We believe that performing these unifications would generally be more expensive than calculating the RPINF of each hypothesis type (which can be cached).
 
 \paragraph{Match Equivalence.}
-Naively, two matches $[h₁, \dots, hₙ]$ and $[h₁', \dots, hₙ']$ would be considered equal if they are structurally equal, i.e.\ $h₁ = h₁'$, \dots, $hₙ = hₙ'$.
+Naively, two matches would be considered equal if they contain the same hypotheses (up to definitional equality, again implemented via RPINF).
 However, consider the rule $r : \All{x~y}{P~x~y → Q~x~z → R~x}$ and the hypotheses $h₁ : P~a~b$, $h₂ : P~a~c$, $h₃ : Q~a~d$ and $h₄ : Q~a~e$.
-The complete matches $m₁ ≔ [h₁, h₃]$ and $m₁' ≔ [h₁, h₄]$ are structurally different, but the difference does not matter because the applications $r~h₁~h₃$ and $r~h₁~h₄$ both have type $R~a$.
+The complete matches $m₁ ≔ [h₁, h₃]$ and $m₁' ≔ [h₁, h₄]$ contain different hypotheses, but the difference does not matter because the applications $r~h₁~h₃$ and $r~h₁~h₄$ both have type $R~a$.
 Generating both $m₁$ and $m₁'$ would therefore be redundant.
-Similarly, the partial matches $m₂ ≔ [h₁]$ and $m₂' ≔ [h₂]$ are redundant because the choice between $h₁$ and $h₂$ affects neither the type of any eventual application derived from $m₂$ and $m₂'$ nor the hypotheses that can be used to complete $m₂$ and $m₂'$.
+Similarly, the partial matches $m₂ ≔ [h₁]$ and $m₂' ≔ [h₂]$ are redundant because the choice between $h₁$ and $h₂$ affects neither the type of any eventual application derived from $m₂$ and $m₂'$ nor the hypotheses that can be used to complete them.
 Hence we should store only one of $m₂$ and $m₂'$ in the rule state of $r$.
 Based on these observations, we consider two matches $m$ and $m'$ of a rule $r$ with conclusion $A$ equal if $\lvl(m) = \lvl(m')$ and for each variable $x$ that appears either in the conclusion of $r$ or in the type of a slot $i$ of $r$ with $i > \lvl(m)$ we have $\sub(m)(x) = \sub(m')(x)$.
-To ensure that we respect definitional equality, the instantiations in the substitutions are kept in RPINF.
 
@@
 We modified the naive forward reasoning implementation to use the same mechanism.
+
+\paragraph{Lazy Insertion.}
+Adding a hypothesis to a rule state is cheap, but not free: it involves one unification for each slot that the hypothesis may match, according to a discrimination tree index, followed by several operations on the variable indices.
+We therefore delay insertions until the index has selected at least one hypothesis for each slot.
+This helps, for example, with the defining rule for involutive functions: if $f : α → α$ is involutive, then for any $a : α$ we have $f (f a) = a$.
+When this rule is used as a forward rule, the premise $a$ matches any non-propositional hypothesis, but with lazy insertion we only process hypotheses for $a$ if there is, in fact, an involutive function in the context.
 

### 2025-02-19 paper.tex
Clarify a sentence in RPINF

@@
 
-With this setup, two expressions $t$ and $u$ are definitionally equal at reducible transparency if and only if their RPINFs, $t'$ and $u'$, are syntactically equal except for those subexpressions marked as proofs.
+With this setup, two expressions $t$ and $u$ are definitionally equal at reducible transparency if and only if their RPINFs are syntactically equal except for the marked proof subexpressions.
 This notion of equality is syntactic, so it can be checked quickly and we can define an effective hash function that is compatible with it.

### 2025-02-19 paper.tex
Reformulate naive algorithm in terms of matches

@@
 
-Our evaluation (Sec.~\ref{sec:evaluation}) shows that this new implementation of forward reasoning in Aesop substantially outperforms the previous, naive implementation on some (though not all) synthetic benchmarks.
-The implementation, benchmarks and evaluation data are available in a supplement to this paper~\cite{supplement}.
-Mainline Aesop already uses the new implementation by default (TODO true?).
+Our evaluation (Sec.~\ref{sec:evaluation}) shows that this new implementation of forward reasoning in Aesop substantially outperforms the previous, naive implementation on some synthetic benchmarks.
+The implementation, benchmarks and evaluation data are available in a supplement to this paper~\cite{supplement} and the new implementation is already available in mainline Aesop.
 
@@
 
-\jcom{Formulate this in terms of matches.}
-Prior to this work, Aesop used what we call naive forward reasoning~\cite{Aesop}.
+Prior to this work, Aesop used what we call \emph{naive forward reasoning}~\cite{Aesop}.
 This approach treats forward rules essentially like any other Aesop rule.
-When Aesop selects a forward rule $r : \All{(x₁ : A₁) \dots (xₘ : Aₘ)}{B}$ as the highest-priority rule currently applicable to a goal $Γ ⊢ T$, we first unify the last premise $Aₘ$ with each hypothesis $h : C$ such that $C$ and $Aₘ$ likely unify (as determined by a discrimination tree index).
-If successful, this results in a substitution $σ₁$ containing $xₘ$ and any variables occurring in $Aₘ$.
-We then unify $A_{m-1}[σ₁]$ with the type of each hypothesis in the context, yielding a substitution $σ₂$, and proceed with $A_{m-2}[σ₂]$, etc., until we obtain a complete match for $r$ (skipping premises $A_{m-i}$ if they are already determined by $σᵢ$).
+Hence, when Aesop selects a forward rule $r$ with $m$ slots of types $A₁, \dots, Aₘ$ as the highest-priority rule currently applicable to a goal $Γ ⊢ T$, we look for all complete matches for $r$ in $Γ$.
+
+To that end, we first unify the type of the last slot, $Aₘ$, with each hypothesis $h : B$ such that $B$ and $Aₘ$ likely unify (as determined by a discrimination tree index).
+If successful, this results in a substitution $σ₁$ containing all variables that occur in $Aₘ$.
+We then unify $A_{m-1}[σ₁]$ with the type of each hypothesis in the context, yielding a substitution $σ₂$, and proceed with $A_{m-2}[σ₂]$, etc., until we obtain a complete match for $r$.
 If a premise $A_{m-i}[σᵢ]$ does not unify with any hypothesis type, we backtrack and consider other potential matching hypotheses for $A_{i+1}$.
 These may lead to a different substitution $τ$ and therefore to matching hypotheses for $A_{i}[τ]$.
-If this process yields at least one full application $r~a₁ \dots aₘ$, we can add a new hypothesis of type $T = B[x₁ ↦ a₁, \dots, xₘ ↦ aₘ]$, meaning we generate the subgoal $Γ, h : T$ (unless $Γ$ already contains a hypothesis of type $T$).
+For each complete match $m$ with substitution $σ$ found during this process, we add a new hypothesis $h : B[σ]$, where $B$ is the conclusion of $r$.
 
-This approach is efficient when the last premise of a rule is not present in the context since the rule is then excluded by indexing and never run.
-If the last premise is present, however, we perform a large number of unifications between premises and hypotheses (though some of these could perhaps be avoided heuristically~\cite{TammetSubsumption}).
+This approach is efficient when the last slot of a rule is not present in the context since the rule is then excluded by indexing and never run.
+Otherwise, we perform a large number of unifications between premises and hypotheses (though some of these could perhaps be avoided heuristically~\cite{TammetSubsumption}).
 Furthermore, the whole process is repeated when we proceed to another goal with a similar context.
@@
 
-% TODO pre- and postcondition?
 \begin{algorithm}
@@
 There are unfortunately no standard benchmarks for Aesop-like tactics, and since Aesop is not expected to solve most goals without an extensive, manually curated set of rules, benchmarks for push-button theorem provers such as TPTP~\cite{TPTP} are not very meaningful.
-Additionally, forward rules are currently little used by Aesop's main client, the Mathlib mathematics library~\cite{Mathlib}.
+Additionally, forward rules are little used by Aesop's main client, the Mathlib library~\cite{Mathlib}.
 We therefore only evaluate our technique against the previous, naive Aesop implementation on several synthetic benchmarks that demonstrate the performance characteristics of incremental forward reasoning.
@@
           };
+          \addlegendentry{naive}
         \addplot[
@@
           };
+          \addlegendentry{incremental}
         \addplot[
@@
         log basis y=2,
-        yticklabels={},
+        %yticklabels={},
         legend pos=north west,

### 2025-02-19 paper.tex
Better framing in intro

@@
 A successful rule reduces the current goal to zero or more subgoals, which are solved recursively until either a full proof is found or no rule applies to any open goal or the search times out.
+A typical rule is the $∧$-introduction lemma $A → B → A ∧ B$, which, when applied backwards, reduces a goal $Γ ⊢ A ∧ B$ to subgoals $Γ ⊢ A$ and $Γ ⊢ B$.
 
-Typical rules would be the $∧$-introduction lemma $A → B → A ∧ B$, which, when applied backwards, reduces a goal $Γ ⊢ A ∧ B$ to subgoals $Γ ⊢ A$ and $Γ ⊢ B$; or left $∨$-introduction, which reduces $Γ ⊢ A ∨ B$ to $Γ ⊢ A$.
-Such \emph{backward rules} are the most common rule type, but most search tactics allow users to use any tactic as a rule.
-These rule tactics can output arbitrary subgoals, subject only to the condition that proofs of the returned subgoals can be assembled into a proof of the goal on which the rule was run.
-\jcom{The framing here is wrong: we want to talk about non-forward rules that mess with the context, e.g. case splitting.}
+In addition to such \emph{backwards rules}, most search tactics can also use arbitrary tactics as rules.
+For example, Aesop uses the \texttt{cases} tactic on disjunctive hypotheses, splitting goals of the form $Γ,\, h : A ∨ B ⊢ C$ into subgoals $Γ,\, h : A ⊢ C$ and $Γ,\, h : B ⊢ C$.
+Tactics implement arbitrary inferences; in particular, they can remove any hypothesis and add any provable proposition as a new hypothesis.
 
@@
 In particular, each forward rule was tried on each goal separately (subject to pre-filtering by a discrimination tree~\cite{DiscriminationTrees} index).
-This is wasteful since in practice, many rules produce subgoals with a similar---or even the same---context as the initial goal.
-For example, the ∧-introduction rule affects only the goal's target, not its context.
-Still, all forward rules applied to the initial $Γ ⊢ A ∧ B$ would be tried again on the subgoals $Γ ⊢ A$ and $Γ ⊢ B$, with predictably little effect.
+This is wasteful: while in principle any rule can modify the context in any way, in practice many rules produce subgoals with a similar---or even the same---context as the initial goal.
+For example, the ∧-introduction rule (and, in general, any backwards rule) does not affect the goal's context.
+Still, all forward rules applied to the initial goal $Γ ⊢ A ∧ B$ would be tried again on the subgoals $Γ ⊢ A$ and $Γ ⊢ B$, with predictably little effect.
 

### 2025-02-19 paper.tex
A sentence

@@
 Benchmark code and data are available in our supplement~\cite{supplement}.
-The results shown below were obtained on a MacBook Pro with an Apple~M2 Pro processor and \SI{32}{GB} of RAM.
-The results are an average of \textcolor{red}{N} (TODO) runs per benchmark.
+The results shown below were obtained on a MacBook Pro with an Apple~M2 Pro processor and \SI{32}{GB} of RAM, averaging over \textcolor{red}{N} (TODO) runs per benchmark.
 

### 2025-02-19 paper.tex
small changes (Jasmin's feedback)

--- 2025-02-19 paper.tex
@@
 
-Our evaluation (Sec.~\ref{sec:evaluation}) shows that this new implementation of forward reasoning in Aesop substantially outperforms the previous, naive implementation on some synthetic benchmarks.
-The implementation, benchmarks and evaluation data are available in a supplement to this paper~\cite{supplement} and the new implementation is already available in mainline Aesop.
+Our evaluation (Sec.~\ref{sec:evaluation}) shows that this new implementation of forward reasoning in Aesop substantially outperforms the previous, naive implementation on some (though not all) synthetic benchmarks.
+The implementation, benchmarks and evaluation data are available in a supplement to this paper~\cite{supplement}.
+Mainline Aesop already uses the new implementation by default (TODO true?).
 
@@
 
-Prior to this work, Aesop used what we call \emph{naive forward reasoning}~\cite{Aesop}.
+\jcom{Formulate this in terms of matches.}
+Prior to this work, Aesop used what we call naive forward reasoning~\cite{Aesop}.
 This approach treats forward rules essentially like any other Aesop rule.
-Hence, when Aesop selects a forward rule $r$ with $m$ slots of types $A₁, \dots, Aₘ$ as the highest-priority rule currently applicable to a goal $Γ ⊢ T$, we look for all complete matches for $r$ in $Γ$.
-
-To that end, we first unify the type of the last slot, $Aₘ$, with each hypothesis $h : B$ such that $B$ and $Aₘ$ likely unify (as determined by a discrimination tree index).
-If successful, this results in a substitution $σ₁$ containing all variables that occur in $Aₘ$.
-We then unify $A_{m-1}[σ₁]$ with the type of each hypothesis in the context, yielding a substitution $σ₂$, and proceed with $A_{m-2}[σ₂]$, etc., until we obtain a complete match for $r$.
+When Aesop selects a forward rule $r : \All{(x₁ : A₁) \dots (xₘ : Aₘ)}{B}$ as the highest-priority rule currently applicable to a goal $Γ ⊢ T$, we first unify the last premise $Aₘ$ with each hypothesis $h : C$ such that $C$ and $Aₘ$ likely unify (as determined by a discrimination tree index).
+If successful, this results in a substitution $σ₁$ containing $xₘ$ and any variables occurring in $Aₘ$.
+We then unify $A_{m-1}[σ₁]$ with the type of each hypothesis in the context, yielding a substitution $σ₂$, and proceed with $A_{m-2}[σ₂]$, etc., until we obtain a complete match for $r$ (skipping premises $A_{m-i}$ if they are already determined by $σᵢ$).
 If a premise $A_{m-i}[σᵢ]$ does not unify with any hypothesis type, we backtrack and consider other potential matching hypotheses for $A_{i+1}$.
 These may lead to a different substitution $τ$ and therefore to matching hypotheses for $A_{i}[τ]$.
-For each complete match $m$ with substitution $σ$ found during this process, we add a new hypothesis $h : B[σ]$, where $B$ is the conclusion of $r$.
+If this process yields at least one full application $r~a₁ \dots aₘ$, we can add a new hypothesis of type $T = B[x₁ ↦ a₁, \dots, xₘ ↦ aₘ]$, meaning we generate the subgoal $Γ, h : T$ (unless $Γ$ already contains a hypothesis of type $T$).
 
-This approach is efficient when the last slot of a rule is not present in the context since the rule is then excluded by indexing and never run.
-Otherwise, we perform a large number of unifications between premises and hypotheses (though some of these could perhaps be avoided heuristically~\cite{TammetSubsumption}).
+This approach is efficient when the last premise of a rule is not present in the context since the rule is then excluded by indexing and never run.
+If the last premise is present, however, we perform a large number of unifications between premises and hypotheses (though some of these could perhaps be avoided heuristically~\cite{TammetSubsumption}).
 Furthermore, the whole process is repeated when we proceed to another goal with a similar context.
@@
 
+% TODO pre- and postcondition?
 \begin{algorithm}

### 2025-02-19 paper.tex
Add auto2 to related work

--- 2025-02-19 paper.tex
@@
 In addition to such \emph{backwards rules}, most search tactics can also use arbitrary tactics as rules.
-For example, Aesop uses the \texttt{cases} tactic on disjunctive hypotheses, splitting goals of the form $Γ,\, h : A ∨ B ⊢ C$ into subgoals $Γ,\, h : A ⊢ C$ and $Γ,\, h : B ⊢ C$.
+For example, Aesop uses the cases tactic on disjunctive hypotheses, splitting goals of the form $Γ,\, h : A ∨ B ⊢ C$ into subgoals $Γ,\, h : A ⊢ C$ and $Γ,\, h : B ⊢ C$.
 Tactics implement arbitrary inferences; in particular, they can remove any hypothesis and add any provable proposition as a new hypothesis.
@@
 
-\paragraph{ACL2 and PVS.}
-Among white-box proof automation tactics, to our knowledge only ACL2's \enquote{waterfall}~\cite{ACL2} and PVS's various proof search strategies~\cite{ACL2,PVS} have special support for forward reasoning.
+\paragraph{White-Box Tactics.}
+Among white-box proof automation tactics, to our knowledge only ACL2's \enquote{waterfall}~\cite{ACL2} and PVS's various proof search strategies~\cite{PVS} have special support for forward reasoning.
 Both use essentially Aesop's naive algorithm, i.e.\ the context of each goal is saturated independently without reusing partial results from other goals.
@@
 Aesop implements a similar mechanism, \emph{rule patterns}, not described here due to space constraints.
-Some other white-box tactics, e.g.\ Rocq's auto, can execute arbitrary tactics including forward rules, but make no special provisions for them.
+
+The auto2 tactic~\cite{auto2} employs an SMT-like search procedure based on forward reasoning and e-matching~\cite{simplify}.
+As such, it does not require incrementality, but it could perhaps benefit from indexing along the lines of our variable indices.
+Many other white-box tactics can execute arbitrary tactics, including forward rules, but make no special provisions for them.
 
 \paragraph{SATCHMO.}
-The SATCHMO first-order theorem prover~\cite{SATCHMO} uses a search procedure that treats a given set of implications as forward rules and uses a single non-forward rule to perform case analysis on disjunctive hypotheses.
+The SATCHMO first-order theorem prover~\cite{SATCHMO} uses a search procedure that treats a given set of implications as forward rules and uses a case-splitting rule to perform case analysis on disjunctive hypotheses.
 SATCHMO-style provers may benefit from our incremental approach if the set of implications is large.
@@
 Nevertheless, our technique bears some resemblance to the RETE algorithm~\cite{RETE} for forward reasoning in large knowledge bases.
-
-\jcom{auto2?}
 

### 2025-02-19 paper.tex
Clarify n = 6 in depth benchmark

@@
   \end{subfigure}
-  \caption{Results of the depth benchmark with precompilation disabled (\subref{fig:erase-noprecomp}) or enabled (\subref{fig:erase-precomp}) and with $a = 0$ (\protect\showpgfsquare) or $a = 100$ (\protect\showpgfcircle).}%
+  \caption{Results of the depth benchmark for $n = 6$ with precompilation disabled (\subref{fig:erase-noprecomp}) or enabled (\subref{fig:erase-precomp}) and with $a = 0$ (\protect\showpgfsquare) or $a = 100$ (\protect\showpgfcircle).}%
   \label{fig:erase}
@@
 We use 100 copies of a rule $r : \All{x}{P_{1}~x → \dots → P_{n}~x → A}$, where the predicates $P_{i}$ are all distinct and $A$ is an arbitrary proposition.
-\jcom{What is $n$?}
 Given a \emph{depth} $k ≤ n$, the initial goal contains hypotheses of types $P_{i}~a$ for each $i ≠ n - k$ with $1 ≤ i ≤ n$, as well as one hypothesis $P_{n-k}~(a + 1)$ (with $a = 0$ or $a = 100$).
@@
 
-This setup is challenging for the incremental approach because (1) no rule is ever applied, so incrementality does not help; and (2) the incremental algorithm expends a non-negligible amount of effort for each of the $n$ hypotheses, whereas the naive algorithm can, at least for small $k$, decide fairly quickly that the rule is not applicable.
+This setup is challenging for the incremental algorithm because (1) no rule is ever applied, so incrementality does not help; and (2) the incremental algorithm expends a non-negligible amount of effort for each of the $n$ hypotheses, whereas the naive algorithm can, at least for small $k$, decide fairly quickly that the rule is not applicable.
 And indeed, Fig.~\ref{fig:erase-noprecomp} shows that while the naive approach slows down for greater $k$, big constant factors cause the incremental approach to remain slower for $k$ up to 5.

### 2025-02-19 paper.tex
Mention forward reasoning in intro (lol)
proposed example diagram change

@@
 
-ACL2, PVS and Aesop provide special support for \emph{forward rules}.
+ACL2, PVS and Aesop additionally provide special support for \emph{forward reasoning}, also known as \emph{forward chaining}, in the form of \emph{forward rules}.
 Given, for example, the transitivity lemma $\All{x~y~z}{x < y → y < z → x < z}$, a forward rule run on the goal $h₁ : a < b,\, h₂ : b < c ⊢ A$ produces the single subgoal $h₁ : a < b,\, h₂ : b < c,\, h₃ : a < c ⊢ A$.
@@
 \]
-At this point, $H$ and $M$ are in the state shown in Fig.~\ref{fig:example1:state1}.
-
-\begin{figure}
-  \centering
-  \begin{minipage}{\textwidth}
-    \begin{subcaptiongroup}
-      \subcaptionlistentry{}
-      \begin{minipage}{0.5\textwidth}
-        \begin{tikzpicture}[outer sep=auto, level distance=4em, scale=0.50]
-          \node at (-2,0) {\captiontext*{}\label{fig:example1:state1}};
-          \node {$H_{y}\colon$};
-          \node at (1.5,0) {2}
-            child {node {$b$}
-              child {node {$\{h₁\}$}}};
-          \node at (5, 0) {$M_y\colon$};
-          \node at (6.5, 0) {1}
-            child {node {$c$}
-              child {node {$\{[h₁]\}$}}};
-        \end{tikzpicture}
-      \end{minipage}
-    \end{subcaptiongroup}
-    \begin{subcaptiongroup}
-      \subcaptionlistentry{}
-      \begin{minipage}{0.5\textwidth}
-        \begin{tikzpicture}[outer sep=auto, level distance=4em, scale=0.50]
-          \node at (-2,0) {\captiontext*{}\label{fig:example1:state2}};
-          \node {$H_y\colon$};
-          \node at (1.5, 0) {2}
-            child {node[xshift=-2] {$b$}
-              child {node {$\{h₁\}$}}}
-            child {node[xshift=2] {$a$}
-              child {node {$\{h₂\}$}}};
-          \node at (5, 0) {$M_y\colon$};
-          \node at (6.5, 0) {1}
-            child {node[xshift=-2] {$c$}
-              child {node {$\{[h₁]\}$}}}
-            child {node[xshift=2] {$b$}
-              child {node {$\{[h₂]\}$}}};
-        \end{tikzpicture}
-      \end{minipage}
-    \end{subcaptiongroup}
-  \end{minipage}
-  \\[1em]
-  \begin{subcaptiongroup}
-    \subcaptionlistentry{}
-    \begin{minipage}{\textwidth}
-      \begin{tikzpicture}[outer sep=auto, level distance=4em, scale=0.50]
-        \node at (-2,0) {\captiontext*{}\label{fig:example2:state1}};
-        \node {$H_x\colon$};
-        \node at (1.5, 0) {2}
-          child {node {$a$}
-            child {node {$\{h_1\}$}}};
-        \node at (5, 0) {$H_y\colon$};
-        \node at (6.5, 0) {2}
-          child {node {$a$}
-            child {node {$\{h_1\}$}}};
-        \node at (10, 0) {$M_x\colon$};
-        \node at (11.5, 0) {1}
-          child {node {$a$}
-            child {node {$\{[h_2]\}$}}};
-        \node at (15, 0) {$M_y\colon$};
-        \node at (16.5, 0) {1}
-          child {node {$b$}
-            child {node {$\{[h_2]\}$}}};
-      \end{tikzpicture}
-    \end{minipage}
-  \end{subcaptiongroup}
-  \\[1em]
-  \begin{subcaptiongroup}
-    \subcaptionlistentry{}
-    \begin{minipage}{\textwidth}
-      \begin{tikzpicture}[outer sep=auto, level distance=4em, scale=0.50]
-        \node at (-2,0) {\captiontext*{}\label{fig:example2:state2}};
-        \node {$H_{x}\colon$};
-        \node at (1.5, 0) {2}
-          child {node {$a$}
-            child {node {$\{h_1, h_3\}$}}};
-        \node at (5, 0) {$H_y\colon$};
-        \node at (6.5, 0) {2}
-          child {node {$a$}
-            child {node {$\{h_1\}$}}}
-          child {node {$b$}
-            child {node {$\{h_3\}$}}};
-        \node at (10, 0) {$M_x\colon$};
-        \node at (11.5, 0) {1}
-          child {node {$a$}
-            child {node {$\{[h_2]\}$}}};
-        \node at (15, 0) {$M_y\colon$};
-        \node at (16.5, 0) {1}
-          child {node {$b$}
-            child {node {$\{[h_2]\}$}}};
-      \end{tikzpicture}
-    \end{minipage}
-  \end{subcaptiongroup}
-  \caption{%
-    Rule states encountered during Examples~1 and~2.
-    The variable indices are represented as trees, with nodes in the first row denoting slots and nodes in the second row denoting variable instantiations.%
-  }
-\end{figure}
+At this point, $H$ and $M$ are in the state shown in Diagram~\eqref{eq:example1:state1}.
 
@@
 For slot~2, we add $h₂$ to $H_{y}(2, a)$.
-There are no compatible matches at level~1 so we are done with $h₂$, yielding the final rule state in Fig.~\ref{fig:example1:state2}.
+There are no compatible matches at level~1 so we are done with $h₂$, yielding the final rule state in Diagram~\eqref{eq:example1:state2}.
+
+    \begin{minipage}{0.40\textwidth}
+      \begin{center}
+      \begin{equation}\label{eq:example1:state1}
+        H_y : 
+        \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)}, outer sep=auto, level distance=4em, scale=0.5]
+              \node at (0,0) {2}
+                child {node {$b$}
+                  child {node {$\{h₁\}$}}};
+        \end{tikzpicture}
+        && \ \
+        M_y : 
+        \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)}, outer sep=auto, level distance=4em, scale=0.5]
+              \node at (0, 0) {1}
+                child {node {$c$}
+                  child {node {$\{[h₁]\}$}}};
+        \end{tikzpicture}
+      \end{equation}
+    \end{center}
+    \end{minipage}
+    \begin{minipage}{0.55\textwidth}
+      \begin{center}
+      \begin{equation}\label{eq:example1:state2}
+        H_y : 
+        \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)}, level distance=4em, scale=0.5]
+              \node at (0,0) {2}
+              child {node[xshift=-2] {$b$}
+                child {node {$\{h₁\}$}}}
+              child {node[xshift=2] {$a$}
+                child {node {$\{h₂\}$}}};
+        \end{tikzpicture}
+        &&
+        M_y : 
+        \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)}, outer sep=auto, level distance=4em, scale=0.5]
+              \node at (-1, 0) {1}
+              child {node[xshift=-3] {$c$}
+                child {node {$\{[h₁]\}$}}}
+              child {node[xshift=1] {$b$}
+                child {node {$\{[h₂]\}$}}}
+        \end{tikzpicture}
+      \end{equation}
+      \end{center}
+    \end{minipage}
 
@@
 There are no matches yet, so there is nothing to extend.
-We then add $h_2 : A~a~b$, which initially generates a match $m = [h₂]$ that is added to $M$, yielding the rule state in Fig.~\ref{fig:example2:state1}.
-% NOTE : uncomment to get full trees.
+We then add $h_2 : A~a~b$, which initially generates a match $m = [h₂]$ that is added to $M$, yielding the rule state in Diagram~\eqref{eq:example2:state1}:
+  \begin{align}\label{eq:example2:state1}
+    H_x : 
+    \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)},outer sep=auto, level distance=4em, scale=0.50]
+      \node at (0, 0) {2}
+        child {node {$a$}
+          child {node {$\{h_1\}$}}};
+    \end{tikzpicture}
+    &&
+    H_y : 
+    \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)},outer sep=auto, level distance=4em, scale=0.50]
+      \node at (0, 0) {2}
+        child {node {$a$}
+          child {node {$\{h_1\}$}}};
+    \end{tikzpicture}
+    &&
+    M_x : 
+    \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)},outer sep=auto, level distance=4em, scale=0.50]
+      \node at (0, 0) {1}
+      child {node {$a$}
+        child {node {$\{[h_2]\}$}}};
+    \end{tikzpicture}
+    &&
+    M_y : 
+    \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)},outer sep=auto, level distance=4em, scale=0.50]
+      \node at (0, 0) {1}
+      child {node {$b$}
+        child {node {$\{[h_2]\}$}}};
+    \end{tikzpicture}
+  \end{align}
+
 To find hypotheses compatible with $m$, we consider the set of variables $\vars(\lvl(m) + 1)$.
@@
 We now turn to the remaining hypothesis $h_3 : B~a~b$, adding it first to $H$.
-This yields the rule state in Fig.~\ref{fig:example2:state2}.
+This yields the rule state in Diagram~\eqref{eq:example2:state2}.
 Since $h₃$ matches slot $i = 2$, the matches that can be extended by $h₃$ are
@@
 This produces the complete match $[h₂, h₃]$, so we can add a new hypothesis of type $C~a~b$.
+
+\begin{align}\label{eq:example2:state2}
+  H_x : 
+  \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)},outer sep=auto, level distance=4em, scale=0.50]
+    \node at (0, 0) {2}
+    child {node {$a$}
+      child {node {$\{h_1, h_3\}$}}};
+  \end{tikzpicture}
+  &&
+  H_y : 
+  \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)},outer sep=auto, level distance=4em, scale=0.50]
+    \node at (0, 0) {2}
+    child {node {$a$}
+      child {node {$\{h_1\}$}}}
+    child {node {$b$}
+      child {node {$\{h_3\}$}}};
+  \end{tikzpicture}
+  &&
+  M_x : 
+  \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)},outer sep=auto, level distance=4em, scale=0.50]
+    \node at (0, 0) {1}
+    child {node {$a$}
+      child {node {$\{[h_2]\}$}}};
+  \end{tikzpicture}
+  &&
+  M_y : 
+  \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)},outer sep=auto, level distance=4em, scale=0.50]
+    \node at (0, 0) {1}
+    child {node {$b$}
+      child {node {$\{[h_2]\}$}}};
+  \end{tikzpicture}
+\end{align}
 

### 2025-02-19 paper.tex
fix errors
Move example figures around

@@
-\documentclass[runningheads]{llncs}
+\documentclass[runningheads,leqno]{llncs}
 
@@
 \]
-At this point, $H$ and $M$ are in the state shown in Diagram~\eqref{eq:example1:state1}.
+At this point, $H$ and $M$ are in state~\eqref{eq:example1:state1}, shown below.
 
-Now we add $h₂ : a ≤ b$, which again unifies with both slots.
-For slot~1, $\sub₁(h₂)(y) = b$, so we add the match $[h₂]$ to $M_{y}(1, b)$.
-We then check for hypotheses in slot~2 that also instantiate $y$ with $b$, and indeed we find $M_{y}(2, b) = \{h₁\}$.
-Hence we return the complete match $[h₂, h₁]$, which induces a new hypothesis $h₃ : a ≤ c$ to be added later.
-For slot~2, we add $h₂$ to $H_{y}(2, a)$.
-There are no compatible matches at level~1 so we are done with $h₂$, yielding the final rule state in Diagram~\eqref{eq:example1:state2}.
-
-    \begin{minipage}{0.40\textwidth}
+\begin{center}
+  \begin{minipage}{0.44\textwidth}
     \begin{center}
     \begin{equation}\label{eq:example1:state1}
-        H_y : 
+      H_y\colon
       \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)}, outer sep=auto, level distance=4em, scale=0.5]
@@
       \end{tikzpicture}
-        && \ \
-        M_y : 
+      M_y\colon
       \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)}, outer sep=auto, level distance=4em, scale=0.5]
@@
     \begin{equation}\label{eq:example1:state2}
-        H_y : 
+      H_y\colon
       \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)}, level distance=4em, scale=0.5]
@@
       \end{tikzpicture}
-        &&
-        M_y : 
+      M_y\colon
       \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)}, outer sep=auto, level distance=4em, scale=0.5]
@@
             child {node[xshift=1] {$b$}
-                child {node {$\{[h₂]\}$}}}
+              child {node {$\{[h₂]\}$}}};
       \end{tikzpicture}
@@
   \end{minipage}
+\end{center}
+
+Now we add $h₂ : a ≤ b$, which again unifies with both slots.
+For slot~1, $\sub₁(h₂)(y) = b$, so we add the match $[h₂]$ to $M_{y}(1, b)$.
+We then check for hypotheses in slot~2 that also instantiate $y$ with $b$, and indeed we find $M_{y}(2, b) = \{h₁\}$.
+Hence we return the complete match $[h₂, h₁]$, which induces a new hypothesis $h₃ : a ≤ c$ to be added later.
+For slot~2, we add $h₂$ to $H_{y}(2, a)$.
+There are no compatible matches at level~1 so we are done with $h₂$, yielding the final rule state~\eqref{eq:example1:state2}.
 
@@
 There are no matches yet, so there is nothing to extend.
-We then add $h_2 : A~a~b$, which initially generates a match $m = [h₂]$ that is added to $M$, yielding the rule state in Diagram~\eqref{eq:example2:state1}:
-  \begin{align}\label{eq:example2:state1}
-    H_x : 
+We then add $h_2 : A~a~b$, which initially generates a match $m = [h₂]$ that is added to $M$, yielding the following rule state:
+\begin{align*}
+  H_x\colon
   \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)},outer sep=auto, level distance=4em, scale=0.50]
@@
   &&
-    H_y : 
+  H_y\colon
   \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)},outer sep=auto, level distance=4em, scale=0.50]
@@
   &&
-    M_x : 
+  M_x\colon
   \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)},outer sep=auto, level distance=4em, scale=0.50]
@@
   &&
-    M_y : 
+  M_y\colon
   \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)},outer sep=auto, level distance=4em, scale=0.50]
@@
   \end{tikzpicture}
-  \end{align}
+\end{align*}
 
@@
 We now turn to the remaining hypothesis $h_3 : B~a~b$, adding it first to $H$.
-This yields the rule state in Diagram~\eqref{eq:example2:state2}.
-Since $h₃$ matches slot $i = 2$, the matches that can be extended by $h₃$ are
-\[
-  \bigcap_{x \in \vars(i)} M_{x}(i - 1, \sub_{i}(x)) = M_{x}(1, a) \cap M_{y}(1, b) = \{[h₂]\}
-\]
-This produces the complete match $[h₂, h₃]$, so we can add a new hypothesis of type $C~a~b$.
-
-\begin{align}\label{eq:example2:state2}
-  H_x : 
+This yields the following rule state:
+\begin{align*}
+  H_x\colon
   \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)},outer sep=auto, level distance=4em, scale=0.50]
@@
   &&
-  H_y : 
+  H_y\colon
   \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)},outer sep=auto, level distance=4em, scale=0.50]
@@
   &&
-  M_x : 
+  M_x\colon
   \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)},outer sep=auto, level distance=4em, scale=0.50]
@@
   &&
-  M_y : 
+  M_y\colon
   \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)},outer sep=auto, level distance=4em, scale=0.50]
@@
   \end{tikzpicture}
-\end{align}
+\end{align*}
+Since $h₃$ matches slot $i = 2$, the matches that can be extended by $h₃$ are
+\[
+  \bigcap_{x \in \vars(i)} M_{x}(i - 1, \sub_{i}(x)) = M_{x}(1, a) \cap M_{y}(1, b) = \{[h₂]\}
+\]
+This produces the complete match $[h₂, h₃]$, so we can add a new hypothesis of type $C~a~b$.
 

### 2025-02-19 paper.tex
Fix figure indentation, spacing
Fix figure spacing
Rewrite a para in intro

@@
 Hence forward rules can be used to establish additional facts that may be relevant for other rules.
-For example, the Aeneas project~\cite{Aeneas} uses Aesop's forward rules \jcom{ensure that this is accurate} in its proof automation to establish facts that are later used by an arithmetic decision procedure.
+For example, the Aeneas project~\cite{Aeneas} uses Aesop's forward rules in its proof automation to establish facts that are later used by an arithmetic decision procedure.
 
-However, Aesop's support for forward rules was, until recently, quite basic.
-In particular, each forward rule was tried on each goal separately (subject to pre-filtering by a discrimination tree~\cite{DiscriminationTrees} index).
-This is wasteful: while in principle any rule can modify the context in any way, in practice many rules produce subgoals with a similar---or even the same---context as the initial goal.
-For example, the ∧-introduction rule (and, in general, any backwards rule) does not affect the goal's context.
-Still, all forward rules applied to the initial goal $Γ ⊢ A ∧ B$ would be tried again on the subgoals $Γ ⊢ A$ and $Γ ⊢ B$, with predictably little effect.
+Until recently, Aesop used what we call the \emph{naive algorithm} to apply forward rules.
+When a forward rule was run on a goal, the rule would search the goal's context for all combinations of hypotheses satisfying its premises.
+This method is simple, but involves much repeated work because in practice, contexts encountered during the search tend to be similar.
+For example, the ∧-introduction rule, when run on $Γ ⊢ A ∧ B$, produces subgoals $Γ ⊢ A$ and $Γ ⊢ B$, so the naive algorithm runs the same forward rules on the same context three times.
+More generally, any backwards rule leaves the context unchanged, and many rules that change the context do so only in limited ways.
 
@@
 Our main contribution is a data structure, the \emph{forward state}, that organises the partial matches in such a way that the update can be done efficiently.
-
 In addition to this \emph{incrementality}, the forward state satisfies two other desiderata:
@@
 \begin{center}
-  \begin{minipage}{0.44\textwidth}
+  \begin{minipage}{0.43\textwidth}
     \begin{center}
-    \begin{equation}\label{eq:example1:state1}
+      \begin{align}\label{eq:example1:state1}
         H_y\colon
@@
         \end{tikzpicture}
-      M_y\colon
+        && M_y\colon
         \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)}, outer sep=auto, level distance=4em, scale=0.5]
@@
         \end{tikzpicture}
-    \end{equation}
+      \end{align}
     \end{center}
   \end{minipage}
-  \begin{minipage}{0.55\textwidth}
+  \begin{minipage}{0.56\textwidth}
     \begin{center}
-    \begin{equation}\label{eq:example1:state2}
+      \begin{align}\label{eq:example1:state2}
         H_y\colon
@@
         \end{tikzpicture}
-      M_y\colon
+        && M_y\colon
         \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)}, outer sep=auto, level distance=4em, scale=0.5]
           \node at (-1, 0) {1}
-            child {node[xshift=-3] {$c$}
+            child {node[xshift=-4] {$c$}
               child {node {$\{[h₁]\}$}}}
-            child {node[xshift=1] {$b$}
+            child {node[xshift=4] {$b$}
               child {node {$\{[h₂]\}$}}};
         \end{tikzpicture}
-    \end{equation}
+      \end{align}
     \end{center}

### 2025-02-19 paper.tex
Discuss possible extension in conclusion

@@
 
+One obvious extension of our method would be to exploit redundancy not just between contexts of parent and child goals, but also between those of siblings or even unrelated goals.
+The forward state supports this since any two goals can be diffed and the old goal's forward state updated to match the new one's context.
+
 \bibliography{lit}

### 2025-02-19 paper.tex
Make paragraph titles bold and upright

@@
 
+\newcommand{\para}[1]{\paragraph{\bfseries\upshape #1}}
 \newcommand{\xcom}[1]{{\textcolor{cyan}{Xavier: #1}} }
@@
 
-\paragraph{Forward State.}
+\para{Forward States.}
 We associate to each goal in Aesop's search tree a forward state that reflects the goal's context.
@@
 
-\paragraph{Rule State.}
+\para{Rule States.}
 Partial matches are specific to each rule, so the forward state naturally decomposes into a set of \emph{rule states}, one per rule.
@@
 
-\paragraph{Variable Indices.}
+\para{Variable Indices.}
 Extending the matches above was possible only because all hypotheses are compatible on the shared variable $x$, i.e.\ they all instantiate $x$ with the same term $a$.
@@
 
-\paragraph{Variable Clusters.}
+\para{Variable Clusters.}
 We have so far assumed that for each slot $i > 1$, the set of shared variables $\vars(i)$ is nonempty, i.e.\ each slot shares at least one variable with at least one of the previous slots.
@@
 
-\paragraph{Deletion.}
+\para{Deletion.}
 When a hypothesis $h$ is removed from a goal's context, we must update the forward state accordingly.
@@
 
-\paragraph{Example 1.}
+\para{Example 1.}
 As a first example, consider a rule state for the transitivity rule $r₁ : \All{x~y~z}{x ≤ y → y ≤ z → x ≤ z}$ to which we add, in order, the hypotheses
@@
 
-\paragraph{Example 2.}
+\para{Example 2.}
 We now consider a rule $r₂ : \All{x~y}{A~x~y → B~x~y → C~x~y}$ with multiple shared variables.
@@
 
-\paragraph{Unification.}
+\para{Unification.}
 In Lean's type theory, types contain programs and unification is expected to respect definitional equality, i.e.\ equality up to evaluation.
@@
 
-\paragraph{RPINF.}
+\para{RPINF.}
 Since the hypothesis index $H$ and the match index $M$ of a rule state have the same domain, we combine them into one map.
@@
 
-\paragraph{Match Equivalence.}
+\para{Match Equivalence.}
 Naively, two matches would be considered equal if they contain the same hypotheses (up to definitional equality, again implemented via RPINF).
@@
 
-\paragraph{Redundant Hypotheses.}
+\para{Redundant Hypotheses.}
 When we find a complete match, we add a new hypothesis of the corresponding type $T$ to the goal, but only if there is no other hypothesis of type $T$ already present.
@@
 
-\paragraph{Lazy Insertion.}
+\para{Lazy Insertion.}
 Adding a hypothesis to a rule state is cheap, but not free: it involves one unification for each slot that the hypothesis may match, according to a discrimination tree index, followed by several operations on the variable indices.
@@
 
-\paragraph{Transitivity Benchmark.}
+\para{Transitivity Benchmark.}
 Our first benchmark is a scaled-up variant of the transitivity example from Sec.~\ref{sec:incremental}.
@@
 
-\paragraph{Independence Benchmark.}
+\para{Independence Benchmark.}
 In the last example, the incremental algorithm performs well partly because each match at level~1 can be combined with many hypotheses at level~2 without processing the level~1 hypothesis repeatedly.
@@
 
-\paragraph{Depth Benchmark.}
+\para{Depth Benchmark.}
 Our final benchmark concerns situations where many rules match almost, but not quite.
@@
 
-\paragraph{White-Box Tactics.}
+\para{White-Box Tactics.}
 Among white-box proof automation tactics, to our knowledge only ACL2's \enquote{waterfall}~\cite{ACL2} and PVS's various proof search strategies~\cite{PVS} have special support for forward reasoning.
@@
 
-\paragraph{SATCHMO.}
+\para{SATCHMO.}
 The SATCHMO first-order theorem prover~\cite{SATCHMO} uses a search procedure that treats a given set of implications as forward rules and uses a case-splitting rule to perform case analysis on disjunctive hypotheses.
@@
 
-\paragraph{Knowledge-Based Systems.}
+\para{Knowledge-Based Systems.}
 Some knowledge-based reasoning systems, e.g.\ Algernon~\cite{Algernon} and EYE~\cite{EYE}, support combined forward and backward reasoning.

### 2025-02-19 paper.tex
Fix some leftover "forward match state" occurrences

@@
 When we find a complete match, we add a new hypothesis of the corresponding type $T$ to the goal, but only if there is no other hypothesis of type $T$ already present.
-To efficiently implement this check, we store in the forward match state a set containing the RPINFs of all hypotheses that have been added to (and not subsequently removed from) the state.
+To efficiently implement this check, we store in the forward state a set containing the RPINFs of all hypotheses that have been added to (and not subsequently removed from) the state.
 We modified the naive forward reasoning implementation to use the same mechanism.
@@
 We have presented a technique that integrates forward reasoning into a tree-based proof search procedure supporting arbitrary proof rules.
-We avoid redundant reasoning steps by caching partial forward rule applications in a data structure, the \emph{forward match state}, that enables efficient updates.
+We avoid redundant reasoning steps by caching partial forward rule applications in a data structure, the \emph{forward state}, that enables efficient updates.
 The technique is independent of the search strategy and logic used.

### 2025-02-20 paper.tex
first pass Jasmin's comments

@@
 \newcommand{\All}[2]{\ensuremath{\forall\, #1,\; #2}}
-\newcommand{\mvar}[1]{\ensuremath{?#1}}
+\newcommand{\mvar}[1]{\ensuremath{{?#1}}}
 \newcommand{\Prop}{\ensuremath{\mathrm{Prop}}}
@@
 
-In many proof assistants, simple \enquote*{white-box} automation tactics based on tableau-style tree search play a substantial role in practice: Isabelle's~\cite{Isabelle} and Rocq's~\cite{Coq} auto~\cite{IsabelleAuto}, ACL2's \enquote{waterfall}~\cite{ACL2}, various PVS~\cite{PVS,PVS-tutorial} tactics, Lean's~\cite{Lean4} Aesop~\cite{Aesop}, etc.
+In many proof assistants, simple \enquote*{white-box} automation tactics based on tableau-style tree search play a substantial role in practice: Isabelle's~\cite{Isabelle} and Rocq's~\cite{Coq} auto~\cite{IsabelleAuto}, ACL2's \enquote{waterfall}~\cite{ACL2}, various PVS~\cite{PVS-tutorial,PVS} tactics, Lean's~\cite{Lean4} Aesop~\cite{Aesop}, etc.
 These tactics operate on \emph{goals} $Γ ⊢ A$, where the \emph{context} $Γ$ is a list of assumptions and $A$ is the proposition to be proved.
@@
 When a forward rule was run on a goal, the rule would search the goal's context for all combinations of hypotheses satisfying its premises.
-This method is simple, but involves much repeated work because in practice, contexts encountered during the search tend to be similar.
+This method is simple but involves much repeated work because in practice, contexts encountered during the search tend to be similar.
 For example, the ∧-introduction rule, when run on $Γ ⊢ A ∧ B$, produces subgoals $Γ ⊢ A$ and $Γ ⊢ B$, so the naive algorithm runs the same forward rules on the same context three times.
@@
 Whenever this process produces a full match---i.e., all premises of a forward rule are satisfied by a set of hypothesis in $Δᵢ$---the rule is applied and adds a corresponding new hypothesis to $Δᵢ$.
-Our main contribution is a data structure, the \emph{forward state}, that organises the partial matches in such a way that the update can be done efficiently.
+
+Our main contribution is a data structure, the \emph{forward state}, that organises the partial matches so that updates can be done efficiently.
 In addition to this \emph{incrementality}, the forward state satisfies two other desiderata:
@@
   \item It reduces the number of \emph{unifications} performed when matching hypotheses against rule premises.
-        In dependent type theory, unification happens up to definitional equality, which makes it quite expensive.
+        In dependent type theory, unification happens up to definitional equality, which makes it expensive.
         We therefore want to avoid redundant unifications.
@@
 
-Our evaluation (Sec.~\ref{sec:evaluation}) shows that this new implementation of forward reasoning in Aesop substantially outperforms the previous, naive implementation on some synthetic benchmarks.
-The implementation, benchmarks and evaluation data are available in a supplement to this paper~\cite{supplement} and the new implementation is already available in mainline Aesop.
+Our evaluation (Sec.~\ref{sec:evaluation}) shows that this new implementation of forward reasoning in Aesop substantially outperforms the previous, naive implementation on synthetic benchmarks.
+The implementation, benchmarks and evaluation data are available in a supplement to this paper~\cite{supplement} and the new implementation is already available for the newest version of Aesop.
 
@@
 Given a lemma $\All{(x₁ : A₁) \dots (xₙ : Aₙ)}{B}$, the $xᵢ$ are \emph{premises} and $B$ is the lemma's \emph{conclusion}.
-The type $Aᵢ$ of the $i$-th premise may depend on (i.e., refer to) any premise $xⱼ$ with $j < i$.
+The type $Aᵢ$ of the $i$th premise may depend on (i.e., refer to) any premise $xⱼ$ with $j < i$.
 Implication $A → B$ is an abbreviation for $\All{x : A}{B}$ where $B$ does not depend on $x$.
-Propositions are (certain) types and we do not distinguish between the two concepts.
+Propositions are (certain) types, and we do not distinguish between the two concepts.
 
@@
 For now, we assume that the input premises are numbered consecutively, so in the above example slot~1 has type $x < y$ and slot~2 has type $y < z$.
-The \emph{shared variables of slot $i$}, $\vars(i)$ are those variables that appear in the type of $i$ and also in at least one type of a slot $j < i$.
+The \emph{shared variables of slot $i$}, $\vars(i)$, are those variables that appear in the type of $i$ and also in at least one type of a slot $j < i$.
 
@@
 Unification induces a substitution $σ$ such that $A[σ]$ is definitionally equal to $B[σ]$; we call this substitution $\sub_{i}(h)$.
-For now we do not specify exactly what \enquote{unifies} means; see Sec.~\ref{sec:implementation}.
+In Sec.~\ref{sec:implementation}., we will specify exactly what \enquote{unifies} means.
 
@@
   \item\label{cond:match:matches} For each slot $i ∈ \dom(m)$, the hypothesis $m(i)$ matches $i$.
+  %\smallspace
+
   \item\label{cond:match:compat} For all slots $i,j ∈ \dom(m)$, the substitutions $\sub_{i}(m(i))$ and $\sub_{j}(m(j))$ are \emph{compatible}.
@@
 The \emph{substitution of a match $m$}, $\sub(m)$, is the union of the substitutions induced by the hypotheses in $m$, i.e. $\sub(m) ≔ ⋃_{i ∈ \dom(m)}\sub_{i}(m(i))$.
-This is well-defined by requirement~\ref{cond:match:compat}.
+This is well defined by requirement~\ref{cond:match:compat}.
 The \textit{level of a match $m$}, $\lvl(m)$, is the length of $m$.
@@
 Note that matches represent only those partial applications where arguments are added without gaps.
-Hence for a rule $r : A → B → C → D$, the partial applications $r~a$ and $r~a~b$ for some $a : A$ and $b : B$ correspond to matches, but not $r~\text{--}~b$.
+Hence for a rule $r : A → B → C → D$, the applications $r~a~\_ : B \to D$ and $r~a~b : D$ for some $a : A$ and $b : B$ correspond to matches, but not $r~\_~b : A \to D$.
 This will prevent us from constructing $r~a~b$ in two different but ultimately redundant ways.
@@
 
-Prior to this work, Aesop used what we call \emph{naive forward reasoning}~\cite{Aesop}.
+Before this work, Aesop used what we call \emph{naive forward reasoning}~\cite{Aesop}.
 This approach treats forward rules essentially like any other Aesop rule.
@@
 \para{Forward States.}
-We associate to each goal in Aesop's search tree a forward state that reflects the goal's context.
+We associate with each goal in Aesop's search tree a forward state that reflects the goal's context.
 Let $\States$ be the set of forward states, $\Hyps$ the set of hypotheses, $\Rules$ the set of rules, $\CMatches$ the set of complete matches and $\powerset(X)$ the powerset of $X$.
@@
 The $\addHyp$ function adds a hypothesis to the forward state.
-Hence, if $s$ is the forward state corresponding to context $Γ$ and $\addHyp(s, h : T) = (s', C)$, then $s'$ is the forward state corresponding to $Γ,\, h : T$.
-Similarly, if $s$ is the forward state corresponding to $Γ, h : T$, then $\delHyp(s, h : T)$ is the forward state corresponding to $Γ$.
+Hence, if $s$ is the forward state corresponding to context $Γ$ and $\addHyp(s,\, h : T) = (s', C)$, then $s'$ is the forward state corresponding to $Γ,\, h : T$.
+Similarly, if $s$ is the forward state corresponding to $Γ, h : T$, then $\delHyp(s,\, h : T)$ is the forward state corresponding to $Γ$.
 The second output of $\addHyp$, $C$, is a set of pairs $(r, m)$ where $r$ is a rule and $m$ is a complete match for $r$.
@@
 
-\para{RPINF.}
+\para{Normal Form.}
 Since the hypothesis index $H$ and the match index $M$ of a rule state have the same domain, we combine them into one map.
@@
 For example, the application of a theorem $\All{x}{P~x → Q~x → R~x}$ to hypotheses of type $P~a$ and $Q~((\Lam{x}{x})~a)$ should succeed even though $x$ is instantiated with different terms.
-We therefore do not store $a$ and $(\Lam{x}{x})~a$ in the variable indices as-is.
+We therefore do not store $a$ and $(\Lam{x}{x})~a$ in the variable indices as is.
 Instead, we bring all instantiations into \emph{reducible proof-irrelevant normal form} (RPINF).
@@
 
-We believe that this scheme is more efficient than alternative schemes relying on repeated unifications.
+There are reasons to believe that this scheme is more efficient than alternative schemes relying on repeated unifications.
 For example, we could use Lean's discrimination trees, which implement a map from expressions to arbitrary data where lookup respects definitional equality.
@@
 
-There are unfortunately no standard benchmarks for Aesop-like tactics, and since Aesop is not expected to solve most goals without an extensive, manually curated set of rules, benchmarks for push-button theorem provers such as TPTP~\cite{TPTP} are not very meaningful.
+There are unfortunately no standard benchmarks for Aesop-like tactics, and since Aesop is not expected to solve most goals without an extensive, manually curated set of rules, benchmarks for push-button theorem provers such as TPTP~\cite{TPTP} are not meaningful.
 Additionally, forward rules are little used by Aesop's main client, the Mathlib library~\cite{Mathlib}.
@@
 \para{Independence Benchmark.}
-In the last example, the incremental algorithm performs well partly because each match at level~1 can be combined with many hypotheses at level~2 without processing the level~1 hypothesis repeatedly.
+In the previous example, the incremental algorithm performs well partly because each match at level~1 can be combined with many hypotheses at level~2 without processing the level~1 hypothesis repeatedly.
 We now consider a benchmark with many independent rules where this kind of reuse does not occur.
@@
 Given a \emph{depth} $k ≤ n$, the initial goal contains hypotheses of types $P_{i}~a$ for each $i ≠ n - k$ with $1 ≤ i ≤ n$, as well as one hypothesis $P_{n-k}~(a + 1)$ (with $a = 0$ or $a = 100$).
-In other words, the hypotheses match all premises with compatible substitutions, except the $k^{\text{th}}$ premise from the back.
-Since the naive algorithm processes the premises back-to-front, $k$ determines how many premises the algorithm matches before it notices that the rule cannot be applied.
+In other words, the hypotheses match all premises with compatible substitutions, except the $k$th premise from the back.
+Since the naive algorithm processes the premises from back to front, $k$ determines how many premises the algorithm matches before it notices that the rule cannot be applied.
 
@@
 The technique is independent of the search strategy and logic used.
-Our evaluation shows substantial speedups on some synthetic benchmarks, relative to Aesop's previous implementation.
+Our evaluation shows substantial speedups on synthetic benchmarks, relative to Aesop's previous implementation.
 
 One obvious extension of our method would be to exploit redundancy not just between contexts of parent and child goals, but also between those of siblings or even unrelated goals.
-The forward state supports this since any two goals can be diffed and the old goal's forward state updated to match the new one's context.
+The forward state supports this since for any two goals, we can adapt a goal's saved forward state to a similar goal.
 

### 2025-02-21 paper.tex
some of massin's comments

@@
 This method is simple but involves much repeated work because in practice, contexts encountered during the search tend to be similar.
-For example, the ∧-introduction rule, when run on $Γ ⊢ A ∧ B$, produces subgoals $Γ ⊢ A$ and $Γ ⊢ B$, so the naive algorithm runs the same forward rules on the same context three times.
+For example, the ∧-introduction rule, when run on $Γ ⊢ A ∧ B$, produces subgoals $Γ ⊢ A$ and $Γ ⊢ B$, so the naive algorithm runs the same forward rules on the same context $Γ$ three times.
 More generally, any backwards rule leaves the context unchanged, and many rules that change the context do so only in limited ways.
@@
 
-A \textit{match} $m$ for a give rule $r$ with $n$ slots is a list containing at least one and at most $n$ hypotheses.
+A \textit{match} $m$ for a given rule $r$ with $n$ slots is a list containing at least one and at most $n$ hypotheses.
 We interpret $m$ as a partial map from slots of $r$ to hypotheses, so $\dom(m) = \{i \mid 1 ≤ i ≤ |m| \}$ and for each slot $i ∈ \dom(m)$, $m(i)$ is the hypothesis assigned to $i$.
@@
         \For{$x$ \KwIn $\vars(i)$}{
-          Insert $h$ into the set $H_{x}(i, σ(h))$
+          Insert $h$ into the set $H_{x}(i, σ(x))$
         }
@@
 For slot~1, $\sub₁(h₂)(y) = b$, so we add the match $[h₂]$ to $M_{y}(1, b)$.
-We then check for hypotheses in slot~2 that also instantiate $y$ with $b$, and indeed we find $M_{y}(2, b) = \{h₁\}$.
+We then check for hypotheses in slot~2 that also instantiate $y$ with $b$, and indeed we find $H_{y}(2, b) = \{h₁\}$.
 Hence we return the complete match $[h₂, h₁]$, which induces a new hypothesis $h₃ : a ≤ c$ to be added later.

### 2025-02-21 paper.tex
Minor edits

@@
 Our evaluation (Sec.~\ref{sec:evaluation}) shows that this new implementation of forward reasoning in Aesop substantially outperforms the previous, naive implementation on synthetic benchmarks.
-The implementation, benchmarks and evaluation data are available in a supplement to this paper~\cite{supplement} and the new implementation is already available for the newest version of Aesop.
+The implementation, benchmarks and evaluation data are available in a supplement to this paper~\cite{supplement} and the new implementation is already available in the newest version of Aesop.
 
@@
 Hence, if $s$ is the forward state corresponding to context $Γ$ and $\addHyp(s,\, h : T) = (s', C)$, then $s'$ is the forward state corresponding to $Γ,\, h : T$.
-Similarly, if $s$ is the forward state corresponding to $Γ, h : T$, then $\delHyp(s,\, h : T)$ is the forward state corresponding to $Γ$.
+Similarly, if $s$ is the forward state corresponding to $Γ,\, h : T$, then $\delHyp(s,\, h : T)$ is the forward state corresponding to $Γ$.
 The second output of $\addHyp$, $C$, is a set of pairs $(r, m)$ where $r$ is a rule and $m$ is a complete match for $r$.

### 2025-02-21 paper.tex
Add Jannis's acknowledgements
Save a line in abstract

@@
   We present an approach to forward rules that avoids some of this repeated work by taking advantage of similarities between successive goals.
-  For each goal, we cache partial applications of forward rules in a custom data structure, the \emph{forward state}, which enables efficient updates.
+  For each goal, we cache partial applications of forward rules in a custom data structure that enables efficient updates.
   Our technique is compatible with any search strategy and most logics.
@@
 
+\para{Acknowledgements.}
+We thank Jasmin Blanchette and Massin Guerdi for their valuable comments on drafts of this paper.
+\jcom{TODO Xavier funding.}
+
 \bibliography{lit}

### 2025-02-21 paper.tex
Add explanation for non-horrible scaling of naive algorithm on depth benchmark

@@
 This setup is challenging for the incremental algorithm because (1) no rule is ever applied, so incrementality does not help; and (2) the incremental algorithm expends a non-negligible amount of effort for each of the $n$ hypotheses, whereas the naive algorithm can, at least for small $k$, decide fairly quickly that the rule is not applicable.
-And indeed, Fig.~\ref{fig:erase-noprecomp} shows that while the naive approach slows down for greater $k$, big constant factors cause the incremental approach to remain slower for $k$ up to 5.
+And indeed, Fig.~\ref{fig:erase-noprecomp} shows that while the naive approach slows down for greater $k$ (though not by much since the context contains only $n = 6$ hypotheses), big constant factors cause the incremental approach to remain slower for $k$ up to 5.
 

### 2025-02-24 paper.tex
xavier acknowl.

@@
 We thank Jasmin Blanchette and Massin Guerdi for their valuable comments on drafts of this paper.
-\jcom{TODO Xavier funding.}
+Généreux's research was co-funded by the European Union (ERC, Nekoka, 101083038) and by the Fonds de recherche du Québec, 330205.
+Views and opinions expressed are however those of the authors only and do not necessarily reflect those of the European Union or the European Research Council.
+Neither the European Union nor the granting authority can be held responsible for them.
 

### 2025-02-24 paper.tex
Clarify terminology around variables in intro

@@
 \newcommand{\Prop}{\ensuremath{\mathrm{Prop}}}
-\newcommand{\vars}{\ensuremath{\mathrm{vars}}}
+\newcommand{\vars}{\ensuremath{\mathrm{svars}}}
 \newcommand{\dom}{\ensuremath{\mathrm{dom}}}
@@
 We first introduce some terminology for matches and for forward rules (which we identify with their underlying lemmas).
-Our foundation in this paper is dependent type theory, but our technique can be adapted to other logics.
+We use dependent type theory as a foundation, but our technique generalises to other logics.
 
@@
 
-Premises on which a later premise depends are \emph{variables}.
-All other premises are \emph{input premises}.
-For example, the lemma $\All{x~y~z}{x < y → y < z → x < z}$ has variables $x, y, z$ and (unnamed) input premises of types $x < y$ and $y < z$.
-If a variable occurs in at least two premises, like $y$ above, it is \emph{shared}.
+Using somewhat non-standard terminology, we call premises on which a later premise depends \emph{variables}.
+These are typically terms of non-propositional types, e.g.\ natural numbers or lists.
+The non-variable premises, whose types are typically propositions, are called \emph{slots}.
+For example, the lemma $\All{(x~y~z : ℕ)}{x < y → y < z → x < z}$ has variables $x, y, z$ and slots of types $x < y$ and $y < z$.
+If more than two premises depend on a variable (e.g.\ $y$), the variable is \emph{shared}.
+Note that premises on which only the conclusion depends are not considered variables.
 
-For a lemma with $n$ input premises, we assign to each input premise a unique number between $1$ and $n$.
-This is the premise's \emph{slot}, which we often identify with the premise itself.
-For now, we assume that the input premises are numbered consecutively, so in the above example slot~1 has type $x < y$ and slot~2 has type $y < z$.
+If a lemma has $n$ slots, each slot is assigned a unique number between $1$ and $n$, and we identify slots with their numbers.
+For now, we assume that the slots are numbered consecutively, so in the example above slot~1 has type $x < y$ and slot~2 has type $y < z$.
 The \emph{shared variables of slot $i$}, $\vars(i)$, are those variables that appear in the type of $i$ and also in at least one type of a slot $j < i$.
@@
 We interpret $m$ as a partial map from slots of $r$ to hypotheses, so $\dom(m) = \{i \mid 1 ≤ i ≤ |m| \}$ and for each slot $i ∈ \dom(m)$, $m(i)$ is the hypothesis assigned to $i$.
-Each match $m$ must satisfy the following requirements:
+A match $m$ must additionally satisfy the following requirements:
 \begin{enumerate}
   \item\label{cond:match:matches} For each slot $i ∈ \dom(m)$, the hypothesis $m(i)$ matches $i$.
-  %\smallspace
 
@@
 Hence we should store only one of $m₂$ and $m₂'$ in the rule state of $r$.
-Based on these observations, we consider two matches $m$ and $m'$ of a rule $r$ with conclusion $A$ equal if $\lvl(m) = \lvl(m')$ and for each variable $x$ that appears either in the conclusion of $r$ or in the type of a slot $i$ of $r$ with $i > \lvl(m)$ we have $\sub(m)(x) = \sub(m')(x)$.
+Based on these observations, we consider two matches $m$ and $m'$ of a rule $r$ with conclusion $A$ equal if $\lvl(m) = \lvl(m')$ and for each premise $x$ that appears either in the conclusion of $r$ or in the type of a slot $i$ of $r$ with $i > \lvl(m)$ we have $\sub(m)(x) = \sub(m')(x)$.
 

### 2025-02-24 paper.tex
Clarify matching vs unification

@@
 
-If $h : A$ is a hypothesis in the context of a goal and $A$ unifies with the type $B$ of slot $i$, we say that $h$ \emph{matches} $i$.
-Unification induces a substitution $σ$ such that $A[σ]$ is definitionally equal to $B[σ]$; we call this substitution $\sub_{i}(h)$.
-In Sec.~\ref{sec:implementation}., we will specify exactly what \enquote{unifies} means.
+If $h : C$ is a hypothesis and $C$ matches the type $A$ of slot $i$, we say that $h$ \emph{matches} $i$.
+Matching induces a substitution $σ$ such that $A[σ]$ is definitionally equal to $C$; we define $\sub_{i}(h) ≔ σ$.
+What exactly \enquote{matching} means is discussed in Sec.~\ref{sec:implementation}.
 
@@
 Hence for a rule $r : A → B → C → D$, the applications $r~a~\_ : B \to D$ and $r~a~b : D$ for some $a : A$ and $b : B$ correspond to matches, but not $r~\_~b : A \to D$.
-This will prevent us from constructing $r~a~b$ in two different but ultimately redundant ways.
+This prevents us from constructing $r~a~b$ in two different but ultimately redundant ways.
 The term orders used in various saturation calculi serve a similar function.
@@
 This approach treats forward rules essentially like any other Aesop rule.
-Hence, when Aesop selects a forward rule $r$ with $m$ slots of types $A₁, \dots, Aₘ$ as the highest-priority rule currently applicable to a goal $Γ ⊢ T$, we look for all complete matches for $r$ in $Γ$.
+Hence, when Aesop selects a forward rule $r$ with $n$ slots of types $A₁, \dots, Aₙ$ as the highest-priority rule currently applicable to a goal $Γ ⊢ T$, we look for all complete matches for $r$ in $Γ$.
 
-To that end, we first unify the type of the last slot, $Aₘ$, with each hypothesis $h : B$ such that $B$ and $Aₘ$ likely unify (as determined by a discrimination tree index).
+To that end, we first match the type of the last slot, $Aₙ$, with each hypothesis $h : B$ (using a discrimination tree index to quickly determine potentially matching hypotheses).
 If successful, this results in a substitution $σ₁$ containing all variables that occur in $Aₘ$.
-We then unify $A_{m-1}[σ₁]$ with the type of each hypothesis in the context, yielding a substitution $σ₂$, and proceed with $A_{m-2}[σ₂]$, etc., until we obtain a complete match for $r$.
-If a premise $A_{m-i}[σᵢ]$ does not unify with any hypothesis type, we backtrack and consider other potential matching hypotheses for $A_{i+1}$.
-These may lead to a different substitution $τ$ and therefore to matching hypotheses for $A_{i}[τ]$.
-For each complete match $m$ with substitution $σ$ found during this process, we add a new hypothesis $h : B[σ]$, where $B$ is the conclusion of $r$.
+We then match $A_{m-1}[σ₁]$ against the type of each hypothesis in the context, yielding a substitution $σ₂$, and proceed with $A_{m-2}[σ₂]$, etc., until we obtain a complete match for $r$.
+If a premise $A_{m-i}[σᵢ]$ does not match any hypothesis type, we backtrack and consider other potential matching hypotheses for $A_{i+1}$.
+These may lead to a different substitution $τ$ and therefore to different matching hypotheses for $A_{i}[τ]$.
+For each complete match $m$ with substitution $σ$ found during this process, we add a new hypothesis of type $B[σ]$, where $B$ is the conclusion of $r$.
 
 This approach is efficient when the last slot of a rule is not present in the context since the rule is then excluded by indexing and never run.
-Otherwise, we perform a large number of unifications between premises and hypotheses (though some of these could perhaps be avoided heuristically~\cite{TammetSubsumption}).
-Furthermore, the whole process is repeated when we proceed to another goal with a similar context.
-In particular, the contexts of subgoals produced by whatever rule is eventually applied to the current goal are likely to be substantially similar, if not identical, to the current context, but these similarities are not exploited by the naive algorithm.
+Otherwise, we match large numbers of premises and hypotheses against each other (though this could likely be sped up heuristically~\cite{TammetSubsumption}).
+Moreover, the whole process is repeated when we proceed to another goal with a similar context.
+In particular, the contexts of subgoals produced by whatever rule is eventually applied to the current goal are likely to be similar to the current context, but the naive algorithm does not exploit these similarities.
 
@@
 
-\para{Unification.}
-In Lean's type theory, types contain programs and unification is expected to respect definitional equality, i.e.\ equality up to evaluation.
+\para{Matching.}
+In type theory, types contain programs and matching (or, more generally, unification) is expected to respect definitional equality, i.e.\ equality up to evaluation.
 For instance, a theorem with premise $P~2$, where $P$ is some predicate, should apply to a hypothesis of type $P~((\Lam{x}{x})~2)$ since the two types are definitionally equal.
-As a result, unifying premises and hypotheses is arbitrarily expensive in theory and a major cost centre in practice.
-Our technique therefore seeks to minimise the number of unifications performed.
+As a result, matching premises and hypotheses is arbitrarily expensive in theory and a major cost centre in practice.
+We therefore seek to avoid redundant invocations of the matching procedure.
 
-To limit the expense of unification, we perform unification at \emph{reducible} transparency.
+To limit the cost of matching, we match with \emph{reducible transparency}.
 Lean has multiple transparency levels that determine which constants are unfolded.
 \enquote{Reducible} is the most restrictive of these, so we unfold few (and, in practice, only non-recursive) definitions.
-Additionally, we use Lean's usual approximation of higher-order unification, which is complete for first-order unification problems and heuristically solves some higher-order problems.
-Either of these limitations render forward reasoning incomplete, but since the same unification is used throughout Lean, users are used to this.
-
-During unification, we treat metavariables in hypotheses as constants, i.e.\ we never assign them.
-This reflects the expected behaviour of forward rules: given a hypothesis $h : \mvar{x} < \mvar{y}$ and rule $1 < 0 → ⊥$, applying the rule to $h$ would replace $\mvar{x}$ with 1 and $\mvar{y}$ with 0 in all goals where $\mvar{x}$ and $\mvar{y}$ appear, even though this may render some goals unprovable.
+Additionally, we use Lean's usual approximation of higher-order matching, which is complete for first-order matching problems and heuristically solves some higher-order problems.
+Either of these limitations render forward reasoning incomplete, but since the same matching algorithm is used throughout Lean, users are used to this source of incompleteness.
 
@@
 For example, we could use Lean's discrimination trees, which implement a map from expressions to arbitrary data where lookup respects definitional equality.
-However, these discrimination trees are imperfect, so the validity of a lookup result must be verified by unifying the query expression with the expression that was originally stored in the discrimination tree.
-We believe that performing these unifications would generally be more expensive than calculating the RPINF of each hypothesis type (which can be cached).
+However, these discrimination trees are imperfect, so the validity of a lookup result must be verified by matching the query expression against the expression that was originally stored in the discrimination tree.
+We believe that this would generally be more expensive than calculating the RPINF of each hypothesis type (which can be cached).
 
@@
 Instead of the natural number type of Lean's standard library, which has special support in the compiler and kernel, we use a custom type of Peano naturals for the benchmarks.
-Unifying two concrete Peano natural numbers $n$ and $m$ involves comparing $\min(n, m)$ constructors.
-We can therefore use moderately large numbers to simulate the bigger unification problems that Aesop encounters, for example, in Mathlib's category theory development.
+Matching two concrete Peano natural numbers $n$ and $m$ involves comparing $\min(n, m)$ constructors.
+We can therefore use moderately large numbers to simulate the bigger matching problems that Aesop encounters, for example, in Mathlib's category theory development.
 

### 2025-02-24 paper.tex
Avoid dubious empty intersection in variable clusters

@@
 \para{Variable Clusters.}
-We have so far assumed that for each slot $i > 1$, the set of shared variables $\vars(i)$ is nonempty, i.e.\ each slot shares at least one variable with at least one of the previous slots.
-This assumption is violated for rules such as $\All{x~y}{P~x → Q~y → R~x~y}$: slot~2 ($Q~y$) shares no variables with slot~1 ($P~x$).
-As a result, our above algorithms fail.
-Suppose we are adding the hypothesis $h₂ : Q~b$, having already added the hypothesis $h₁ : P~a$ and hence the partial match $[h₁]$.
-The set of matches compatible with $h₂$ is supposedly $\bigcap_{x \in \vars(2)} M_{x}(i-1, σ(x)) = ∅$, but the match $[h₁]$ can be extended with $h₂$.
-In fact, any match at level~1 can be extended with any hypothesis matching slot~2 since the two slots are independent.
+We have so far only considered rules where for each slot $i > 1$, the set of shared variables $\vars(i)$ is nonempty, i.e.\ each slot shares at least one variable with at least one of the previous slots.
+However, there are also rules such as $\All{x~y}{P~x → Q~y → R~x~y}$, where slot~2 ($Q~y$) shares no variables with slot~1 ($P~x$).
+In this case, any match at level 1 can be extended with any hypothesis at level 2, since the two slots are independent.
 
@@
 The variable clusters share no variables with each other, so each can be treated as a separate rule with its own rule state.
-Finding a complete match for a rule amounts to finding one complete match for each of its clusters.
+Finding a complete match for a rule then amounts to finding one complete match for each of its clusters.
 The slots within each cluster can be ordered to ensure that each slot shares at least one variable with the previous slots, satisfying our assumption.

### 2025-02-24 paper.tex
Use λ x. t instead of λ x, t
Weaken claims about ACL2 and PVS; clarify trigger terms

@@
 
-\newcommand{\Lam}[2]{\ensuremath{\lambda\, #1,\; #2}}
+\newcommand{\Lam}[2]{\ensuremath{\lambda\, #1.\; #2}}
 \newcommand{\All}[2]{\ensuremath{\forall\, #1,\; #2}}
@@
 Among white-box proof automation tactics, to our knowledge only ACL2's \enquote{waterfall}~\cite{ACL2} and PVS's various proof search strategies~\cite{PVS} have special support for forward reasoning.
-Both use essentially Aesop's naive algorithm, i.e.\ the context of each goal is saturated independently without reusing partial results from other goals.
-ACL2 additionally supports \emph{trigger terms}, which allow forward rules to be selected based on the presence of a particular term anywhere in the goal, rather than based on particular hypotheses.
+Both use techniques similar to Aesop's naive algorithm; in particular, the context of each goal is saturated independently.
+ACL2 additionally allows forward rules to be annotated with \emph{trigger terms}, which cause the rules to be run if the trigger terms are matched anywhere in a goal.
 Aesop implements a similar mechanism, \emph{rule patterns}, not described here due to space constraints.
@@
 As such, it does not require incrementality, but it could perhaps benefit from indexing along the lines of our variable indices.
-Many other white-box tactics can execute arbitrary tactics, including forward rules, but make no special provisions for them.
 

### 2025-02-24 paper.tex
Clarify naive and fix indices

@@
 This approach treats forward rules essentially like any other Aesop rule.
-Hence, when Aesop selects a forward rule $r$ with $n$ slots of types $A₁, \dots, Aₙ$ as the highest-priority rule currently applicable to a goal $Γ ⊢ T$, we look for all complete matches for $r$ in $Γ$.
+Hence, when Aesop selects a forward rule $r$ with $n$ slots of types $A₁, \dots, Aₙ$ as the highest-priority rule currently applicable to a goal $Γ ⊢ T$, we look for all complete matches for $r$ in $Γ$, using an approach reminiscent of backtracking subsumption algorithms.
 
-To that end, we first match the type of the last slot, $Aₙ$, with each hypothesis $h : B$ (using a discrimination tree index to quickly determine potentially matching hypotheses).
-If successful, this results in a substitution $σ₁$ containing all variables that occur in $Aₘ$.
-We then match $A_{m-1}[σ₁]$ against the type of each hypothesis in the context, yielding a substitution $σ₂$, and proceed with $A_{m-2}[σ₂]$, etc., until we obtain a complete match for $r$.
-If a premise $A_{m-i}[σᵢ]$ does not match any hypothesis type, we backtrack and consider other potential matching hypotheses for $A_{i+1}$.
-These may lead to a different substitution $τ$ and therefore to different matching hypotheses for $A_{i}[τ]$.
+More specifically, we first match the type of the last slot, $Aₙ$, with each hypothesis $h : B$ (using a discrimination tree index to quickly determine potentially matching hypotheses).
+If successful, this results in a substitution $σ₁$ containing all variables that occur in $Aₙ$.
+We then match $A_{ₙ-1}[σ₁]$ against the type of each hypothesis in the context, yielding a substitution $σ₂$, and proceed with $A_{ₙ-2}[σ₂]$, etc., until we obtain a complete match for $r$.
+If a premise $A_{ₙ-i}[σᵢ]$ does not match any hypothesis type, we backtrack and consider other potential matching hypotheses for $A_{ₙ-i+1}$.
+These may lead to a different substitution $τ$ and therefore to different matching hypotheses for $A_{ₙ-i}[τ]$.
 For each complete match $m$ with substitution $σ$ found during this process, we add a new hypothesis of type $B[σ]$, where $B$ is the conclusion of $r$.
@@
 This approach is efficient when the last slot of a rule is not present in the context since the rule is then excluded by indexing and never run.
-Otherwise, we match large numbers of premises and hypotheses against each other (though this could likely be sped up heuristically~\cite{TammetSubsumption}).
+Otherwise, we match large numbers of premises and hypotheses against each other, though this process could likely be sped up heuristically~\cite{TammetSubsumption}.
 Moreover, the whole process is repeated when we proceed to another goal with a similar context.

### 2025-02-24 paper.tex
Minor edit

@@
 This prevents us from constructing $r~a~b$ in two different but ultimately redundant ways.
-The term orders used in various saturation calculi serve a similar function.
+The term orders used in various saturation calculi have a similar effect.
 
@@
 Finding a complete match for a rule then amounts to finding one complete match for each of its clusters.
-The slots within each cluster can be ordered to ensure that each slot shares at least one variable with the previous slots, satisfying our assumption.
+The slots within each cluster can be ordered to ensure that each slot shares at least one variable with the previous slots.
 

### 2025-02-24 paper.tex
Clarify future work
fix number of benchmark

@@
 Benchmark code and data are available in our supplement~\cite{supplement}.
-The results shown below were obtained on a MacBook Pro with an Apple~M2 Pro processor and \SI{32}{GB} of RAM, averaging over \textcolor{red}{N} (TODO) runs per benchmark.
+The results shown below were obtained on a MacBook Pro with an Apple~M2 Pro processor and \SI{32}{GB} of RAM, averaging over 20 runs per benchmark.
 
@@
 One obvious extension of our method would be to exploit redundancy not just between contexts of parent and child goals, but also between those of siblings or even unrelated goals.
-The forward state supports this since for any two goals, we can adapt a goal's saved forward state to a similar goal.
+This would require no changes to the forward state data structure; we would only need to find pairs of goals that produce minimal diffs.
 

### 2025-02-25 paper.tex
update tests (without precomp)

@@
           coordinates {
-            (1, 7.663472) (2, 9.354750) (4, 15.768194) (8, 104.921791) (16, 1882.683111) (32, 45220.919667)  
+            (1, 6.133137) (2, 7.651537) (4, 13.290939) (8, 91.869091) (16, 1718.359150) (32, 43017.778968) 
           };
@@
           coordinates {
-            (1, 10.308541) (2, 10.704638) (4, 16.274305) (8, 38.800833) (16, 179.697681) (32, 1175.888750)
+            (1, 7.088877) (2, 9.455166) (4, 13.889670) (8, 33.293693) (16, 153.769004) (32, 1018.813208)
           };
@@
           coordinates {
-            (1, 10.591138) (2, 23.881333) (4, 142.013389) (8, 1676.148375) (16, 24327.359417) (32, 392049.849708) 
+            (1, 8.589029) (2, 19.014450) (4, 126.356247) (8, 1560.866835) (16, 22985.414025) (32, 373397.722281) 
           };
@@
           coordinates {
-            (1, 13.974694) (2, 26.216833) (4, 68.953444) (8, 245.601028) (16, 1028.670680) (32, 4388.917500) 
+            (1, 11.729506) (2, 22.231345) (4, 58.294695) (8, 207.863068) (16, 888.210662) (32, 3804.700212)
           };
@@
           coordinates {
-            (1, 10.399958) (2, 16.426541) (4, 43.461403) (8, 205.595541) (16, 1238.090889) (32, 8927.857819)
+            (1, 8.481945) (2, 13.017566) (4, 36.759979) (8, 180.629616) (16, 1156.813791) (32, 8343.946414)
           };
@@
           coordinates {
-            (1, 11.231875) (2, 12.879500) (4, 16.851930) (8, 24.666485) (16, 42.054055) (32, 76.369778)
+            (1, 10.116120) (2, 11.520718) (4, 14.605806) (8, 21.079808) (16, 34.858708) (32, 64.986725)
           };
@@
           coordinates {
-            (1, 19.405222) (2, 33.487472) (4, 80.107597) (8, 284.954791) (16, 1536.089624) (32, 10721.729861)
+            (1, 13.743712) (2, 23.221681) (4, 58.603629) (8, 235.183885) (16, 1389.365902) (32, 9687.141454) 
           };
@@
           coordinates {
-            (1, 25.892069) (2, 40.031930) (4, 69.446902) (8, 130.640986) (16, 260.653917) (32, 481.827902)
+            (1, 19.886189) (2, 30.294954) (4, 51.187760) (8, 93.739397) (16, 178.442899) (32, 351.595918)
           };
@@
           coordinates {
-            (1, 35.466292) (2, 40.918638) (3, 46.936125) (4, 54.933124) (5, 61.269861)
+            (1, 29.953518) (2, 36.796839) (3, 43.879808) (4, 50.748666) (5, 57.841152) 
           };
@@
           coordinates {
-            (1, 135.967736) (2, 127.324444) (3, 136.891250) (4, 118.720444) (5, 116.293222)
+            (1, 130.215285) (2, 115.646322) (3, 125.256110) (4, 111.167837) (5, 110.668739)
           };
@@
           coordinates {
-            (1, 107.978069) (2, 115.900527) (3, 122.454958) (4, 137.732764) (5, 144.037681)
+            (1, 99.132787) (2, 106.834506) (3, 114.863343) (4, 122.601670) (5, 130.616379) 
           };
@@
           coordinates {
-            (1, 619.932611) (2, 601.456805) (3, 600.926110) (4, 589.144389) (5, 595.091180)
+            (1, 573.555837) (2, 559.131666) (3, 568.973041) (4, 550.852189) (5, 551.364258)
           };

### 2025-02-25 paper.tex
fix big O and broken indices

@@
 If successful, this results in a substitution $σ₁$ containing all variables that occur in $Aₙ$.
-We then match $A_{ₙ-1}[σ₁]$ against the type of each hypothesis in the context, yielding a substitution $σ₂$, and proceed with $A_{ₙ-2}[σ₂]$, etc., until we obtain a complete match for $r$.
-If a premise $A_{ₙ-i}[σᵢ]$ does not match any hypothesis type, we backtrack and consider other potential matching hypotheses for $A_{ₙ-i+1}$.
-These may lead to a different substitution $τ$ and therefore to different matching hypotheses for $A_{ₙ-i}[τ]$.
+We then match $A_{n-1}[σ₁]$ against the type of each hypothesis in the context, yielding a substitution $σ₂$, and proceed with $A_{n-2}[σ₂]$, etc., until we obtain a complete match for $r$.
+If a premise $A_{n-i}[σᵢ]$ does not match any hypothesis type, we backtrack and consider other potential matching hypotheses for $A_{n-i+1}$.
+These may lead to a different substitution $τ$ and therefore to different matching hypotheses for $A_{n-i}[τ]$.
 For each complete match $m$ with substitution $σ$ found during this process, we add a new hypothesis of type $B[σ]$, where $B$ is the conclusion of $r$.
@@
 The naive algorithm goes over the context, containing initially $n$ hypotheses, and compares all hypotheses pairwise.
-This is repeated approximately for each of the $n(n-1)/2$ added hypotheses, so the overall complexity is $O(n^4)$.
-By contrast, the incremental algorithm considers each hypothesis only once, matching it against both slots and then comparing it with at most $n$ other hypotheses, so the complexity is $O(n^{2})$.
+This is repeated approximately for each of the $n(n-1)/2$ added hypotheses, so the overall number of unifications performed is of the order of $n^4$.
+By contrast, the incremental algorithm considers each hypothesis only once, matching it against both slots and then comparing it with at most $n$ other hypotheses.
+In this case, the number of unifications performed is of the order of $n^{2}$.
 
@@
 The incremental algorithm avoids this source of redundancy.
-Accordingly, the naive algorithm has complexity $O(n^{3})$ on this benchmark, versus $O(n)$ for the incremental algorithm.
+Accordingly, the naive algorithm performs $n^{3}$ unifications on this benchmark, versus $n$ for the incremental algorithm.
 

### 2025-02-25 paper.tex
Acknowledge Massin's suggested optimisation

@@
 Adding a hypothesis to a rule state is cheap, but not free: it involves one unification for each slot that the hypothesis may match, according to a discrimination tree index, followed by several operations on the variable indices.
-We therefore delay insertions until the index has selected at least one hypothesis for each slot.
-This helps, for example, with the defining rule for involutive functions: if $f : α → α$ is involutive, then for any $a : α$ we have $f (f a) = a$.
-When this rule is used as a forward rule, the premise $a$ matches any non-propositional hypothesis, but with lazy insertion we only process hypotheses for $a$ if there is, in fact, an involutive function in the context.
+We therefore delay insertions in two ways.
+First, we observe that complete matches can only be generated once all slots have been filled.
+Therefore, hypotheses are matched against slots and inserted into the variable indices only once we have found, for each slot, a hypothesis that may match the slot.
+Second, we observe that the variable indices are only queried for hypotheses in a slot $i > 1$ when a match is added to slot $i - 1$.
+Hence we delay the insertion of hypotheses at slot $i > 1$ until there is at least one match in slot $i - 1$.%
+\footnote{This latter optimisation was suggested by Massin Guerdi.}
 
@@
 One obvious extension of our method would be to exploit redundancy not just between contexts of parent and child goals, but also between those of siblings or even unrelated goals.
-This would require no changes to the forward state data structure; we would only need to find pairs of goals that produce minimal diffs.
+This would require no changes to the forward state data structure; we would only need to find pairs of goals that produce small diffs.
 
 \para{Acknowledgements.}
-We thank Jasmin Blanchette and Massin Guerdi for their valuable comments on drafts of this paper.
+We thank Jasmin Blanchette and Massin Guerdi for their very insightful comments on drafts of this paper.
 Généreux's research was co-funded by the European Union (ERC, Nekoka, 101083038) and by the Fonds de recherche du Québec, 330205.

### 2025-02-25 paper.tex
Revert "Acknowledge Massin's suggested optimisation"
This reverts commit ca7e2c414adda5ed7d700b66f18f9f46f147dc91.

Turns out this was actually Xavier's idea. ^^
Minor edit

@@
 Therefore, hypotheses are matched against slots and inserted into the variable indices only once we have found, for each slot, a hypothesis that may match the slot.
-Second, we observe that the variable indices are only queried for hypotheses in a slot $i > 1$ when a match is added to slot $i - 1$.
-Hence we delay the insertion of hypotheses at slot $i > 1$ until there is at least one match in slot $i - 1$.%
-\footnote{This latter optimisation was suggested by Massin Guerdi.}
+Second, we observe that the variable indices are only queried for hypotheses at a slot $i > 1$ when a match is added at slot $i - 1$.
+Hence we delay the insertion of hypotheses at $i$ until there is at least one match in slot $i - 1$.
 

### 2025-02-25 paper.tex
new depth data (temp)

@@
           coordinates {
-            (1, 29.953518) (2, 36.796839) (3, 43.879808) (4, 50.748666) (5, 57.841152) 
+            (1, 32.192292) (2, 39.246833) (3, 47.596875) (4, 53.228625) (5, 60.162625)
           };
@@
           coordinates {
-            (1, 130.215285) (2, 115.646322) (3, 125.256110) (4, 111.167837) (5, 110.668739)
+            (1, 75.599833) (2, 94.688208) (3, 113.647292) (4, 131.776291) (5, 148.492917) 
           };
@@
           coordinates {
-            (1, 99.132787) (2, 106.834506) (3, 114.863343) (4, 122.601670) (5, 130.616379) 
+            (1, 101.828250) (2, 111.354459) (3, 115.116500) (4, 125.602792) (5, 130.459292) 
           };
@@
           coordinates {
-            (1, 573.555837) (2, 559.131666) (3, 568.973041) (4, 550.852189) (5, 551.364258)
+            (1, 235.018833) (2, 319.038292) (3, 421.737083) (4, 522.937000) (5, 604.252875) 
           };
@@
           coordinates {
-            (1, 13.728805) (2, 17.909194) (3, 23.324666) (4, 28.030819) (5, 33.279611)
+            (1, 12.507986) (2, 17.695611) (3, 22.835264) (4, 27.568305) (5, 32.235500) 
           };
@@
           coordinates {
-            (1, 13.841305) (2, 12.801791) (3, 13.358389) (4, 13.200611) (5, 13.251902)
+            (1, 7.969861) (2, 8.940027) (3, 10.796555) (4, 11.887569) (5, 13.822249) 
           };
@@
           coordinates {
-            (1, 81.546486) (2, 85.345444) (3, 90.532347) (4, 98.247764) (5, 105.423875)
+            (1, 76.545402) (2, 82.492861) (3, 88.691375) (4, 94.362653) (5, 98.254972) 
           };
@@
           coordinates {
-            (1, 25.539097) (2, 24.432485) (3, 25.629666) (4, 23.935250) (5, 23.911139)
+            (1, 11.640499) (2, 15.205833) (3, 18.346361) (4, 21.828027) (5, 24.862680) 
           };

### 2025-02-25 paper.tex
Adjust depth benchmark discussion to new data
... with lazy insertion optimisation.

@@
 Since the naive algorithm processes the premises from back to front, $k$ determines how many premises the algorithm matches before it notices that the rule cannot be applied.
+The incremental algorithm would do a constant amount of work per hypothesis if not for the lazy insertion optimisation, which, in combination with a heuristic that orders the slots such that later premises become earlier slots, produces a similar effect.
 
-This setup is challenging for the incremental algorithm because (1) no rule is ever applied, so incrementality does not help; and (2) the incremental algorithm expends a non-negligible amount of effort for each of the $n$ hypotheses, whereas the naive algorithm can, at least for small $k$, decide fairly quickly that the rule is not applicable.
-And indeed, Fig.~\ref{fig:erase-noprecomp} shows that while the naive approach slows down for greater $k$ (though not by much since the context contains only $n = 6$ hypotheses), big constant factors cause the incremental approach to remain slower for $k$ up to 5.
+This setup is challenging for the incremental algorithm because no rule is ever applied, so incrementality does not help.
+And indeed, Fig.~\ref{fig:erase-noprecomp} shows that while both approaches slow down for greater $k$ (though not by much since the context contains only $n = 6$ hypotheses), the incremental approach remains slower for $k$ up to 5.
 
-Fig.~\ref{fig:erase-precomp}, however, shows a different picture, with the incremental approach leading for all $k > 1$.
+Fig.~\ref{fig:erase-precomp}, however, shows a different picture, with the incremental approach leading for all values of $k$.
 The figure visualises the results of the depth benchmark with \emph{precompilation} enabled.
 This is a feature of Lean that compiles tactics ahead of time instead of interpreting them.
-Our technique appears to benefit from this much more than the naive approach, perhaps because the RPINF calculation incurs a large interpretation overhead.
+Our technique benefits from this much more than the naive approach, likely because calculating RPINFs incurs a large interpretation overhead.
 We observe the same effect when enabling precompilation for the other benchmarks.

### 2025-02-25 paper.tex
Fix formulation of complexity analysis

@@
 \newcommand{\powerset}{\ensuremath{\mathcal{P}}}
+\newcommand{\bigO}{\ensuremath{\mathcal{O}}}
 
@@
 
-These results are unsurprising, given the complexity of both algorithms on this benchmark.
-The naive algorithm goes over the context, containing initially $n$ hypotheses, and compares all hypotheses pairwise.
-This is repeated approximately for each of the $n(n-1)/2$ added hypotheses, so the overall number of unifications performed is of the order of $n^4$.
+These results are explained by the number of times either algorithm matches a premise against a hypothesis.
+The naive algorithm iterates over the context, which initially contains $n$ hypotheses, and compares all hypotheses pairwise.
+This is repeated approximately for each of the $n(n-1)/2$ added hypotheses, so $\bigO(n^4)$ match operations are performed.
 By contrast, the incremental algorithm considers each hypothesis only once, matching it against both slots and then comparing it with at most $n$ other hypotheses.
-In this case, the number of unifications performed is of the order of $n^{2}$.
+It therefore performs only $\bigO(n^2)$ match operations.
 
@@
 If only the longest such chain, from $a$ to $a + n$, is desired, the transitivity rule can be registered as a \emph{destruct} rule, a special sort of forward rule that removes from the context any hypotheses it is applied to.
-This reduces the complexity of the incremental algorithm to $O(n)$.
+The incremental algorithm then needs only $\bigO(n)$ match operations.
 
@@
 The incremental algorithm avoids this source of redundancy.
-Accordingly, the naive algorithm performs $n^{3}$ unifications on this benchmark, versus $n$ for the incremental algorithm.
+Accordingly, the incremental algorithm performs $\bigO(n)$ match operations, versus $\bigO(n^3)$ for the naive algorithm.
 

### 2025-02-25 paper.tex
Add some TODOs
Fix ordered partial application example

@@
 % TODO: Title Case for paragraphs
+% TODO: check over/underfull boxes
+% TODO: ensure that numbers in evaluation discussion match the final data
+% TODO: process Massin's minor feedback points
 
@@
 Note that matches represent only those partial applications where arguments are added without gaps.
-Hence for a rule $r : A → B → C → D$, the applications $r~a~\_ : B \to D$ and $r~a~b : D$ for some $a : A$ and $b : B$ correspond to matches, but not $r~\_~b : A \to D$.
+Hence for a rule $r : A → B → C$, the applications $r~a~\_ : B \to C$ and $r~a~b : C$ for some $a : A$ and $b : B$ correspond to matches, but not $r~\_~b : A \to C$.
 This prevents us from constructing $r~a~b$ in two different but ultimately redundant ways.

### 2025-02-26 paper.tex
Incorporate Massin's minor feedback points

@@
 
-% TODO: standardise indices vs indexes
-% TODO: standardise dots after equations
-% TODO: rename variable maps to variable indices everywhere
-% TODO: Title Case for paragraphs
 % TODO: check over/underfull boxes
+% TODO: move references to new page
 % TODO: ensure that numbers in evaluation discussion match the final data
-% TODO: process Massin's minor feedback points
 
@@
 
-More specifically, we first match the type of the last slot, $Aₙ$, with each hypothesis $h : B$ (using a discrimination tree index to quickly determine potentially matching hypotheses).
+More specifically, we first match the type of the last slot, $Aₙ$, with each hypothesis $h : B$ (using an imperfect discrimination tree index to quickly determine potentially matching hypotheses).
 If successful, this results in a substitution $σ₁$ containing all variables that occur in $Aₙ$.
@@
 These may lead to a different substitution $τ$ and therefore to different matching hypotheses for $A_{n-i}[τ]$.
-For each complete match $m$ with substitution $σ$ found during this process, we add a new hypothesis of type $B[σ]$, where $B$ is the conclusion of $r$.
+For each complete match $m$ with substitution $σ$ found during this process, we add a new hypothesis of type $C[σ]$, where $C$ is the conclusion of $r$.
 
-This approach is efficient when the last slot of a rule is not present in the context since the rule is then excluded by indexing and never run.
+This approach is efficient when the last slot of a rule does not match any hypothesis in the context since the rule is then usually excluded by indexing and never run.
 Otherwise, we match large numbers of premises and hypotheses against each other, though this process could likely be sped up heuristically~\cite{TammetSubsumption}.
@@
 The forward state's interface then consists of two functions
-\begin{equation*}
+\[
   \addHyp : \States → \Hyps → \States × \powerset(\Rules × \CMatches) \qquad \delHyp : \States → \Hyps → \States
-\end{equation*}
+\]
 The $\addHyp$ function adds a hypothesis to the forward state.
@@
 Hence we should store only one of $m₂$ and $m₂'$ in the rule state of $r$.
-Based on these observations, we consider two matches $m$ and $m'$ of a rule $r$ with conclusion $A$ equal if $\lvl(m) = \lvl(m')$ and for each premise $x$ that appears either in the conclusion of $r$ or in the type of a slot $i$ of $r$ with $i > \lvl(m)$ we have $\sub(m)(x) = \sub(m')(x)$.
+Based on these observations, we consider two matches $m$ and $m'$ of a rule $r$ equal if $\lvl(m) = \lvl(m')$ and for each premise $x$ that appears either in the conclusion of $r$ or in the type of a slot $i$ of $r$ with $i > \lvl(m)$ we have $\sub(m)(x) = \sub(m')(x)$.
 
@@
 First, we observe that complete matches can only be generated once all slots have been filled.
-Therefore, hypotheses are matched against slots and inserted into the variable indices only once we have found, for each slot, a hypothesis that may match the slot.
+Therefore, hypotheses are matched against slots and inserted into the variable indices only once we have found a potentially matching hypothesis for each slot.
 Second, we observe that the variable indices are only queried for hypotheses at a slot $i > 1$ when a match is added at slot $i - 1$.

### 2025-02-27 paper.tex
update bench

@@
           coordinates {
-            (1, 6.133137) (2, 7.651537) (4, 13.290939) (8, 91.869091) (16, 1718.359150) (32, 43017.778968) 
+            (1, 6.168695) (2, 7.647181) (4, 13.331245) (8, 91.160610) (16, 1686.694560) (32, 42247.399772) 
           };
@@
           coordinates {
-            (1, 7.088877) (2, 9.455166) (4, 13.889670) (8, 33.293693) (16, 153.769004) (32, 1018.813208)
+            (1, 7.279874) (2, 9.579929) (4, 14.167385) (8, 33.641056) (16, 152.281068) (32, 1003.538712)
           };
@@
           coordinates {
-            (1, 8.589029) (2, 19.014450) (4, 126.356247) (8, 1560.866835) (16, 22985.414025) (32, 373397.722281) 
+            (1, 8.599714) (2, 19.510758) (4, 123.130497) (8, 1535.950129) (16, 22642.477537) (32, 371185.740641)
           };
@@
           coordinates {
-            (1, 11.729506) (2, 22.231345) (4, 58.294695) (8, 207.863068) (16, 888.210662) (32, 3804.700212)
+            (1, 11.817681) (2, 22.297554) (4, 58.571268) (8, 204.596366) (16, 870.644077) (32, 3748.262977)
           };
@@
           coordinates {
-            (1, 8.481945) (2, 13.017566) (4, 36.759979) (8, 180.629616) (16, 1156.813791) (32, 8343.946414)
+            (1, 8.376025) (2, 12.775183) (4, 36.042362) (8, 179.360097) (16, 1149.268666) (32, 8300.053437) 
           };
@@
           coordinates {
-            (1, 10.116120) (2, 11.520718) (4, 14.605806) (8, 21.079808) (16, 34.858708) (32, 64.986725)
+            (1, 10.400941) (2, 12.097460) (4, 15.495423) (8, 22.644318) (16, 37.950312) (32, 70.298812)
           };
@@
           coordinates {
-            (1, 13.743712) (2, 23.221681) (4, 58.603629) (8, 235.183885) (16, 1389.365902) (32, 9687.141454) 
+            (1, 13.713279) (2, 23.140410) (4, 57.941920) (8, 234.314904) (16, 1378.349248) (32, 9612.427347) 
           };
@@
           coordinates {
-            (1, 19.886189) (2, 30.294954) (4, 51.187760) (8, 93.739397) (16, 178.442899) (32, 351.595918)
+            (1, 20.117926) (2, 30.537418) (4, 51.648758) (8, 95.003862) (16, 180.553745) (32, 357.379087)
           };
@@
           coordinates {
-            (1, 32.192292) (2, 39.246833) (3, 47.596875) (4, 53.228625) (5, 60.162625)
+            (1, 29.711047) (2, 36.485449) (3, 43.350398) (4, 50.436345) (5, 57.259252) 
           };
@@
           coordinates {
-            (1, 75.599833) (2, 94.688208) (3, 113.647292) (4, 131.776291) (5, 148.492917) 
+            (1, 74.253768) (2, 92.385454) (3, 110.028423) (4, 127.629943) (5, 144.888551) 
           };
@@
           coordinates {
-            (1, 101.828250) (2, 111.354459) (3, 115.116500) (4, 125.602792) (5, 130.459292) 
+            (1, 98.821245) (2, 106.618691) (3, 114.094895) (4, 122.223695) (5, 129.960526) 
           };
@@
           coordinates {
-            (1, 235.018833) (2, 319.038292) (3, 421.737083) (4, 522.937000) (5, 604.252875) 
+            (1, 225.777493) (2, 318.982995) (3, 410.668802) (4, 501.413283) (5, 592.961541) 
           };
@@
           coordinates {
-            (1, 12.507986) (2, 17.695611) (3, 22.835264) (4, 27.568305) (5, 32.235500) 
+            (1, 11.508160) (2, 16.016310) (3, 20.662793) (4, 25.250989) (5, 29.880510) 
           };
@@
           coordinates {
-            (1, 7.969861) (2, 8.940027) (3, 10.796555) (4, 11.887569) (5, 13.822249) 
+            (1, 6.451668) (2, 7.927170) (3, 9.389370) (4, 10.853781) (5, 12.284477)
           };
@@
           coordinates {
-            (1, 76.545402) (2, 82.492861) (3, 88.691375) (4, 94.362653) (5, 98.254972) 
+            (1, 72.542545) (2, 78.261291) (3, 83.286783) (4, 88.887660) (5, 93.587293) 
           };
@@
           coordinates {
-            (1, 11.640499) (2, 15.205833) (3, 18.346361) (4, 21.828027) (5, 24.862680) 
+            (1, 9.969747) (2, 13.145151) (3, 16.236820) (4, 19.391110) (5, 22.623412)
           };

### 2025-02-27 paper.tex
Update numbers in evaluation to match new data

@@
 % TODO: move references to new page
-% TODO: ensure that numbers in evaluation discussion match the final data
 
@@
 The scales are logarithmic, with each dotted horizontal line representing an 8-fold increase.
-Hence, for $2^{5} = 32$ hypotheses and $a = 0$, incremental forward reasoning is over 38 times faster than naive forward reasoning.
-The difference for $a = 100$ is even larger.
+Hence, for $2^{5} = 32$ hypotheses and $a = 0$, incremental forward reasoning is over 42 times faster than naive forward reasoning.
+The speedup for $a = 100$ is around 99x, with naive forward reasoning taking over 6 minutes.
 
-These results are explained by the number of times either algorithm matches a premise against a hypothesis.
+These results are largely explained by the number of times either algorithm matches a premise against a hypothesis.
 The naive algorithm iterates over the context, which initially contains $n$ hypotheses, and compares all hypotheses pairwise.

### 2025-02-27 paper.tex
Fix overfull hboxes

@@
 
-In many proof assistants, simple \enquote*{white-box} automation tactics based on tableau-style tree search play a substantial role in practice: Isabelle's~\cite{Isabelle} and Rocq's~\cite{Coq} auto~\cite{IsabelleAuto}, ACL2's \enquote{waterfall}~\cite{ACL2}, various PVS~\cite{PVS-tutorial,PVS} tactics, Lean's~\cite{Lean4} Aesop~\cite{Aesop}, etc.
+In many proof assistants, simple \enquote*{white-box} proof automation tactics based on tableau-style tree search play a substantial role in practice: Isabelle's~\cite{Isabelle} and Rocq's~\cite{Coq} auto~\cite{IsabelleAuto}, ACL2's \enquote{waterfall}~\cite{ACL2}, various PVS~\cite{PVS-tutorial,PVS} tactics, Lean's~\cite{Lean4} Aesop~\cite{Aesop}, etc.
 These tactics operate on \emph{goals} $Γ ⊢ A$, where the \emph{context} $Γ$ is a list of assumptions and $A$ is the proposition to be proved.
@@
 Among white-box proof automation tactics, to our knowledge only ACL2's \enquote{waterfall}~\cite{ACL2} and PVS's various proof search strategies~\cite{PVS} have special support for forward reasoning.
-Both use techniques similar to Aesop's naive algorithm; in particular, the context of each goal is saturated independently.
+Both use approaches similar to Aesop's naive algorithm; in particular, the context of each goal is saturated independently.
 ACL2 additionally allows forward rules to be annotated with \emph{trigger terms}, which cause the rules to be run if the trigger terms are matched anywhere in a goal.

### 2025-02-28 paper.tex
Move bibliography to separate page
Apply languagetool suggestions

@@
 \documentclass[runningheads,leqno]{llncs}
-
-% TODO: check over/underfull boxes
-% TODO: move references to new page
 
@@
 Given, for example, the transitivity lemma $\All{x~y~z}{x < y → y < z → x < z}$, a forward rule run on the goal $h₁ : a < b,\, h₂ : b < c ⊢ A$ produces the single subgoal $h₁ : a < b,\, h₂ : b < c,\, h₃ : a < c ⊢ A$.
-Hence forward rules can be used to establish additional facts that may be relevant for other rules.
+Hence, forward rules can be used to establish additional facts that may be relevant for other rules.
 For example, the Aeneas project~\cite{Aeneas} uses Aesop's forward rules in its proof automation to establish facts that are later used by an arithmetic decision procedure.
@@
 When a non-forward rule then generates subgoals $Δ₁ ⊢ A₁$, \dots, $Δₙ ⊢ Aₙ$, the new contexts $Δᵢ$ are likely similar to $Γ$.
-Hence it is generally cheaper (and often much cheaper) to start with the partial matches for $Γ$ and update them for each of the $Δᵢ$.
+Hence, it is generally cheaper (and often much cheaper) to start with the partial matches for $Γ$ and update them for each of the $Δᵢ$.
 The update first deletes partial matches that include hypotheses no longer present in $Δᵢ$.
@@
 The \emph{substitution of a match $m$}, $\sub(m)$, is the union of the substitutions induced by the hypotheses in $m$, i.e. $\sub(m) ≔ ⋃_{i ∈ \dom(m)}\sub_{i}(m(i))$.
-This is well defined by requirement~\ref{cond:match:compat}.
+This is well-defined by requirement~\ref{cond:match:compat}.
 The \textit{level of a match $m$}, $\lvl(m)$, is the length of $m$.
@@
 The application is type-correct by requirements~\ref{cond:match:matches} and~\ref{cond:match:compat}.
-If $m$ is complete, then $\sub(m)$ contains a term for every slot and every variable---hence for every premise---of $r$ and we have a full application.
+If $m$ is complete, then $\sub(m)$ contains a term for every slot and every variable---hence for every premise---of $r$, and we have a full application.
 
 Note that matches represent only those partial applications where arguments are added without gaps.
-Hence for a rule $r : A → B → C$, the applications $r~a~\_ : B \to C$ and $r~a~b : C$ for some $a : A$ and $b : B$ correspond to matches, but not $r~\_~b : A \to C$.
+Hence, for a rule $r : A → B → C$, the applications $r~a~\_ : B \to C$ and $r~a~b : C$ for some $a : A$ and $b : B$ correspond to matches, but not $r~\_~b : A \to C$.
 This prevents us from constructing $r~a~b$ in two different but ultimately redundant ways.
@@
 Given a hypothesis $h : T$ that matches slot $i > 1$ with substitution $σ = \sub_{i}(h)$, the partial matches that can be extended by $h$ are those in slot $i - 1$ whose substitutions are compatible with $σ$ on the variables $\vars(i)$ (i.e., the variables shared between $i$ and any earlier slot).
-Assuming for the moment that $\vars(i)$ is nonempty for all slots $i > 1$, these matches are
+Assuming for the moment that $\vars(i)$ is non-empty for all slots $i > 1$, these matches are
 \[
@@
 If $m$ is already complete, it is returned.
-Otherwise $m$ is inserted into $M$ and we query $H$ for hypotheses that can extend $m$.
+Otherwise, $m$ is inserted into $M$ and we query $H$ for hypotheses that can extend $m$.
 Any extended matches thus produced are added to the rule state recursively.
@@
 \para{Variable Clusters.}
-We have so far only considered rules where for each slot $i > 1$, the set of shared variables $\vars(i)$ is nonempty, i.e.\ each slot shares at least one variable with at least one of the previous slots.
+We have so far only considered rules where for each slot $i > 1$, the set of shared variables $\vars(i)$ is non-empty, i.e.\ each slot shares at least one variable with at least one of the previous slots.
 However, there are also rules such as $\All{x~y}{P~x → Q~y → R~x~y}$, where slot~2 ($Q~y$) shares no variables with slot~1 ($P~x$).
@@
 We then check for hypotheses in slot~2 that also instantiate $y$ with $b$, and indeed we find $H_{y}(2, b) = \{h₁\}$.
-Hence we return the complete match $[h₂, h₁]$, which induces a new hypothesis $h₃ : a ≤ c$ to be added later.
+Hence, we return the complete match $[h₂, h₁]$, which induces a new hypothesis $h₃ : a ≤ c$ to be added later.
 For slot~2, we add $h₂$ to $H_{y}(2, a)$.
-There are no compatible matches at level~1 so we are done with $h₂$, yielding the final rule state~\eqref{eq:example1:state2}.
+There are no compatible matches at level~1, so we are done with $h₂$, yielding the final rule state~\eqref{eq:example1:state2}.
 
@@
 These are the variables shared between slot~$\lvl(m) + 1 = 2$ and all previous slots, i.e.\ $x$ and $y$.
-Hence the compatible hypotheses are
+Hence, the compatible hypotheses are
 \[
@@
 With this setup, two expressions $t$ and $u$ are definitionally equal at reducible transparency if and only if their RPINFs are syntactically equal except for the marked proof subexpressions.
-This notion of equality is syntactic, so it can be checked quickly and we can define an effective hash function that is compatible with it.
+This notion of equality is syntactic, so it can be checked quickly, and we can define an effective hash function that is compatible with it.
 This allows us to use a hash map for the instantiations while still comparing them up to definitional equality.
@@
 Similarly, the partial matches $m₂ ≔ [h₁]$ and $m₂' ≔ [h₂]$ are redundant because the choice between $h₁$ and $h₂$ affects neither the type of any eventual application derived from $m₂$ and $m₂'$ nor the hypotheses that can be used to complete them.
-Hence we should store only one of $m₂$ and $m₂'$ in the rule state of $r$.
+Hence, we should store only one of $m₂$ and $m₂'$ in the rule state of $r$.
 Based on these observations, we consider two matches $m$ and $m'$ of a rule $r$ equal if $\lvl(m) = \lvl(m')$ and for each premise $x$ that appears either in the conclusion of $r$ or in the type of a slot $i$ of $r$ with $i > \lvl(m)$ we have $\sub(m)(x) = \sub(m')(x)$.
@@
 Second, we observe that the variable indices are only queried for hypotheses at a slot $i > 1$ when a match is added at slot $i - 1$.
-Hence we delay the insertion of hypotheses at $i$ until there is at least one match in slot $i - 1$.
+Hence, we delay the insertion of hypotheses at $i$ until there is at least one match in slot $i - 1$.
 
@@
 
-For each benchmark, we set up certain forward rules and a goal with handpicked hypotheses.
+For each benchmark, we set up certain forward rules and a goal with hand-picked hypotheses.
 We then run the \texttt{saturate} tactic, which uses either naive or incremental forward reasoning to exhaustively apply the forward rules.
@@
 
+\newpage
 \bibliography{lit}
